# Cluster Sampling

## Key Formulas for Cluster Sampling

Cluster sampling divides the population into $N$ clusters, out of which a simple random sample of $n$ clusters is selected. This design is highly practical when a sampling frame of individual elements is unavailable, or when individuals are geographically grouped.

#### 1. One-Stage Cluster Sampling {-}
In a one-stage design, every element $M_i$ within the sampled cluster $i$ is surveyed. 

*   **Total in Cluster $i$:** $t_i = \sum_{j=1}^{M_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:** The ratio estimator leverages cluster sizes to estimate the overall element mean:
    $$
\overline{y}_r = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n M_i}.
$$

*   **Estimated Variance of the Mean:**
    $$
\hat{V}(\overline{y}_r) = \left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2} \frac{\sum_{i=1}^n (t_i - \overline{y}_r M_i)^2}{n-1}.
$$

#### 2. Two-Stage Cluster Sampling {-}
In a two-stage design, we first sample $n$ clusters, then take a subsample of $m_i$ elements from the $M_i$ elements within each selected cluster. Because we don't observe the true cluster total $t_i$, we estimate it:

*   **Estimated Total in Cluster $i$:** $\hat{t}_i = M_i \overline{y}_i$, where $\overline{y}_i = \frac{1}{m_i} \sum_{j=1}^{m_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:**
    $$
\overline{y}_r = \frac{\sum_{i=1}^n \hat{t}_i}{\sum_{i=1}^n M_i}
$$

*   **Estimated Variance of the Mean:** because $\hat{t}_i$ is now itself an
    estimate, the variance picks up a second-stage term on top of the
    between-cluster term:
    $$
\hat{V}(\overline{y}_r) =
\underbrace{\left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2}
  \frac{\sum_{i=1}^n (\hat{t}_i - \overline{y}_r M_i)^2}{n-1}}_{\text{between clusters}}
+ \underbrace{\frac{1}{nN\overline{M}^2} \sum_{i=1}^n \hat{v}_i}_{\text{within clusters}},
\qquad
\hat{v}_i = \left(1 - \frac{m_i}{M_i}\right) M_i^2 \frac{s_i^2}{m_i},
$$
    where $s_i^2$ is the sample variance of the $m_i$ elements measured in
    cluster $i$.

Why is the second term there at all? The between-cluster sum of squares is
computed from the *estimated* totals $\hat{t}_i$, so it already carries some of
the sub-sampling noise --- but only the fraction $1 - n/N$ of it. The
within-cluster term adds back the missing part. It vanishes in two familiar
cases: a one-stage sample ($m_i = M_i$, so $1 - m_i/M_i = 0$), and a population
with $N$ unknown or effectively infinite (where $1/(nN) \to 0$ and $s_e^2$
alone already does the whole job). It matters most when an appreciable fraction
of the clusters is sampled and the elements within a cluster are heterogeneous.

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `cluster_ratio`, used below for one-stage and two-stage cluster samples) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Estimating for One-stage cluster sampling: analysis of algebra.csv

Consider a population of 187 high school algebra classes in a city. An investigator takes an SRS of 12 of those classes (the clusters). Because this is a one-stage sample, every student in the 12 sampled classes takes the test to evaluate their function knowledge.

We read the raw data and display it using `gt`'s interactive pagination feature to keep the document clean.

```{r tbl-algebra-raw-data}
#| tbl-cap: "Raw data: algebra classes (sample)"
algebra <- read.csv("data/algebra.csv")

algebra |>
  gt() |>
  tab_header(title = "Raw Data: Algebra Classes") |>
  cols_label(Mi = html("M<sub>i</sub>")) |>
  opt_interactive(use_pagination = TRUE, page_size_default = 10)
```
**Output Note:** The table above displays the raw, unaggregated survey data. The interactive pagination allows you to flip through the dataset compactly.

Next, we run our updated function to compute both the cluster-level working table and the overall ratio estimate for the average student score.

```{r algebra-cluster-ratio}
alg_res <- cluster_ratio(data = algebra,
                             cname = "class",
                             csize = "Mi",
                             yvar = "score",
                             N = 187)

# Display the cluster-level working table
alg_res$table
```
**Output Note:** The working table above displays, for each of the 12 randomly selected classes, the class size ($M_i$), the total class score ($t_i$), the fitted value $\hat B M_i$, and the residual and squared residual used to estimate $s_e^2$.

Finally, we extract the ratio estimate from the function output list to assess the overall population mean.

```{r algebra-ratio-mean}
# Display the overall ratio estimate
alg_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** Using the ratio estimator, the estimated mean score for a student across all 187 algebra classes is `r round(alg_res$estimate["Est."], 2)`, with a 95% confidence interval of [`r round(alg_res$estimate["ci.low"], 2)`, `r round(alg_res$estimate["ci.upp"], 2)`]. The standard error accounts for the clustering effect, which typically increases variance compared to a pure SRS of students.

## Estimating for two-stage cluster sampling: analysis of coots.csv

In two-stage cluster sampling, we sample clusters, and then subsample elements within them. In this dataset, researcher Arnold investigated coot eggs. He first selected a random sample of clutches (the clusters), and then measured the volume of a random subsample of eggs within each selected clutch.


```{r tbl-coots-raw-data}
#| tbl-cap: "Raw data: coot egg clutches (sample)"
coots <- read.csv("data/coots.csv")

coots |>
  gt() |>
  tab_header(title = "Raw Data: Coots") |>
  tab_options(
    container.height = px(400), # Adjust height as needed
    container.overflow.y = TRUE # Forces the vertical scroll bar
  )
```


We apply the exact same ratio estimator formulation, though mathematically, the cluster total $t_i$ is now an *estimated* total $\hat{t}_i$.

```{r coots-cluster-ratio}
coots_res <- cluster_ratio(data = coots,
                               cname = "clutch",
                               csize = "csize",
                               yvar = "volume")

coots_res$table
```
**Output Note:** Unlike the previous example, we only measured a subset of eggs per clutch, and the working table now carries an $m_i$ column recording how many. The function automatically estimates the total volume of each clutch ($\hat{t}_i$) by multiplying the true clutch size ($M_i$) by the observed subsample mean ($\overline{y}_i$); the working table (truncated to the first 6 rows + Sum) shows these estimated totals alongside the ratio fit and residuals. No within-cluster variance column appears here: with $N$ unknown (`N = Inf`), the factor $1/(nN)$ in front of the second-stage term is zero, and $s_e^2$ already carries all of the sub-sampling noise on its own.

```{r coots-ratio-mean}
# Display the overall ratio estimate
coots_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimated overall average egg volume is `r round(coots_res$estimate["Est."], 3)` units. Because the total number of clutches in the population ($N$) is not explicitly provided (set to `Inf` by default), the finite population correction factor is safely omitted from the standard error calculation.

## Two-stage cluster sampling with a finite $N$ for Estimating Agriculture Acreages for Farming

Neither example above actually exercises the within-cluster term --- `algebra`
is one-stage, and `coots` has an unknown $N$. To see it at work we need a
population whose clusters we can count. The `agpop.csv` file provides one: its
3,059 counties (dropping those coded `-99`) sit inside exactly $N = 50$ states,
so the states are clusters of known number and known sizes $M_i$.

### Drawing the sample

We take a two-stage sample in the obvious way: an SRS of $n = 10$ states,
followed by an SRS of $m_i = 30\%$ of the counties in each selected state.

```{r agpop-two-stage-sample}
agpop <- read.csv("data/agpop.csv")
agpop$acres92[agpop$acres92 == -99] <- NA
agpop <- agpop[!is.na(agpop$acres92), ]

## cluster sizes M_i: the number of counties in each state
Mi_all <- table(agpop$state)
agpop$Mi <- as.integer(Mi_all[agpop$state])

set.seed(2026)
states <- sample(names(Mi_all), 10)                 # stage 1: n = 10 clusters
agclus <- do.call(rbind, lapply(states, function(st) {
  rows <- which(agpop$state == st)                  # stage 2: m_i = 0.3 M_i
  agpop[sample(rows, max(1, round(0.30 * length(rows)))), ]
}))

c(clusters = length(states), counties = nrow(agclus))
```

### Computing the estimate directly

Everything the estimator needs is a cluster-level summary: the known size $M_i$
of each sampled state, the number of counties actually measured in it ($m_i$),
their mean $\overline{y}_i$ and variance $s_i^2$, and the estimated state total
$\hat{t}_i = M_i \overline{y}_i$.

```{r tbl-agpop-cluster-summary}
#| tbl-cap: "Cluster-level summaries for the 10 sampled states"
n  <- length(states)                                            # clusters drawn
N  <- 50                                                        # clusters in all

Mi    <- tapply(agclus$Mi,      agclus$state, function(x) x[1])  # cluster sizes
mi    <- tapply(agclus$acres92, agclus$state, length)            # counties measured
ybari <- tapply(agclus$acres92, agclus$state, mean)              # subsample means
si2   <- tapply(agclus$acres92, agclus$state, var)               # within-cluster vars
thati <- Mi * ybari                                              # estimated totals

data.frame(Mi, mi, ybari, si2, thati) |>
  gt(rownames_to_stub = TRUE) |>
  cols_label(Mi    = md("$M_i$"),    mi  = md("$m_i$"),
             ybari = md("$\\bar y_i$"), si2 = md("$s_i^2$"),
             thati = md("$\\hat t_i$")) |>
  fmt_number(columns = c(ybari, thati), decimals = 0) |>
  fmt_scientific(columns = si2, decimals = 2)
```
**Output Note:** The states drawn range from `r min(Mi)` to `r max(Mi)` counties,
of which `r min(mi)` to `r max(mi)` were visited. $s_i^2$ is large throughout:
county acreages vary enormously even inside a single state, and that is the
variability the second-stage term will account for.

The ratio estimate itself divides the estimated totals by the cluster sizes.

```{r agpop-ratio-estimate}
Mbar  <- mean(Mi)                  # average size of the sampled clusters
ybarr <- sum(thati) / sum(Mi)      # ybar_r = sum(t_i hat) / sum(M_i)

c(Mbar = Mbar, ybar_r = ybarr)
```

Its variance has two pieces. The first measures how the estimated totals scatter
about the fitted line $\overline{y}_r M_i$; the second collects what each state
contributes through sub-sampling.

```{r agpop-variance-between}
## between-cluster term
ei  <- thati - ybarr * Mi          # ratio residuals
se2 <- sum(ei^2) / (n - 1)         # s_e^2
V_between <- (1 - n / N) * se2 / (n * Mbar^2)

c(s_e2 = se2, V_between = V_between)
```

```{r agpop-variance-within}
## within-cluster term
vi <- (1 - mi / Mi) * Mi^2 * si2 / mi        # v_i hat, one per sampled state
V_within <- sum(vi) / (n * N * Mbar^2)

c(sum_vi = sum(vi), V_within = V_within)
```

Adding them gives the standard error, and the interval uses $n - 1$ degrees of
freedom --- the clusters, not the counties, are the units that were randomised.

```{r agpop-se-ci}
SE  <- sqrt(V_between + V_within)
mem <- qt(0.975, df = n - 1) * SE

c(Est. = ybarr, S.E. = SE, ci.low = ybarr - mem, ci.upp = ybarr + mem)
```
**Output Note:** The estimated mean 1992 acreage per county is
`r format(round(ybarr), big.mark = ",")` acres, with a 95% confidence interval
of [`r format(round(ybarr - mem), big.mark = ",")`,
`r format(round(ybarr + mem), big.mark = ",")`]. The interval is wide: only 10
states were visited, so $n - 1 = 9$ degrees of freedom is all the between-cluster
scatter provides.

### Using the function `cluster_ratio`

The function performs exactly these steps, provided $N$ is supplied.

```{r agpop-cluster-ratio}
clus_res <- cluster_ratio(data = agclus,
                          cname = "state",
                          csize = "Mi",
                          yvar = "acres92",
                          N = 50)

clus_res$table
```
**Output Note:** Because this sample is two-stage *and* $N$ is finite, the
working table carries the three extra columns computed above: $m_i$, the number
of counties measured in each state; $s_i^2$, the variance among them; and
$\hat v_i = (1 - m_i/M_i) M_i^2 s_i^2/m_i$, that state's contribution to the
second-stage variance, whose Sum row feeds the $1/(nN)$ term. The source note
under the table spells the standard error out as the sum of the two pieces.

```{r agpop-cluster-estimate}
clus_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimate, standard error, and interval reproduce the direct
computation above.

## Interactive Demonstration: Clustering by State or at Random {#sec-app-cluster}

The app below repeats this exercise many times over. At every step it draws a
**two-stage cluster sample** --- an SRS of $n$ clusters, then an SRS of
$m_i = p \, M_i$ counties inside each --- together with an ordinary **SRS of
counties** of the same size, from the full `agpop.csv` population (with the
`-99` codes dropped). It estimates the mean of `acres92` from each: the ratio
estimator $\overline{y}_r = \sum \hat{t}_i / \sum M_i$ for the cluster sample,
and the plain sample average for the SRS.

A radio button chooses the clustering variable:

*   **state** --- the 50 real states, the natural clusters of `agpop.csv`.
*   **Random** --- the state labels randomly permuted, which keeps the cluster
    sizes $M_i$ exactly as they are but scrambles which counties belong
    together.

Two sliders set the design: $n$, the number of clusters drawn, and
$p = m_i/M_i$, the percentage of counties subsampled from each selected cluster
($p = 100\%$ gives a one-stage sample). Play advances the simulation one draw at
a time, Pause stops it, Skip jumps straight to the end, and changing any setting
clears the accumulated history.

The left panel shows the population split into its clusters, with the
highlighted sample in red. For the cluster design the selected clusters are
shaded and labelled $m_i/M_i$, a bar marks each subsample mean
$\overline{y}_i$, and the ratio estimate $\overline{y}_r$ runs across the
panel; for the SRS a single bar marks the sample mean.

The right panel accumulates the estimates of both designs, one column each,
drawn as a violin once enough draws have accumulated. The bar beside each
column is the current confidence interval, green when it covers
$\mu$ and red when it does not. The legends report the running CI coverage and
the design effect --- the variance of the ratio estimator relative to the SRS
average, simulated and theoretical --- and the subtitle gives the
between-cluster $R^2$ of `acres92`.

::: {#fig-cluster-app}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 820

## file: app.R
library(shiny)

data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/data/agpop.csv"
download.file(data_url, "agpop.csv")
dat <- read.csv("agpop.csv")

## -99 is the missing-value code in agpop.csv; recode to NA and drop the
## counties with no acres92, so that a single population underlies both
## clustering choices below.
dat$acres92[dat$acres92 == -99] <- NA
dat <- dat[!is.na(dat$acres92), ]

pop   <- dat$acres92
Nel   <- length(pop)          # number of elements (counties) in the population
ybarU <- mean(pop)
S     <- sd(pop)

## Build the clustering: "state" groups the counties as they really are,
## "random" permutes the state labels, so the clusters keep exactly the same
## sizes M_i but lose every trace of within-cluster homogeneity. Comparing the
## two isolates the effect of homogeneity inside clusters, and nothing else.
make_clusters <- function(method) {
  lab <- factor(dat$state)
  if (method == "random") {
    set.seed(7)
    lab <- sample(lab)                       # same sizes, scrambled membership
    levels(lab) <- sprintf("C%02d", seq_len(nlevels(lab)))
  }
  H     <- nlevels(lab)                      # N, the number of clusters
  idx   <- split(seq_len(Nel), lab)
  Mi    <- sapply(idx, length)
  ti    <- sapply(idx, function(i) sum(pop[i]))
  Si2   <- sapply(idx, function(i) if (length(i) > 1) var(pop[i]) else 0)

  set.seed(42)
  x_pos <- as.integer(lab) + runif(Nel, -0.32, 0.32)

  list(lev = levels(lab), H = H, idx = idx, Mi = Mi, ti = ti, Si2 = Si2,
       M0 = sum(Mi), Mbar = mean(Mi), x_pos = x_pos,
       R2 = 1 - sum(sapply(idx, function(i) sum((pop[i] - mean(pop[i]))^2))) /
                sum((pop - ybarU)^2))
}

CLUST_CHOICES <- c("state (real clusters)" = "state",
                   "Random (scrambled labels)" = "random")

## subsample size in cluster i: m_i = p * M_i, at least 1 and at most M_i
subsize <- function(Mi, p) pmax(1, pmin(Mi, round(p * Mi)))

## Anticipated variances from the FULL population, for the "theory" figures.
## Two-stage ratio estimator (between-cluster term + within-cluster term):
##   V = (1 - n/N) S_t^2 / (n Mbar^2)
##       + sum_U M_i^2 (1 - m_i/M_i) S_i^2 / m_i / (n N Mbar^2),
## with S_t^2 the population variance of the ratio residuals t_i - ybarU M_i.
## SRS is compared at the sample size the cluster design delivers on average.
theory <- function(s, n, p) {
  mi  <- subsize(s$Mi, p)
  St2 <- sum((s$ti - ybarU * s$Mi)^2) / (s$H - 1)
  Vc  <- (1 - n / s$H) * St2 / (n * s$Mbar^2) +
         sum(s$Mi^2 * (1 - mi / s$Mi) * s$Si2 / mi) / (n * s$H * s$Mbar^2)
  nel <- max(2, round(n * mean(mi)))
  Vs  <- (1 - nel / s$M0) * S^2 / nel
  list(Vc = Vc, Vs = Vs, nel = nel)
}

## One two-stage cluster sample: an SRS of n clusters, then an SRS of
## m_i = p M_i counties inside each sampled cluster. The estimate is the ratio
## estimator with M_i as the auxiliary variable, ybar_r = sum(t_i hat)/sum(M_i),
## and its variance is estimated exactly as cluster_ratio() does: the
## between-cluster term plus the within-cluster (second-stage) term
##   sum_i (1 - m_i/M_i) M_i^2 s_i^2 / m_i / (n N Mbar^2),
## which the between-cluster sum of squares misses a fraction n/N of.
clu_draw <- function(s, n, p, a) {
  ci  <- sort(sample(s$H, n))
  Mis <- s$Mi[ci]
  mi  <- subsize(Mis, p)
  ii  <- unlist(lapply(seq_len(n), function(k) sample(s$idx[[ci[k]]], mi[k])))
  yl    <- split(pop[ii], rep(seq_len(n), mi))
  ybari <- sapply(yl, mean)
  si2   <- sapply(yl, function(v) if (length(v) > 1) var(v) else 0)
  that  <- Mis * ybari                       # t_i hat = M_i ybar_i
  Mbar  <- mean(Mis)
  m     <- sum(that) / sum(Mis)
  v1    <- (1 - n / s$H) * sum((that - m * Mis)^2) / (n - 1) / (n * Mbar^2)
  v2    <- sum((1 - mi / Mis) * Mis^2 * si2 / mi) / (n * s$H * Mbar^2)
  list(ci = ci, mi = mi, Mi = Mis, ii = ii, ybari = ybari,
       est = c(m, qt(a, df = n - 1) * sqrt(v1 + v2)))
}

## One simple random sample of counties, of the same size as the cluster
## sample just drawn, estimated by the plain sample average.
srs_draw <- function(nel, a) {
  ii  <- sample(Nel, nel)
  y   <- pop[ii]
  m   <- mean(y)
  se  <- sqrt(1 - nel / Nel) * sd(y) / sqrt(nel)
  list(ii = ii, est = c(m, qt(a, df = nel - 1) * se))
}

## ---- broken y-axis, shared by the left panel --------------------------------
## Drawn region [0, 2 * ybarU]: visible scale [0, cut] plus a compressed lane
## for the right tail, so the population mean sits at the midpoint of the panel.
lane_frac <- 0.10
cut       <- 2 * ybarU / (1 + lane_frac)
lane_w    <- lane_frac * cut
ymax_top  <- cut + lane_w

py      <- pop
tail_id <- which(pop > cut)
n_tail  <- length(tail_id)
if (n_tail > 0) {
  r <- rank(pop[tail_id], ties.method = "first") / (n_tail + 1)
  py[tail_id] <- cut + lane_w * (0.10 + 0.80 * r)
}

METHOD <- c("SRS average", "Cluster ratio")
MCOL   <- c("grey35", "#2ca02c")
MAR    <- c(5.0, 4.5, 3.0, 1.0)

## symmetric density polygon, drawn vertically about x = at
violin <- function(x, at, col, w = 0.30) {
  if (length(x) < 10 || diff(range(x)) <= 0) return(invisible())
  d  <- density(x)
  hw <- w * d$y / max(d$y)
  polygon(c(at + hw, rev(at - hw)), c(d$x, rev(d$x)),
          col = adjustcolor(col, 0.35), border = col)
  m <- median(x)
  segments(at - w * 0.6, m, at + w * 0.6, m, col = col, lwd = 2)
}

pct <- function(v) if (is.na(v)) "--" else sprintf("%.1f%%", 100 * v)
fmt <- function(x) format(round(x), big.mark = ",")

## ---- left panel: the population, cluster by cluster -------------------------
## clu is the last cluster draw, srs the last SRS draw (either may be NULL);
## j = 1 highlights the SRS sample, j = 2 the two-stage cluster sample.
panel_pop <- function(s, j, clu, srs, p, lab) {
  H <- s$H
  par(mar = MAR, yaxs = "i")
  plot(NA, xlim = c(0.3, H + 0.7), ylim = c(0, ymax_top),
       xlab = "", ylab = "acres92 (thousands)", xaxt = "n", yaxt = "n", bty = "n",
       main = sprintf("Population by %s, with the %s sample", lab, METHOD[j]))

  ## one band per cluster; the clusters that made it into the two-stage
  ## sample are shaded, which is the whole point of the design
  inS <- rep(FALSE, H)
  if (j == 2 && !is.null(clu)) inS[clu$ci] <- TRUE
  for (h in seq_len(H))
    rect(h - 0.38, 0, h + 0.38, ymax_top,
         col = if (inS[h]) "#fff0c2" else adjustcolor("grey92", 0.6), border = NA)

  points(s$x_pos, py, pch = 16, cex = 0.42, col = adjustcolor("grey55", 0.5))
  smp <- if (j == 1) srs else clu
  if (!is.null(smp))
    points(s$x_pos[smp$ii], py[smp$ii], pch = 16, cex = 0.55,
           col = adjustcolor("red", 0.85))

  ## sample mean(s): one bar per sampled cluster for the two-stage design
  ## (the ybar_i that build t_i hat), a single bar for SRS
  if (!is.null(smp)) {
    if (j == 1) {
      segments(0.3, srs$est[1], H + 0.7, srs$est[1], col = MCOL[1], lwd = 3)
    } else {
      segments(clu$ci - 0.38, clu$ybari, clu$ci + 0.38, clu$ybari,
               col = MCOL[2], lwd = 3)
      abline(h = clu$est[1], col = MCOL[2], lwd = 1.5, lty = 4)
    }
    leg <- if (j == 1)
             list(txt = c("Population mean", "Sample mean"),
                  col = c("black", MCOL[1]), lty = c(2, 1), lwd = c(2, 3))
           else
             list(txt = c("Population mean", "Cluster means (ybar_i)",
                          "Ratio estimate (ybar_r)"),
                  col = c("black", MCOL[2], MCOL[2]), lty = c(2, 1, 4),
                  lwd = c(2, 3, 1.5))
    legend("topleft", bty = "o", bg = "white", box.col = NA, cex = 0.75,
           inset = c(0.005, 0.01), legend = leg$txt,
           col = leg$col, lty = leg$lty, lwd = leg$lwd)
  }

  ## cluster labels: sampled clusters in red, with m_i / M_i when there is room
  cx <- if (H > 40) 0.45 else 0.6
  for (h in seq_len(H)) {
    mtext(s$lev[h], side = 1, at = h, line = 0.2, las = 2, cex = cx,
          col = if (inS[h]) "red" else "grey45", font = if (inS[h]) 2 else 1)
  }
  if (j == 2 && !is.null(clu) && length(clu$ci) <= 20)
    mtext(sprintf("%d/%d", clu$mi, clu$Mi), side = 1, at = clu$ci, line = 1.9,
          las = 2, cex = 0.5, col = "red")
  mtext(if (j == 2) sprintf("clusters (sampled ones shaded%s; p = %.0f%%)",
                            if (length(clu$ci) <= 20) ", labelled m_i/M_i" else "",
                            100 * p)
        else "clusters",
        side = 1, line = 3.6, cex = 0.72, col = "grey30")

  at <- pretty(c(0, cut), 6); at <- at[at <= cut]
  axis(2, at = at, labels = format(at / 1000, big.mark = ",", trim = TRUE), las = 1)
  segments(0.3, cut, H + 0.7, cut, col = "grey60", lty = 3)
  text(H + 0.6, cut + lane_w / 2, sprintf("compressed: > %s", fmt(cut / 1000)),
       adj = c(1, 0), col = "grey40", cex = 0.7)
  abline(h = ybarU, col = "black", lwd = 2, lty = 2)
}

## ---- right panel: the two sampling distributions ----------------------------
panel_dist <- function(s, rv, est_range, th, n, p, conf, nrep) {
  par(mar = MAR, yaxs = "i")
  plot(NA, xlim = c(0.4, 2.6), ylim = est_range, xaxt = "n", yaxt = "n", bty = "n",
       xlab = "", ylab = "Estimated mean of acres92 (thousands)",
       main = "Sampling distributions of the two estimators")
  abline(h = ybarU, col = "black", lwd = 2, lty = 2)
  axis(1, at = 1:2, labels = METHOD, tick = FALSE, line = -0.5)
  at_r <- pretty(est_range, 6)
  at_r <- at_r[at_r >= est_range[1] & at_r <= est_range[2]]
  axis(2, at = at_r, labels = format(at_r / 1000, big.mark = ",", trim = TRUE), las = 1)

  if (is.null(rv$M)) {
    text(1.5, ybarU, "Click 'Draw new samples' or Play to begin", col = "grey50")
    return(invisible())
  }

  k <- NROW(rv$M)
  for (jj in 1:2) {
    violin(rv$M[, jj], at = jj, col = MCOL[jj])

    ci_col <- if (rv$HIT[k, jj]) "forestgreen" else "red"
    xb     <- jj + 0.40
    segments(xb, rv$est[jj, 2], xb, rv$est[jj, 3], col = ci_col, lwd = 3)
    segments(xb - 0.06, rv$est[jj, 2:3], xb + 0.06, rv$est[jj, 2:3], col = ci_col, lwd = 3)
    points(xb, rv$est[jj, 1], pch = 19, cex = 1.1, col = ci_col)
  }

  legend("topleft", bty = "n", cex = 0.85, inset = c(0.02, 0.01),
         title = sprintf("%.0f%% CI coverage", 100 * conf), fill = MCOL, border = NA,
         legend = paste(METHOD, vapply(1:2, function(jj) pct(mean(rv$HIT[, jj])), "")))

  ## design effect: variance of the cluster estimator relative to SRS at the
  ## same sample size, simulated so far and anticipated from the population
  if (k > 1) {
    deff <- var(rv$M[, 2]) / var(rv$M[, 1])
    legend("bottomleft", bty = "n", cex = 0.85, inset = c(0.02, 0.01),
           title = "Variance of cluster ratio vs. SRS", fill = MCOL[2], border = NA,
           text.col = if (deff > 1) "red" else "black",
           legend = sprintf("deff = %.2f  (theory %.2f)", deff, th$Vc / th$Vs))
  }

  outside <- sum(rv$M < est_range[1] | rv$M > est_range[2])
  sub <- sprintf("n = %d clusters, %s counties  |  R2 between clusters = %s  |  %d of %d draws%s",
                 n, fmt(rv$nel), pct(s$R2), k, nrep,
                 if (outside > 0) sprintf("  |  %d off axis", outside) else "")
  mtext(sub, side = 3, line = -0.2, cex = 0.7, col = "grey30")
}

## ---- inline SVG button icons (no font or icon-library dependency) ----------
svg_icon <- function(paths)
  HTML(sprintf(paste0('<svg width="16" height="16" viewBox="0 0 16 16" ',
                      'fill="currentColor" style="vertical-align:-2px">%s</svg>'), paths))
icon_play  <- svg_icon('<path d="M3 2v12l10-6z"/>')
icon_pause <- svg_icon('<rect x="3" y="2" width="4" height="12"/><rect x="9" y="2" width="4" height="12"/>')
icon_skip  <- svg_icon('<path d="M1 2v12l7-6zM8 2v12l6-6z"/><rect x="14" y="2" width="2" height="12"/>')

ui <- fluidPage(
  titlePanel("Shinylive Simulation for Two-Stage Cluster Sampling"),
  tags$style(HTML(
    ".well {padding: 10px 12px; margin-bottom: 10px;}
     .form-group {margin-bottom: 6px;}
     .radio {margin-top: 2px; margin-bottom: 2px;}
     .irs {margin-bottom: 0;}
     .btn {margin-bottom: 6px;}"
  )),
  wellPanel(
    fluidRow(
      column(
        3,
        radioButtons("clustvar", "Cluster variable", choices = CLUST_CHOICES,
                     selected = "state"),
        radioButtons("show", "Highlight sample from", choiceNames = METHOD,
                     choiceValues = as.character(1:2), selected = "2")
      ),
      column(
        3,
        sliderInput("n", "Number of clusters drawn (n)", min = 2, max = 50,
                    value = 10, step = 1),
        sliderInput("p", "Percentage sampled per cluster (p = m_i/M_i)",
                    min = 0.05, max = 1, value = 0.30, step = 0.05)
      ),
      column(
        3,
        sliderInput("conf", "Confidence level", min = 0.80, max = 0.99,
                    value = 0.95, step = 0.01),
        sliderInput("speed", "Samples per second", min = 1, max = 10, value = 3, step = 1)
      ),
      column(
        3,
        numericInput("nrep", "Stop after this many samples",
                     value = 500, min = 10, max = 20000, step = 50),
        actionButton("draw", "Draw new samples", class = "btn-primary", width = "100%"),
        br(), br(),
        fluidRow(
          column(4, actionButton("play",  icon_play,  class = "btn-success", width = "100%",
                                 title = "Play")),
          column(4, actionButton("pause", icon_pause, class = "btn-warning", width = "100%",
                                 title = "Pause")),
          column(4, actionButton("skip",  icon_skip,  class = "btn-info",    width = "100%",
                                 title = "Skip to the end"))
        ),
        actionButton("reset", "Reset", width = "100%")
      )
    )
  ),
  plotOutput("mainplot", height = "530px")
)

server <- function(input, output, session) {

  rv <- reactiveValues(
    clu = NULL,   # last two-stage cluster draw
    srs = NULL,   # last simple random sample
    est = NULL,   # matrix 2 x 3: estimate, lower, upper (last draw)
    nel = NA,     # number of counties in the last sample
    M   = NULL,   # accumulated estimates, one column per design
    HIT = NULL    # accumulated coverage indicators
  )
  running <- reactiveVal(FALSE)

  nrep <- reactive({
    v <- input$nrep
    if (is.null(v) || is.na(v) || v < 1) 1L else min(as.integer(v), 20000L)
  })

  si     <- reactive(make_clusters(input$clustvar))
  clab   <- reactive(names(CLUST_CHOICES)[CLUST_CHOICES == input$clustvar])
  th     <- reactive(theory(si(), input$n, input$p))

  ## a window wide enough for whichever design is the more variable
  est_range <- reactive({
    t0 <- th()
    ybarU + c(-1, 1) * 4 * sqrt(max(t0$Vc, t0$Vs))
  })

  ## draw k replications; each one draws a two-stage cluster sample and an SRS
  ## of the same realised size. Only the last replication's sampled units are
  ## kept, for the left panel; "Skip" calls this with the full remaining count.
  draw_many <- function(k) {
    n    <- isolate(input$n)
    p    <- isolate(input$p)
    conf <- isolate(input$conf)
    a    <- 1 - (1 - conf) / 2
    s    <- isolate(si())

    Mnew   <- matrix(0, k, 2)
    HITnew <- matrix(FALSE, k, 2)

    for (i in seq_len(k)) {
      cl <- clu_draw(s, n, p, a)
      sr <- srs_draw(length(cl$ii), a)

      est <- rbind(c(sr$est[1], sr$est[1] - sr$est[2], sr$est[1] + sr$est[2]),
                   c(cl$est[1], cl$est[1] - cl$est[2], cl$est[1] + cl$est[2]))

      Mnew[i, ]   <- est[, 1]
      HITnew[i, ] <- est[, 2] <= ybarU & ybarU <= est[, 3]

      if (i == k) {
        rv$clu <- cl; rv$srs <- sr; rv$est <- est; rv$nel <- length(cl$ii)
      }
    }
    rv$M   <- rbind(rv$M,   Mnew)
    rv$HIT <- rbind(rv$HIT, HITnew)
  }
  draw_one <- function() draw_many(1)

  observeEvent(input$draw,  { running(FALSE); draw_one() })
  observeEvent(input$play,  running(TRUE))
  observeEvent(input$pause, running(FALSE))

  observeEvent(input$skip, {
    running(FALSE)
    k <- nrep() - NROW(rv$M)
    if (k > 0) draw_many(k)
  })

  observe({
    if (!isTRUE(running())) return()
    if (NROW(isolate(rv$M)) >= isolate(nrep())) { running(FALSE); return() }
    invalidateLater(1000 / isolate(input$speed), session)
    isolate(draw_one())
  })

  reset_history <- function() {
    running(FALSE)
    rv$clu <- NULL; rv$srs <- NULL; rv$est <- NULL; rv$nel <- NA
    rv$M <- NULL; rv$HIT <- NULL
  }
  observeEvent(input$reset,    reset_history())
  observeEvent(input$n,        reset_history())
  observeEvent(input$p,        reset_history())
  observeEvent(input$conf,     reset_history())
  observeEvent(input$clustvar, reset_history())

  output$mainplot <- renderPlot({
    layout(matrix(1:2, nrow = 1), widths = c(1.3, 1))
    panel_pop(si(), as.integer(input$show), rv$clu, rv$srs, input$p, clab())
    panel_dist(si(), rv, est_range(), th(), input$n, input$p, input$conf, nrep())
  })
}

shinyApp(ui, server)
```

**Shinylive App Comparing Two-Stage Cluster Sampling (ratio estimator) with SRS**

:::
