# Cluster Sampling

## Estimation for Cluster Sampling

Cluster sampling divides the population into $N$ clusters, out of which a simple random sample of $n$ clusters is selected. This design is highly practical when a sampling frame of individual elements is unavailable, or when individuals are geographically grouped.

#### 1. One-Stage Cluster Sampling {-}
In a one-stage design, every element $M_i$ within the sampled cluster $i$ is surveyed. 

*   **Total in Cluster $i$:** $t_i = \sum_{j=1}^{M_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:** The ratio estimator leverages cluster sizes to estimate the overall element mean:
    $$
\overline{y}_r = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n M_i}.
$$

*   **Estimated Variance of the Mean:**
    $$
\hat{V}(\overline{y}_r) = \left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2} \frac{\sum_{i=1}^n (t_i - \overline{y}_r M_i)^2}{n-1}.
$$

#### 2. Two-Stage Cluster Sampling {-}
In a two-stage design, we first sample $n$ clusters, then take a subsample of $m_i$ elements from the $M_i$ elements within each selected cluster. Because we don't observe the true cluster total $t_i$, we estimate it:

*   **Estimated Total in Cluster $i$:** $\hat{t}_i = M_i \overline{y}_i$, where $\overline{y}_i = \frac{1}{m_i} \sum_{j=1}^{m_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:**
    $$
\overline{y}_r = \frac{\sum_{i=1}^n \hat{t}_i}{\sum_{i=1}^n M_i}
$$

*   **Estimated Variance of the Mean:** it splits into two recognisable pieces
    --- the variance of a **ratio** estimate computed over the $n$ sampled
    clusters, plus the cluster sampling fraction $n/N$ times the variance of a
    **stratified** sample in which those same clusters are the strata:
    $$
\hat{V}(\overline{y}_r) =
\underbrace{\left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2}
  \frac{\sum_{i=1}^n (\hat{t}_i - \overline{y}_r M_i)^2}{n-1}}_{\textstyle \hat{V}_{\mathrm{ratio}}}
\;+\;
\frac{n}{N}
\underbrace{\sum_{i=1}^n \left(\frac{M_i}{M}\right)^{2}
  \left(1 - \frac{m_i}{M_i}\right)\frac{s_i^2}{m_i}}_{\textstyle \hat{V}_{\mathrm{str}}},
\qquad M = \sum_{i=1}^n M_i = n\overline{M} .
$$
    Here $s_i^2$ is the sample variance of the $m_i$ elements measured in
    cluster $i$, and $M$ is the number of elements the sampled clusters
    contain. $\hat{V}_{\mathrm{ratio}}$ is exactly the ratio-estimate variance
    for $\sum \hat{t}_i / \sum M_i$, and $\hat{V}_{\mathrm{str}}$ is exactly the
    stratified variance $\sum_h W_h^2(1 - n_h/N_h)s_h^2/n_h$ with the sampled
    clusters as strata: stratum sizes $N_h = M_i$, sample sizes $n_h = m_i$, and
    weights $W_h = M_i/M$. Each term
    $$\hat{v}_i = \left(\frac{M_i}{M}\right)^{2}\left(1 - \frac{m_i}{M_i}\right)\frac{s_i^2}{m_i}$$
    is one stratum's contribution; the working tables below carry them in a
    $\hat{v}_i$ column whose Sum is $\hat{V}_{\mathrm{str}}$.

Why is the second piece there at all? If we could see the true cluster totals
$t_i$, the ratio piece alone would be the whole variance. Sub-sampling replaces
$t_i$ by the estimate $\hat{t}_i$, and the between-cluster sum of squares
absorbs only the fraction $1 - n/N$ of the noise that creates; the stratified
piece pays the rest. Three special cases follow straight from the formula:

*   **One-stage sampling** ($m_i = M_i$): every $\hat{v}_i = 0$, and only the
    ratio piece survives.
*   **$N$ unknown or effectively infinite**: $n/N \to 0$, the stratified piece
    drops out, and $s_e^2$ alone does the whole job.
*   **A census of clusters** ($n = N$): the factor $1 - n/N$ kills the ratio
    piece and $n/N = 1$ leaves $\hat{V}(\overline{y}_r) = \hat{V}_{\mathrm{str}}$
    exactly --- drawing every cluster *is* stratified sampling, and the formula
    says so.

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `cluster_ratio`, used below for one-stage and two-stage cluster samples) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Estimating for One-stage cluster sampling: analysis of algebra.csv

Consider a population of 187 high school algebra classes in a city. An investigator takes an SRS of 12 of those classes (the clusters). Because this is a one-stage sample, every student in the 12 sampled classes takes the test to evaluate their function knowledge.

We read the raw data and display it using `gt`'s interactive pagination feature to keep the document clean.

```{r tbl-algebra-raw-data}
#| tbl-cap: "Raw data: algebra classes (sample)"
algebra <- read.csv("data/algebra.csv")

algebra |>
  gt() |>
  tab_header(title = "Raw Data: Algebra Classes") |>
  cols_label(Mi = html("M<sub>i</sub>")) |>
  opt_interactive(use_pagination = TRUE, page_size_default = 10)
```
**Output Note:** The table above displays the raw, unaggregated survey data. The interactive pagination allows you to flip through the dataset compactly.

Next, we run our updated function to compute both the cluster-level working table and the overall ratio estimate for the average student score.

```{r algebra-cluster-ratio}
alg_res <- cluster_ratio(data = algebra,
                             cname = "class",
                             csize = "Mi",
                             yvar = "score",
                             N = 187)

# Display the cluster-level working table
alg_res$table
```
**Output Note:** The working table above displays, for each of the 12 randomly selected classes, the class size ($M_i$), the total class score ($t_i$), the fitted value $\hat B M_i$, and the residual and squared residual used to estimate $s_e^2$.

Finally, we extract the ratio estimate from the function output list to assess the overall population mean.

```{r algebra-ratio-mean}
# Display the overall ratio estimate
alg_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** Using the ratio estimator, the estimated mean score for a student across all 187 algebra classes is `r round(alg_res$estimate["Est."], 2)`, with a 95% confidence interval of [`r round(alg_res$estimate["ci.low"], 2)`, `r round(alg_res$estimate["ci.upp"], 2)`]. The standard error accounts for the clustering effect, which typically increases variance compared to a pure SRS of students.

## Estimating for two-stage cluster sampling: analysis of coots.csv

In two-stage cluster sampling, we sample clusters, and then subsample elements within them. In this dataset, researcher Arnold investigated coot eggs. He first selected a random sample of clutches (the clusters), and then measured the volume of a random subsample of eggs within each selected clutch.


```{r tbl-coots-raw-data}
#| tbl-cap: "Raw data: coot egg clutches (sample)"
coots <- read.csv("data/coots.csv")

coots |>
  gt() |>
  tab_header(title = "Raw Data: Coots") |>
  tab_options(
    container.height = px(400), # Adjust height as needed
    container.overflow.y = TRUE # Forces the vertical scroll bar
  )
```


We apply the exact same ratio estimator formulation, though mathematically, the cluster total $t_i$ is now an *estimated* total $\hat{t}_i$.

```{r coots-cluster-ratio}
coots_res <- cluster_ratio(data = coots,
                               cname = "clutch",
                               csize = "csize",
                               yvar = "volume")

coots_res$table
```
**Output Note:** Unlike the previous example, we only measured a subset of eggs per clutch, and the working table now carries an $m_i$ column recording how many. The function automatically estimates the total volume of each clutch ($\hat{t}_i$) by multiplying the true clutch size ($M_i$) by the observed subsample mean ($\overline{y}_i$); the working table (truncated to the first 6 rows + Sum) shows these estimated totals alongside the ratio fit and residuals. No within-cluster variance column appears here: with $N$ unknown (`N = Inf`), the factor $1/(nN)$ in front of the second-stage term is zero, and $s_e^2$ already carries all of the sub-sampling noise on its own.

```{r coots-ratio-mean}
# Display the overall ratio estimate
coots_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimated overall average egg volume is `r round(coots_res$estimate["Est."], 3)` units. Because the total number of clutches in the population ($N$) is not explicitly provided (set to `Inf` by default), the finite population correction factor is safely omitted from the standard error calculation.

## Two-stage cluster sampling with a finite $N$ for Estimating Agriculture Acreages for Farming

Neither example above actually exercises the within-cluster term --- `algebra`
is one-stage, and `coots` has an unknown $N$. To see it at work we need a
population whose clusters we can count. The `agpop.csv` file provides one: its
3,059 counties (dropping those coded `-99`) sit inside exactly $N = 50$ states,
so the states are clusters of known number and known sizes $M_i$.

### Drawing the sample

We take a two-stage sample in the obvious way: an SRS of $n = 10$ states,
followed by an SRS of $m_i = 30\%$ of the counties in each selected state.

```{r agpop-two-stage-sample}
agpop <- acres_in_thousands(read.csv("data/agpop.csv"))
agpop <- agpop[!is.na(agpop$acres92), ]

## cluster sizes M_i: the number of counties in each state
Mi_all <- table(agpop$state)
agpop$Mi <- as.integer(Mi_all[agpop$state])

set.seed(2026)
states <- sample(names(Mi_all), 10)                 # stage 1: n = 10 clusters
agclus <- do.call(rbind, lapply(states, function(st) {
  rows <- which(agpop$state == st)                  # stage 2: m_i = 0.3 M_i
  agpop[sample(rows, max(1, round(0.30 * length(rows)))), ]
}))

c(clusters = length(states), counties = nrow(agclus))
```

### Computing the estimate directly

Everything the estimator needs is a cluster-level summary: the known size $M_i$
of each sampled state, the number of counties actually measured in it ($m_i$),
their mean $\overline{y}_i$ and variance $s_i^2$, and the estimated state total
$\hat{t}_i = M_i \overline{y}_i$.

```{r tbl-agpop-cluster-summary}
#| tbl-cap: "Cluster-level summaries for the 10 sampled states"
n  <- length(states)                                            # clusters drawn
N  <- 50                                                        # clusters in all

Mi    <- tapply(agclus$Mi,      agclus$state, function(x) x[1])  # cluster sizes
mi    <- tapply(agclus$acres92, agclus$state, length)            # counties measured
ybari <- tapply(agclus$acres92, agclus$state, mean)              # subsample means
si2   <- tapply(agclus$acres92, agclus$state, var)               # within-cluster vars
thati <- Mi * ybari                                              # estimated totals

clus_summary <- data.frame(Mi, mi, ybari, si2, thati)

clus_summary |>
  gt(rownames_to_stub = TRUE) |>
  cols_label(Mi    = md("$M_i$"),    mi  = md("$m_i$"),
             ybari = md("$\\bar y_i$"), si2 = md("$s_i^2$"),
             thati = md("$\\hat t_i$")) |>
  fmt_auto(data = clus_summary, columns = c("ybari", "si2", "thati"))
```
**Output Note:** The states drawn range from `r min(Mi)` to `r max(Mi)` counties,
of which `r min(mi)` to `r max(mi)` were visited. $s_i^2$ is large throughout:
county acreages vary enormously even inside a single state, and that is the
variability the second-stage term will account for.

The ratio estimate itself divides the estimated totals by the cluster sizes.

```{r agpop-ratio-estimate}
Mbar  <- mean(Mi)                  # average size of the sampled clusters
ybarr <- sum(thati) / sum(Mi)      # ybar_r = sum(t_i hat) / sum(M_i)

c(Mbar = Mbar, ybar_r = ybarr)
```

Its variance has the two pieces from the formula above. The **ratio** piece
measures how the estimated totals scatter about the fitted line
$\overline{y}_r M_i$:

```{r agpop-variance-ratio}
## ratio piece: the variance of the ratio estimate over the 10 sampled states
ei  <- thati - ybarr * Mi          # ratio residuals
se2 <- sum(ei^2) / (n - 1)         # s_e^2
V_ratio <- (1 - n / N) * se2 / (n * Mbar^2)

c(s_e2 = se2, V_ratio = V_ratio)
```

The **stratified** piece treats those same 10 states as strata, of sizes $M_i$
inside the $M = \sum_i M_i$ counties they contain, and is then scaled by the
cluster sampling fraction $n/N$:

```{r agpop-variance-str}
## stratified piece: strata = the sampled states, weights W_i = M_i / M
M     <- sum(Mi)                                  # elements in the 10 states
vi    <- (Mi / M)^2 * (1 - mi / Mi) * si2 / mi    # each stratum's contribution
V_str <- sum(vi)                                  # the stratified variance

c(M = M, n_Mbar = n * Mbar, V_str = V_str, scaled = (n / N) * V_str)
```

Adding the two gives the standard error, and the interval uses $n - 1$ degrees
of freedom --- the clusters, not the counties, are the units that were
randomised.

```{r agpop-se-ci}
SE  <- sqrt(V_ratio + (n / N) * V_str)
mem <- qt(0.975, df = n - 1) * SE

c(Est. = ybarr, S.E. = SE, ci.low = ybarr - mem, ci.upp = ybarr + mem)
```
**Output Note:** The estimated mean 1992 acreage per county is
`r round(ybarr, 1)` thousand acres, with a 95% confidence interval
of [`r round(ybarr - mem, 1)`, `r round(ybarr + mem, 1)`] thousand acres. The
interval is wide: only 10
states were visited, so $n - 1 = 9$ degrees of freedom is all the between-cluster
scatter provides.

### Using the function `cluster_ratio`

The function performs exactly these steps, provided $N$ is supplied.

```{r agpop-cluster-ratio}
clus_res <- cluster_ratio(data = agclus,
                          cname = "state",
                          csize = "Mi",
                          yvar = "acres92",
                          N = 50)

clus_res$table
```
**Output Note:** Because this sample is two-stage *and* $N$ is finite, the
working table carries the three extra columns computed above: $m_i$, the number
of counties measured in each state; $s_i^2$, the variance among them; and
$\hat v_i = (M_i/M)^2 (1 - m_i/M_i) s_i^2/m_i$, that state's contribution to the
stratified variance. The Sum of the $\hat v_i$ column is
$\hat{V}_{\mathrm{str}}$ itself, and the source note under the table spells the
standard error out as $\hat{V}_{\mathrm{ratio}} + (n/N)\hat{V}_{\mathrm{str}}$.

```{r agpop-cluster-estimate}
clus_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimate, standard error, and interval reproduce the direct
computation above.

## Interactive Demonstration: Clustering by State or at Random {#sec-app-cluster}

The simulation for this section runs as a standalone app:
[Two-Stage Cluster Sampling (ratio estimator) with SRS](app-cluster.html){target="_blank"}.
It opens in a new tab, together with its documentation and the full
[list of apps](shinyliveapps_sampling.html){target="_blank"}.

