# Cluster Sampling

## Key Formulas for Cluster Sampling

Cluster sampling divides the population into $N$ clusters, out of which a simple random sample of $n$ clusters is selected. This design is highly practical when a sampling frame of individual elements is unavailable, or when individuals are geographically grouped.

#### 1. One-Stage Cluster Sampling {-}
In a one-stage design, every element $M_i$ within the sampled cluster $i$ is surveyed. 

*   **Total in Cluster $i$:** $t_i = \sum_{j=1}^{M_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:** The ratio estimator leverages cluster sizes to estimate the overall element mean:
    $$
\overline{y}_r = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n M_i}.
$$

*   **Estimated Variance of the Mean:**
    $$
\hat{V}(\overline{y}_r) = \left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2} \frac{\sum_{i=1}^n (t_i - \overline{y}_r M_i)^2}{n-1}.
$$

#### 2. Two-Stage Cluster Sampling {-}
In a two-stage design, we first sample $n$ clusters, then take a subsample of $m_i$ elements from the $M_i$ elements within each selected cluster. Because we don't observe the true cluster total $t_i$, we estimate it:

*   **Estimated Total in Cluster $i$:** $\hat{t}_i = M_i \overline{y}_i$, where $\overline{y}_i = \frac{1}{m_i} \sum_{j=1}^{m_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:**
    $$
\overline{y}_r = \frac{\sum_{i=1}^n \hat{t}_i}{\sum_{i=1}^n M_i}
$$

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `cluster_ratio`, used below for one-stage and two-stage cluster samples) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Estimating for One-stage cluster sampling: analysis of algebra.csv

Consider a population of 187 high school algebra classes in a city. An investigator takes an SRS of 12 of those classes (the clusters). Because this is a one-stage sample, every student in the 12 sampled classes takes the test to evaluate their function knowledge.

We read the raw data and display it using `gt`'s interactive pagination feature to keep the document clean.

```{r tbl-algebra-raw-data}
#| tbl-cap: "Raw data: algebra classes (sample)"
algebra <- read.csv("data/algebra.csv")

algebra |>
  gt() |>
  tab_header(title = "Raw Data: Algebra Classes") |>
  cols_label(Mi = html("M<sub>i</sub>")) |>
  opt_interactive(use_pagination = TRUE, page_size_default = 10)
```
**Output Note:** The table above displays the raw, unaggregated survey data. The interactive pagination allows you to flip through the dataset compactly.

Next, we run our updated function to compute both the cluster-level working table and the overall ratio estimate for the average student score.

```{r algebra-cluster-ratio}
alg_res <- cluster_ratio(data = algebra,
                             cname = "class",
                             csize = "Mi",
                             yvar = "score",
                             N = 187)

# Display the cluster-level working table
alg_res$table
```
**Output Note:** The working table above displays, for each of the 12 randomly selected classes, the class size ($M_i$), the total class score ($t_i$), the fitted value $\hat B M_i$, and the residual and squared residual used to estimate $s_e^2$.

Finally, we extract the ratio estimate from the function output list to assess the overall population mean.

```{r algebra-ratio-mean}
# Display the overall ratio estimate
alg_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** Using the ratio estimator, the estimated mean score for a student across all 187 algebra classes is `r round(alg_res$estimate["Est."], 2)`, with a 95% confidence interval of [`r round(alg_res$estimate["ci.low"], 2)`, `r round(alg_res$estimate["ci.upp"], 2)`]. The standard error accounts for the clustering effect, which typically increases variance compared to a pure SRS of students.

## Estimating for two-stage cluster sampling: analysis of coots.csv

In two-stage cluster sampling, we sample clusters, and then subsample elements within them. In this dataset, researcher Arnold investigated coot eggs. He first selected a random sample of clutches (the clusters), and then measured the volume of a random subsample of eggs within each selected clutch.


```{r tbl-coots-raw-data}
#| tbl-cap: "Raw data: coot egg clutches (sample)"
coots <- read.csv("data/coots.csv")

coots |>
  gt() |>
  tab_header(title = "Raw Data: Coots") |>
  tab_options(
    container.height = px(400), # Adjust height as needed
    container.overflow.y = TRUE # Forces the vertical scroll bar
  )
```


We apply the exact same ratio estimator formulation, though mathematically, the cluster total $t_i$ is now an *estimated* total $\hat{t}_i$.

```{r coots-cluster-ratio}
coots_res <- cluster_ratio(data = coots,
                               cname = "clutch",
                               csize = "csize",
                               yvar = "volume")

coots_res$table
```
**Output Note:** Unlike the previous example, we only measured a subset of eggs per clutch. The function automatically estimates the total volume of each clutch ($\hat{t}_i$) by multiplying the true clutch size ($M_i$) by the observed subsample mean ($\overline{y}_i$); the working table (truncated to the first 6 rows + Sum) shows these estimated totals alongside the ratio fit and residuals.

```{r coots-ratio-mean}
# Display the overall ratio estimate
coots_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimated overall average egg volume is `r round(coots_res$estimate["Est."], 3)` units. Because the total number of clutches in the population ($N$) is not explicitly provided (set to `Inf` by default), the finite population correction factor is safely omitted from the standard error calculation.