# Cluster Sampling

## Key Formulas for Cluster Sampling

Cluster sampling divides the population into $N$ clusters, out of which a simple random sample of $n$ clusters is selected. This design is highly practical when a sampling frame of individual elements is unavailable, or when individuals are geographically grouped.

#### 1. One-Stage Cluster Sampling {-}
In a one-stage design, every element $M_i$ within the sampled cluster $i$ is surveyed. 

*   **Total in Cluster $i$:** $t_i = \sum_{j=1}^{M_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:** The ratio estimator leverages cluster sizes to estimate the overall element mean:
    $$
\overline{y}_r = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n M_i}.
$$

*   **Estimated Variance of the Mean:**
    $$
\hat{V}(\overline{y}_r) = \left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2} \frac{\sum_{i=1}^n (t_i - \overline{y}_r M_i)^2}{n-1}.
$$

#### 2. Two-Stage Cluster Sampling {-}
In a two-stage design, we first sample $n$ clusters, then take a subsample of $m_i$ elements from the $M_i$ elements within each selected cluster. Because we don't observe the true cluster total $t_i$, we estimate it:

*   **Estimated Total in Cluster $i$:** $\hat{t}_i = M_i \overline{y}_i$, where $\overline{y}_i = \frac{1}{m_i} \sum_{j=1}^{m_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:**
    $$
\overline{y}_r = \frac{\sum_{i=1}^n \hat{t}_i}{\sum_{i=1}^n M_i}
$$

---

## Functions and packages for Analyzing Data

The updated `cluster_ratio_est` function analyzes both one-stage and two-stage cluster samples. It computes the cluster-level summary statistics and returns them in a list alongside the final ratio estimation output. The robust `format_est_gt` function guarantees cleanly formatted mathematical headers for the final output.

```{r}
#| code-fold: true
library(latex2exp)
library(gt)

## Formats a gt table column as ordinary fixed-point, UNLESS its values are
## too small (< `small` in absolute value) or too large (>= `large`), in
## which case it switches that column to scientific notation instead.
fmt_auto <- function (gt_tbl, data, columns, decimals = 3, small = 1e-3, large = 1e5)
{
    for (col in columns)
    {
        x <- data[[col]]
        x <- x[is.finite (x) & x != 0]
        if (length (x) == 0) next
        if (any (abs (x) < small) || any (abs (x) >= large))
            gt_tbl <- fmt_scientific (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
        else
            gt_tbl <- fmt_number (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
    }
    gt_tbl
}

## Truncates a row-level working data.frame whose LAST row is a summary row
## to the first head_n data rows + "..." + the summary row.
truncate_working <- function (working, id_col, head_n = 6)
{
    n <- nrow (working) - 1
    if (n <= head_n + 1) return (working)

    dots <- working[1, ]
    dots[1, ] <- NA
    dots[1, id_col] <- "..."

    rbind (working[seq_len (head_n), ], dots, working[nrow (working), ])
}

## ydata --- observations of the variable of interest
## xdata --- observations of the auxiliary variable
## xbarU --- population mean of the auxiliary variable; required when
##           estimate is "mean" or "total"; ignored when estimate = "model"
## N --- population size
## estimate --- "mean" (default) returns the ratio estimate of the
##              population MEAN of y, Bhat*xbarU; "total" returns the ratio
##              estimate of the population TOTAL of y (requires finite N);
##              "model" returns the raw ratio coefficient Bhat = ybar/xbar
##              itself (with its own SE)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values y_i, x_i, fitted yhat_i =
##                  Bhat*x_i, residual e_i, and e_i^2, with a Sum row and a
##                  footnote showing Bhat and s_e^2; NULL otherwise
## col_labels --- named list (y, x, yhat) of RAW latex for those
##                working-table columns, so this function can be reused
##                (e.g. by cluster_ratio_est()) with cluster-appropriate labels
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_ratio_est <- function (ydata, xdata, xbarU = NULL, N = Inf,
                            estimate = c ("mean", "total", "model"), show.details = TRUE,
                            col_labels = list (y = "y_i", x = "x_i", yhat = "\\hat y_i"))
{
  estimate <- match.arg (estimate)

  n <- length (xdata)
  xbar <- mean (xdata)
  ybar <- mean (ydata)
  B_hat <- ybar / xbar
  yhat <- B_hat * xdata
  e <- ydata - yhat
  var_e <- sum (e^2) / (n - 1)
  sd_B_hat <- sqrt ((1 - n/N) * var_e / n) / xbar

  if (estimate == "model")
  {
      est <- B_hat
      sd_est <- sd_B_hat
  }
  else
  {
      if (is.null (xbarU))
          stop ("xbarU (population mean of x) is required when estimate is \"mean\" or \"total\"")
      if (estimate == "total")
      {
          if (!is.finite (N))
              stop ("N must be finite to compute estimate = \"total\"")
          est <- B_hat * xbarU * N
          sd_est <- sd_B_hat * xbarU * N
      }
      else
      {
          est <- B_hat * xbarU
          sd_est <- sd_B_hat * xbarU
      }
  }

  mem <- qt (0.975, df = n - 1) * sd_est
  estimate_vec <- c (Est. = est, S.E. = sd_est, ci.low = est - mem, ci.upp = est + mem)

  if (!show.details)
      return (list (estimate = estimate_vec, table = NULL))

  working <- data.frame (
      i    = c (seq_len (n), "Sum"),
      y    = c (ydata, sum (ydata)),
      x    = c (xdata, sum (xdata)),
      yhat = c (yhat, sum (yhat)),
      e    = c (e, sum (e)),
      e2   = c (e^2, sum (e^2))
  )
  working <- truncate_working (working, id_col = "i")

  table <- gt (working) |>
      tab_header (title = "Ratio Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
      cols_label (
          i    = md ("$i$"),
          y    = md (sprintf ("$%s$", col_labels$y)),
          x    = md (sprintf ("$%s$", col_labels$x)),
          yhat = md (sprintf ("$%s$", col_labels$yhat)),
          e    = md ("$e_i$"),
          e2   = md ("$e_i^2$")
      ) |>
      cols_width (i ~ px (40), everything () ~ px (100)) |>
      fmt_auto (data = working, columns = c ("y", "x", "yhat", "e", "e2"), decimals = 3) |>
      sub_missing (missing_text = "...") |>
      tab_style (
          style     = cell_text (weight = "bold"),
          locations = cells_body (rows = i == "Sum")
      ) |>
      tab_footnote (
          footnote  = md (sprintf ("$\\hat B = \\sum_i %s / \\sum_i %s = %.4f$, $%s = \\hat B\\, %s$, $s_e^2 = %.4f$.",
                                    col_labels$y, col_labels$x, B_hat, col_labels$yhat, col_labels$x, var_e)),
          locations = cells_column_labels (columns = yhat)
      )

  list (estimate = estimate_vec, table = table)
}

## Cluster ratio estimation (one-stage if Mi = mi, i.e. every element in the
## sampled cluster is measured; two-stage/ratio-to-size if only mi < Mi
## elements are subsampled, so t_hat_i = Mi * ybari is an ESTIMATED total).
## data --- original data frame for holding data
## cname --- variable recording cluster (psu) identity
## csize --- variable recording cluster (psu) population size (not sample size)
## yvar --- variable of interest
## N --- total number of clusters (psus) in the population
## estimate --- "mean" (default) or "model" both return the raw ratio
##              Bhat = sum(t_hat_i)/sum(Mi) --- for this ratio-to-size
##              cluster estimator, that raw B IS the population MEAN per
##              element, so the two coincide; "total" returns the
##              population TOTAL of y, Bhat*Mtotal_U (requires Mtotal_U,
##              the population total of the cluster-size variable)
## Mtotal_U --- population total of the cluster-size variable; required
##              when estimate = "total"
## show.details --- if TRUE, also return a gt table of the cluster-level
##                  working values (t_hat_i, M_i, fitted, e_i, e_i^2), via
##                  srs_ratio_est()'s own working table
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
cluster_ratio_est <- function (data, cname, csize, yvar, N = Inf,
                                estimate = c ("mean", "total", "model"),
                                show.details = TRUE, Mtotal_U = NULL)
{
  estimate <- match.arg (estimate)

  clust <- data[, cname]
  ydata <- data[, yvar]

  ybari <- tapply (ydata, clust, mean)
  Mi    <- tapply (data[, csize], clust, function (x) x[1])
  ## same as the cluster total if all Mi elements were measured (mi = Mi)
  t_hat_cls <- ybari * Mi

  if (estimate == "total")
  {
      if (is.null (Mtotal_U))
          stop ("Mtotal_U (population total of the cluster-size variable) is required when estimate = \"total\"")
      srs_ratio_est (t_hat_cls, Mi, xbarU = Mtotal_U, N = N, estimate = "mean", show.details = show.details,
                     col_labels = list (y = "\\hat t_i", x = "M_i", yhat = "\\hat B M_i"))
  }
  else
  {
      srs_ratio_est (t_hat_cls, Mi, N = N, estimate = "model", show.details = show.details,
                     col_labels = list (y = "\\hat t_i", x = "M_i", yhat = "\\hat B M_i"))
  }
}

## sdata --- a vector of original survey data
## N --- population size
## estimate --- "mean" (default) returns the population MEAN estimate;
##              "total" returns the population TOTAL estimate (requires
##              finite N)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values, with a Sum row and a
##                  footnote showing s_y^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_mean_est <- function (sdata, N = Inf, estimate = c ("mean", "total"),
                          show.details = TRUE, col_labels = list (y = "y_i"))
{
    estimate <- match.arg (estimate)

    n <- length (sdata)
    ybar <- mean (sdata)
    dev <- sdata - ybar
    s2y <- sum (dev^2) / (n - 1)
    se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)

    if (estimate == "total")
    {
        if (!is.finite (N))
            stop ("N must be finite to compute estimate = \"total\"")
        est <- N * ybar
        se.est <- N * se.ybar
    }
    else
    {
        est <- ybar
        se.est <- se.ybar
    }
    mem <- qt (0.975, df = n - 1) * se.est
    estimate_vec <- c (Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return (list (estimate = estimate_vec, table = NULL))

    working <- data.frame (
        i    = c (seq_len (n), "Sum"),
        y    = c (sdata, sum (sdata)),
        dev  = c (dev, sum (dev)),
        dev2 = c (dev^2, sum (dev^2))
    )
    working <- truncate_working (working, id_col = "i")

    table <- gt (working) |>
        tab_header (title = "SRS Mean Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
        cols_label (
            i    = md ("$i$"),
            y    = md (sprintf ("$%s$", col_labels$y)),
            dev  = md (sprintf ("$%s-\\bar y$", col_labels$y)),
            dev2 = md (sprintf ("$(%s-\\bar y)^2$", col_labels$y))
        ) |>
        cols_width (i ~ px (40), everything () ~ px (110)) |>
        fmt_auto (data = working, columns = c ("y", "dev", "dev2"), decimals = 3) |>
        sub_missing (missing_text = "...") |>
        tab_style (
            style     = cell_text (weight = "bold"),
            locations = cells_body (rows = i == "Sum")
        ) |>
        tab_footnote (
            footnote  = md (sprintf ("$s_y^2 = \\sum_i (%s-\\bar y)^2/(n-1) = %.4f$.", col_labels$y, s2y)),
            locations = cells_column_labels (columns = dev2)
        )

    list (estimate = estimate_vec, table = table)
}

# Helper function to format estimation outputs with appropriate mathematical headers
format_est_gt <- function(est_data, est_type = c("mean", "total", "ratio", "reg_mean", "reg_total", "ratio_mean", "domain_mean")) {
  est_type <- match.arg(est_type)
  
  sym_map <- list(
    "mean"        = c(est = "$\\bar{y}$",         se = "$\\mathrm{SE}(\\bar{y})$"),
    "total"       = c(est = "$\\hat{t}$",         se = "$\\mathrm{SE}(\\hat{t})$"),
    "ratio"       = c(est = "$\\hat{B}$",         se = "$\\mathrm{SE}(\\hat{B})$"),
    "reg_mean"    = c(est = "$\\bar{y}_{reg}$",   se = "$\\mathrm{SE}(\\bar{y}_{reg})$"),
    "reg_total"   = c(est = "$\\hat{t}_{reg}$",   se = "$\\mathrm{SE}(\\hat{t}_{reg})$"),
    "ratio_mean"  = c(est = "$\\bar{y}_{r}$",     se = "$\\mathrm{SE}(\\bar{y}_{r})$"),
    "domain_mean" = c(est = "$\\bar{y}_{d}$",     se = "$\\mathrm{SE}(\\bar{y}_{d})$")
  )
  
  est_sym <- sym_map[[est_type]]["est"]
  se_sym  <- sym_map[[est_type]]["se"]
  
  # Convert to data frame and strip hidden names to prevent gt() errors
  if (is.vector(est_data) && !is.list(est_data)) {
    df <- as.data.frame(t(unname(est_data)))
  } else {
    df <- as.data.frame(est_data)
  }
  
  colnames(df) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  use_stub <- !is.null(rownames(df)) && !all(rownames(df) == as.character(1:nrow(df)))
  
  df |>
    gt(rownames_to_stub = use_stub) |>
    cols_label(
      Est.   = md(est_sym),
      S.E.   = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(
      columns = everything(),
      decimals = 4
    ) |>
    tab_options(table.width = pct(70))
}
```

## Estimating for One-stage cluster sampling: analysis of algebra.csv

Consider a population of 187 high school algebra classes in a city. An investigator takes an SRS of 12 of those classes (the clusters). Because this is a one-stage sample, every student in the 12 sampled classes takes the test to evaluate their function knowledge.

We read the raw data and display it using `gt`'s interactive pagination feature to keep the document clean.

```{r}
algebra <- read.csv("data/algebra.csv")

algebra |>
  gt() |>
  tab_header(title = "Raw Data: Algebra Classes") |>
  cols_label(Mi = html("M<sub>i</sub>")) |>
  opt_interactive(use_pagination = TRUE, page_size_default = 10)
```
**Output Note:** The table above displays the raw, unaggregated survey data. The interactive pagination allows you to flip through the dataset compactly.

Next, we run our updated function to compute both the cluster-level working table and the overall ratio estimate for the average student score.

```{r}
alg_res <- cluster_ratio_est(data = algebra,
                             cname = "class",
                             csize = "Mi",
                             yvar = "score",
                             N = 187)

# Display the cluster-level working table
alg_res$table
```
**Output Note:** The working table above displays, for each of the 12 randomly selected classes, the class size ($M_i$), the total class score ($t_i$), the fitted value $\hat B M_i$, and the residual and squared residual used to estimate $s_e^2$.

Finally, we extract the ratio estimate from the function output list to assess the overall population mean.

```{r}
# Display the overall ratio estimate
alg_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** Using the ratio estimator, the estimated mean score for a student across all 187 algebra classes is `r round(alg_res$estimate["Est."], 2)`, with a 95% confidence interval of [`r round(alg_res$estimate["ci.low"], 2)`, `r round(alg_res$estimate["ci.upp"], 2)`]. The standard error accounts for the clustering effect, which typically increases variance compared to a pure SRS of students.

## Estimating for two-stage cluster sampling: analysis of coots.csv

In two-stage cluster sampling, we sample clusters, and then subsample elements within them. In this dataset, researcher Arnold investigated coot eggs. He first selected a random sample of clutches (the clusters), and then measured the volume of a random subsample of eggs within each selected clutch.


```{r}
coots <- read.csv("data/coots.csv")

coots |>
  gt() |>
  tab_header(title = "Raw Data: Coots") |>
  tab_options(
    container.height = px(400), # Adjust height as needed
    container.overflow.y = TRUE # Forces the vertical scroll bar
  )
```


We apply the exact same ratio estimator formulation, though mathematically, the cluster total $t_i$ is now an *estimated* total $\hat{t}_i$.

```{r}
coots_res <- cluster_ratio_est(data = coots,
                               cname = "clutch",
                               csize = "csize",
                               yvar = "volume")

coots_res$table
```
**Output Note:** Unlike the previous example, we only measured a subset of eggs per clutch. The function automatically estimates the total volume of each clutch ($\hat{t}_i$) by multiplying the true clutch size ($M_i$) by the observed subsample mean ($\overline{y}_i$); the working table (truncated to the first 6 rows + Sum) shows these estimated totals alongside the ratio fit and residuals.

```{r}
# Display the overall ratio estimate
coots_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimated overall average egg volume is `r round(coots_res$estimate["Est."], 3)` units. Because the total number of clutches in the population ($N$) is not explicitly provided (set to `Inf` by default), the finite population correction factor is safely omitted from the standard error calculation.