# Cluster Sampling

## Key Formulas for Cluster Sampling

Cluster sampling divides the population into $N$ clusters, out of which a simple random sample of $n$ clusters is selected. This design is highly practical when a sampling frame of individual elements is unavailable, or when individuals are geographically grouped.

#### 1. One-Stage Cluster Sampling {-}
In a one-stage design, every element $M_i$ within the sampled cluster $i$ is surveyed. 

*   **Total in Cluster $i$:** $t_i = \sum_{j=1}^{M_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:** The ratio estimator leverages cluster sizes to estimate the overall element mean:
    $$
\overline{y}_r = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n M_i}.
$$

*   **Estimated Variance of the Mean:**
    $$
\hat{V}(\overline{y}_r) = \left(1 - \frac{n}{N}\right) \frac{1}{n\overline{M}^2} \frac{\sum_{i=1}^n (t_i - \overline{y}_r M_i)^2}{n-1}.
$$

#### 2. Two-Stage Cluster Sampling {-}
In a two-stage design, we first sample $n$ clusters, then take a subsample of $m_i$ elements from the $M_i$ elements within each selected cluster. Because we don't observe the true cluster total $t_i$, we estimate it:

*   **Estimated Total in Cluster $i$:** $\hat{t}_i = M_i \overline{y}_i$, where $\overline{y}_i = \frac{1}{m_i} \sum_{j=1}^{m_i} y_{ij}$
*   **Ratio Estimator of the Population Mean:**
    $$
\overline{y}_r = \frac{\sum_{i=1}^n \hat{t}_i}{\sum_{i=1}^n M_i}
$$

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `cluster_ratio`, used below for one-stage and two-stage cluster samples) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Estimating for One-stage cluster sampling: analysis of algebra.csv

Consider a population of 187 high school algebra classes in a city. An investigator takes an SRS of 12 of those classes (the clusters). Because this is a one-stage sample, every student in the 12 sampled classes takes the test to evaluate their function knowledge.

We read the raw data and display it using `gt`'s interactive pagination feature to keep the document clean.

```{r tbl-algebra-raw-data}
#| tbl-cap: "Raw data: algebra classes (sample)"
algebra <- read.csv("data/algebra.csv")

algebra |>
  gt() |>
  tab_header(title = "Raw Data: Algebra Classes") |>
  cols_label(Mi = html("M<sub>i</sub>")) |>
  opt_interactive(use_pagination = TRUE, page_size_default = 10)
```
**Output Note:** The table above displays the raw, unaggregated survey data. The interactive pagination allows you to flip through the dataset compactly.

Next, we run our updated function to compute both the cluster-level working table and the overall ratio estimate for the average student score.

```{r algebra-cluster-ratio}
alg_res <- cluster_ratio(data = algebra,
                             cname = "class",
                             csize = "Mi",
                             yvar = "score",
                             N = 187)

# Display the cluster-level working table
alg_res$table
```
**Output Note:** The working table above displays, for each of the 12 randomly selected classes, the class size ($M_i$), the total class score ($t_i$), the fitted value $\hat B M_i$, and the residual and squared residual used to estimate $s_e^2$.

Finally, we extract the ratio estimate from the function output list to assess the overall population mean.

```{r algebra-ratio-mean}
# Display the overall ratio estimate
alg_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** Using the ratio estimator, the estimated mean score for a student across all 187 algebra classes is `r round(alg_res$estimate["Est."], 2)`, with a 95% confidence interval of [`r round(alg_res$estimate["ci.low"], 2)`, `r round(alg_res$estimate["ci.upp"], 2)`]. The standard error accounts for the clustering effect, which typically increases variance compared to a pure SRS of students.

## Estimating for two-stage cluster sampling: analysis of coots.csv

In two-stage cluster sampling, we sample clusters, and then subsample elements within them. In this dataset, researcher Arnold investigated coot eggs. He first selected a random sample of clutches (the clusters), and then measured the volume of a random subsample of eggs within each selected clutch.


```{r tbl-coots-raw-data}
#| tbl-cap: "Raw data: coot egg clutches (sample)"
coots <- read.csv("data/coots.csv")

coots |>
  gt() |>
  tab_header(title = "Raw Data: Coots") |>
  tab_options(
    container.height = px(400), # Adjust height as needed
    container.overflow.y = TRUE # Forces the vertical scroll bar
  )
```


We apply the exact same ratio estimator formulation, though mathematically, the cluster total $t_i$ is now an *estimated* total $\hat{t}_i$.

```{r coots-cluster-ratio}
coots_res <- cluster_ratio(data = coots,
                               cname = "clutch",
                               csize = "csize",
                               yvar = "volume")

coots_res$table
```
**Output Note:** Unlike the previous example, we only measured a subset of eggs per clutch. The function automatically estimates the total volume of each clutch ($\hat{t}_i$) by multiplying the true clutch size ($M_i$) by the observed subsample mean ($\overline{y}_i$); the working table (truncated to the first 6 rows + Sum) shows these estimated totals alongside the ratio fit and residuals.

```{r coots-ratio-mean}
# Display the overall ratio estimate
coots_res$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The estimated overall average egg volume is `r round(coots_res$estimate["Est."], 3)` units. Because the total number of clutches in the population ($N$) is not explicitly provided (set to `Inf` by default), the finite population correction factor is safely omitted from the standard error calculation.

## Interactive Demonstration: Clustering by State or at Random {#sec-app-cluster}

The app below draws two samples of the same size at every step. The first is a
**two-stage cluster sample**: an SRS of $n$ clusters, followed by an SRS of
$m_i = p \, M_i$ counties inside each selected cluster. The second is an
ordinary **SRS of counties**, drawn directly from the whole population with
exactly as many counties as the cluster sample happened to collect, so the two
designs are compared at equal sample size. The population is the full
`agpop.csv` file (with the `-99` missing-value codes recoded to `NA` and
dropped), and the variable of interest is again `acres92`.

A radio button chooses the clustering variable:

*   **state** --- the 50 real states, the natural clusters of `agpop.csv`.
    Counties in the same state resemble one another, so these clusters are
    internally homogeneous.
*   **Random** --- the state labels randomly permuted. The clusters keep
    *exactly* the same sizes $M_i$, but membership is now arbitrary, so all
    homogeneity inside clusters is destroyed. Comparing the two settings
    isolates the effect of within-cluster homogeneity and nothing else.

Two sliders set the design: $n$, the number of clusters drawn, and $p =
m_i/M_i$, the percentage of counties subsampled from each selected cluster.
Setting $p = 100\%$ measures every county in the chosen clusters, which is
one-stage cluster sampling.

The two estimators shown are the ones from the formulas above: the ratio
estimator $\overline{y}_r = \sum \hat{t}_i / \sum M_i$ with the cluster sizes
$M_i$ as the auxiliary variable (with $\hat{t}_i = M_i \overline{y}_i$ and
$\hat{V}(\overline{y}_r) = (1 - n/N) \sum (\hat{t}_i - \overline{y}_r
M_i)^2 / [n(n-1)\overline{M}^2]$), and the plain sample average of the SRS.

The left panel shows the population split into its clusters along the x-axis,
each drawn as a column of jittered points on a broken `acres92` axis. When the
cluster sample is highlighted, the selected clusters are shaded and labelled
with $m_i/M_i$, and a horizontal bar marks each subsample mean
$\overline{y}_i$, with the overall ratio estimate $\overline{y}_r$ drawn across
the panel. When the SRS is highlighted instead, the red points are scattered
over every cluster and a single bar marks the sample mean --- the visual
signature of the difference between the two designs.

The right panel accumulates the estimates of both designs on a common axis, one
column each, with a violin summarising each cloud once enough draws have
accumulated. The bar beside each column is the current confidence interval,
green when it covers $\mu$ and red when it does not. The legends report the
running CI coverage and the **design effect**, the simulated variance of the
ratio estimator divided by that of the SRS average, alongside its theoretical
counterpart computed from the true population $M_i$, $t_i$, and $S_i^2$. The
subtitle also gives the between-cluster $R^2$ of `acres92`.

With `state` clusters the design effect is far above 1 --- cluster sampling
loses badly to an SRS of the same size, because counties within a state carry
largely the same information. With `Random` clusters the design effect drops to
about 1: once the clusters are internally as variable as the population,
cluster sampling costs nothing statistically. With `state` clusters, raising
$n$ helps a great deal, while raising $p$ at fixed $n$ barely moves the
variance --- going from $p = 30\%$ to a full census of each sampled state
($p = 100\%$) triples the fieldwork and buys only a few percent. This is the
central lesson of cluster sampling: when clusters are homogeneous, precision is
bought with *clusters*, not with elements inside them.

Play advances the simulation one draw at a time; Pause stops it; Skip jumps
straight to the final violin plots by drawing every remaining replication up to
the cap in a single step.

::: {#fig-cluster-app}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 820

## file: app.R
library(shiny)

data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/data/agpop.csv"
download.file(data_url, "agpop.csv")
dat <- read.csv("agpop.csv")

## -99 is the missing-value code in agpop.csv; recode to NA and drop the
## counties with no acres92, so that a single population underlies both
## clustering choices below.
dat$acres92[dat$acres92 == -99] <- NA
dat <- dat[!is.na(dat$acres92), ]

pop   <- dat$acres92
Nel   <- length(pop)          # number of elements (counties) in the population
ybarU <- mean(pop)
S     <- sd(pop)

## Build the clustering: "state" groups the counties as they really are,
## "random" permutes the state labels, so the clusters keep exactly the same
## sizes M_i but lose every trace of within-cluster homogeneity. Comparing the
## two isolates the effect of homogeneity inside clusters, and nothing else.
make_clusters <- function(method) {
  lab <- factor(dat$state)
  if (method == "random") {
    set.seed(7)
    lab <- sample(lab)                       # same sizes, scrambled membership
    levels(lab) <- sprintf("C%02d", seq_len(nlevels(lab)))
  }
  H     <- nlevels(lab)                      # N, the number of clusters
  idx   <- split(seq_len(Nel), lab)
  Mi    <- sapply(idx, length)
  ti    <- sapply(idx, function(i) sum(pop[i]))
  Si2   <- sapply(idx, function(i) if (length(i) > 1) var(pop[i]) else 0)

  set.seed(42)
  x_pos <- as.integer(lab) + runif(Nel, -0.32, 0.32)

  list(lev = levels(lab), H = H, idx = idx, Mi = Mi, ti = ti, Si2 = Si2,
       M0 = sum(Mi), Mbar = mean(Mi), x_pos = x_pos,
       R2 = 1 - sum(sapply(idx, function(i) sum((pop[i] - mean(pop[i]))^2))) /
                sum((pop - ybarU)^2))
}

CLUST_CHOICES <- c("state (real clusters)" = "state",
                   "Random (scrambled labels)" = "random")

## subsample size in cluster i: m_i = p * M_i, at least 1 and at most M_i
subsize <- function(Mi, p) pmax(1, pmin(Mi, round(p * Mi)))

## Anticipated variances from the FULL population, for the "theory" figures.
## Two-stage ratio estimator (between-cluster term + within-cluster term):
##   V = (1 - n/N) S_t^2 / (n Mbar^2)
##       + sum_U M_i^2 (1 - m_i/M_i) S_i^2 / m_i / (n N Mbar^2),
## with S_t^2 the population variance of the ratio residuals t_i - ybarU M_i.
## SRS is compared at the sample size the cluster design delivers on average.
theory <- function(s, n, p) {
  mi  <- subsize(s$Mi, p)
  St2 <- sum((s$ti - ybarU * s$Mi)^2) / (s$H - 1)
  Vc  <- (1 - n / s$H) * St2 / (n * s$Mbar^2) +
         sum(s$Mi^2 * (1 - mi / s$Mi) * s$Si2 / mi) / (n * s$H * s$Mbar^2)
  nel <- max(2, round(n * mean(mi)))
  Vs  <- (1 - nel / s$M0) * S^2 / nel
  list(Vc = Vc, Vs = Vs, nel = nel)
}

## One two-stage cluster sample: an SRS of n clusters, then an SRS of
## m_i = p M_i counties inside each sampled cluster. The estimate is the ratio
## estimator with M_i as the auxiliary variable, ybar_r = sum(t_i hat)/sum(M_i).
clu_draw <- function(s, n, p, a) {
  ci  <- sort(sample(s$H, n))
  Mis <- s$Mi[ci]
  mi  <- subsize(Mis, p)
  ii  <- unlist(lapply(seq_len(n), function(k) sample(s$idx[[ci[k]]], mi[k])))
  ybari <- sapply(split(pop[ii], rep(seq_len(n), mi)), mean)
  that  <- Mis * ybari                       # t_i hat = M_i ybar_i
  Mbar  <- mean(Mis)
  m     <- sum(that) / sum(Mis)
  v     <- (1 - n / s$H) * sum((that - m * Mis)^2) / (n - 1) / (n * Mbar^2)
  list(ci = ci, mi = mi, Mi = Mis, ii = ii, ybari = ybari,
       est = c(m, qt(a, df = n - 1) * sqrt(v)))
}

## One simple random sample of counties, of the same size as the cluster
## sample just drawn, estimated by the plain sample average.
srs_draw <- function(nel, a) {
  ii  <- sample(Nel, nel)
  y   <- pop[ii]
  m   <- mean(y)
  se  <- sqrt(1 - nel / Nel) * sd(y) / sqrt(nel)
  list(ii = ii, est = c(m, qt(a, df = nel - 1) * se))
}

## ---- broken y-axis, shared by the left panel --------------------------------
## Drawn region [0, 2 * ybarU]: visible scale [0, cut] plus a compressed lane
## for the right tail, so the population mean sits at the midpoint of the panel.
lane_frac <- 0.10
cut       <- 2 * ybarU / (1 + lane_frac)
lane_w    <- lane_frac * cut
ymax_top  <- cut + lane_w

py      <- pop
tail_id <- which(pop > cut)
n_tail  <- length(tail_id)
if (n_tail > 0) {
  r <- rank(pop[tail_id], ties.method = "first") / (n_tail + 1)
  py[tail_id] <- cut + lane_w * (0.10 + 0.80 * r)
}

METHOD <- c("SRS average", "Cluster ratio")
MCOL   <- c("grey35", "#2ca02c")
MAR    <- c(5.0, 4.5, 3.0, 1.0)

## symmetric density polygon, drawn vertically about x = at
violin <- function(x, at, col, w = 0.30) {
  if (length(x) < 10 || diff(range(x)) <= 0) return(invisible())
  d  <- density(x)
  hw <- w * d$y / max(d$y)
  polygon(c(at + hw, rev(at - hw)), c(d$x, rev(d$x)),
          col = adjustcolor(col, 0.35), border = col)
  m <- median(x)
  segments(at - w * 0.6, m, at + w * 0.6, m, col = col, lwd = 2)
}

pct <- function(v) if (is.na(v)) "--" else sprintf("%.1f%%", 100 * v)
fmt <- function(x) format(round(x), big.mark = ",")

## ---- left panel: the population, cluster by cluster -------------------------
## clu is the last cluster draw, srs the last SRS draw (either may be NULL);
## j = 1 highlights the SRS sample, j = 2 the two-stage cluster sample.
panel_pop <- function(s, j, clu, srs, p, lab) {
  H <- s$H
  par(mar = MAR, yaxs = "i")
  plot(NA, xlim = c(0.3, H + 0.7), ylim = c(0, ymax_top),
       xlab = "", ylab = "acres92 (thousands)", xaxt = "n", yaxt = "n", bty = "n",
       main = sprintf("Population by %s, with the %s sample", lab, METHOD[j]))

  ## one band per cluster; the clusters that made it into the two-stage
  ## sample are shaded, which is the whole point of the design
  inS <- rep(FALSE, H)
  if (j == 2 && !is.null(clu)) inS[clu$ci] <- TRUE
  for (h in seq_len(H))
    rect(h - 0.38, 0, h + 0.38, ymax_top,
         col = if (inS[h]) "#fff0c2" else adjustcolor("grey92", 0.6), border = NA)

  points(s$x_pos, py, pch = 16, cex = 0.42, col = adjustcolor("grey55", 0.5))
  smp <- if (j == 1) srs else clu
  if (!is.null(smp))
    points(s$x_pos[smp$ii], py[smp$ii], pch = 16, cex = 0.55,
           col = adjustcolor("red", 0.85))

  ## sample mean(s): one bar per sampled cluster for the two-stage design
  ## (the ybar_i that build t_i hat), a single bar for SRS
  if (!is.null(smp)) {
    if (j == 1) {
      segments(0.3, srs$est[1], H + 0.7, srs$est[1], col = MCOL[1], lwd = 3)
    } else {
      segments(clu$ci - 0.38, clu$ybari, clu$ci + 0.38, clu$ybari,
               col = MCOL[2], lwd = 3)
      abline(h = clu$est[1], col = MCOL[2], lwd = 1.5, lty = 4)
    }
    leg <- if (j == 1)
             list(txt = c("Population mean", "Sample mean"),
                  col = c("black", MCOL[1]), lty = c(2, 1), lwd = c(2, 3))
           else
             list(txt = c("Population mean", "Cluster means (ybar_i)",
                          "Ratio estimate (ybar_r)"),
                  col = c("black", MCOL[2], MCOL[2]), lty = c(2, 1, 4),
                  lwd = c(2, 3, 1.5))
    legend("topleft", bty = "o", bg = "white", box.col = NA, cex = 0.75,
           inset = c(0.005, 0.01), legend = leg$txt,
           col = leg$col, lty = leg$lty, lwd = leg$lwd)
  }

  ## cluster labels: sampled clusters in red, with m_i / M_i when there is room
  cx <- if (H > 40) 0.45 else 0.6
  for (h in seq_len(H)) {
    mtext(s$lev[h], side = 1, at = h, line = 0.2, las = 2, cex = cx,
          col = if (inS[h]) "red" else "grey45", font = if (inS[h]) 2 else 1)
  }
  if (j == 2 && !is.null(clu) && length(clu$ci) <= 20)
    mtext(sprintf("%d/%d", clu$mi, clu$Mi), side = 1, at = clu$ci, line = 1.9,
          las = 2, cex = 0.5, col = "red")
  mtext(if (j == 2) sprintf("clusters (sampled ones shaded, labelled m_i/M_i; p = %.0f%%)",
                            100 * p)
        else "clusters",
        side = 1, line = 3.6, cex = 0.72, col = "grey30")

  at <- pretty(c(0, cut), 6); at <- at[at <= cut]
  axis(2, at = at, labels = format(at / 1000, big.mark = ",", trim = TRUE), las = 1)
  segments(0.3, cut, H + 0.7, cut, col = "grey60", lty = 3)
  text(H + 0.6, cut + lane_w / 2, sprintf("compressed: > %s", fmt(cut / 1000)),
       adj = c(1, 0), col = "grey40", cex = 0.7)
  abline(h = ybarU, col = "black", lwd = 2, lty = 2)
}

## ---- right panel: the two sampling distributions ----------------------------
panel_dist <- function(s, rv, est_range, th, n, p, conf, nrep) {
  par(mar = MAR, yaxs = "i")
  plot(NA, xlim = c(0.4, 2.6), ylim = est_range, xaxt = "n", yaxt = "n", bty = "n",
       xlab = "", ylab = "Estimated mean of acres92 (thousands)",
       main = "Sampling distributions of the two estimators")
  abline(h = ybarU, col = "black", lwd = 2, lty = 2)
  axis(1, at = 1:2, labels = METHOD, tick = FALSE, line = -0.5)
  at_r <- pretty(est_range, 6)
  at_r <- at_r[at_r >= est_range[1] & at_r <= est_range[2]]
  axis(2, at = at_r, labels = format(at_r / 1000, big.mark = ",", trim = TRUE), las = 1)

  if (is.null(rv$M)) {
    text(1.5, ybarU, "Click 'Draw new samples' or Play to begin", col = "grey50")
    return(invisible())
  }

  k <- NROW(rv$M)
  for (jj in 1:2) {
    points(jj + rv$jit, rv$M[, jj], pch = 16, cex = 0.45,
           col = adjustcolor(MCOL[jj], 0.30))
    violin(rv$M[, jj], at = jj, col = MCOL[jj])

    ci_col <- if (rv$HIT[k, jj]) "forestgreen" else "red"
    xb     <- jj + 0.40
    segments(xb, rv$est[jj, 2], xb, rv$est[jj, 3], col = ci_col, lwd = 3)
    segments(xb - 0.06, rv$est[jj, 2:3], xb + 0.06, rv$est[jj, 2:3], col = ci_col, lwd = 3)
    points(xb, rv$est[jj, 1], pch = 19, cex = 1.1, col = ci_col)
  }

  legend("topleft", bty = "n", cex = 0.85, inset = c(0.02, 0.01),
         title = sprintf("%.0f%% CI coverage", 100 * conf), fill = MCOL, border = NA,
         legend = paste(METHOD, vapply(1:2, function(jj) pct(mean(rv$HIT[, jj])), "")))

  ## design effect: variance of the cluster estimator relative to SRS at the
  ## same sample size, simulated so far and anticipated from the population
  if (k > 1) {
    deff <- var(rv$M[, 2]) / var(rv$M[, 1])
    legend("bottomleft", bty = "n", cex = 0.85, inset = c(0.02, 0.01),
           title = "Variance of cluster ratio vs. SRS", fill = MCOL[2], border = NA,
           text.col = if (deff > 1) "red" else "black",
           legend = sprintf("deff = %.2f  (theory %.2f)", deff, th$Vc / th$Vs))
  }

  outside <- sum(rv$M < est_range[1] | rv$M > est_range[2])
  sub <- sprintf("n = %d clusters, %s counties  |  R2 between clusters = %s  |  %d of %d draws%s",
                 n, fmt(rv$nel), pct(s$R2), k, nrep,
                 if (outside > 0) sprintf("  |  %d off axis", outside) else "")
  mtext(sub, side = 3, line = -0.2, cex = 0.7, col = "grey30")
}

## ---- inline SVG button icons (no font or icon-library dependency) ----------
svg_icon <- function(paths)
  HTML(sprintf(paste0('<svg width="16" height="16" viewBox="0 0 16 16" ',
                      'fill="currentColor" style="vertical-align:-2px">%s</svg>'), paths))
icon_play  <- svg_icon('<path d="M3 2v12l10-6z"/>')
icon_pause <- svg_icon('<rect x="3" y="2" width="4" height="12"/><rect x="9" y="2" width="4" height="12"/>')
icon_skip  <- svg_icon('<path d="M1 2v12l7-6zM8 2v12l6-6z"/><rect x="14" y="2" width="2" height="12"/>')

ui <- fluidPage(
  titlePanel("Shinylive Simulation for Two-Stage Cluster Sampling"),
  tags$style(HTML(
    ".well {padding: 10px 12px; margin-bottom: 10px;}
     .form-group {margin-bottom: 6px;}
     .radio {margin-top: 2px; margin-bottom: 2px;}
     .irs {margin-bottom: 0;}
     .btn {margin-bottom: 6px;}"
  )),
  wellPanel(
    fluidRow(
      column(
        3,
        radioButtons("clustvar", "Cluster variable", choices = CLUST_CHOICES,
                     selected = "state"),
        radioButtons("show", "Highlight sample from", choiceNames = METHOD,
                     choiceValues = as.character(1:2), selected = "2")
      ),
      column(
        3,
        sliderInput("n", "Number of clusters drawn (n)", min = 2, max = 40,
                    value = 10, step = 1),
        sliderInput("p", "Percentage sampled per cluster (p = m_i/M_i)",
                    min = 0.05, max = 1, value = 0.30, step = 0.05)
      ),
      column(
        3,
        sliderInput("conf", "Confidence level", min = 0.80, max = 0.99,
                    value = 0.95, step = 0.01),
        sliderInput("speed", "Samples per second", min = 1, max = 10, value = 3, step = 1)
      ),
      column(
        3,
        numericInput("nrep", "Stop after this many samples",
                     value = 500, min = 10, max = 20000, step = 50),
        actionButton("draw", "Draw new samples", class = "btn-primary", width = "100%"),
        br(), br(),
        fluidRow(
          column(4, actionButton("play",  icon_play,  class = "btn-success", width = "100%",
                                 title = "Play")),
          column(4, actionButton("pause", icon_pause, class = "btn-warning", width = "100%",
                                 title = "Pause")),
          column(4, actionButton("skip",  icon_skip,  class = "btn-info",    width = "100%",
                                 title = "Skip to the end"))
        ),
        actionButton("reset", "Reset", width = "100%")
      )
    )
  ),
  plotOutput("mainplot", height = "530px")
)

server <- function(input, output, session) {

  rv <- reactiveValues(
    clu = NULL,   # last two-stage cluster draw
    srs = NULL,   # last simple random sample
    est = NULL,   # matrix 2 x 3: estimate, lower, upper (last draw)
    nel = NA,     # number of counties in the last sample
    M   = NULL,   # accumulated estimates, one column per design
    HIT = NULL,   # accumulated coverage indicators
    jit = numeric(0)
  )
  running <- reactiveVal(FALSE)

  nrep <- reactive({
    v <- input$nrep
    if (is.null(v) || is.na(v) || v < 1) 1L else min(as.integer(v), 20000L)
  })

  si     <- reactive(make_clusters(input$clustvar))
  clab   <- reactive(names(CLUST_CHOICES)[CLUST_CHOICES == input$clustvar])
  th     <- reactive(theory(si(), input$n, input$p))

  ## a window wide enough for whichever design is the more variable
  est_range <- reactive({
    t0 <- th()
    ybarU + c(-1, 1) * 4 * sqrt(max(t0$Vc, t0$Vs))
  })

  ## draw k replications; each one draws a two-stage cluster sample and an SRS
  ## of the same realised size. Only the last replication's sampled units are
  ## kept, for the left panel; "Skip" calls this with the full remaining count.
  draw_many <- function(k) {
    n    <- isolate(input$n)
    p    <- isolate(input$p)
    conf <- isolate(input$conf)
    a    <- 1 - (1 - conf) / 2
    s    <- isolate(si())

    Mnew   <- matrix(0, k, 2)
    HITnew <- matrix(FALSE, k, 2)
    jitnew <- numeric(k)

    for (i in seq_len(k)) {
      cl <- clu_draw(s, n, p, a)
      sr <- srs_draw(length(cl$ii), a)

      est <- rbind(c(sr$est[1], sr$est[1] - sr$est[2], sr$est[1] + sr$est[2]),
                   c(cl$est[1], cl$est[1] - cl$est[2], cl$est[1] + cl$est[2]))

      Mnew[i, ]   <- est[, 1]
      HITnew[i, ] <- est[, 2] <= ybarU & ybarU <= est[, 3]
      jitnew[i]   <- runif(1, -0.09, 0.09)

      if (i == k) {
        rv$clu <- cl; rv$srs <- sr; rv$est <- est; rv$nel <- length(cl$ii)
      }
    }
    rv$M   <- rbind(rv$M,   Mnew)
    rv$HIT <- rbind(rv$HIT, HITnew)
    rv$jit <- c(rv$jit, jitnew)
  }
  draw_one <- function() draw_many(1)

  observeEvent(input$draw,  { running(FALSE); draw_one() })
  observeEvent(input$play,  running(TRUE))
  observeEvent(input$pause, running(FALSE))

  observeEvent(input$skip, {
    running(FALSE)
    k <- nrep() - NROW(rv$M)
    if (k > 0) draw_many(k)
  })

  observe({
    if (!isTRUE(running())) return()
    if (NROW(isolate(rv$M)) >= isolate(nrep())) { running(FALSE); return() }
    invalidateLater(1000 / isolate(input$speed), session)
    isolate(draw_one())
  })

  reset_history <- function() {
    running(FALSE)
    rv$clu <- NULL; rv$srs <- NULL; rv$est <- NULL; rv$nel <- NA
    rv$M <- NULL; rv$HIT <- NULL; rv$jit <- numeric(0)
  }
  observeEvent(input$reset,    reset_history())
  observeEvent(input$n,        reset_history())
  observeEvent(input$p,        reset_history())
  observeEvent(input$conf,     reset_history())
  observeEvent(input$clustvar, reset_history())

  output$mainplot <- renderPlot({
    layout(matrix(1:2, nrow = 1), widths = c(1.3, 1))
    panel_pop(si(), as.integer(input$show), rv$clu, rv$srs, input$p, clab())
    panel_dist(si(), rv, est_range(), th(), input$n, input$p, input$conf, nrep())
  })
}

shinyApp(ui, server)
```

**Shinylive App Comparing Two-Stage Cluster Sampling (ratio estimator) with SRS**

:::
