---
title: "STAT 348 Sampling Techniques"
subtitle: "Stratified Random Sampling"
format: banner_slides-revealjs
engine: knitr
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center"
)
```

# What Is Stratified Sampling?

Textbook sections:

- 3.1--3.5 in Lohr

- 5.1--5.9 in Scheaffer et al.

## What Is Stratified Sampling?

If the variable we are interested in takes on different mean values in different
subpopulations, we may be able to obtain more precise estimates of population
quantities by taking a stratified random sample. The word *stratify* comes from
Latin words meaning "to make layers"; we divide the population into $H$
subpopulations, called **strata**. The strata do not overlap, and they constitute
the whole population, so that each sampling unit belongs to exactly one stratum.

We draw an independent probability sample from each stratum, then pool the
information to obtain overall population estimates.

## A Population Divided into Strata

```{r}
#| label: fig-strata-pop
#| fig-width: 8
#| fig-height: 4
#| out-width: "80%"
set.seed(11)
strata <- list(
  list(xmin = 0,   xmax = 2.2, n = 6, k = 2, label = "1"),
  list(xmin = 2.3, xmax = 4.9, n = 7, k = 2, label = "2"),
  list(xmin = 5.0, xmax = 6.6, n = 5, k = 1, label = "3"),
  list(xmin = 6.7, xmax = 9.4, n = 7, k = 2, label = "…"),
  list(xmin = 9.5, xmax = 12,  n = 6, k = 2, label = "H")
)
par(mar = c(0.2, 0.2, 0.2, 0.2))
plot(NA, xlim = c(0, 12), ylim = c(0, 5), asp = 1,
     xaxt = "n", yaxt = "n", xlab = "", ylab = "", bty = "o")
for (s in strata) {
  rect(s$xmin, 0, s$xmax, 5, border = "blue4", lwd = 1.5)
  text((s$xmin + s$xmax) / 2, 4.6, s$label, col = "blue4", font = 2, cex = 1.8)
  px <- runif(s$n, s$xmin + 0.3, s$xmax - 0.3)
  py <- runif(s$n, 0.3, 3.8)
  points(px, py, pch = 19, col = "blue4", cex = 1.1)
  sel <- sample(s$n, s$k)
  points(px[sel], py[sel], pch = 21, col = "firebrick", lwd = 2, cex = 2.4)
}
```

Each stratum $h$ has $N_h$ units in the population; we sample $n_h$ of them
(circled). $H$ is the number of strata.

## Example: Stratifying by Census Region

We use the four U.S. census regions as strata. The SRS in an earlier example
sampled about 10% of all 3078 counties; to compare stratified sampling with SRS
directly, we also sample about 10% of the counties in *each* stratum:

| Stratum | Counties in Stratum | Counties in Sample |
|:---|---:|---:|
| Northeast | 220 | 21 |
| North Central | 1054 | 103 |
| South | 1382 | 135 |
| West | 422 | 41 |
| **Total** | **3078** | **300** |

We select four separate, independent SRSs, one per stratum. Knowing which
counties are in the sample from the Northeast tells us nothing about which
counties are in the sample from the South.

## Example: Data Summary from Stratified Sampling

Summary statistics for `acres92` in each stratum (data file `agstrat.dat`):

| Region ($h$) | Sample Size ($n_h$) | Average ($\overline y_h$) | Variance ($s_h^2$) |
|:---|---:|---:|---:|
| Northeast | 21 | 97,629.8 | 7,647,472,708 |
| North Central | 103 | 300,504.2 | 29,618,183,543 |
| South | 135 | 211,315.0 | 53,587,487,856 |
| West | 41 | 662,295.5 | 396,185,950,266 |

# Estimation with Stratified Samples

## Population Quantities

For unit $j$ in stratum $h$, let $y_{hj}$ denote its value, and let
$$
\pi_h = \frac{N_h}{N},
$$
denote the proportion of population units belonging to stratum $h$ (so
$\sum_{h=1}^H \pi_h = 1$). Then,
$$
t_h = \sum_{j=1}^{N_h} y_{hj} \quad(\text{stratum total}) \qquad
t = \sum_{h=1}^H t_h \quad(\text{population total})
$$
$$
\overline y_{hU} = \frac{\sum_{j=1}^{N_h} y_{hj}}{N_h} \quad(\text{stratum mean}) \qquad
\overline y_U = \frac tN = \sum_{h=1}^H \pi_h\, \overline y_{hU} \quad(\text{overall mean})
$$
$$
S_h^2 = \frac{1}{N_h-1}\sum_{j=1}^{N_h}(y_{hj}-\overline y_{hU})^2 \quad(\text{stratum variance}).
$$

## SRS Estimators Within Each Stratum

Since an SRS is taken within each stratum, we estimate the stratum quantities
using ordinary SRS formulas, applied separately to each stratum:
$$
\overline y_h = \frac{1}{n_h}\sum_{j\in\mathcal S_h} y_{hj}
\qquad
\hat t_h = \frac{N_h}{n_h}\sum_{j\in\mathcal S_h} y_{hj} = N_h \overline y_h
$$
$$
s_h^2 = \sum_{j\in\mathcal S_h} \frac{(y_{hj}-\overline y_h)^2}{n_h - 1}.
$$

## Estimators for the Population Total and Mean
$$
\hat t_{\text{str}} = \sum_{h=1}^H \hat t_h = \sum_{h=1}^H N_h \overline y_h
$$
$$
\overline y_{\text{str}} = \frac{\hat t_{\text{str}}}{N} = \sum_{h=1}^H \pi_h\overline y_h
$$

$\overline y_{\text{str}}$ is a weighted average of the sample stratum averages, with
$\overline y_h$ weighted by $\pi_h = N_h/N$, the fraction of population units in
stratum $h$.
To use stratified sampling, the population sizes of the strata must be known.

## Standard Errors for Stratified Samples

Since we sample independently across strata, variances add:
$$
V(\hat t_{\text{str}}) = \sum_{h=1}^H V(\hat t_h) = \sum_{h=1}^H \Big(1-\frac{n_h}{N_h}\Big)N_h^2\frac{S_h^2}{n_h}.
$$

Substituting sample variances $s_h^2$ for $S_h^2$ gives unbiased estimators:
$$
\hat V(\hat t_{\text{str}}) = \sum_{h=1}^H \Big(1-\frac{n_h}{N_h}\Big)N_h^2\frac{s_h^2}{n_h}
$$
$$
\hat V(\overline y_{\text{str}}) = \frac{1}{N^2}\hat V(\hat t_{\text{str}}) = \sum_{h=1}^H \Big(1-\frac{n_h}{N_h}\Big)\pi_h^2\frac{s_h^2}{n_h}.
$$

Note we need at least two sampled units per stratum to estimate the variances.

## Unbiasedness of $\overline y_{\text{str}}$
$$
E\left[\sum_{h=1}^H \frac{N_h}{N}\overline y_h\right]
 = \sum_{h=1}^H \frac{N_h}{N}E[\overline y_h]
 = \sum_{h=1}^H \frac{N_h}{N}\overline y_{hU}
 = \sum_{h=1}^H \frac{N_h}{N}\cdot \frac{\sum_{j=1}^{N_h} y_{hj}}{N_h}
$$
$$
= \sum_{h=1}^H \frac{\sum_{j=1}^{N_h}y_{hj}}{N} = \overline y_U.
$$

Since an SRS is taken in each stratum, $E[\overline y_h]=\overline y_{hU}$; averaging these
unbiased stratum estimates with weights $N_h/N$ keeps the overall estimate
unbiased.

## Confidence Intervals for Stratified Samples

If either (1) the sample sizes within each stratum are large, or (2) the
sampling design has a large number of strata, an approximate
$100(1-\alpha)\%$ CI for $\overline y_U$ is
$$
\overline y_{\text{str}} \pm z_{\alpha/2}\,\text{SE}(\overline y_{\text{str}}).
$$

When $n$ is not large, it is common to use the quantile of a $t$ distribution
with $n-H$ degrees of freedom instead of the normal quantile:
$$
\hat t_{\text{str}} \pm t_{\alpha/2,\,n-H}\cdot \text{SE}(\hat t_{\text{str}}).
$$

## Example 1: The Nelchina Caribou Herd

Siniff and Skoog (1964) used stratified random sampling to estimate the size of
the Nelchina herd of Alaska caribou in February 1962. Preliminary caribou
density estimates were used to divide the area of interest into six strata; each
stratum was divided into a grid of 4-mi$^2$ sampling units. Stratum A, for
example, contained $N_1=400$ sampling units, of which $n_1=98$ were randomly
selected for the survey.

## Example 1: Strata Layout

```{r}
#| label: fig-caribou-strata
#| fig-width: 10
#| fig-height: 4
#| out-width: "95%"
set.seed(348)
strata <- data.frame(
  label = c("A", "B", "C", "D", "E", "F"),
  N = c(400, 30, 61, 18, 70, 120),
  n = c(98, 10, 37, 6, 39, 21)
)
strata$ncol <- ceiling(sqrt(strata$N))
strata$nrow <- ceiling(strata$N / strata$ncol)

gap <- 2
x0 <- 0
par(mar = c(2, 0.2, 2, 0.2))
plot(NA, xlim = c(0, sum(strata$ncol) + gap * (nrow(strata) - 1)),
     ylim = c(-2, max(strata$nrow) + 2), asp = 1,
     xaxt = "n", yaxt = "n", xlab = "", ylab = "", bty = "n")

for (i in seq_len(nrow(strata))) {
  s <- strata[i, ]
  ncol_i <- s$ncol
  nrow_i <- s$nrow
  idx <- expand.grid(col = 1:ncol_i, row = 1:nrow_i)
  idx <- idx[order(idx$row, idx$col), ][1:s$N, ]
  sampled <- sample(s$N, s$n)
  cell_col <- ifelse(seq_len(s$N) %in% sampled, "firebrick", "grey85")
  rect(x0 + idx$col - 1, idx$row - 1, x0 + idx$col, idx$row,
       col = cell_col, border = "white", lwd = 0.3)
  rect(x0, 0, x0 + ncol_i, nrow_i, border = "blue4", lwd = 1.5)
  text(x0 + ncol_i / 2, nrow_i + 1, s$label, col = "blue4", font = 2, cex = 1.6)
  text(x0 + ncol_i / 2, -1.2, paste0("N=", s$N, ", n=", s$n), cex = 0.75)
  x0 <- x0 + ncol_i + gap
}
```

Each grid represents one stratum's $N_h$ sampling units; the highlighted
(red) cells are the $n_h$ units randomly selected in that stratum.

## Example 1: Spreadsheet for Calculations

Here $\pi_h = N_h/N$ with $N=699$, and the last two columns give each stratum's
contribution to $\overline y_{\text{str}}=\sum_h \pi_h\overline y_h$ and to
$\hat V(\overline y_{\text{str}}) = \sum_h\big(1-\tfrac{n_h}{N_h}\big)\pi_h^2\tfrac{s_h^2}{n_h}$:

| Stratum | $N_h$ | $n_h$ | $\pi_h=N_h/N$ | $\overline y_h$ | $s_h^2$ | $\pi_h\overline y_h$ | $\hat v_h$[^vh] |
|:---:|---:|---:|-------:|---:|---:|---:|---:|
| A | 400 | 98 | 0.5722 | 24.1 | 5,575 | 13.7911 | 14.0648 |
| B | 30 | 10 | 0.0429 | 25.6 | 4,064 | 1.0987 | 0.4991 |
| C | 61 | 37 | 0.0873 | 267.6 | 347,556 | 23.3528 | 28.1455 |
| D | 18 | 6 | 0.0258 | 179.0 | 22,798 | 4.6094 | 1.6798 |
| E | 70 | 39 | 0.1001 | 293.7 | 123,578 | 29.4120 | 14.0728 |
| F | 120 | 21 | 0.1717 | 33.2 | 9,795 | 5.6996 | 11.3409 |
| **Total** | **699** | **211** | **1.0000** | | | **77.9636** | **69.8029** |

[^vh]: $\hat v_h = \big(1-\tfrac{n_h}{N_h}\big)\pi_h^2\tfrac{s_h^2}{n_h}$, the stratum-$h$ contribution to $\hat V(\overline y_{\text{str}}) = \sum_{h=1}^H \hat v_h$.

## Example 1: Results

The estimated mean number of caribou per sampling unit is
$$
\overline y_{\text{str}} = 77.96, 
$$
$$
 \text{SE}(\overline y_{\text{str}}) = \sqrt{69.8029} = 8.35.
$$

The estimated total number of caribou:
$$
\hat t_{\text{str}} = N\overline y_{\text{str}} = 699(77.96) = 54{,}497,
$$
$$
\text{SE}(\hat t_{\text{str}}) = N\cdot\text{SE}(\overline y_{\text{str}}) = 699(8.35) = 5{,}840.
$$

An approximate 95% CI for the total number of caribou:
$$
54{,}497 \pm 1.96(5840) = [43{,}051,\ 65{,}943].
$$

This CI reflects only sampling error; if the field procedure for counting
caribou tends to miss animals, the entire CI will be biased low.

## Example 2: Estimated Farm Acreage by Region

Using the same census-region strata for `acres92`:

| Stratum | Estimated Total of Farm Acres | Estimated Variance of Total |
|:---|---:|---:|
| Northeast | 21,478,558.2 | $1.5943\times10^{13}$ |
| North Central | 316,731,379.4 | $2.8823\times10^{14}$ |
| South | 292,037,390.8 | $6.8408\times10^{14}$ |
| West | 279,488,706.1 | $1.5537\times10^{15}$ |
| **Total** | **909,736,034.4** | $2.5419\times10^{15}$ |

Since sampling was independent across strata, the variance of the U.S. total is
the sum of the four stratum variances.

## Example 2: Results
$$
\hat t_{\text{str}} = 909{,}736{,}034, \qquad
\text{SE}(\hat t_{\text{str}}) = \sqrt{2.5419\times10^{15}} = 50{,}417{,}248.
$$

The average number of acres devoted to farming per county:
$$
\overline y_{\text{str}} = \frac{909{,}736{,}034}{3078} = 295{,}560.76, \qquad
\text{SE}(\overline y_{\text{str}}) = \frac{50{,}417{,}248}{3078} = 16{,}379.87.
$$

## Comparison of Stratified Sampling and SRS

Only the West had sample variance larger than the overall SRS sample variance;
the sample variance in the Northeast was much smaller than the population as a
whole. Observations within many strata tend to be more homogeneous than
observations in the population overall.

The relative gain from stratification can be estimated by the ratio
$$
\frac{\text{estimated variance from stratified sample},\ n=300}{\text{estimated variance from SRS},\ n=300}
 = \frac{2.5419\times10^{15}}{3.3837\times10^{15}} = 0.75.
$$

Percentage of variance reduction $= 1-0.75 = 0.25$. If these were the population
variances, we would need only $(300)(0.75)=225$ observations with a stratified
sample to match the precision of an SRS of 300.

# Stratified Sampling for Proportions

## Estimators for a Population Proportion $p$

A proportion is the mean of a 0/1 variable, so we apply the same formulas with
$\overline y_h = \hat p_h$ and $s_h^2 = \frac{n_h}{n_h-1}\hat p_h(1-\hat p_h)$:
$$
\hat p_{\text{str}} = \sum_{h=1}^H \frac{N_h}{N}\hat p_h
$$
$$
\hat V(\hat p_{\text{str}}) = \sum_{h=1}^H \Big(1-\frac{n_h}{N_h}\Big)\Big(\frac{N_h}{N}\Big)^2\frac{\hat p_h(1-\hat p_h)}{n_h-1}.
$$

## Example: ACLS Survey of Scholars

The American Council of Learned Societies (ACLS) used a stratified random
sample of member societies in seven disciplines to study publication patterns
and computer/library use (Morton and Price, 1989).

| Discipline | Membership ($N_h$) | Valid Returns ($n_h$) | Female Members (%) |
|:---|---:|---:|---:|
| Literature | 9,100 | 636 | 38 |
| Classics | 1,950 | 451 | 27 |
| Philosophy | 5,500 | 481 | 18 |
| History | 10,850 | 611 | 19 |
| Linguistics | 2,100 | 493 | 36 |
| Political Science | 5,500 | 575 | 13 |
| Sociology | 9,000 | 588 | 26 |
| **Totals** | **44,000** | **3,835** | |

## Example: ACLS Spreadsheet Calculation

Here $\pi_h = N_h/N$ with $N=44{,}000$, and the last two columns give each
stratum's contribution to $\hat p_{\text{str}}=\sum_h \pi_h\hat p_h$ and to
$\hat V(\hat p_{\text{str}})=\sum_h\big(1-\tfrac{n_h}{N_h}\big)\pi_h^2\tfrac{\hat p_h(1-\hat p_h)}{n_h-1}$:

| Discipline | $N_h$ | $n_h$ | $\pi_h=N_h/N$ | $\hat p_h$ | $\pi_h\hat p_h$ | $\hat v_h$ ($\times10^{-5}$)[^vhp] |
|:---|---:|---:|------:|---:|---:|-----:|
| Literature | 9,100 | 636 | 0.2068 | 0.38 | 0.0786 | 1.476 |
| Classics | 1,950 | 451 | 0.0443 | 0.27 | 0.0120 | 0.066 |
| Philosophy | 5,500 | 481 | 0.1250 | 0.18 | 0.0225 | 0.439 |
| History | 10,850 | 611 | 0.2466 | 0.19 | 0.0469 | 1.448 |
| Linguistics | 2,100 | 493 | 0.0477 | 0.36 | 0.0172 | 0.082 |
| Political Science | 5,500 | 575 | 0.1250 | 0.13 | 0.0163 | 0.276 |
| Sociology | 9,000 | 588 | 0.2045 | 0.26 | 0.0532 | 1.282 |
| **Total** | **44,000** | **3,835** | **1.0000** | | **0.2465** | **5.068** |

[^vhp]: $\hat v_h = \big(1-\tfrac{n_h}{N_h}\big)\pi_h^2\tfrac{\hat p_h(1-\hat p_h)}{n_h-1}$, the stratum-$h$ contribution to $\hat V(\hat p_{\text{str}}) = \sum_{h=1}^7 \hat v_h$.

## Example: ACLS Estimates

Let $N_h$ be the membership figures and $n_h$ the number of valid surveys.
Ignoring nonresponse for now, and estimating the proportion of these scholars
who are female:
$$
\hat p_{\text{str}} = \sum_{h=1}^7 \frac{N_h}{N}\hat p_h
 = \frac{9100}{44{,}000}(0.38) + \cdots + \frac{9000}{44{,}000}(0.26) = 0.2465
$$
$$
\text{SE}(\hat p_{\text{str}}) = \sqrt{\sum_{h=1}^7 \Big(1-\frac{n_h}{N_h}\Big)\Big(\frac{N_h}{N}\Big)^2\frac{\hat p_h(1-\hat p_h)}{n_h-1}} = 0.0071.
$$

To estimate the total number of female scholars, multiply $N=44{,}000$ by these
two numbers.

# Comparison with SRS Formulas

In contrast, if we had taken an SRS of size $n=n_1+\cdots+n_H$ from the whole
population, the sample mean would be
$$
\overline y_{\text{SRS}} = \frac{\sum_{(h,j)\in\mathcal S} y_{hj}}{n}
 = \frac{\overline y_1 n_1 + \cdots + \overline y_H n_H}{n}
 = \overline y_1\frac{n_1}{n} + \cdots + \overline y_H\frac{n_H}{n},
$$

compared with the stratified estimator
$$
\overline y_{\text{str}} = \overline y_1\frac{N_1}{N} + \cdots + \overline y_H\frac{N_H}{N}.
$$

## Comparison with SRS Formulas (cont'd)

Similarly, for proportions:
$$
\hat p_{\text{SRS}} = \sum_{h=1}^H \frac{n_h}{n}\hat p_h = \sum_{h=1}^H \frac{n_h \hat p_h}{n}.
$$

The key difference: $\overline y_{\text{SRS}}$ weights each stratum's sample mean by
its *realized sample fraction* $n_h/n$, which may be influenced by sampling
variation and by non-response rates. $\overline y_{\text{str}}$ instead uses the
*known, fixed* population fraction $N_h/N$ for each stratum, which is what
guarantees unbiasedness for $\overline y_U$.

# Sampling Weights for Stratified Sampling

## Sampling Probabilities for Each Sample

Write the overall stratified sample as $\mathcal S = (\mathcal S_1,\ldots,\mathcal S_H)$,
where $\mathcal S_h$ is the SRS drawn from stratum $h$. Since the strata are
sampled independently,
$$
P(\mathcal S) = P(\mathcal S_1)\times P(\mathcal S_2)\times\cdots\times P(\mathcal S_H)
 = \frac{1}{\binom{N_1}{n_1}}\cdot\frac{1}{\binom{N_2}{n_2}}\cdots\frac{1}{\binom{N_H}{n_H}}.
$$

## Sampling Weights

We introduced the sampling weight $w_i = 1/\pi_i$ for SRS. For unit $j$ in
stratum $h$, the inclusion probability is $\pi_{hj} = n_h/N_h$ (the sampling
fraction in stratum $h$), so the sampling weight is,
$$
w_{hj} = \frac{1}{\pi_{hj}} = \frac{N_h}{n_h}.
$$

Stratified sampling is an example of **unequal-probability sampling**.

The population total can then be written as a weighted sum over *all* sampled
units, regardless of stratum:
$$
\hat t_{\text{str}} = \sum_{h=1}^H N_h \overline y_h = \sum_{h=1}^H \sum_{j\in\mathcal S_h} w_{hj}\, y_{hj}.
$$


## Sampling Weights (cont'd)

The sampling weight can be thought of as the number of population units
represented by the sampled unit. 

The weights sum to the population size:
$$
\sum_{h=1}^H\sum_{j\in\mathcal S_h} w_{hj} = N_1+N_2+\cdots+N_H = N,
$$

so the population mean can also be written
$$
\overline y_{\text{str}} = \frac{\displaystyle\sum_{h=1}^H\sum_{j\in\mathcal S_h} w_{hj}y_{hj}}{\displaystyle\sum_{h=1}^H\sum_{j\in\mathcal S_h} w_{hj}}.
$$

## Example: Weights for the Caribou Survey

| Stratum | $N_h$ | $n_h$ | $w_{hj} = N_h/n_h$ |
|:---:|---:|---:|---:|
| A | 400 | 98 | 4.08 |
| B | 30 | 10 | 3.00 |
| C | 61 | 37 | 1.65 |
| D | 18 | 6 | 3.00 |
| E | 70 | 39 | 1.79 |
| F | 120 | 21 | 5.71 |

Note that using only the weights $w_{hj}$ and sample sizes $n_h$ is not enough
to estimate variances — that requires computing $s_h^2$ separately within each
stratum, since the weights alone don't identify stratum membership.

# Efficiency of Proportional Allocation

## Proportional Allocation

A special case, useful for comparing with SRS: **proportional allocation**
takes the same sampling fraction in every stratum,
$$
\frac{n_h}{N_h} = \frac nN \quad \text{for all } h.
$$

This is not an *optimal* allocation in general, but it is a convenient
benchmark.

## Decomposition of Population Variance

| Source | df | Sum of Squares |
|:---|:---:|:---|
| Between strata | $H-1$ | $\text{SSB} = \sum_{h=1}^H N_h(\overline y_{hU}-\overline y_U)^2$ |
| Within strata | $N-H$ | $\text{SSW} = \sum_{h=1}^H (N_h-1)S_h^2$ |
| Total, about $\overline y_U$ | $N-1$ | $\text{SSTO} = (N-1)S^2$ |
$$
\text{SSTO} = \text{SSB} + \text{SSW}.
$$

## Illustration of SSB and SSW

```{r}
#| label: fig-ssb-ssw
#| fig-width: 7
#| fig-height: 4.2
#| out-width: "75%"
set.seed(21)
means <- c(2.2, 4.6, 7.0)
n_each <- 12
x <- rep(1:3, each = n_each) + runif(3 * n_each, -0.18, 0.18)
y <- rep(means, each = n_each) + rnorm(3 * n_each, 0, 0.6)
grand_mean <- mean(y)

par(mar = c(3.5, 4, 0.5, 0.5))
plot(x, y, pch = 19, col = "blue4", xlim = c(0.5, 3.5), ylim = c(0, 9),
     xaxt = "n", xlab = "Stratum", ylab = "y", cex.lab = 1.2)
axis(1, at = 1:3, labels = c("1", "2", "3"))
for (h in 1:3) {
  segments(h - 0.25, means[h], h + 0.25, means[h], col = "firebrick", lwd = 3)
}
abline(h = grand_mean, col = "darkgreen", lwd = 2, lty = 2)
text(3.4, grand_mean + 0.3, expression(bar(y)[U]), col = "darkgreen", pos = 2)
text(3.35, means[3] + 0.4, expression(bar(y)[HU]), col = "firebrick")
```

Red segments show stratum means (SSW is variability *around* these); the green
dashed line is the overall mean (SSB is variability of the stratum means
*around* it).

## Comparing Variances: Proportional Allocation vs. SRS

For a stratified sample of size $n$ with proportional allocation,
$$
V_{\text{prop}}(\hat t_{\text{str}}) = \sum_{h=1}^H \Big(1-\frac nN\Big)\frac Nn\, N_h S_h^2
 = \Big(1-\frac nN\Big)\frac Nn\left(\text{SSW} + \sum_{h=1}^H S_h^2\right).
$$

Since $\text{SSTO}=\text{SSW}+\text{SSB}$, the variance of the total from an SRS
of size $n$ is.
$$
V_{\text{SRS}}(\hat t) = \Big(1-\frac nN\Big)\frac{N^2 S^2}{n}
 = \Big(1-\frac nN\Big)\frac{N^2}{n(N-1)}(\text{SSW}+\text{SSB}).
$$

## When Does Stratification Help?

Comparing the two expressions, proportional allocation gives smaller variance
than SRS whenever
$$
\text{SSB} > \sum_{h=1}^H \Big(1-\frac{N_h}{N}\Big)S_h^2.
$$

This typically holds because SSB is a sum of $N$ squared deviations, while
$\sum_h S_h^2$ is a sum of only $H$ (typically much fewer) terms.

## The Variance Ratio and $R^2$
$$
\frac{V_{\text{prop}}(\hat t_{\text{str}})}{V(\hat t_{\text{SRS}})}
 \approx \frac{\text{SSW}+\sum_h S_h^2}{\text{SSW}+\text{SSB}} \approx \frac{\text{SSW}}{\text{SSW}+\text{SSB}}
$$
$$
\text{percentage of variance reduction} = \frac{\text{SSB}}{\text{SSW}+\text{SSB}}
 = R^2_{y\mid\text{strata}},
$$

the proportion of variance in $y$ that can be explained by strata membership.

## Coefficient of Determination $R^2$
$$
R^2 = 1-\frac{\text{SSW}}{\text{SST}} = \frac{\text{SSB}}{\text{SST}}
$$

$R^2$ generalizes the squared Pearson correlation coefficient: it measures the
correlation between a response $y$ and one or more predictors
$x_1,\ldots,x_p$ — here, "stratum membership" plays the role of the predictor.
$R^2$ is the proportion of the variance in $y$ explained by the predictors.

## $R^2$ Large vs. $R^2$ Small

```{r}
#| label: fig-r2-compare
#| fig-width: 9
#| fig-height: 4
#| out-width: "90%"
set.seed(7)
n_each <- 15

par(mfrow = c(1, 2), mar = c(3, 3.5, 2, 0.5))

means_hi <- c(2, 5, 8)
x1 <- rep(1:3, each = n_each) + runif(3 * n_each, -0.15, 0.15)
y1 <- rep(means_hi, each = n_each) + rnorm(3 * n_each, 0, 0.35)
plot(x1, y1, pch = 19, col = "blue4", xaxt = "n", xlim = c(0.5, 3.5),
     ylim = c(0, 10), xlab = "", ylab = "y", main = expression(R^2~"large"))
axis(1, at = 1:3, labels = c("1", "2", "3"))
for (h in 1:3) segments(h - 0.25, means_hi[h], h + 0.25, means_hi[h], col = "firebrick", lwd = 3)

means_lo <- c(4.6, 5, 5.4)
x2 <- rep(1:3, each = n_each) + runif(3 * n_each, -0.15, 0.15)
y2 <- rep(means_lo, each = n_each) + rnorm(3 * n_each, 0, 1.4)
plot(x2, y2, pch = 19, col = "blue4", xaxt = "n", xlim = c(0.5, 3.5),
     ylim = c(0, 10), xlab = "", ylab = "y", main = expression(R^2~"small"))
axis(1, at = 1:3, labels = c("1", "2", "3"))
for (h in 1:3) segments(h - 0.25, means_lo[h], h + 0.25, means_lo[h], col = "firebrick", lwd = 3)
```

When the strata means differ widely relative to the within-stratum spread
(left), stratification explains a large share of the variance in $y$.

# Optimal Allocation

## Setting Up the Cost Function

The objective in optimal allocation is to gain the most information for the
least cost. Let $C$ be total cost, $c_0$ overhead cost, and $c_h$ the cost of
sampling one unit in stratum $h$:
$$
C = c_0 + \sum_{h=1}^H c_h n_h, \qquad n = n_1+\cdots+n_H.
$$

We want to allocate $n_1,\ldots,n_H$ to minimize $V(\overline y_{\text{str}})$ for a
fixed total cost $C$ (or equivalently minimize $C$ for fixed $V$).

## Optimal Allocation Formula

Using the method of Lagrange multipliers, the optimal $n_h$ is proportional to
$N_hS_h/\sqrt{c_h}$:
$$
n_h = \left(\frac{N_hS_h/\sqrt{c_h}}{\sum_{l=1}^H N_lS_l/\sqrt{c_l}}\right)n
\qquad\text{i.e.}\qquad n_h \propto \frac{N_hS_h}{\sqrt{c_h}}.
$$

We sample more heavily within a stratum if: the stratum is a large part of the
population ($N_h$ large); the stratum is more variable ($S_h$ large); or
sampling in the stratum is cheap ($c_h$ small).

**Special cases:** if $c_1=\cdots=c_H$, then $n_h\propto N_hS_h$
(**Neyman allocation**); if additionally $S_1=\cdots=S_H$, then $n_h\propto N_h$
(**proportional allocation**).

## Example: Neyman Allocation for the Caribou Survey

The investigators used rough estimates of caribou counts as a proxy for $S_h$,
with total sample size $n=225$:

| Stratum | $N_h$ | $s_h$ | $N_hs_h$ | $n_h$ (Neyman) | Sample size used |
|:---:|---:|---:|---:|---:|---:|
| A | 400 | 3,000 | 1,200,000 | 96.26 | 98 |
| B | 30 | 2,000 | 60,000 | 4.81 | 10 |
| C | 61 | 9,000 | 549,000 | 44.04 | 37 |
| D | 18 | 2,000 | 36,000 | 2.89 | 6 |
| E | 70 | 12,000 | 840,000 | 67.38 | 39 |
| F | 120 | 1,000 | 120,000 | 9.63 | 21 |
| **Total** | **699** | | **2,805,000** | **225** | **211** |

They wanted the sampling fraction to be at least $1/3$ in the smaller strata,
so the Neyman values (column $n_h$) were used only as a guideline.

## Example: Optimal Allocation for Polls

For estimating a proportion, $S_h^2 = p_h(1-p_h)$. As a function of $p$,
$S^2(p)=p(1-p)$ is a quadratic maximized at $p=1/2$. So election regions (swing
states) with $p$ near 0.5 should be assigned relatively larger sample sizes,
since they contribute the most variance per unit sampled.

## Example: Dollar Stratification

**Dollar stratification** is common in accounting audits. Recorded book
amounts stratify the population: stratum 1 might be all loans over \$1 million,
stratum 2 loans between \$500,000 and \$999,999, and so on down to loans under
\$10,000.

Optimal allocation is efficient here: $S_h$ is much larger in strata with large
loan amounts, so optimal allocation prescribes a higher sampling fraction for
those strata. An error in a \$3,000,000 loan is likely to contribute far more
to the total audited discrepancy than an error in a \$3,000 loan.

# Choice of Stratifying Variables

## Using an Auxiliary Variable

Stratification is most efficient when the stratum means differ widely, since
then SSB is large and within-stratum variability is small. Ideally we would
stratify directly by the values of $y$; since $y$ is unknown before sampling,
we instead look for an auxiliary variable $x$ that is highly correlated with
(or predictive of) $y$.

```{r}
#| label: fig-auxiliary
#| fig-width: 9
#| fig-height: 4.2
#| out-width: "90%"
set.seed(4)
n <- 60
x <- runif(n, 0, 10)
y <- 2 * x + rnorm(n, 0, 3)
cuts <- quantile(x, probs = c(1/3, 2/3))
strat <- cut(x, breaks = c(-Inf, cuts, Inf), labels = c("1", "2", "3"))

par(mfrow = c(1, 2), mar = c(4, 4, 1, 0.5))
plot(x, y, pch = 19, col = "blue4", xlab = "x", ylab = "y")
abline(lm(y ~ x), col = "firebrick", lwd = 2)
abline(v = cuts, col = "darkgreen", lty = 2, lwd = 2)
text(cuts, rep(max(y), 2), c("St 1|2", "St 2|3"), col = "darkgreen", pos = 2, cex = 0.8)

boxplot(y ~ strat, col = "lightblue", xlab = "Stratum", ylab = "y")
```

Cutting on $x$ (left) produces strata with visibly different distributions of
$y$ (right) — exactly the goal of stratification.

## Example: Choosing an Auxiliary Variable

- To estimate total business expenditures on advertising, we would like
  businesses that spend the most on advertising in stratum 1, the next highest
  spenders in stratum 2, and so on. We might stratify by number of employees
  or size of business, and by type of product or service.

- For farm income, we might use farm size as the stratifying variable, since we
  expect larger farms to have higher incomes.





## Limitations of Stratified Sampling

- **Difficulty in implementation.** We need to know the stratum sizes $N_h$
  (and, for optimal allocation, rough estimates of $S_h$) *before* sampling.

- **Multiple variables of interest.** A single stratifying variable that works
  well for one $y$ may not work well for another $y$ measured in the same
  survey — a variable highly correlated with one outcome may be nearly
  uncorrelated with a different outcome of interest.

