---
title: "Prediction Theory for Sampling Survey"
author: "Longhai Li"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
engine: knitr
execute:
  warning: false
  message: false
  cache: false
  freeze: false
---

Sampling is a prediction problem. The sample plays the role of a training set, the non-sampled units play the role of test cases, and a population total is the sum of the observed values and the predictions for the test cases. This chapter states the framework once, in general, and then derives the mean squared prediction error of the estimator of a total for simple random, stratified, ratio and regression estimation as special cases of a single formula.

## Sampling as prediction {#sec-setup}

Let $U=\{1,\dots,N\}$ index a finite population with study variable $Y_i$, let $S\subset U$ with $|S|=n$ denote the observed sample, and let $R=U\setminus S$ denote the $N-n$ non-sampled units. The quantities of interest are
$$
T=\sum_{i\in U}Y_i=\sum_{i\in S}Y_i+\sum_{i\in R}Y_i,\qquad \bar Y_U=T/N .
$$
Only the second sum is unknown, so the task is to predict $Y_i$ for each $i\in R$.

The prediction approach treats $Y_1,\dots,Y_N$ as random variables and $S$ as fixed. All expectations are taken with respect to a model $\xi$ rather than the randomization distribution. The sampling design enters only through the requirement that selection be uninformative, so that $\xi$ governs the units in $R$ as well as those in $S$. Any auxiliary variables $x_i$ are treated as known constants. The general working model is
$$
\xi:\quad E(Y_i)=\mu_i(\theta),\quad \operatorname{Var}(Y_i)=\sigma_i^2,\quad
\operatorname{Cov}(Y_i,Y_j)=0\ (i\neq j),
$$ {#eq-model}
where $\theta$ is an unknown parameter and $\mu_i(\theta)$ is the mean function, which may depend on $x_i$. The model errors are
$$
e_i=Y_i-\mu_i(\theta),\qquad E(e_i)=0,\quad E(e_i^2)=\sigma_i^2,\quad E(e_ie_j)=0\ (i\neq j).
$$ {#eq-errors}

A prediction rule estimates $\theta$ from the sample by $\hat\theta=\hat\theta(Y_S)$ and predicts
$$
\hat Y_i=
\begin{cases}
Y_i, & i\in S,\\
\mu_i(\hat\theta), & i\in R,
\end{cases}
\qquad
\hat T=\sum_{i\in U}\hat Y_i=\sum_{i\in S}Y_i+\sum_{i\in R}\mu_i(\hat\theta).
$$ {#eq-predictor}
Observed values are kept as they are, and the fitted mean function is used only for the unobserved units. We take $\hat\theta$ to be a linear function of $Y_S$ (given the $x_i$), as it is for every estimator in this chapter, so that $\hat\theta$ is uncorrelated with $e_i$ for each $i\in R$.

## Two ways to judge a prediction {#sec-two-metrics}

Add and subtract $\mu_i(\theta)$ in the error for an unobserved unit $i\in R$:
$$
\hat Y_i-Y_i=\underbrace{\mu_i(\hat\theta)-\mu_i(\theta)}_{\delta_i}-\underbrace{\big(Y_i-\mu_i(\theta)\big)}_{e_i}.
$$ {#eq-error-individual}
There are two sources of error. The term $e_i$ is the gap between the unobserved $Y_i$ and its true mean, which no predictor can remove. The term $\delta_i$ is the gap between the fitted mean and the true mean, caused by the uncertainty in $\hat\theta$.

**Traditional machine learning.** A predictive model is judged by the mean squared prediction error of a single case, $E(\hat Y_i-Y_i)^2$, and by its sum or average over the test cases. Because $\delta_i$ is uncorrelated with $e_i$,
$$
E(\hat Y_i-Y_i)^2=\sigma_i^2+E(\delta_i^2),
\qquad
\frac1{N-n}\sum_{i\in R}E(\hat Y_i-Y_i)^2=\frac1{N-n}\sum_{i\in R}\sigma_i^2+\frac1{N-n}\sum_{i\in R}E(\delta_i^2).
$$ {#eq-mse-individual}
Only the diagonal terms $E(\delta_i^2)$ enter. This is the familiar split into irreducible noise and estimation error.

**Sampling survey.** The target is the total, and the quantity of interest is the mean squared prediction error of $\hat T$,
$$
\operatorname{MSPE}(\hat T)=E\big[(\hat T-T)^2\big]=E\Big[\Big(\sum_{i\in R}\hat Y_i-Y_i\Big)^{2}\Big],
\qquad
\operatorname{MSPE}(\hat{\bar y}_U)=\frac{\operatorname{MSPE}(\hat T)}{N^2}.
$$ {#eq-mspe-def}
The observed units contribute no error, so by @eq-error-individual
$$
\hat T-T=\sum_{i\in R}\delta_i-\sum_{i\in R}e_i .
$$ {#eq-error-decomp}
The cross term between $\sum_R\delta_i$ and $\sum_Re_i$ vanishes, and the $e_i$ are uncorrelated, so
$$
\operatorname{MSPE}(\hat T)
=\underbrace{\sum_{i\in R}\sigma_i^2}_{\text{scatter of the unobserved }Y_i}
+\underbrace{E\Big[\Big(\sum_{i\in R}\delta_i\Big)^{2}\Big]}_{\text{uncertainty in }\hat\theta} .
$$ {#eq-mspe}

The difference from @eq-mse-individual is that the square is taken *after* summing. Expanding,
$$
E\Big[\Big(\sum_{i\in R}\delta_i\Big)^{2}\Big]=\sum_{i\in R}E(\delta_i^2)+\sum_{i\neq j\in R}E(\delta_i\delta_j),
$$
and the cross terms are the new ingredient. They are not negligible: all $\delta_i$ are functions of the same $\hat\theta$, so they are correlated. If the average pairwise correlation is $\bar\rho$, the cross terms are of order $(N-n)^2\bar\rho\,E(\delta^2)$ against $(N-n)E(\delta^2)$ for the diagonal, and $\bar\rho$ does not shrink with $N$ for a model with a fixed number of parameters. The estimation error of a total is therefore dominated by the shared error in $\hat\theta$, which does not average out across units.

## The general result {#sec-general}

Suppose $\mu_i(\theta)=x_i^{\top}\theta$ is linear in $\theta$, which covers all the cases below. Then
$$
\sum_{i\in R}\delta_i=\Big(\sum_{i\in R}x_i\Big)^{\top}(\hat\theta-\theta)=t_{x,R}^{\top}(\hat\theta-\theta),
\qquad t_{x,R}=\sum_{i\in R}x_i=t_x-t_{x,S},\quad t_x=\sum_{i\in U}x_i,\ t_{x,S}=\sum_{i\in S}x_i ,
$$
a single linear contrast of the estimation error, and @eq-mspe becomes
$$
\operatorname{MSPE}(\hat T)=\sum_{i\in R}\sigma_i^2+t_{x,R}^{\top}\operatorname{MSE}(\hat\theta)\,t_{x,R} .
$$ {#eq-mspe-general}
For a nonlinear mean function the same holds approximately, with $t_{x,R}$ replaced by $\sum_R\partial\mu_i/\partial\theta$ (delta method).

For the working model $\sigma_i^2=\sigma^2v_i$ with known weights $v_i$, the weighted least squares estimator
$$
\hat\theta=\big(X_S^{\top}V_S^{-1}X_S\big)^{-1}X_S^{\top}V_S^{-1}y_S,\qquad V_S=\operatorname{diag}(v_i)_{i\in S},
$$
is unbiased with $\operatorname{Var}(\hat\theta)=\sigma^2(X_S^{\top}V_S^{-1}X_S)^{-1}$, and
$$
\boxed{\;\operatorname{MSPE}(\hat T)=\sigma^2\Big[\sum_{i\in R}v_i+t_{x,R}^{\top}\big(X_S^{\top}V_S^{-1}X_S\big)^{-1}t_{x,R}\Big]\;}
$$ {#eq-mspe-wls}
The first term is the scatter of the unobserved units, and the second is the cost of estimating $\theta$ from the sample. The cost is a quadratic form in $t_{x,R}$, so it is small only when the sample supplies a precise estimate in the direction in which the unobserved units need it. A precise $\hat\theta$ is not enough: the sample must also resemble the population in the directions $t_{x,R}$ points to. Each estimator below corresponds to a choice of $x_i$ and $v_i$.

In all of these cases $\hat T=\sum_SY_i+\sum_Rx_i^{\top}\hat\theta$, and the estimate of $\sigma^2$ is the weighted residual mean square $\hat\sigma^2=\sum_Sv_i^{-1}(Y_i-x_i^{\top}\hat\theta)^2/(n-p)$ with $p=\dim\theta$, which gives the variance estimator on substituting into @eq-mspe-wls.

## Simple random sampling {#sec-srs}

With no auxiliary variable every unit has the same mean,
$$
\mu_i(\theta)=\mu,\quad x_i=1,\quad v_i=1,\qquad e_i=Y_i-\mu,\quad \bar e_A=|A|^{-1}\sum_{i\in A}e_i,
$$
and $\hat\mu=\bar y_S$. The predictor is
$$
\hat Y_i=
\begin{cases}
Y_i, & i\in S,\\
\bar y_S, & i\in R,
\end{cases}
\qquad
\hat T=\sum_{i\in S}Y_i+(N-n)\bar y_S=N\bar y_S ,
$$ {#eq-expansion}
so the expansion estimator is a prediction rule that fills in every unobserved unit with the sample mean. The estimation error is common to all units, $\delta_i=\bar y_S-\mu=\bar e_S$, so the pairwise correlation of the $\delta_i$ is exactly one and
$$
\hat T-T=(N-n)\,\bar e_S-\sum_{i\in R}e_i .
$$ {#eq-error-srs}
Both terms have mean zero, so $\hat T$ is model-unbiased. With $t_{x,R}=N-n$ and $X_S^{\top}X_S=n$, @eq-mspe-wls gives
$$
\operatorname{MSPE}(\hat T)=\sigma^2\Big[(N-n)+\frac{(N-n)^2}{n}\Big]=\frac{N(N-n)}{n}\,\sigma^2 ,
$$ {#eq-mspe-total}
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\frac{\sigma^2}{n}\left(1-\frac{n}{N}\right).
$$ {#eq-mspe-mean}

Expression @eq-mspe-mean is the classical simple-random-sampling variance with $\sigma^2$ in place of the population variance $S^2$, obtained without appeal to randomization. Several consequences follow.

1. The two terms of @eq-mspe-total are the two sources of error. The term $(N-n)\sigma^2$ is the irreducible scatter of the unobserved units; the term $(N-n)^2\sigma^2/n$ is the cost of estimating $\mu$ from $n$ observations and alone decreases with $n$.
2. The per-unit error of @eq-mse-individual is $\sigma^2(1+1/n)$, barely above $\sigma^2$ for moderate $n$, so the estimation error looks negligible for individual prediction. For the total it is not: it is shared by all $N-n$ unobserved units and accumulates to a factor $(N-n)^2$, whereas the scatter accumulates only to $N-n$. The ratio of the two terms is $(N-n)/n$.
3. The finite-population correction is not an artifact of sampling without replacement. The $n$ observed units contribute exactly zero to @eq-error-srs, leaving $N-n$ terms to predict; at $n=N$ the MSPE is zero.
4. If $\mu$ is known, $\delta_i=0$ and the predictor $\hat Y_i=\mu$ for $i\in R$ has $\operatorname{MSPE}(\hat T)=(N-n)\sigma^2$, the irreducible error. The cost of not knowing $\mu$ is the factor $N/n$ applied to it.

@fig-pred-index displays one realization of a population under @eq-model with $\mu_i=\mu$. Since the labels are uninformative, the sample is placed on the left, shaded grey, and the units to be predicted on the right.

```{r}
#| echo: false
#| label: fig-pred-index
#| fig-cap: "One realized population under the model. The sample occupies the shaded region on the left; the remaining units are to be predicted. Horizontal lines: the model mean $\\mu$ (orange), the estimand $\\bar Y_U$ (red), and the predictor $\\bar y_S$ (blue, dashed)."
#| fig-width: 7.5
#| fig-height: 4.2

set.seed(20000)
N <- 100; n <- 10; mu <- 50; sigma <- 20

y <- rnorm(N, mean = mu, sd = sigma)   # one realization of Y_1, ..., Y_N
S <- 1:n                               # sample, placed on the left
R <- (n + 1):N                         # non-sampled units

ybar_S <- mean(y[S])                   # predictor of mu
ybar_U <- mean(y)                      # the estimand

ylim <- range(y) + c(-1, 1) * 0.10 * diff(range(y))
plot(seq_len(N), y, type = "n", ylim = ylim,
     xlab = "unit index", ylab = expression(y[i]))
rect(0.5, ylim[1], n + 0.5, ylim[2], col = "grey90", border = NA)
points(S, y[S], pch = 16, cex = 0.7)
points(R, y[R], pch = 1,  cex = 0.7, col = "grey45")
abline(h = mu,     col = "orange", lwd = 2)
abline(h = ybar_U, col = "red",    lwd = 2)
abline(h = ybar_S, col = "blue",   lwd = 2, lty = 2)
text(n / 2,       ylim[2], "observed (S)",        pos = 1, cex = 0.85)
text((n + N) / 2, ylim[2], "to be predicted (R)", pos = 1, cex = 0.85)
legend("bottomright", bty = "n", cex = 0.85, lwd = 2,
       col = c("orange", "red", "blue"), lty = c(1, 1, 2),
       legend = c(expression(mu), expression(bar(Y)[U]), expression(bar(y)[S])))
box()
```

The scatter of the open points about the orange line is the irreducible error. The gap between the orange and blue lines is the estimation error $\delta_i=\bar e_S$, with root mean square $\sigma/\sqrt{n}$, the same for every unit in $R$. The gap between the blue and red lines, scaled by $N$, is the realized prediction error of the total; @eq-mspe-total states that its root mean square is $\sigma\sqrt{N(N-n)/n}$.

## Stratified sampling {#sec-strat}

Let the population be divided into strata $h=1,\dots,H$ of sizes $N_h$, with $n_h$ units sampled from stratum $h$, $R_h$ the non-sampled units and $S_h$ the sampled units. The model gives each stratum its own mean and variance,
$$
\mu_i(\theta)=\mu_h,\quad \sigma_i^2=\sigma_h^2,\qquad i\in\text{stratum }h,
$$
so $x_i$ is the vector of stratum indicators and $\theta=(\mu_1,\dots,\mu_H)$. Then $\hat\mu_h=\bar y_{S_h}$, and
$$
\hat T=\sum_{h=1}^HN_h\bar y_{S_h},\qquad
\hat{\bar y}_U=\sum_{h=1}^HW_h\bar y_{S_h},\quad W_h=N_h/N .
$$
The vector $t_{x,R}$ has components $N_h-n_h$, the matrix $X_S^{\top}X_S$ is diagonal with entries $n_h$, and the $\hat\mu_h$ are independent across strata because they depend on disjoint, uncorrelated data. The correlated errors $\delta_i$ therefore occur only *within* a stratum, and the cross terms between strata vanish. Applying @eq-mspe-general stratum by stratum,
$$
\operatorname{MSPE}(\hat T)=\sum_{h=1}^H\sigma_h^2\Big[(N_h-n_h)+\frac{(N_h-n_h)^2}{n_h}\Big]=\sum_{h=1}^H\frac{N_h(N_h-n_h)}{n_h}\,\sigma_h^2 ,
$$ {#eq-mspe-strat}
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\sum_{h=1}^HW_h^2\Big(1-\frac{n_h}{N_h}\Big)\frac{\sigma_h^2}{n_h}.
$$ {#eq-mspe-strat-mean}
This is the classical stratified variance, with $\sigma_h^2$ playing the role of $S_h^2$. It is the sum of the $H$ separate SRS results of @eq-mspe-total, each with its own shared estimation error. Substituting $\hat\sigma_h^2=s_h^2$ gives the usual variance estimator.

Two allocations follow immediately from @eq-mspe-strat-mean. Under proportional allocation $n_h=nW_h$,
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\Big(\frac1n-\frac1N\Big)\sum_{h=1}^HW_h\sigma_h^2 ,
$$
the SRS result with $\sigma^2$ replaced by the average within-stratum variance, so that stratification helps whenever the strata means differ. Minimizing @eq-mspe-strat-mean over $n_h$ subject to $\sum n_h=n$ gives Neyman allocation $n_h\propto N_h\sigma_h$, with $\operatorname{MSPE}(\hat{\bar y}_U)=\frac1n\big(\sum_hW_h\sigma_h\big)^2-\frac1N\sum_hW_h\sigma_h^2$.

## Ratio estimation {#sec-ratio}

Let $x_i>0$ be an auxiliary variable whose population total $t_x=\sum_Ux_i$ is known. The ratio model is
$$
\mu_i(\beta)=\beta x_i,\qquad \sigma_i^2=\sigma^2x_i ,
$$
that is $x_i$ in the mean and $v_i=x_i$ in the variance. Then $X_S^{\top}V_S^{-1}X_S=\sum_Sx_i=t_{x,S}$, and
$$
\hat\beta=\frac{\sum_{i\in S}Y_i}{\sum_{i\in S}x_i}=\frac{\bar y_S}{\bar x_S},
\qquad
\hat T=\sum_{i\in S}Y_i+\hat\beta\,t_{x,R}=\hat\beta\,t_x ,
$$
because $\sum_SY_i=\hat\beta\,t_{x,S}$. This is the ratio estimator of the total. The shared error is $\delta_i=x_i(\hat\beta-\beta)$, so $\sum_R\delta_i=t_{x,R}(\hat\beta-\beta)$, with $t_{x,R}=t_x-t_{x,S}$ and $\sum_Rv_i=t_{x,R}$. @eq-mspe-wls gives
$$
\operatorname{MSPE}(\hat T)=\sigma^2\Big[t_{x,R}+\frac{t_{x,R}^2}{t_{x,S}}\Big]
=\sigma^2\,\frac{t_x\,(t_x-t_{x,S})}{t_{x,S}} .
$$ {#eq-mspe-ratio}
Write $t_{x,S}=n\bar x_S$ and $t_x=N\bar x_U$. Then
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\frac{\sigma^2\,\bar x_U}{n\,\bar x_S}\Big(\bar x_U-\frac nN\bar x_S\Big).
$$ {#eq-mspe-ratio-mean}
In a balanced sample, $\bar x_S=\bar x_U$, this reduces to $\frac{\sigma^2\bar x_U}{n}(1-n/N)$: the SRS form with $\sigma^2\bar x_U$, the variance of an average-size unit, in place of $\sigma^2$. When $\bar x_S$ is below $\bar x_U$ the MSPE is inflated, because the estimate of $\beta$ is then extrapolated to the larger unobserved units. The variance estimator uses $\hat\sigma^2=\sum_S(Y_i-\hat\beta x_i)^2/x_i\big/(n-1)$.

## Regression estimation {#sec-reg}

With the same auxiliary variable, the straight-line model with constant variance is
$$
\mu_i(\theta)=\beta_0+\beta_1x_i,\qquad \sigma_i^2=\sigma^2,
$$
so $x_i=(1,x_i)^{\top}$ and $v_i=1$. The least squares fit gives
$$
\hat T=\sum_{i\in S}Y_i+\sum_{i\in R}(\hat\beta_0+\hat\beta_1x_i)=N\big[\bar y_S+\hat\beta_1(\bar x_U-\bar x_S)\big],
$$
the regression estimator, because the residuals in $S$ sum to zero. Here $t_{x,R}=(N-n)(1,\bar x_R)^{\top}$ and $X_S^{\top}X_S$ has the inverse that gives the standard leverage formula
$$
(1,\bar x_R)\,(X_S^{\top}X_S)^{-1}(1,\bar x_R)^{\top}=\frac1n+\frac{(\bar x_R-\bar x_S)^2}{S_{xx,S}},
\qquad S_{xx,S}=\sum_{i\in S}(x_i-\bar x_S)^2 .
$$
Moreover $(N-n)(\bar x_R-\bar x_S)=N(\bar x_U-\bar x_S)$, so @eq-mspe-wls becomes
$$
\operatorname{MSPE}(\hat T)=\sigma^2\Big[(N-n)+\frac{(N-n)^2}{n}+\frac{N^2(\bar x_U-\bar x_S)^2}{S_{xx,S}}\Big]
=\sigma^2\Big[\frac{N(N-n)}{n}+\frac{N^2(\bar x_U-\bar x_S)^2}{S_{xx,S}}\Big],
$$ {#eq-mspe-reg}
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\sigma^2\Big[\frac1n\Big(1-\frac nN\Big)+\frac{(\bar x_U-\bar x_S)^2}{S_{xx,S}}\Big].
$$ {#eq-mspe-reg-mean}
The first term is the SRS result @eq-mspe-mean with $\sigma^2$ now the *residual* variance about the line, which is smaller than the marginal variance of $Y$ when $x$ is informative. The second term is a penalty for an unbalanced sample, which vanishes when $\bar x_S=\bar x_U$. The variance estimator uses $\hat\sigma^2=\sum_S(Y_i-\hat\beta_0-\hat\beta_1x_i)^2/(n-2)$.

## Summary {#sec-summary}

| Estimator | $\mu_i(\theta)$ | $v_i$ | $t_{x,R}$ | $\operatorname{MSPE}(\hat T)/\sigma^2$ |
|:--|:--|:--|:--|:--|
| SRS | $\mu$ | $1$ | $N-n$ | $\dfrac{N(N-n)}{n}$ |
| Stratified | $\mu_h$ | $1$ (per stratum $\sigma_h^2$) | $(N_h-n_h)_h$ | $\sum_h\dfrac{N_h(N_h-n_h)}{n_h}$ |
| Ratio | $\beta x_i$ | $x_i$ | $t_{x,R}$ | $\dfrac{t_x(t_x-t_{x,S})}{t_{x,S}}$ |
| Regression | $\beta_0+\beta_1x_i$ | $1$ | $(N-n)(1,\bar x_R)$ | $\dfrac{N(N-n)}{n}+\dfrac{N^2(\bar x_U-\bar x_S)^2}{S_{xx,S}}$ |

: Special cases of @eq-mspe-wls. {#tbl-summary}

In every case the MSPE is the scatter of the unobserved units, $\sum_R\sigma_i^2$, plus a cost of estimating $\theta$ that is a quadratic form in the total auxiliary information $t_{x,R}$ still to be predicted. The ratio and regression estimators do not remove the shared error; they change the model so that $\sigma^2$ is the residual variance rather than the marginal one, and they pay a penalty when the sample is unrepresentative in $x$.

## Numerical check {#sec-check}

The formulas can be checked by simulating many populations from the model, drawing a sample, and comparing the empirical mean squared error of $\hat T-T$ with the formula. The sample is fixed across replications, which is the conditioning of the prediction approach.

```{r}
set.seed(1)
N <- 200; n <- 20; B <- 20000
x   <- runif(N, 1, 10)                 # known auxiliary variable
S   <- sample(N, n)                    # one fixed sample
R   <- setdiff(seq_len(N), S)
strat <- rep(1:2, each = N / 2)        # two strata of equal size
sg  <- c(1, 3)                         # stratum sd's
nh  <- table(factor(strat[S], 1:2)); Nh <- table(strat)

sim <- function(gen, est) {
  err <- replicate(B, { y <- gen(); est(y) - sum(y) })
  mean(err^2)
}

## SRS: Y ~ (mu, sigma^2)
sigma <- 2
srs_emp <- sim(function() rnorm(N, 10, sigma), function(y) N * mean(y[S]))
srs_th  <- sigma^2 * N * (N - n) / n

## Stratified: stratum means and sd's differ
str_emp <- sim(function() rnorm(N, c(10, 20)[strat], sg[strat]),
               function(y) sum(Nh * tapply(y[S], factor(strat[S], 1:2), mean)))
str_th  <- sum(sg^2 * Nh * (Nh - nh) / nh)

## Ratio: mean beta * x, variance sigma^2 * x
beta <- 2
rat_emp <- sim(function() rnorm(N, beta * x, sigma * sqrt(x)),
               function(y) sum(x) * sum(y[S]) / sum(x[S]))
rat_th  <- sigma^2 * sum(x) * (sum(x) - sum(x[S])) / sum(x[S])

## Regression: straight line, constant variance
reg_emp <- sim(function() rnorm(N, 3 + beta * x, sigma),
               function(y) { f <- lm(y[S] ~ x[S]); N * (mean(y[S]) + coef(f)[2] * (mean(x) - mean(x[S]))) })
reg_th  <- sigma^2 * (N * (N - n) / n + N^2 * (mean(x) - mean(x[S]))^2 / sum((x[S] - mean(x[S]))^2))

round(rbind(SRS = c(simulated = srs_emp, formula = srs_th),
            Stratified = c(str_emp, str_th),
            Ratio = c(rat_emp, rat_th),
            Regression = c(reg_emp, reg_th)))
```
