# Ratio, Regression, and Domain Estimation

## Key Formulas

When auxiliary information (a variable $x$) is highly correlated with our target variable of interest ($y$), we can use **Ratio** or **Regression** estimation to improve precision without increasing the sample size. We can also use auxiliary population structures to adjust our estimates after the fact via **Post-stratification**, or estimate means for specific sub-populations (**Domains**).

#### 1. Ratio Estimation {-}
If the relationship between $y$ and $x$ passes through the origin, we estimate the ratio $B$ and use it to adjust the mean:

*   **Ratio:** $\hat{B} = \frac{\overline{y}}{\overline{x}}$
*   **Mean Estimator:** $\overline{y}_r = \hat{B} \overline{x}_{\text{U}}$ (where $\overline{x}_{\text{U}}$ is the true population mean of $x$)

*   **Estimated Variance:** $\hat{V}(\overline{y}_r) = \left(1 - \frac{n}{N}\right) \frac{s_e^2}{n}$, where $s_e^2$ is the sample variance of the residuals $e_i = y_i - \hat{B}x_i$.

#### 2. Regression Estimation {-}
If the relationship does not pass through the origin, a linear regression is more appropriate:

*   **Mean Estimator:** $\overline{y}_{reg} = \hat{B}_0 + \hat{B}_1 \overline{x}_{\text{U}}$ (where $\hat{B}_0$ and $\hat{B}_1$ are OLS coefficients)
*   **Estimated Variance:** $\hat{V}(\overline{y}_{reg}) = \left(1 - \frac{n}{N}\right) \frac{s_e^2}{n}$, where $s_e^2$ is the Mean Square Error (Residuals) from the regression.

#### 3. Domain Estimation {-}
To estimate the mean of a subgroup (domain $d$) when the domain sample size $n_d$ is a random variable:

*   **Mean Estimator:** $\overline{y}_d = \frac{1}{n_d} \sum_{i \in U_d} y_i$

#### 4. Post-stratification {-}
If we know the true population sizes ($N_h$) of strata, but we used a Simple Random Sample, we can post-stratify to reduce variance. We substitute $n_h$ with expected proportional sample sizes:

*   $n_h^{post-strat} = n \frac{N_h}{N}$

---

## Functions and packages for Analyzing Data

We begin by loading the necessary libraries, including `gt` for formatting mathematical tables, and defining our custom functions for Ratio, Regression, Domain, and Post-stratified estimation. We also define a versatile `gt` helper to render our estimation outputs cleanly.

### General functions for ratio and regression estimation

```{r}
#| code-fold: true
library(latex2exp)
library(gt)

## Formats a gt table column as ordinary fixed-point, UNLESS its values are
## too small (< `small` in absolute value) or too large (>= `large`), in
## which case it switches that column to scientific notation instead.
fmt_auto <- function (gt_tbl, data, columns, decimals = 3, small = 1e-3, large = 1e5)
{
    for (col in columns)
    {
        x <- data[[col]]
        x <- x[is.finite (x) & x != 0]
        if (length (x) == 0) next
        if (any (abs (x) < small) || any (abs (x) >= large))
            gt_tbl <- fmt_scientific (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
        else
            gt_tbl <- fmt_number (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
    }
    gt_tbl
}

## Formats a single scalar for inline formula display (fixed-point with
## thousands separators, switching to scientific notation for very small or
## very large magnitudes, mirroring fmt_auto()) --- used to show worked-example
## steps such as \bar y = sum/n with the actual numbers plugged in.
fmt_num <- function (x, decimals = 3, small = 1e-3, large = 1e5)
{
    use_sci <- is.finite (x) & x != 0 & (abs (x) < small | abs (x) >= large)
    ifelse (use_sci,
            formatC (x, format = "e", digits = decimals),
            formatC (x, format = "f", digits = decimals, big.mark = ","))
}

## Truncates a row-level working data.frame whose LAST row is a summary row
## (e.g. "Sum"/"Total") to the first head_n data rows + "..." + the summary
## row, so long working tables stay readable. id_col --- row-label column.
truncate_working <- function (working, id_col, head_n = 6)
{
    n <- nrow (working) - 1
    if (n <= head_n + 1) return (working)

    dots <- working[1, ]
    dots[1, ] <- NA
    dots[1, id_col] <- "..."

    rbind (working[seq_len (head_n), ], dots, working[nrow (working), ])
}

## ydata --- observations of the variable of interest
## xdata --- observations of the auxiliary variable
## xbarU --- population mean of the auxiliary variable; required when
##           estimate is "mean" or "total"; ignored when estimate = "model"
## N --- population size
## estimate --- "mean" (default) returns the regression estimate of the
##              population MEAN of y, Bhat0 + Bhat1*xbarU; "total" returns
##              the regression estimate of the population TOTAL of y
##              (requires finite N); "model" returns the raw fitted slope
##              Bhat1 itself (with its own SE from the lm fit)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values y_i, x_i, fitted yhat_i, e_i,
##                  e_i^2, with a Sum row and a footnote showing Bhat0,
##                  Bhat1, and s_e^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_reg_est_mean <- function (ydata, xdata, xbarU = NULL, N = Inf,
                               estimate = c ("mean", "total", "model"), show.details = TRUE)
{
  estimate <- match.arg (estimate)

  n <- length (ydata)
  lmfit <- lm (ydata ~ xdata)
  Bhat <- lmfit$coefficients
  yhat <- lmfit$fitted.values
  e <- lmfit$residuals
  SSe <- sum (e^2) / (n - 2)

  if (estimate == "model")
  {
      est <- unname (Bhat[2])
      sd_est <- summary (lmfit)$coefficients[2, "Std. Error"]
  }
  else
  {
      if (is.null (xbarU))
          stop ("xbarU (population mean of x) is required when estimate is \"mean\" or \"total\"")
      yhat_reg <- unname (Bhat[1] + Bhat[2] * xbarU)
      se_yhat_reg <- sqrt ((1-n/N) * SSe / n)
      if (estimate == "total")
      {
          if (!is.finite (N))
              stop ("N must be finite to compute estimate = \"total\"")
          est <- yhat_reg * N
          sd_est <- se_yhat_reg * N
      }
      else
      {
          est <- yhat_reg
          sd_est <- se_yhat_reg
      }
  }

  mem <- qt (0.975, df = n - 2) * sd_est
  estimate_vec <- c (Est. = est, S.E. = sd_est, ci.low = est - mem, ci.upp = est + mem)

  if (!show.details)
      return (list (estimate = estimate_vec, table = NULL))

  Sxx <- sum ((xdata - mean (xdata))^2)
  Sxy <- sum ((xdata - mean (xdata)) * (ydata - mean (ydata)))

  working <- data.frame (
      i    = c (seq_len (n), "Sum"),
      y    = c (ydata, sum (ydata)),
      x    = c (xdata, sum (xdata)),
      yhat = c (yhat, sum (yhat)),
      e    = c (e, sum (e)),
      e2   = c (e^2, sum (e^2))
  )
  working <- truncate_working (working, id_col = "i")

  table <- gt (working) |>
      tab_header (title = "Regression Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
      cols_label (
          i    = md ("$i$"),
          y    = md ("$y_i$"),
          x    = md ("$x_i$"),
          yhat = md ("$\\hat y_i$"),
          e    = md ("$e_i$"),
          e2   = md ("$e_i^2$")
      ) |>
      cols_width (i ~ px (40), everything () ~ px (100)) |>
      fmt_auto (data = working, columns = c ("y", "x", "yhat", "e", "e2"), decimals = 3) |>
      sub_missing (missing_text = "...") |>
      tab_style (
          style     = cell_text (weight = "bold"),
          locations = cells_body (rows = i == "Sum")
      ) |>
      tab_footnote (
          footnote  = md (sprintf ("$\\hat B_0 = %.4f$, $\\hat B_1 = %.4f$, $\\hat y_i = \\hat B_0 + \\hat B_1 x_i$, $s_e^2 = %.4f$.",
                                    Bhat[1], Bhat[2], SSe)),
          locations = cells_column_labels (columns = yhat)
      ) |>
      tab_source_note (
          source_note = md (if (estimate == "model")
              sprintf ("**Point estimate:** $\\hat B_1 = \\dfrac{\\sum_i(x_i-\\bar x)(y_i-\\bar y)}{\\sum_i(x_i-\\bar x)^2} = \\dfrac{%s}{%s} = %s$",
                       fmt_num (Sxy), fmt_num (Sxx), fmt_num (unname (Bhat[2])))
          else if (estimate == "total")
              sprintf ("**Point estimate:** $\\hat t_{reg} = N\\bar y_{reg} = %s \\times %s = %s$",
                       fmt_num (N, 0), fmt_num (yhat_reg), fmt_num (est))
          else
              sprintf ("**Point estimate:** $\\bar y_{reg} = \\hat B_0+\\hat B_1\\bar x_U = %s + %s \\times %s = %s$",
                       fmt_num (Bhat[1]), fmt_num (Bhat[2]), fmt_num (xbarU), fmt_num (yhat_reg)))
      ) |>
      tab_source_note (
          source_note = md (if (estimate == "model")
              sprintf ("**Standard error:** $\\mathrm{SE}(\\hat B_1) = \\sqrt{\\dfrac{s_e^2}{\\sum_i(x_i-\\bar x)^2}} = \\sqrt{\\dfrac{%s}{%s}} = %s$",
                       fmt_num (SSe), fmt_num (Sxx), fmt_num (sd_est))
          else if (estimate == "total")
              sprintf ("**Standard error:** $\\mathrm{SE}(\\hat t_{reg}) = N\\times\\mathrm{SE}(\\bar y_{reg}) = %s \\times %s = %s$",
                       fmt_num (N, 0), fmt_num (se_yhat_reg), fmt_num (sd_est))
          else if (is.finite (N))
              sprintf ("**Standard error:** $\\mathrm{SE}(\\bar y_{reg}) = \\sqrt{\\left(1-\\dfrac{n}{N}\\right)\\dfrac{s_e^2}{n}} = \\sqrt{\\left(1-\\dfrac{%d}{%d}\\right)\\dfrac{%s}{%d}} = %s$",
                       n, N, fmt_num (SSe), n, fmt_num (se_yhat_reg))
          else
              sprintf ("**Standard error:** $\\mathrm{SE}(\\bar y_{reg}) = \\sqrt{\\dfrac{s_e^2}{n}} = \\sqrt{\\dfrac{%s}{%d}} = %s$",
                       fmt_num (SSe), n, fmt_num (se_yhat_reg)))
      )

  list (estimate = estimate_vec, table = table)
}
srs_reg_est <- srs_reg_est_mean

## ydata --- observations of the variable of interest
## xdata --- observations of the auxiliary variable
## xbarU --- population mean of the auxiliary variable; required when
##           estimate is "mean" or "total"; ignored when estimate = "model"
## N --- population size
## estimate --- "mean" (default) returns the ratio estimate of the
##              population MEAN of y, Bhat*xbarU; "total" returns the ratio
##              estimate of the population TOTAL of y (requires finite N);
##              "model" returns the raw ratio coefficient Bhat = ybar/xbar
##              itself (with its own SE)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values y_i, x_i, fitted yhat_i =
##                  Bhat*x_i, residual e_i, and e_i^2, with a Sum row and a
##                  footnote showing Bhat and s_e^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_ratio_est <- function (ydata, xdata, xbarU = NULL, N = Inf,
                            estimate = c ("mean", "total", "model"), show.details = TRUE,
                            col_labels = list (y = "y_i", x = "x_i", yhat = "\\hat y_i"))
{
  estimate <- match.arg (estimate)

  n <- length (xdata)
  xbar <- mean (xdata)
  ybar <- mean (ydata)
  B_hat <- ybar / xbar
  yhat <- B_hat * xdata
  e <- ydata - yhat
  var_e <- sum (e^2) / (n - 1)
  sd_B_hat <- sqrt ((1 - n/N) * var_e / n) / xbar

  if (estimate == "model")
  {
      est <- B_hat
      sd_est <- sd_B_hat
  }
  else
  {
      if (is.null (xbarU))
          stop ("xbarU (population mean of x) is required when estimate is \"mean\" or \"total\"")
      if (estimate == "total")
      {
          if (!is.finite (N))
              stop ("N must be finite to compute estimate = \"total\"")
          est <- B_hat * xbarU * N
          sd_est <- sd_B_hat * xbarU * N
      }
      else
      {
          est <- B_hat * xbarU
          sd_est <- sd_B_hat * xbarU
      }
  }

  mem <- qt (0.975, df = n - 1) * sd_est
  estimate_vec <- c (Est. = est, S.E. = sd_est, ci.low = est - mem, ci.upp = est + mem)

  if (!show.details)
      return (list (estimate = estimate_vec, table = NULL))

  working <- data.frame (
      i    = c (seq_len (n), "Sum"),
      y    = c (ydata, sum (ydata)),
      x    = c (xdata, sum (xdata)),
      yhat = c (yhat, sum (yhat)),
      e    = c (e, sum (e)),
      e2   = c (e^2, sum (e^2))
  )
  working <- truncate_working (working, id_col = "i")

  table <- gt (working) |>
      tab_header (title = "Ratio Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
      cols_label (
          i    = md ("$i$"),
          y    = md (sprintf ("$%s$", col_labels$y)),
          x    = md (sprintf ("$%s$", col_labels$x)),
          yhat = md (sprintf ("$%s$", col_labels$yhat)),
          e    = md ("$e_i$"),
          e2   = md ("$e_i^2$")
      ) |>
      cols_width (i ~ px (40), everything () ~ px (100)) |>
      fmt_auto (data = working, columns = c ("y", "x", "yhat", "e", "e2"), decimals = 3) |>
      sub_missing (missing_text = "...") |>
      tab_style (
          style     = cell_text (weight = "bold"),
          locations = cells_body (rows = i == "Sum")
      ) |>
      tab_footnote (
          footnote  = md (sprintf ("$\\hat B = \\sum_i %s / \\sum_i %s = %.4f$, $%s = \\hat B\\, %s$, $s_e^2 = %.4f$.",
                                    col_labels$y, col_labels$x, B_hat, col_labels$yhat, col_labels$x, var_e)),
          locations = cells_column_labels (columns = yhat)
      ) |>
      tab_source_note (
          source_note = md (if (estimate == "model")
              sprintf ("**Point estimate:** $\\hat B = \\dfrac{\\sum_i %s}{\\sum_i %s} = \\dfrac{%s}{%s} = %s$",
                       col_labels$y, col_labels$x, fmt_num (sum (ydata)), fmt_num (sum (xdata)), fmt_num (B_hat))
          else if (estimate == "total")
              sprintf ("**Point estimate:** $\\hat t = \\hat B\\,\\bar x_U\\,N = %s \\times %s \\times %s = %s$",
                       fmt_num (B_hat), fmt_num (xbarU), fmt_num (N, 0), fmt_num (est))
          else
              sprintf ("**Point estimate:** $\\bar y_r = \\hat B\\,\\bar x_U = %s \\times %s = %s$",
                       fmt_num (B_hat), fmt_num (xbarU), fmt_num (est)))
      ) |>
      tab_source_note (
          source_note = md (if (estimate == "model")
              (if (is.finite (N))
                  sprintf ("**Standard error:** $\\mathrm{SE}(\\hat B) = \\dfrac{1}{\\bar x}\\sqrt{\\left(1-\\dfrac{n}{N}\\right)\\dfrac{s_e^2}{n}} = \\dfrac{1}{%s}\\sqrt{\\left(1-\\dfrac{%d}{%d}\\right)\\dfrac{%s}{%d}} = %s$",
                           fmt_num (xbar), n, N, fmt_num (var_e), n, fmt_num (sd_B_hat))
              else
                  sprintf ("**Standard error:** $\\mathrm{SE}(\\hat B) = \\dfrac{1}{\\bar x}\\sqrt{\\dfrac{s_e^2}{n}} = \\dfrac{1}{%s}\\sqrt{\\dfrac{%s}{%d}} = %s$",
                           fmt_num (xbar), fmt_num (var_e), n, fmt_num (sd_B_hat)))
          else if (estimate == "total")
              sprintf ("**Standard error:** $\\mathrm{SE}(\\hat t) = \\mathrm{SE}(\\hat B)\\,\\bar x_U\\,N = %s \\times %s \\times %s = %s$",
                       fmt_num (sd_B_hat), fmt_num (xbarU), fmt_num (N, 0), fmt_num (sd_est))
          else
              sprintf ("**Standard error:** $\\mathrm{SE}(\\bar y_r) = \\mathrm{SE}(\\hat B)\\,\\bar x_U = %s \\times %s = %s$",
                       fmt_num (sd_B_hat), fmt_num (xbarU), fmt_num (sd_est)))
      )

  list (estimate = estimate_vec, table = table)
}

## sdata --- a vector of original survey data in a domain
## N --- population size
## n --- total sample size (not the sample size in the domain)
## to find total, multiply domain size N_d to the estimate returned by this function
srs_domain_mean_est <- function (sdata, n, N = Inf)
{
    n_d <- length (sdata)
    ybar <- mean (sdata)
    se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n_d)
    mem <- qt (0.975, df = n_d - 1) * se.ybar
    c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

## sdata --- a vector of original survey data
## N --- population size
## estimate --- "mean" (default) returns the population MEAN estimate;
##              "total" returns the population TOTAL estimate (requires
##              finite N)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values, with a Sum row and a
##                  footnote showing s_y^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_mean_est <- function (sdata, N = Inf, estimate = c ("mean", "total"),
                          show.details = TRUE, col_labels = list (y = "y_i"))
{
    estimate <- match.arg (estimate)

    n <- length (sdata)
    ybar <- mean (sdata)
    dev <- sdata - ybar
    s2y <- sum (dev^2) / (n - 1)
    se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)

    if (estimate == "total")
    {
        if (!is.finite (N))
            stop ("N must be finite to compute estimate = \"total\"")
        est <- N * ybar
        se.est <- N * se.ybar
    }
    else
    {
        est <- ybar
        se.est <- se.ybar
    }
    mem <- qt (0.975, df = n - 1) * se.est
    estimate_vec <- c (Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return (list (estimate = estimate_vec, table = NULL))

    working <- data.frame (
        i    = c (seq_len (n), "Sum"),
        y    = c (sdata, sum (sdata)),
        dev  = c (dev, sum (dev)),
        dev2 = c (dev^2, sum (dev^2))
    )
    working <- truncate_working (working, id_col = "i")

    table <- gt (working) |>
        tab_header (title = "SRS Mean Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
        cols_label (
            i    = md ("$i$"),
            y    = md (sprintf ("$%s$", col_labels$y)),
            dev  = md (sprintf ("$%s-\\bar y$", col_labels$y)),
            dev2 = md (sprintf ("$(%s-\\bar y)^2$", col_labels$y))
        ) |>
        cols_width (i ~ px (40), everything () ~ px (110)) |>
        fmt_auto (data = working, columns = c ("y", "dev", "dev2"), decimals = 3) |>
        sub_missing (missing_text = "...") |>
        tab_style (
            style     = cell_text (weight = "bold"),
            locations = cells_body (rows = i == "Sum")
        ) |>
        tab_footnote (
            footnote  = md (sprintf ("$s_y^2 = \\sum_i (%s-\\bar y)^2/(n-1) = %.4f$.", col_labels$y, s2y)),
            locations = cells_column_labels (columns = dev2)
        ) |>
        tab_source_note (
            source_note = md (if (estimate == "total")
                sprintf ("**Point estimate:** $\\hat t = N\\bar y = %s \\times %s = %s$",
                         fmt_num (N, 0), fmt_num (ybar), fmt_num (est))
            else
                sprintf ("**Point estimate:** $\\bar y = \\dfrac{\\sum_i %s}{n} = \\dfrac{%s}{%d} = %s$",
                         col_labels$y, fmt_num (sum (sdata)), n, fmt_num (ybar)))
        ) |>
        tab_source_note (
            source_note = md (if (estimate == "total")
                sprintf ("**Standard error:** $\\mathrm{SE}(\\hat t) = N \\times \\mathrm{SE}(\\bar y) = %s \\times %s = %s$",
                         fmt_num (N, 0), fmt_num (se.ybar), fmt_num (se.est))
            else if (is.finite (N))
                sprintf ("**Standard error:** $\\mathrm{SE}(\\bar y) = \\sqrt{1-\\dfrac{n}{N}}\\,\\dfrac{s}{\\sqrt n} = \\sqrt{1-\\dfrac{%d}{%d}}\\,\\dfrac{%s}{\\sqrt{%d}} = %s$",
                         n, N, fmt_num (sqrt (s2y)), n, fmt_num (se.ybar))
            else
                sprintf ("**Standard error:** $\\mathrm{SE}(\\bar y) = \\dfrac{s}{\\sqrt n} = \\dfrac{%s}{\\sqrt{%d}} = %s$",
                         fmt_num (sqrt (s2y)), n, fmt_num (se.ybar)))
        )

    list (estimate = estimate_vec, table = table)
}

# Helper function to format estimation outputs with appropriate mathematical headers
format_est_gt <- function(est_data, est_type = c("mean", "total", "ratio", "reg_mean", "reg_total", "ratio_mean", "domain_mean")) {
  est_type <- match.arg(est_type)
  
  sym_map <- list(
    "mean"        = c(est = "$\\bar{y}$",         se = "$\\mathrm{SE}(\\bar{y})$"),
    "total"       = c(est = "$\\hat{t}$",         se = "$\\mathrm{SE}(\\hat{t})$"),
    "ratio"       = c(est = "$\\hat{B}$",         se = "$\\mathrm{SE}(\\hat{B})$"),
    "reg_mean"    = c(est = "$\\bar{y}_{reg}$",   se = "$\\mathrm{SE}(\\bar{y}_{reg})$"),
    "reg_total"   = c(est = "$\\hat{t}_{reg}$",   se = "$\\mathrm{SE}(\\hat{t}_{reg})$"),
    "ratio_mean"  = c(est = "$\\bar{y}_{r}$",     se = "$\\mathrm{SE}(\\bar{y}_{r})$"),
    "domain_mean" = c(est = "$\\bar{y}_{d}$",     se = "$\\mathrm{SE}(\\bar{y}_{d})$")
  )
  
  est_sym <- sym_map[[est_type]]["est"]
  se_sym  <- sym_map[[est_type]]["se"]
  
  # Convert to data frame. Use unname() to strip hidden lm() coefficient names
  if (is.vector(est_data) && !is.list(est_data)) {
    df <- as.data.frame(t(unname(est_data)))
  } else {
    df <- as.data.frame(est_data)
  }
  
  # Force standard column names so gt() never gets confused
  colnames(df) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  
  # Check if we should use rownames as a stub
  use_stub <- !is.null(rownames(df)) && !all(rownames(df) == as.character(1:nrow(df)))
  
  df |>
    gt(rownames_to_stub = use_stub) |>
    cols_label(
      Est.   = md(est_sym),
      S.E.   = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(
      columns = everything(),
      decimals = 4
    ) |>
    tab_options(table.width = pct(70))
}
```

### Post-stratification

For post-stratification, we use the standard stratified estimator, but we plug in the expected sample sizes $n_h = n \times \frac{N_h}{N}$ instead of the actual observed random $n_h$.

```{r}
## Given per-stratum summaries, compute the (post-)stratified mean or total.
## estimate --- "mean" (default) returns the stratified MEAN estimate;
##              "total" returns the stratified TOTAL estimate
## show.details --- if TRUE (default), also return a gt table of the
##                  stratum-by-stratum working values, with a Total row
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
str_mean_estimate <- function (ybarh, sh, nh, Nh, estimate = c ("mean", "total"), show.details = TRUE)
{
    estimate <- match.arg (estimate)

    N    <- sum (Nh)
    Pi_h <- Nh / N
    v_h  <- (1 - nh / Nh) * Pi_h^2 * sh^2 / nh

    ybar   <- sum (ybarh * Pi_h)
    seybar <- sqrt (sum (v_h))

    if (estimate == "total")
    {
        est    <- N * ybar
        se.est <- N * seybar
    }
    else
    {
        est    <- ybar
        se.est <- seybar
    }
    mem          <- 1.96 * se.est
    estimate_vec <- c (Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return (list (estimate = estimate_vec, table = NULL))

    stratum <- names (Nh)
    if (is.null (stratum)) stratum <- seq_along (Nh)

    working <- data.frame (
        Stratum       = c (stratum, "Total"),
        Nh            = c (Nh, N),
        nh            = c (nh, sum (nh)),
        Pi_h          = c (Pi_h, sum (Pi_h)),
        ybarh         = c (ybarh, NA),
        sh2           = c (sh^2, NA),
        weighted_mean = c (Pi_h * ybarh, ybar),
        v_h           = c (v_h, seybar^2)
    )

    table <- gt (working) |>
        tab_header (title = "Stratified Mean Estimation: Working Table") |>
        cols_label (
            Stratum       = md ("Stratum"),
            Nh            = md ("$N_h$"),
            nh            = md ("$n_h$"),
            Pi_h          = md ("$\\pi_h$"),
            ybarh         = md ("$\\bar y_h$"),
            sh2           = md ("$s_h^2$"),
            weighted_mean = md ("$\\pi_h\\bar y_h$"),
            v_h           = md ("$v_h$")
        ) |>
        cols_width (c (Nh, nh) ~ px (60), everything () ~ px (110)) |>
        fmt_number (columns = c (Nh, nh), decimals = 0) |>
        fmt_number (columns = Pi_h, decimals = 2) |>
        fmt_auto (data = working, columns = c ("ybarh", "sh2", "weighted_mean", "v_h"), decimals = 2) |>
        sub_missing (missing_text = "") |>
        tab_style (
            style     = cell_text (weight = "bold"),
            locations = cells_body (rows = Stratum == "Total")
        ) |>
        tab_footnote (
            footnote  = md ("$\\pi_h = N_h/N$ is the stratum weight."),
            locations = cells_column_labels (columns = Pi_h)
        ) |>
        tab_footnote (
            footnote  = md ("$v_h = (1-n_h/N_h)\\,\\pi_h^2\\, s_h^2/n_h$ is stratum $h$'s contribution to $V(\\bar y_{str}) = \\sum_h v_h$."),
            locations = cells_column_labels (columns = v_h)
        ) |>
        tab_source_note (
            source_note = md (if (estimate == "total")
                sprintf ("**Point estimate:** $\\hat t_{str} = N\\bar y_{str} = %s \\times %s = %s$",
                         fmt_num (N, 0), fmt_num (ybar), fmt_num (est))
            else
                sprintf ("**Point estimate:** $\\bar y_{str} = \\sum_h \\pi_h\\bar y_h = %s$", fmt_num (ybar)))
        ) |>
        tab_source_note (
            source_note = md (if (estimate == "total")
                sprintf ("**Standard error:** $\\mathrm{SE}(\\hat t_{str}) = N \\times \\mathrm{SE}(\\bar y_{str}) = %s \\times %s = %s$",
                         fmt_num (N, 0), fmt_num (seybar), fmt_num (se.est))
            else
                sprintf ("**Standard error:** $\\mathrm{SE}(\\bar y_{str}) = \\sqrt{\\sum_h v_h} = \\sqrt{%s} = %s$",
                         fmt_num (sum (v_h)), fmt_num (seybar)))
        )

    list (estimate = estimate_vec, table = table)
}
```

## Ratio Estimation for cherry.csv dataset 

We want to estimate the volume of wood in cherry trees using their diameter as an auxiliary variable. 

### Importing data

First, let's load the data and verify that a strong linear relationship passing near the origin exists between volume and diameter.

```{r}
cherry <- read.csv("data/cherry.csv", header = T)
head(cherry) |> gt()

plot(cherry$volume ~ cherry$diameter, main = "Volume vs Diameter")
summary(lm(cherry$volume ~ 0 + cherry$diameter))
```
**Output Note:** The plot and the linear model output (which forces a 0 intercept) confirm a very strong relationship, making this a prime candidate for Ratio estimation.

### SRS estimate 

To establish a baseline, we first estimate the mean and total volume using standard Simple Random Sampling (ignoring the auxiliary variable).

```{r}
N <- 2967
## estimating the mean of volume
res_srs_vol <- srs_mean_est(cherry$volume, N = N, estimate = "mean")
res_srs_vol$table
srs_mean_vol_raw <- res_srs_vol$estimate
srs_mean_vol_raw |> format_est_gt("mean")

## estimating the total of volume
res_srs_total_vol <- srs_mean_est(cherry$volume, N = N, estimate = "total")
srs_total_vol_raw <- res_srs_total_vol$estimate
srs_total_vol_raw |> format_est_gt("total")
```
**Output Note:** Under SRS, the standard error for the mean volume is `r round(srs_mean_vol_raw["S.E."], 2)`.

### Ratio Estimation Using the Function

We calculate the ratio $\hat{B} = \overline{y}/\overline{x}$; the function's `$table` shows the row-level working values $x_i$, $y_i$, fitted $\hat{y}_i = \hat{B}x_i$, residuals $e_i = y_i - \hat{y}_i$, and $e_i^2$ (with a Sum row and a footnote giving $\hat B$ and $s_e^2$), so we no longer need to build this table by hand.

```{r}
ydata <- cherry$volume
xdata <- cherry$diameter
N <- 2967

res_ratio_model <- srs_ratio_est(ydata, xdata, N = N, estimate = "model")
res_ratio_model$table

plot(cherry$volume ~ cherry$diameter, main="Ratio Estimation Fit")
abline(a = 0, b = res_ratio_model$estimate["Est."], col="blue", lwd=2)
```
**Output Note:** The blue line represents our estimated ratio $\hat{B} \approx `r round(res_ratio_model$estimate["Est."], 2)`$. The working table above shows the individual residuals used to estimate $s_e^2$.

```{r}
B_v2d <- res_ratio_model$estimate
B_v2d |> format_est_gt("ratio")
```
**Output Note:** The ratio $\hat{B}$ is estimated at `r round(B_v2d["Est."], 3)` with a standard error of `r round(B_v2d["S.E."], 4)`.

#### Estimating the mean volume of wood

If we know the true total of the population diameters is 41,835, we can derive the population mean diameter $\overline{x}_{\text{U}}$ and pass it as `xbarU` with `estimate = "mean"`.

```{r}
mean_diameters <- 41835 / N
ratio_mean_volume_raw <- srs_ratio_est(ydata, xdata, xbarU = mean_diameters, N = N, estimate = "mean")$estimate
ratio_mean_volume_raw |> format_est_gt("ratio_mean")
```

#### Estimating the total volume of wood

Using `estimate = "total"` with the known population total diameter gives the estimated population total volume directly.

```{r}
t_diameters <- 41835
srs_ratio_est(ydata, xdata, xbarU = t_diameters / N, N = N, estimate = "total")$estimate |>
  format_est_gt("total")
```

#### Percentage of Variance Reduction

How much did ratio estimation help compared to ignoring the auxiliary variable?

```{r}
variance_reduction <- 1 - (ratio_mean_volume_raw["S.E."] / srs_mean_vol_raw["S.E."])^2
unname(variance_reduction)
```
**Output Note:** Using the diameter as an auxiliary variable reduced the variance of our estimate by `r round(unname(variance_reduction) * 100, 1)`% compared to simple random sampling. 

## Simulation study with agpop.csv

Let's test this practically by drawing 2000 random samples and seeing how SRS and Ratio estimators behave under repeated sampling.

### Information of population data 

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na

# sample size
n <- 300
# population size
N <- nrow(agpop)

# true values that we want to estimate
tyU <- sum(agpop[, "acres92"])
# suppose known for ratio estimate
txU <- sum(agpop[, "acres87"]) 
B <- tyU / txU

plot(agpop[, "acres87"], agpop[, "acres92"], main="Acres 1992 vs Acres 1987")
abline(a = 0, b = B, col="red")

# linear model output
summary(lm(agpop$acres92 ~ agpop$acres87))

# expected reduction of variance of ratio estimate to srs
exp_var_red <- 1 - var(agpop$acres92 - B * agpop$acres87) / var(agpop$acres92)
exp_var_red
```
**Output Note:** Past acreage (1987) strongly predicts current acreage (1992). The theoretically expected variance reduction is roughly `r round(exp_var_red * 100, 1)`%.

### Simulation Studies

```{r cache=TRUE}
nsim <- 2000
sim_rat <- data.frame(simple = rep(0, nsim), ratio = rep(0, nsim), B = rep(0, nsim))

for (i in 1:nsim) {
  srs <- sample(N, n)
  y_srs <- agpop[srs, "acres92"]
  x_srs <- agpop[srs, "acres87"]

  # SRS estimate (Total)
  sim_rat$simple[i] <- mean(y_srs) * N 
  # ratio estimate (Total)
  sim_rat$B[i] <- mean(y_srs) / mean(x_srs)
  sim_rat$ratio[i] <-  sim_rat$B[i] * txU
}

boxplot(sim_rat[, 1:2], main="Estimator Variance: SRS vs Ratio")
abline(h = tyU, col = "red", lwd=2)
```

We now calculate the actual variance across our 2000 simulations.

```{r}
sim_var <- sapply(sim_rat[, 1:2], var)
ratio_sim_var <- sim_var / sim_var[1]
percentage_reduction <- 1 - ratio_sim_var

data.frame(sim_var, ratio_sim_var, percentage_reduction) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(
    sim_var = "Variance",
    ratio_sim_var = "Relative Variance",
    percentage_reduction = "Variance Reduction"
  ) |>
  fmt_number(columns = c(sim_var), decimals = 0) |> 
  fmt_percent(columns = c(percentage_reduction), decimals = 2) |>
  fmt_number(columns = c(ratio_sim_var), decimals = 4)
```
**Output Note:** The boxplot dramatically illustrates the precision gain. The variance of the ratio estimator is only `r round(ratio_sim_var["ratio"] * 100, 1)`% of the SRS variance, confirming our theoretical expectations.

## Estimating Domain Means and Post-stratification with agsrs.csv example

### Estimating Domain Means

When estimating the mean for a specific subgroup (like a region), the sample size in that subgroup ($n_d$) is a random variable, which increases our uncertainty.

```{r}
agsrs <- read.csv("data/agsrs.csv")
n <- nrow(agsrs)

## estimating domain means
region_agsrs_domain <- data.frame(rbind(
    NC = srs_domain_mean_est(agsrs$acres92[agsrs$region=="NC"], n=n, N = 3078),
    NE = srs_domain_mean_est(agsrs$acres92[agsrs$region=="NE"], n=n, N = 3078),
    S  = srs_domain_mean_est(agsrs$acres92[agsrs$region=="S"], n=n, N = 3078),
    W  = srs_domain_mean_est(agsrs$acres92[agsrs$region=="W"], n=n, N = 3078)
))

region_agsrs_domain |> format_est_gt("domain_mean")
```

If we incorrectly used the standard SRS formula (pretending $n_d$ was fixed), our standard errors would be artificially low:

```{r}
## naively applying SRS estimation to each domain (incorrect analysis)
region_agsrs_srs <- data.frame(rbind(
    NC = srs_mean_est(agsrs$acres92[agsrs$region=="NC"], N = 1054, estimate = "mean", show.details = FALSE)$estimate,
    NE = srs_mean_est(agsrs$acres92[agsrs$region=="NE"], N = 220,  estimate = "mean", show.details = FALSE)$estimate,
    S  = srs_mean_est(agsrs$acres92[agsrs$region=="S"],  N = 1382, estimate = "mean", show.details = FALSE)$estimate,
    W  = srs_mean_est(agsrs$acres92[agsrs$region=="W"],  N = 422,  estimate = "mean", show.details = FALSE)$estimate
))

region_agsrs_srs |> format_est_gt("mean")
```
**Output Note:** Comparing the two tables, you can see the correct `domain_mean` Standard Errors are larger, appropriately penalizing us for the uncertainty in how many counties from each region randomly fell into our sample.

### Post-stratification Analysis

If we know the true regional population sizes ($N_h$), we can post-stratify our SRS data.

```{r}
nh <- tapply(agsrs[, "acres92"], agsrs[,"region"], length)
n <- sum(nh)
sh <- tapply(agsrs[, "acres92"], agsrs[,"region"], sd)
ybarh <- tapply(agsrs[, "acres92"], agsrs[,"region"], mean)
# create a vector with external information
Nh <- c(NC = 1054, NE = 220, S = 1382, W = 422)
```

**Create $n_h^{\text{post-strat}} = n \times \frac{N_h}{N}$**, instead of using the randomly observed $n_h$.

```{r}
nh_post <- Nh / sum(Nh) * n

## regional summary for stratified sampling estimation
data.frame(
  ybarh = ybarh, 
  sh = sh, 
  nh = nh, 
  pi_obs = nh/n, 
  Nh = Nh, 
  pi_pop = Nh/sum(Nh),  
  nh_post = nh_post
) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$"),
    nh = md("$n_h^{\\text{obs}}$"),
    pi_obs = md("$\\pi^{\\text{obs}}_h$"),
    Nh = md("$N_h$"),
    pi_pop = md("$\\pi^{\\text{pop}}_h$"),
    nh_post = md("$n_h^{\\text{post}}$")
  ) |> 
  fmt_number(columns = c(ybarh, sh, pi_obs, pi_pop, nh_post), decimals = 3)
```
**Output Note:** The table above shows the difference between the sample proportion ($\pi^{obs}_h$) and the true population proportion ($\pi^{pop}_h$). Post-stratification forces the weights to reflect the true population structure.

```{r}
## strata mean estimate
res_post_strat <- str_mean_estimate(ybarh, sh, nh_post, Nh, estimate = "mean")
res_post_strat$table
post_strat_est <- res_post_strat$estimate
post_strat_est |> format_est_gt("mean")

## compare with SRS estimate
res_srs_est <- srs_mean_est(agsrs$acres92, N = 3087, estimate = "mean")
srs_est_raw <- res_srs_est$estimate
srs_est_raw |> format_est_gt("mean")

## true mean check
mean(agpop$acres92)
```
**Output Note:** The post-stratified estimate gives a Standard Error of `r round(post_strat_est["S.E."], 1)`, which is more precise than the unadjusted SRS standard error of `r round(srs_est_raw["S.E."], 1)`.

## Estimating Domain Means and Post-stratification with teacher example

### Importing Data and Calculating Contingency Table

```{r}
teacher <- read.csv("data/college_teacher.csv")

table(teacher[, 2:1])
addmargins(table(teacher[, 2:1]), 2)
# Proportions of students who want to become a teacher in two domains (female and male)
prop.table(table(teacher[, 2:1]), margin = "gender")
```

### Domain Summary

```{r}
n <- nrow(teacher)
ybarh <- tapply(teacher$teacher, INDEX = teacher$gender, FUN = mean)
sh <- tapply(teacher$teacher, INDEX = teacher$gender, FUN = sd)
nh <- tapply(teacher$teacher, INDEX = teacher$gender, FUN = length)
prop_h_obs <- nh / sum(nh)

data.frame(
  ybarh = ybarh, 
  sh = sh, 
  nh = nh, 
  pi_obs = prop_h_obs
) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$"),
    nh = md("$n^{\\text{obs}}_h$"),
    pi_obs = md("$\\pi^{\\text{obs}}_h$")
  ) |> 
  fmt_number(columns = everything(), decimals = 3)
```

### Estimating Domain Means

```{r}
Nh <- c(3000, 1000)
n <- nrow(teacher)
domain_teacher_est <- data.frame(rbind(
    female = srs_domain_mean_est(subset(teacher, gender==1)$teacher, n=n, N=4000),
    male   = srs_domain_mean_est(subset(teacher, gender==2)$teacher, n=n, N=4000)
))

domain_teacher_est |> format_est_gt("domain_mean")
```

### Post-stratification Analysis

```{r}
Nh <- c(3000, 1000)
prop_h_pop <- Nh / sum(Nh)
prop_h_pop
```
#### Create $n_h^{\text{post-strat}} = n \times \frac{N_h}{N}$ {-}
```{r}
nh_post <- Nh / sum(Nh) * n

## show the differences
data.frame(
  phat = ybarh, 
  sh = sh, 
  nh = nh, 
  pi_obs = nh/n, 
  Nh = Nh, 
  pi_pop = Nh/sum(Nh),  
  nh_post = nh_post
) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(
    phat = md("$\\hat{p}_h$"),
    sh = md("$s_h$"),
    nh = md("$n_h$"),
    pi_obs = md("$\\pi^{\\text{obs}}_h$"),
    Nh = md("$N_h$"),
    pi_pop = md("$\\pi^{\\text{pop}}_h$"),
    nh_post = md("$n_h^{\\text{post-strat}}$")
  ) |> 
  fmt_number(columns = c(phat, sh, pi_obs, pi_pop, nh_post), decimals = 3)
```

Calculate the post-stratification estimation of the mean:

```{r}
## poststratification estimation of mean
res_teacher_post_strat <- str_mean_estimate(ybarh, sh, nh_post, Nh, estimate = "mean")
res_teacher_post_strat$table
teacher_post_strat <- res_teacher_post_strat$estimate
teacher_post_strat |> format_est_gt("mean")
```

This post-stratification estimate explicitly corrects for the sampling imbalance using the true strata sizes:

```{r}
# 0.35 * 0.75 + 0.25 * 0.25 
(ybarh[1] * prop_h_pop[1]) + (ybarh[2] * prop_h_pop[2])
```

### Comparing to the Analysis with SRS

If we didn't post-stratify, the simple mean is heavily weighted toward males because they were over-sampled by chance (40% in sample vs 25% in population).

```{r}
res_teacher_srs <- srs_mean_est(teacher$teacher, N=sum(Nh), estimate = "mean")
res_teacher_srs$table
teacher_srs <- res_teacher_srs$estimate
teacher_srs |> format_est_gt("mean")
```

This equals the naive sample proportion:

```{r}
# 0.35 * 0.6 + 0.25 * 0.4
mean(teacher$teacher == 1)
```

## Regression Estimation for the cherry.csv dataset 

When the relationship between $y$ and $x$ doesn't pass exactly through the origin, we use Regression Estimation.

### Importing data

```{r}
cherry <- read.csv("data/cherry.csv", header = T)
ydata <- cherry$volume 
xdata <- cherry$diameter
t_diameters <- 41835
xbarU <- t_diameters / 2967 
N <- 2967
```

### Regression Estimation Using the Function

#### Fitting a linear regression model

```{r}
n <- length(ydata)
lmfit <- lm(ydata ~ xdata)
summary(lmfit)

plot(xdata, ydata, main="Linear Regression: Volume vs Diameter")
abline(lmfit, col="green", lwd=2)
```

The function's `$table` shows the row-level working values $x_i$, $y_i$, fitted $\hat{y}_i = \hat{B}_0 + \hat{B}_1 x_i$, residuals $e_i = y_i - \hat{y}_i$, and $e_i^2$ (with a Sum row and a footnote giving $\hat B_0$, $\hat B_1$, and $s_e^2$):

```{r}
res_reg_model <- srs_reg_est(ydata, xdata, N = N, estimate = "model")
res_reg_model$table
```

#### Estimate the mean

We compute the regression estimate $\overline{y}_{reg}$ by passing `xbarU` with `estimate = "mean"`.

```{r}
res_reg_mean <- srs_reg_est(ydata, xdata, xbarU = xbarU, N = N, estimate = "mean")
res_reg_mean$table

output_reg <- res_reg_mean$estimate
output_reg |> format_est_gt("reg_mean")
```
**Output Note:** The Regression estimate for the mean volume is `r round(output_reg["Est."], 3)` with a standard error of `r round(output_reg["S.E."], 4)`.

#### Estimate the total

```{r}
reg_total_volume_raw <- srs_reg_est(ydata, xdata, xbarU = xbarU, N = N, estimate = "total")$estimate
reg_total_volume_raw |> format_est_gt("reg_total")
```

#### Percentage of Variance Reduction

```{r}
reg_mean_volume_raw <- output_reg
var_reduction_reg <- 1 - (reg_mean_volume_raw["S.E."] / srs_mean_vol_raw["S.E."])^2
unname(var_reduction_reg)
```
**Output Note:** The variance reduction for the regression estimator is `r round(unname(var_reduction_reg) * 100, 1)`%, which is similar to the ratio estimator since the true intercept is very close to zero.

## Regression estimation for photo counts of dead trees

To estimate the number of dead trees in an area, we divide the area into 100 square plots and count dead trees on photographs. We select an SRS of 25 plots for exact field counts. We know the true mean photo count is 11.3.

```{r}
photocounts <- c(10, 12, 7, 13, 13, 6, 17, 16, 15, 10, 14, 12, 10, 5, 12, 10,
                 10, 9, 6, 11, 7, 9, 11, 10, 10)

fieldcounts <- c(15, 14, 9, 14, 8, 5, 18, 15, 13, 15, 11, 15, 12, 8, 13,
                 9, 11, 12, 9, 12, 13, 11, 10, 9, 8)

lmfit <- lm(fieldcounts ~ photocounts)
summary(lmfit)

plot(photocounts, fieldcounts, main="Field Counts vs Photo Counts")
abline(lmfit, col="purple", lwd=2)
```

Let's compare the three different estimation methods to see which provides the lowest Standard Error for the population mean.

```{r}
# 1. Regression estimate for the mean
res_dt_reg <- srs_reg_est(ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, estimate = "mean")
res_dt_reg$table
dt_reg_raw <- res_dt_reg$estimate
dt_reg_raw |> format_est_gt("reg_mean")

# (And estimating the total)
srs_reg_est(ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, estimate = "total")$estimate |>
  format_est_gt("reg_total")

# 2. Simple random sampling estimate for the mean (Ignoring photo counts)
res_dt_srs <- srs_mean_est(sdata = fieldcounts, N = 100, estimate = "mean")
dt_srs_raw <- res_dt_srs$estimate
dt_srs_raw |> format_est_gt("mean")

# 3. Ratio estimate for the mean
res_dt_ratio <- srs_ratio_est(ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, estimate = "mean")
res_dt_ratio$table
dt_ratio_raw <- res_dt_ratio$estimate
dt_ratio_raw |> format_est_gt("ratio_mean")
```
#### Output Note: {-}
*   **SRS Mean:** SE = `r round(dt_srs_raw["S.E."], 3)`
*   **Ratio Mean:** SE = `r round(dt_ratio_raw["S.E."], 3)`

*   **Regression Mean:** SE = `r round(dt_reg_raw["S.E."], 3)`

Because the intercept of the true line is non-zero, the **Regression estimator** provides the most precise estimate in this scenario.