# Ratio, Regression, and Domain Estimation

:::{.callout-note}

### Key formulae

When auxiliary information (a variable $x$) is highly correlated with our target variable of interest ($y$), we can use **Ratio** or **Regression** estimation to improve precision without increasing the sample size. We can also use auxiliary population structures to adjust our estimates after the fact via **Post-stratification**, or estimate means for specific sub-populations (**Domains**).

#### 1. Ratio Estimation {-}
If the relationship between $y$ and $x$ passes through the origin, we estimate the ratio $B$ and use it to adjust the mean:

*   **Ratio:** $\hat{B} = \frac{\overline{y}}{\overline{x}}$
*   **Mean Estimator:** $\overline{y}_r = \hat{B} \overline{x}_{\text{U}}$ (where $\overline{x}_{\text{U}}$ is the true population mean of $x$)

*   **Estimated Variance:** $\hat{V}(\overline{y}_r) = \left(\frac{\overline{x}_{\text{U}}}{\overline{x}}\right)^2 \left(1 - \frac{n}{N}\right) \frac{s_e^2}{n}$, where $s_e^2$ is the sample variance of the residuals $e_i = y_i - \hat{B}x_i$.

#### 2. Regression Estimation {-}
If the relationship does not pass through the origin, a linear regression is more appropriate:

*   **Mean Estimator:** $\overline{y}_{reg} = \hat{B}_0 + \hat{B}_1 \overline{x}_{\text{U}}$ (where $\hat{B}_0$ and $\hat{B}_1$ are OLS coefficients)
*   **Estimated Variance:** $\hat{V}(\overline{y}_{reg}) = \left(1 - \frac{n}{N}\right) \frac{s_e^2}{n}$, where $s_e^2$ is the Mean Square Error (Residuals) from the regression.

#### 3. Domain Estimation {-}
To estimate the mean of a subgroup (domain $d$) when the domain sample size $n_d$ is a random variable:

*   **Mean Estimator:** $\overline{y}_d = \frac{1}{n_d} \sum_{i \in U_d} y_i$
*   **Estimated Variance:** $\hat{V}(\overline{y}_d) = \left(1 - \frac{n}{N}\right) \frac{s_d^2}{n_d}$, where $n$ is the overall sample size, $N$ is the population size, and $s_d^2$ is the sample variance of $y$ among the $n_d$ units observed in domain $d$.

#### 4. Post-stratification {-}
If we know the true population sizes ($N_h$) of strata, but we used a Simple Random Sample, we can post-stratify to reduce variance. We substitute $n_h$ with expected proportional sample sizes:

*   $n_h^{post-strat} = n \frac{N_h}{N}$

:::

## Functions and packages for Analyzing Data

We begin by loading the necessary libraries, including `gt` for formatting mathematical tables, and defining our custom functions for Ratio, Regression, Domain, and Post-stratified estimation. We also define a versatile `gt` helper to render our estimation outputs cleanly.

### General functions for ratio and regression estimation

The estimating functions used throughout this book (including `reg_est`, `ratio_est`, `domain_est`, `srs_est`, and the `format_est_gt` table-formatting helper used below) live in a single shared file, `samplingestimate.r`, which we source below.

For post-stratification, we reuse the shared `str_est()` stratified estimator, but we plug in the expected sample sizes $n_h = n \times \frac{N_h}{N}$ instead of the actual observed random $n_h$.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::


## Ratio Estimation for cherry.csv dataset 

We want to estimate the volume of wood in cherry trees using their diameter as an auxiliary variable. 

### Importing data

First, let's load the data and verify that a strong linear relationship passing near the origin exists between volume and diameter.

```{r tbl-cherry-sample-data}
#| tbl-cap: "Sample data: cherry tree volume and diameter"
cherry <- read.csv("data/cherry.csv", header = T)
head(cherry) |> gt()
```

```{r fig-cherry-volume-diameter}
#| fig-cap: "Cherry tree volume vs. diameter"
plot(cherry$volume ~ cherry$diameter, main = "Volume vs Diameter")
```

```{r cherry-load-anova}
anova(lm(cherry$volume ~ 0 + cherry$diameter))
```
**Output Note:** The plot and the linear model output (which forces a 0 intercept) confirm a very strong relationship, making this a prime candidate for Ratio estimation.

### SRS estimate 

To establish a baseline, we first estimate the mean and total volume using standard Simple Random Sampling (ignoring the auxiliary variable).

```{r cherry-srs-est}
N <- 2967
## estimating the mean of volume
res_srs_vol <- srs_est(cherry$volume, N = N, estimate = "mean", show.details = FALSE)
srs_mean_vol_raw <- res_srs_vol$estimate
srs_mean_vol_raw |> format_est_gt("mean")

## estimating the total of volume
res_srs_total_vol <- srs_est(cherry$volume, N = N, estimate = "total", show.details = FALSE)
srs_total_vol_raw <- res_srs_total_vol$estimate
srs_total_vol_raw |> format_est_gt("total")
```
**Output Note:** Under SRS, the standard error for the mean volume is `r round(srs_mean_vol_raw["S.E."], 2)`.

### Ratio Estimation Using the Function

We calculate the ratio $\hat{B} = \overline{y}/\overline{x}$; the function's `$table` shows the row-level working values $x_i$, $y_i$, fitted $\hat{y}_i = \hat{B}x_i$, residuals $e_i = y_i - \hat{y}_i$, and $e_i^2$ (with a Sum row and a footnote giving $\hat B$ and $s_e^2$), so we no longer need to build this table by hand.

```{r cherry-ratio-est}
ydata <- cherry$volume
xdata <- cherry$diameter
N <- 2967

res_ratio_model <- ratio_est(ydata, xdata, N = N, estimate = "model")
res_ratio_model$table
```

```{r fig-cherry-ratio-fit}
#| fig-cap: "Cherry tree volume vs. diameter with the fitted ratio line"
plot(cherry$volume ~ cherry$diameter, main="Ratio Estimation Fit")
abline(a = 0, b = res_ratio_model$estimate["Est."], col="blue", lwd=2)
```
**Output Note:** The blue line represents our estimated ratio $\hat{B} \approx `r round(res_ratio_model$estimate["Est."], 2)`$. The working table above shows the individual residuals used to estimate $s_e^2$.

```{r cherry-ratio-b}
B_v2d <- res_ratio_model$estimate
B_v2d |> format_est_gt("ratio")
```
**Output Note:** The ratio $\hat{B}$ is estimated at `r round(B_v2d["Est."], 3)` with a standard error of `r round(B_v2d["S.E."], 4)`.

#### Estimating the mean volume of wood

If we know the true total of the population diameters is 41,835, we can derive the population mean diameter $\overline{x}_{\text{U}}$ and pass it as `xbarU` with `estimate = "mean"`.

```{r cherry-ratio-mean}
mean_diameters <- 41835 / N
res_ratio_mean <- ratio_est(ydata, xdata, xbarU = mean_diameters, N = N, estimate = "mean")
res_ratio_mean$table
ratio_mean_volume_raw <- res_ratio_mean$estimate
ratio_mean_volume_raw |> format_est_gt("ratio_mean")
```

#### Estimating the total volume of wood

Using `estimate = "total"` with the known population total diameter gives the estimated population total volume directly.

```{r cherry-ratio-total}
t_diameters <- 41835

res_ratio_total <- ratio_est(ydata, xdata, xbarU = t_diameters / N, N = N, estimate = "total")
res_ratio_total$table
ratio_total_volume <- res_ratio_total$estimate
ratio_total_volume |> format_est_gt("total")
```

#### Percentage of Variance Reduction

How much did ratio estimation help compared to ignoring the auxiliary variable?

```{r cherry-ratio-varred}
variance_reduction <- 1 - (ratio_mean_volume_raw["S.E."] / srs_mean_vol_raw["S.E."])^2
unname(variance_reduction)
```
**Output Note:** Using the diameter as an auxiliary variable reduced the variance of our estimate by `r round(unname(variance_reduction) * 100, 1)`% compared to simple random sampling. 

## Simulation study with agpop.csv

Let's test this practically by drawing 2000 random samples and seeing how SRS and Ratio estimators behave under repeated sampling.

### Information of population data 

```{r agpop-load}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na

# sample size
n <- 300
# population size
N <- nrow(agpop)

# true values that we want to estimate
tyU <- sum(agpop[, "acres92"])
# suppose known for ratio estimate
txU <- sum(agpop[, "acres87"]) 
B <- tyU / txU
```

```{r fig-agpop-acres8792-ratio-line}
#| fig-cap: "1992 vs. 1987 county acreage with the true ratio line"
plot(agpop[, "acres87"], agpop[, "acres92"], main="Acres 1992 vs Acres 1987")
abline(a = 0, b = B, col="red")
```

```{r agpop-load-lm}
# linear model output
lm_gt <- format_lm_gt(lm(agpop$acres92 ~ agpop$acres87))
lm_gt$coef
lm_gt$fit

# expected reduction of variance of ratio estimate to srs
exp_var_red <- 1 - var(agpop$acres92 - B * agpop$acres87) / var(agpop$acres92)
exp_var_red
```
**Output Note:** Past acreage (1987) strongly predicts current acreage (1992). The theoretically expected variance reduction is roughly `r round(exp_var_red * 100, 1)`%.

### Simulation Studies

```{r fig-agpop-ratio-simulation-boxplot, cache=TRUE}
#| fig-cap: "Estimator variance under repeated sampling: SRS vs. ratio estimator (2000 simulations)"
nsim <- 2000
sim_rat <- data.frame(simple = rep(0, nsim), ratio = rep(0, nsim), B = rep(0, nsim))

for (i in 1:nsim) {
  srs <- sample(N, n)
  y_srs <- agpop[srs, "acres92"]
  x_srs <- agpop[srs, "acres87"]

  # SRS estimate (Total)
  sim_rat$simple[i] <- mean(y_srs) * N 
  # ratio estimate (Total)
  sim_rat$B[i] <- mean(y_srs) / mean(x_srs)
  sim_rat$ratio[i] <-  sim_rat$B[i] * txU
}

boxplot(sim_rat[, 1:2], main="Estimator Variance: SRS vs Ratio")
abline(h = tyU, col = "red", lwd=2)
```

We now calculate the actual variance across our 2000 simulations.

```{r tbl-agpop-ratio-variance-reduction}
#| tbl-cap: "Simulated variance and variance reduction: ratio estimator vs. SRS"
sim_var <- sapply(sim_rat[, 1:2], var)
ratio_sim_var <- sim_var / sim_var[1]
percentage_reduction <- 1 - ratio_sim_var

data.frame(sim_var, ratio_sim_var, percentage_reduction) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(
    sim_var = "Variance",
    ratio_sim_var = "Relative Variance",
    percentage_reduction = "Variance Reduction"
  ) |>
  fmt_number(columns = c(sim_var), decimals = 0) |> 
  fmt_percent(columns = c(percentage_reduction), decimals = 2) |>
  fmt_number(columns = c(ratio_sim_var), decimals = 4)
```
**Output Note:** The boxplot dramatically illustrates the precision gain. The variance of the ratio estimator is only `r round(ratio_sim_var["ratio"] * 100, 1)`% of the SRS variance, confirming our theoretical expectations.

## The app

```{shinylive-r}
#| standalone: true
#| viewerHeight: 720

## file: app.R
library(shiny)

## ---------------------------------------------------------------- population
data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/agpop.csv"

pop <- tryCatch({
  download.file(data_url, "agpop.csv")
  d <- read.csv("agpop.csv")
  d <- d[d$acres92 != -99 & d$acres87 != -99, ]
  data.frame(x = d$acres87, y = d$acres92)
}, error = function(e) NULL)

if (is.null(pop)) {            # fallback so the app runs before the file is committed
  set.seed(1)
  Nf <- 3000
  xf <- rgamma(Nf, shape = 1.2, scale = 1.8e5)
  pop <- data.frame(x = xf, y = pmax(0, 0.93 * xf + rnorm(Nf, 0, 0.5 * xf^0.85)))
}

N     <- nrow(pop)
ybarU <- mean(pop$y)
B     <- sum(pop$y) / sum(pop$x)
Vy    <- var(pop$y)                  # SRS population variance
Ve    <- var(pop$y - B * pop$x)      # ratio (residual) population variance

## display-only vertical jitter, fixed once so the cloud does not shimmer
set.seed(11)
ypl <- pop$y + rnorm(N, 0, 0.012 * diff(range(pop$y)))

se_theory <- function(n, type) {
  sqrt((1 - n / N) * (if (type == "ratio") Ve else Vy) / n)
}

## predictions for the non-sampled units, under either working model
predict_rest <- function(s, type) {
  if (type == "ratio") {
    (sum(pop$y[s]) / sum(pop$x[s])) * pop$x[-s]
  } else {
    rep(mean(pop$y[s]), N - length(s))
  }
}

col_pop  <- "#9aa0a640"
col_pred <- "#1f77b4"
col_samp <- "#d62728"

## ---------------------------------------------------------------------- ui
ui <- fluidPage(
  tags$style(HTML(
    ".well {padding: 10px 12px; margin-bottom: 10px;}
     .form-group {margin-bottom: 6px;}
     .irs {margin-bottom: 0;}
     .btn {margin-bottom: 6px;}"
  )),
  wellPanel(
    fluidRow(
      column(
        2,
        radioButtons("est", "Estimator",
                     c("Ratio" = "ratio", "Average" = "srs"),
                     selected = "ratio")
      ),
      column(3, sliderInput("n", "Sample size n", 10, 500, 100, step = 10)),
      column(3, sliderInput("reps", "Replications", 50, 2000, 500, step = 50)),
      column(
        2,
        actionButton("new", "Draw a new sample", width = "100%"),
        actionButton("play", "Play", class = "btn-primary", width = "100%")
      ),
      column(
        2,
        actionButton("reset", "Clear the histogram", width = "100%"),
        div(style = "font-size: 90%;", textOutput("counter"))
      )
    )
  ),
  plotOutput("mainplot", height = "520px")
)

## ------------------------------------------------------------------- server
server <- function(input, output, session) {

  rv <- reactiveValues(s = NULL, type = "ratio", ests = numeric(0),
                       playing = FALSE)

  draw_sample <- function() {
    type <- input$est
    s    <- sample.int(N, input$n)
    rv$s    <- s
    rv$type <- type
    rv$ests <- c(rv$ests, (sum(pop$y[s]) + sum(predict_rest(s, type))) / N)
  }

  ## changing n or the estimator invalidates the accumulated replications
  observeEvent(list(input$n, input$est), {
    rv$ests <- numeric(0)
    draw_sample()
  }, ignoreInit = FALSE)

  observeEvent(input$new,   draw_sample())
  observeEvent(input$reset, { rv$ests <- numeric(0); draw_sample() })

  observeEvent(input$play, {
    rv$playing <- !rv$playing
    updateActionButton(session, "play", label = if (rv$playing) "Pause" else "Play")
  })

  observe({
    if (!rv$playing) return()
    invalidateLater(150, session)
    isolate({
      if (length(rv$ests) >= input$reps) {
        rv$playing <- FALSE
        updateActionButton(session, "play", label = "Play")
      } else {
        draw_sample()
      }
    })
  })

  output$counter <- renderText(
    sprintf("%d of %d replications", length(rv$ests), input$reps)
  )

  output$mainplot <- renderPlot({
    s <- rv$s
    req(s)
    type <- rv$type
    n    <- length(s)
    pred <- predict_rest(s, type)
    est  <- (sum(pop$y[s]) + sum(pred)) / N

    layout(matrix(1:2, nrow = 1), widths = c(2.5, 1))

    ## ---- left: population, sample, predictions ----
    par(mar = c(4.2, 4.6, 3.0, 0.8))
    plot(pop$x, ypl, pch = 16, cex = 0.45, col = col_pop,
         xlab = "x: acres87", ylab = "y: acres92",
         main = sprintf("%s estimator, n = %d",
                        if (type == "ratio") "Ratio" else "Average", n))
    if (type == "ratio") {
      abline(0, sum(pop$y[s]) / sum(pop$x[s]), col = col_pred, lwd = 2)
    } else {
      abline(h = mean(pop$y[s]), col = col_pred, lwd = 2)
    }
    points(pop$x[-s], pred, pch = 16, cex = 0.5, col = col_pred)
    points(pop$x[s], ypl[s], pch = 16, cex = 0.9, col = col_samp)
    abline(h = ybarU, col = col_samp, lty = 3)
    abline(h = est, col = col_pred, lty = 2)
    points(par("usr")[1], est, pch = 17, cex = 1.5, col = col_pred, xpd = NA)
    legend("topleft", bty = "n", pch = c(16, 16, NA), lty = c(NA, NA, 3),
           col = c(col_samp, col_pred, col_samp), cex = 0.9,
           legend = c("sampled y", "predicted y", "population mean"))

    ## ---- right: sampling distribution of the estimate ----
    se <- se_theory(n, type)
    lo <- ybarU - 4 * se
    hi <- ybarU + 4 * se
    br <- seq(lo, hi, length.out = 41)
    yy <- seq(lo, hi, length.out = 200)
    dn <- dnorm(yy, ybarU, se)

    par(mar = c(4.2, 4.6, 3.0, 0.8))
    mcsd <- if (length(rv$ests) > 2) sd(rv$ests) else NA
    plot(NA, xlim = c(0, max(dn) * 1.15), ylim = c(lo, hi),
         xlab = "density", ylab = "estimate of the population mean",
         main = sprintf("SE: %s (Monte Carlo) / %s (theory)",
                        if (is.na(mcsd)) "--" else format(round(mcsd), big.mark = ","),
                        format(round(se), big.mark = ",")))
    if (length(rv$ests) > 0) {
      h <- hist(pmin(pmax(rv$ests, lo), hi), breaks = br, plot = FALSE)
      rect(0, br[-41], h$density, br[-1], col = "#1f77b455", border = "white")
    }
    lines(dn, yy, col = "#2ca02c", lwd = 2)
    abline(h = ybarU, col = col_samp, lty = 3)
    abline(h = est, col = col_pred, lty = 2)
  })
}

shinyApp(ui, server)
```


## Estimating Domain Means and Post-stratification with agsrs.csv example

### Estimating Domain Means

When estimating the mean for a specific subgroup (like a region), the sample size in that subgroup ($n_d$) is a random variable, which increases our uncertainty.

```{r agsrs-domain-est}
agsrs <- read.csv("data/agsrs.csv")
n <- nrow(agsrs)

## estimating domain means
region_agsrs_domain <- data.frame(rbind(
    NC = domain_est(agsrs$acres92[agsrs$region=="NC"], n=n, N = 3078),
    NE = domain_est(agsrs$acres92[agsrs$region=="NE"], n=n, N = 3078),
    S  = domain_est(agsrs$acres92[agsrs$region=="S"], n=n, N = 3078),
    W  = domain_est(agsrs$acres92[agsrs$region=="W"], n=n, N = 3078)
))

region_agsrs_domain |> format_est_gt("domain_mean")
```

If we incorrectly used the standard SRS formula (pretending $n_d$ was fixed), our standard errors would be artificially low:

```{r agsrs-domain-srs}
## naively applying SRS estimation to each domain (incorrect analysis)
region_agsrs_srs <- data.frame(rbind(
    NC = srs_est(agsrs$acres92[agsrs$region=="NC"], N = 1054, estimate = "mean", show.details = FALSE)$estimate,
    NE = srs_est(agsrs$acres92[agsrs$region=="NE"], N = 220,  estimate = "mean", show.details = FALSE)$estimate,
    S  = srs_est(agsrs$acres92[agsrs$region=="S"],  N = 1382, estimate = "mean", show.details = FALSE)$estimate,
    W  = srs_est(agsrs$acres92[agsrs$region=="W"],  N = 422,  estimate = "mean", show.details = FALSE)$estimate
))

region_agsrs_srs |> format_est_gt("mean")
```
**Output Note:** Comparing the two tables, you can see the correct `domain_mean` Standard Errors are larger, appropriately penalizing us for the uncertainty in how many counties from each region randomly fell into our sample.

### Post-stratification Analysis

If we know the true regional population sizes ($N_h$), we can post-stratify our SRS data.

```{r agsrs-poststrat-setup}
nh <- tapply(agsrs[, "acres92"], agsrs[,"region"], length)
n <- sum(nh)
sh <- tapply(agsrs[, "acres92"], agsrs[,"region"], sd)
ybarh <- tapply(agsrs[, "acres92"], agsrs[,"region"], mean)
# create a vector with external information
Nh <- c(NC = 1054, NE = 220, S = 1382, W = 422)
```

**Create $n_h^{\text{post-strat}} = n \times \frac{N_h}{N}$**, instead of using the randomly observed $n_h$.

```{r tbl-agsrs-poststrat-summary}
#| tbl-cap: "Observed vs. population stratum proportions and post-stratified sample sizes"
nh_post <- Nh / sum(Nh) * n

## regional summary for stratified sampling estimation
data.frame(
  ybarh = ybarh, 
  sh = sh, 
  nh = nh, 
  pi_obs = nh/n, 
  Nh = Nh, 
  pi_pop = Nh/sum(Nh),  
  nh_post = nh_post
) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$"),
    nh = md("$n_h^{\\text{obs}}$"),
    pi_obs = md("$\\pi^{\\text{obs}}_h$"),
    Nh = md("$N_h$"),
    pi_pop = md("$\\pi^{\\text{pop}}_h$"),
    nh_post = md("$n_h^{\\text{post}}$")
  ) |> 
  fmt_number(columns = c(ybarh, sh, pi_obs, pi_pop, nh_post), decimals = 3)
```
**Output Note:** The table above shows the difference between the sample proportion ($\pi^{obs}_h$) and the true population proportion ($\pi^{pop}_h$). Post-stratification forces the weights to reflect the true population structure.

```{r agsrs-poststrat-est}
## strata mean estimate
res_post_strat <- str_est(ybarh, sh, nh_post, Nh, estimate = "mean")
res_post_strat$table
post_strat_est <- res_post_strat$estimate
post_strat_est |> format_est_gt("mean")

## compare with SRS estimate
res_srs_est <- srs_est(agsrs$acres92, N = 3087, estimate = "mean", show.details = FALSE)
srs_est_raw <- res_srs_est$estimate
srs_est_raw |> format_est_gt("mean")

## true mean check
mean(agpop$acres92)
```
**Output Note:** The post-stratified estimate gives a Standard Error of `r round(post_strat_est["S.E."], 1)`, which is more precise than the unadjusted SRS standard error of `r round(srs_est_raw["S.E."], 1)`.

## Estimating Domain Means and Post-stratification with teacher example

### Importing Data and Calculating Contingency Table

```{r teacher-load}
teacher <- read.csv("data/college_teacher.csv")

table(teacher[, 2:1])
addmargins(table(teacher[, 2:1]), 2)
# Proportions of students who want to become a teacher in two domains (female and male)
prop.table(table(teacher[, 2:1]), margin = "gender")
```

### Domain Summary

```{r tbl-teacher-domain-summary}
#| tbl-cap: "Domain summary statistics for the teacher-aspiration dataset, by gender"
n <- nrow(teacher)
ybarh <- tapply(teacher$teacher, INDEX = teacher$gender, FUN = mean)
sh <- tapply(teacher$teacher, INDEX = teacher$gender, FUN = sd)
nh <- tapply(teacher$teacher, INDEX = teacher$gender, FUN = length)
prop_h_obs <- nh / sum(nh)

data.frame(
  ybarh = ybarh, 
  sh = sh, 
  nh = nh, 
  pi_obs = prop_h_obs
) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$"),
    nh = md("$n^{\\text{obs}}_h$"),
    pi_obs = md("$\\pi^{\\text{obs}}_h$")
  ) |> 
  fmt_number(columns = everything(), decimals = 3)
```

### Estimating Domain Means

```{r teacher-domain-est}
Nh <- c(3000, 1000)
n <- nrow(teacher)
domain_teacher_est <- data.frame(rbind(
    female = domain_est(subset(teacher, gender==1)$teacher, n=n, N=4000),
    male   = domain_est(subset(teacher, gender==2)$teacher, n=n, N=4000)
))

domain_teacher_est |> format_est_gt("domain_mean")
```

### Post-stratification Analysis

```{r teacher-poststrat-prop}
Nh <- c(3000, 1000)
prop_h_pop <- Nh / sum(Nh)
prop_h_pop
```
#### Create $n_h^{\text{post-strat}} = n \times \frac{N_h}{N}$ {-}
```{r tbl-teacher-poststrat-summary}
#| tbl-cap: "Observed vs. population gender proportions and post-stratified sample sizes"
nh_post <- Nh / sum(Nh) * n

## show the differences
data.frame(
  phat = ybarh, 
  sh = sh, 
  nh = nh, 
  pi_obs = nh/n, 
  Nh = Nh, 
  pi_pop = Nh/sum(Nh),  
  nh_post = nh_post
) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(
    phat = md("$\\hat{p}_h$"),
    sh = md("$s_h$"),
    nh = md("$n_h$"),
    pi_obs = md("$\\pi^{\\text{obs}}_h$"),
    Nh = md("$N_h$"),
    pi_pop = md("$\\pi^{\\text{pop}}_h$"),
    nh_post = md("$n_h^{\\text{post-strat}}$")
  ) |> 
  fmt_number(columns = c(phat, sh, pi_obs, pi_pop, nh_post), decimals = 3)
```

Calculate the post-stratification estimation of the mean:

```{r teacher-poststrat-est}
## poststratification estimation of mean
res_teacher_post_strat <- str_est(ybarh, sh, nh_post, Nh, estimate = "mean")
res_teacher_post_strat$table
teacher_post_strat <- res_teacher_post_strat$estimate
teacher_post_strat |> format_est_gt("mean")
```

This post-stratification estimate explicitly corrects for the sampling imbalance using the true strata sizes:

```{r teacher-poststrat-check}
# 0.35 * 0.75 + 0.25 * 0.25 
(ybarh[1] * prop_h_pop[1]) + (ybarh[2] * prop_h_pop[2])
```

### Comparing to the Analysis with SRS

If we didn't post-stratify, the simple mean is heavily weighted toward males because they were over-sampled by chance (40% in sample vs 25% in population).

```{r teacher-srs-est}
res_teacher_srs <- srs_est(teacher$teacher, N=sum(Nh), estimate = "mean", show.details = FALSE)
res_teacher_srs$table
teacher_srs <- res_teacher_srs$estimate
teacher_srs |> format_est_gt("mean")
```

This equals the naive sample proportion:

```{r teacher-srs-check}
# 0.35 * 0.6 + 0.25 * 0.4
mean(teacher$teacher == 1)
```

## Regression Estimation for the cherry.csv dataset 

When the relationship between $y$ and $x$ doesn't pass exactly through the origin, we use Regression Estimation.

### Importing data

```{r cherry-reg-load}
cherry <- read.csv("data/cherry.csv", header = T)
ydata <- cherry$volume 
xdata <- cherry$diameter
t_diameters <- 41835
xbarU <- t_diameters / 2967 
N <- 2967
```

### Regression Estimation Using the Function

#### Fitting a linear regression model

```{r cherry-reg-fit}
n <- length(ydata)
lmfit <- lm(ydata ~ xdata)
lmfit_gt <- format_lm_gt(lmfit)
lmfit_gt$coef
lmfit_gt$fit
```

```{r fig-cherry-regression-fit}
#| fig-cap: "Cherry tree volume vs. diameter with the fitted regression line"
plot(xdata, ydata, main="Linear Regression: Volume vs Diameter")
abline(lmfit, col="green", lwd=2)
```

The function's `$table` shows the row-level working values $x_i$, $y_i$, fitted $\hat{y}_i = \hat{B}_0 + \hat{B}_1 x_i$, residuals $e_i = y_i - \hat{y}_i$, and $e_i^2$ (with a Sum row and a footnote giving $\hat B_0$, $\hat B_1$, and $s_e^2$):

```{r cherry-reg-model}
res_reg_model <- reg_est(ydata, xdata, N = N, estimate = "model")
res_reg_model$table
```

#### Estimate the mean

We compute the regression estimate $\overline{y}_{reg}$ by passing `xbarU` with `estimate = "mean"`.

```{r cherry-reg-mean}
res_reg_mean <- reg_est(ydata, xdata, xbarU = xbarU, N = N, estimate = "mean")
res_reg_mean$table

output_reg <- res_reg_mean$estimate
output_reg |> format_est_gt("reg_mean")
```
**Output Note:** The Regression estimate for the mean volume is `r round(output_reg["Est."], 3)` with a standard error of `r round(output_reg["S.E."], 4)`.

#### Estimate the total

```{r cherry-reg-total}
reg_total_volume_raw <- reg_est(ydata, xdata, xbarU = xbarU, N = N, estimate = "total")$estimate
reg_total_volume_raw |> format_est_gt("reg_total")
```

#### Percentage of Variance Reduction

```{r cherry-reg-varred}
reg_mean_volume_raw <- output_reg
var_reduction_reg <- 1 - (reg_mean_volume_raw["S.E."] / srs_mean_vol_raw["S.E."])^2
unname(var_reduction_reg)
```
**Output Note:** The variance reduction for the regression estimator is `r round(unname(var_reduction_reg) * 100, 1)`%, which is similar to the ratio estimator since the true intercept is very close to zero.

## Regression estimation for photo counts of dead trees

To estimate the number of dead trees in an area, we divide the area into 100 square plots and count dead trees on photographs. We select an SRS of 25 plots for exact field counts. We know the true mean photo count is 11.3.

```{r deadtrees-data}
photocounts <- c(10, 12, 7, 13, 13, 6, 17, 16, 15, 10, 14, 12, 10, 5, 12, 10,
                 10, 9, 6, 11, 7, 9, 11, 10, 10)

fieldcounts <- c(15, 14, 9, 14, 8, 5, 18, 15, 13, 15, 11, 15, 12, 8, 13,
                 9, 11, 12, 9, 12, 13, 11, 10, 9, 8)

lmfit <- lm(fieldcounts ~ photocounts)
lmfit_gt <- format_lm_gt(lmfit)
lmfit_gt$coef
lmfit_gt$fit
```

```{r fig-deadtrees-regression-fit}
#| fig-cap: "Field counts vs. photo counts of dead trees with the fitted regression line"
plot(photocounts, fieldcounts, main="Field Counts vs Photo Counts")
abline(lmfit, col="purple", lwd=2)
```

Let's compare the three different estimation methods to see which provides the lowest Standard Error for the population mean.

```{r deadtrees-compare}
# 1. Regression estimate for the mean
res_dt_reg <- reg_est(ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, estimate = "mean")
res_dt_reg$table
dt_reg_raw <- res_dt_reg$estimate
dt_reg_raw |> format_est_gt("reg_mean")

# (And estimating the total)
reg_est(ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, estimate = "total")$estimate |>
  format_est_gt("reg_total")

# 2. Simple random sampling estimate for the mean (Ignoring photo counts)
res_dt_srs <- srs_est(sdata = fieldcounts, N = 100, estimate = "mean", show.details = FALSE)
dt_srs_raw <- res_dt_srs$estimate
dt_srs_raw |> format_est_gt("mean")

# 3. Ratio estimate for the mean
res_dt_ratio <- ratio_est(ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, estimate = "mean")
res_dt_ratio$table
dt_ratio_raw <- res_dt_ratio$estimate
dt_ratio_raw |> format_est_gt("ratio_mean")
```
#### Output Note: {-}
*   **SRS Mean:** SE = `r round(dt_srs_raw["S.E."], 3)`
*   **Ratio Mean:** SE = `r round(dt_ratio_raw["S.E."], 3)`

*   **Regression Mean:** SE = `r round(dt_reg_raw["S.E."], 3)`

Because the intercept of the true line is non-zero, the **Regression estimator** provides the most precise estimate in this scenario.