---
title: "Simple Random Sampling"
format: html
execute:
  warning: false
  message: false
---


## Analysis of agsrs.csv Data

### Step by step calculation without using a function

To evaluate a Simple Random Sample (SRS), we calculate the point estimate, the standard error (incorporating the Finite Population Correction factor), and the margin of error to construct a 95% confidence interval.

**Key Formulas:**

* **Sample Mean:** $\bar{y} = \frac{1}{n} \sum_{i=1}^n y_i$
* **Standard Error:** $\mathrm{SE}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left( \frac{s}{\sqrt{n}} \right)$
* **Margin of Error:** $\mathrm{ME} = t_{\alpha/2, n-1} \times \mathrm{SE}(\bar{y})$
* **Population Total:** $\hat{t} = N\bar{y}$

```{r}
## read survey data
agsrs <- read.csv ("data/agsrs.csv")
head(agsrs)

## extract the variable of interest
sdata <- agsrs$acres92
N <- 3078

## do calculation
n <- length (sdata)
ybar <- mean (sdata)
se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
mem <- qt (0.975, df = n - 1) * se.ybar

## return estimate vector for pop mean
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)

## return estimate vector for pop total
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem) * N

```

### Write a function for repeated use

#### A function for doing data analysis for srs sample

Wrap the previous manual calculations into a custom R function to quickly estimate means, standard errors, and confidence intervals for any variable of interest. We will also create a helper function using `gt` to format the numerical vector into a nice table with mathematical expressions in the headers.

```{r}
#| code-fold: true

library(gt)

## Formats a gt table column as ordinary fixed-point, UNLESS its values are
## too small (< `small` in absolute value) or too large (>= `large`), in
## which case it switches that column to scientific notation instead.
fmt_auto <- function (gt_tbl, data, columns, decimals = 3, small = 1e-3, large = 1e5)
{
    for (col in columns)
    {
        x <- data[[col]]
        x <- x[is.finite (x) & x != 0]
        if (length (x) == 0) next
        if (any (abs (x) < small) || any (abs (x) >= large))
            gt_tbl <- fmt_scientific (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
        else
            gt_tbl <- fmt_number (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
    }
    gt_tbl
}

## Truncates a row-level working data.frame whose LAST row is a summary row
## (e.g. "Sum"/"Total"): if there are more than head_n data rows (excluding
## that summary row), keeps only the first head_n data rows, inserts one
## "..." placeholder row, then the summary row --- so long working tables
## stay readable. id_col --- name of the row-label column.
truncate_working <- function (working, id_col, head_n = 6)
{
    n <- nrow (working) - 1
    if (n <= head_n + 1) return (working)

    dots <- working[1, ]
    dots[1, ] <- NA
    dots[1, id_col] <- "..."

    rbind (working[seq_len (head_n), ], dots, working[nrow (working), ])
}

# Helper function to format estimation outputs with appropriate mathematical headers
format_est_gt <- function(est_data, est_type = c("mean", "total", "ratio", "reg_mean", "reg_total", "ratio_mean", "domain_mean")) {
  est_type <- match.arg(est_type)

  sym_map <- list(
    "mean"        = c(est = "$\\bar{y}$",         se = "$\\mathrm{SE}(\\bar{y})$"),
    "total"       = c(est = "$\\hat{t}$",         se = "$\\mathrm{SE}(\\hat{t})$"),
    "ratio"       = c(est = "$\\hat{B}$",         se = "$\\mathrm{SE}(\\hat{B})$"),
    "reg_mean"    = c(est = "$\\bar{y}_{reg}$",   se = "$\\mathrm{SE}(\\bar{y}_{reg})$"),
    "reg_total"   = c(est = "$\\hat{t}_{reg}$",   se = "$\\mathrm{SE}(\\hat{t}_{reg})$"),
    "ratio_mean"  = c(est = "$\\bar{y}_{r}$",     se = "$\\mathrm{SE}(\\bar{y}_{r})$"),
    "domain_mean" = c(est = "$\\bar{y}_{d}$",     se = "$\\mathrm{SE}(\\bar{y}_{d})$")
  )

  est_sym <- sym_map[[est_type]]["est"]
  se_sym  <- sym_map[[est_type]]["se"]

  # Convert to data frame. Use unname() to strip hidden lm() coefficient names
  if (is.vector(est_data) && !is.list(est_data)) {
    df <- as.data.frame(t(unname(est_data)))
  } else {
    df <- as.data.frame(est_data)
  }

  # Force standard column names so gt() never gets confused
  colnames(df) <- c("Est.", "S.E.", "ci.low", "ci.upp")

  # Check if we should use rownames as a stub
  use_stub <- !is.null(rownames(df)) && !all(rownames(df) == as.character(1:nrow(df)))

  df |>
    gt(rownames_to_stub = use_stub) |>
    cols_label(
      Est.   = md(est_sym),
      S.E.   = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(
      columns = everything(),
      decimals = 4
    ) |>
    tab_options(table.width = pct(70))
}

## sdata --- a vector of original survey data
## N --- population size
## estimate --- "mean" (default) returns the population MEAN estimate;
##              "total" returns the population TOTAL estimate (N * mean,
##              with SE/CI scaled accordingly); requires finite N
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values y_i, y_i-\bar y, (y_i-\bar y)^2,
##                  with a Sum row and a footnote showing s_y^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_mean_est <- function (sdata, N = Inf, estimate = c ("mean", "total"),
                          show.details = TRUE, col_labels = list (y = "y_i"))
{
    estimate <- match.arg (estimate)

    n <- length (sdata)
    ybar <- mean (sdata)
    dev <- sdata - ybar
    s2y <- sum (dev^2) / (n - 1)
    se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)

    if (estimate == "total")
    {
        if (!is.finite (N))
            stop ("N must be finite to compute estimate = \"total\"")
        est <- N * ybar
        se.est <- N * se.ybar
    }
    else
    {
        est <- ybar
        se.est <- se.ybar
    }
    mem <- qt (0.975, df = n - 1) * se.est
    estimate_vec <- c (Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return (list (estimate = estimate_vec, table = NULL))

    working <- data.frame (
        i    = c (seq_len (n), "Sum"),
        y    = c (sdata, sum (sdata)),
        dev  = c (dev, sum (dev)),
        dev2 = c (dev^2, sum (dev^2))
    )
    working <- truncate_working (working, id_col = "i")

    table <- gt (working) |>
        tab_header (title = "SRS Mean Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
        cols_label (
            i    = md ("$i$"),
            y    = md (sprintf ("$%s$", col_labels$y)),
            dev  = md (sprintf ("$%s-\\bar y$", col_labels$y)),
            dev2 = md (sprintf ("$(%s-\\bar y)^2$", col_labels$y))
        ) |>
        cols_width (i ~ px (40), everything () ~ px (110)) |>
        fmt_auto (data = working, columns = c ("y", "dev", "dev2"), decimals = 3) |>
        sub_missing (missing_text = "...") |>
        tab_style (
            style     = cell_text (weight = "bold"),
            locations = cells_body (rows = i == "Sum")
        ) |>
        tab_footnote (
            footnote  = md (sprintf ("$s_y^2 = \\sum_i (%s-\\bar y)^2/(n-1) = %.4f$.", col_labels$y, s2y)),
            locations = cells_column_labels (columns = dev2)
        )

    list (estimate = estimate_vec, table = table)
}
```

#### Apply srs_mean_est to agsrs.csv data

*Import Data*

Load the survey sample data representing US counties.

```{r}
agsrs <- read.csv ("data/agsrs.csv")

```

*Estimating the mean of acre92*

Calculate the estimated average farm acreage per county using the custom function. The function's `$table` shows the row-level working values (truncated to the first 6 rows + Sum), and `$estimate` holds the final point estimate, SE, and CI, formatted with `format_est_gt()`.

```{r}
res_mean <- srs_mean_est(agsrs[,"acres92"], N = 3078, estimate = "mean")
res_mean$table
res_mean$estimate |> format_est_gt("mean")

```

*Estimating the total of acre92*

Set `estimate = "total"` to have the function itself scale the mean and confidence interval by the population size ($N = 3078$) to estimate the total farm acreage across all counties.

```{r}
res_total <- srs_mean_est(agsrs[,"acres92"], N = 3078, estimate = "total")
res_total$table
res_total$estimate |> format_est_gt("total")

```

*Estimating the proportion of counties with fewer than 200K acres for farming in 1992*

To estimate a proportion under SRS, create an indicator variable ($y_i = 1$ if true, $0$ if false). The mean of this indicator variable serves as the sample proportion ($\hat{p} = \bar{y}_{indicator}$).

```{r}
acres92.is.fewer.200k <- as.numeric (agsrs[,"acres92"] < 200000)
head(acres92.is.fewer.200k)

res_prop <- srs_mean_est(acres92.is.fewer.200k, N = 3078, estimate = "mean")
res_prop$table
res_prop$estimate |> format_est_gt("mean")

```

*Estimating the total number of counties with fewer than 200K acres for farming in 1992*

Set `estimate = "total"` to estimate the absolute count of counties meeting the criteria.

```{r}
res_prop_total <- srs_mean_est(acres92.is.fewer.200k, N = 3078, estimate = "total")
res_prop_total$table
res_prop_total$estimate |> format_est_gt("total")

```

### Comparing with true value

Load the complete census (population) dataset to calculate the true parameters. This allows us to assess the accuracy of our previous SRS estimates against the exact population values.

```{r}
agpop <- read.csv ("data/agpop.csv", na = "-99")
#true mean
mean (agpop[, "acres92"], na.rm = T)
# true total
sum (agpop[, "acres92"], na.rm = T)

# true proportion of counties with less than 200K acres for farming
mean (agpop[, "acres92"] < 200000, na.rm = T)
# true number of counties with less than 200K acres for farming
sum (agpop[, "acres92"] < 200000, na.rm = T)

```

## A Simulation Demonstration of SRS Inference

Load the full population data and filter out missing values to serve as a complete, known finite population for resampling experiments.

```{r}
# read population data 
agpop <- read.csv ("data/agpop.csv")
# remove those counties with na
agpop <- subset( agpop, acres92 != -99)

```

### True Values

Define the target sample size and compute the true population mean ($\mu$) and theoretical standard error ($\mathrm{SE}_{true}$) using the known population standard deviation ($S$).

**Theoretical Standard Error:**


$$\mathrm{SE}_{true}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left(\frac{S}{\sqrt{n}}\right)$$

```{r}
# sample size
n <- 300
# population size
N <- nrow (agpop); N

# true value of population mean
ybarU <- mean (agpop[,"acres92"]); ybarU
# true value of deviation of sample mean
true.se.ybar <- sqrt (1- n/N) * sd (agpop[,"acres92"]) / sqrt (n); true.se.ybar

```

### One SRS sampling

Draw a single random sample of size $n=300$ from the population and run our estimator function to see a single instance of survey inference.

```{r}
##
# srs sampling
srs <- sample (1:N,n)
head(agpop [srs, ])
# get data of variable "acres92"
sdata <- agpop [srs, "acres92"]
# analysis
res_one <- srs_mean_est(sdata, N, estimate = "mean")
res_one$table
res_one$estimate |> format_est_gt("mean")

```

### Repeating SRS sampling 5000 times

To demonstrate the Central Limit Theorem and unbiasedness of the estimator, draw 5,000 independent samples, storing the estimates from each iteration. We plot the resulting sampling distribution against the true population mean.

*(Note: We use `show.details = FALSE` and pull out just `$estimate` here, so the loop skips building 5,000 working tables and stays fast.)*

```{r fig.height=10}
nres <- 5000 # number of repeated sampling
simulation.results <- matrix (0, nres, 4) # matrix recording repeated results
colnames(simulation.results) <- c( "Est.",   "S.E.",   "ci.low", "ci.upp")

for (i in 1:nres)
{
    srs <- sample (N, n)
   sdata <- agpop [srs, "acres92"]
   simulation.results [i,] <- srs_mean_est (sdata, N, estimate = "mean", show.details = FALSE)$estimate
}

head(simulation.results)
# look at the distribution of sample mean
par (mfrow= c(2,2))
hist (agpop$acres92,main = "Population Distribution of acre92")
hist (simulation.results[,1], main = "Sampling Distribution of Sample Mean for acre92")
abline (v = ybarU, col = "red")
qqnorm (simulation.results[,1], main="QQ plot of Sample Mean"); qqline(simulation.results[,1])
boxplot (simulation.results[,1], main = "Boxplot of Sample Mean")
abline (h = ybarU, col = "red")

mean (simulation.results[,1])
ybarU
sd (simulation.results [,1])
true.se.ybar

```

### Empirical Coverage Rate of CIs

Verify the reliability of the 95% confidence intervals by checking whether the true population mean falls within the calculated bounds for each of the 5,000 samples. The empirical coverage rate should closely approximate 0.95. The plot highlights the first 100 intervals, flagging those that fail to capture the true mean in a distinct color.

```{r fig.height=6}
library(dplyr)
library(plotrix) 

# Convert to data frame (if it was a matrix) and use mutate to evaluate coverage
simulation.results <- as.data.frame(simulation.results) |>
  mutate(
    `Covered?` = as.numeric(ci.low < ybarU & ybarU < ci.upp)
  )

head(simulation.results)

par(mfrow=c(1,1))
plotCI(
  x = 1:100,
  y = simulation.results$Est.[1:100], 
  li = simulation.results$ci.low[1:100],
  ui = simulation.results$ci.upp[1:100],
  col = 2 - simulation.results$`Covered?`[1:100]
)
abline(h = ybarU, col = "blue")

# Empirical coverage rate 
mean(simulation.results$`Covered?`)

```
