---
title: "Simple Random Sampling"
filters:
  - shinylive
execute:
  warning: false
  message: false
  cache: true
---


## Analysis of agsrs.csv Data

### Step by step calculation without using a function

To evaluate a Simple Random Sample (SRS), we calculate the point estimate, the standard error (incorporating the Finite Population Correction factor), and the margin of error to construct a 95% confidence interval.

**Key Formulas:**

* **Sample Mean:** $\bar{y} = \frac{1}{n} \sum_{i=1}^n y_i$
* **Standard Error:** $\mathrm{SE}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left( \frac{s}{\sqrt{n}} \right)$
* **Margin of Error:** $\mathrm{ME} = t_{\alpha/2, n-1} \times \mathrm{SE}(\bar{y})$
* **Population Total:** $\hat{t} = N\bar{y}$

```{r srs-manual-calc}
## read survey data
agsrs <- read.csv ("data/agsrs.csv")
head(agsrs)

## extract the variable of interest
sdata <- agsrs$acres92
N <- 3078

## do calculation
n <- length (sdata)
ybar <- mean (sdata)
se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
mem <- qt (0.975, df = n - 1) * se.ybar

## return estimate vector for pop mean
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)

## return estimate vector for pop total
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem) * N

```

### Write a function for repeated use

#### A function for doing data analysis for srs sample

Wrap the previous manual calculations into a custom R function to quickly estimate means, standard errors, and confidence intervals for any variable of interest. To avoid repeating this code in every chapter, all estimating functions (and their `gt`-table formatting helpers) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

#### Apply srs_est to agsrs.csv data

*Import Data*

Load the survey sample data representing US counties.

```{r agsrs-load}
agsrs <- read.csv ("data/agsrs.csv")

```

*Estimating the mean of acre92*

Calculate the estimated average farm acreage per county using the custom function. The function's `$table` shows the row-level working values (truncated to the first 6 rows + Sum), and `$estimate` holds the final point estimate, SE, and CI, formatted with `format_est_gt()`.

```{r agsrs-mean-est}
res_mean <- srs_est(agsrs[,"acres92"], N = 3078, estimate = "mean")
res_mean$table
res_mean$estimate |> format_est_gt("mean")

```

*Estimating the total of acre92*

Set `estimate = "total"` to have the function itself scale the mean and confidence interval by the population size ($N = 3078$) to estimate the total farm acreage across all counties.

```{r agsrs-total-est}
res_total <- srs_est(agsrs[,"acres92"], N = 3078, estimate = "total", show.details = FALSE)
res_total$estimate |> format_est_gt("total")

```

*Estimating the proportion of counties with fewer than 200K acres for farming in 1992*

To estimate a proportion under SRS, create an indicator variable ($y_i = 1$ if true, $0$ if false). The mean of this indicator variable serves as the sample proportion ($\hat{p} = \bar{y}_{indicator}$).

```{r agsrs-prop-est}
acres92.is.fewer.200k <- as.numeric (agsrs[,"acres92"] < 200000)
head(acres92.is.fewer.200k)

res_prop <- srs_est(acres92.is.fewer.200k, N = 3078, estimate = "mean", show.details = FALSE)
res_prop$estimate |> format_est_gt("mean")

```

*Estimating the total number of counties with fewer than 200K acres for farming in 1992*

Set `estimate = "total"` to estimate the absolute count of counties meeting the criteria.

```{r agsrs-prop-total-est}
res_prop_total <- srs_est(acres92.is.fewer.200k, N = 3078, estimate = "total", show.details = FALSE)
res_prop_total$estimate |> format_est_gt("total")

```

### Comparing with true value

Load the complete census (population) dataset to calculate the true parameters. This allows us to assess the accuracy of our previous SRS estimates against the exact population values.

```{r agpop-true-values}
agpop <- read.csv ("data/agpop.csv", na = "-99")
#true mean
mean (agpop[, "acres92"], na.rm = T)
# true total
sum (agpop[, "acres92"], na.rm = T)

# true proportion of counties with less than 200K acres for farming
mean (agpop[, "acres92"] < 200000, na.rm = T)
# true number of counties with less than 200K acres for farming
sum (agpop[, "acres92"] < 200000, na.rm = T)

```

## A Simulation Demonstration of SRS Inference

Load the full population data and filter out missing values to serve as a complete, known finite population for resampling experiments.

```{r agpop-load-clean}
# read population data 
agpop <- read.csv ("data/agpop.csv")
# remove those counties with na
agpop <- subset( agpop, acres92 != -99)

```

### True Values

Define the target sample size and compute the true population mean ($\mu$) and theoretical standard error ($\mathrm{SE}_{true}$) using the known population standard deviation ($S$).

**Theoretical Standard Error:**


$$\mathrm{SE}_{true}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left(\frac{S}{\sqrt{n}}\right)$$

```{r srs-true-values}
# sample size
n <- 300
# population size
N <- nrow (agpop); N

# true value of population mean
ybarU <- mean (agpop[,"acres92"]); ybarU
# true value of deviation of sample mean
true.se.ybar <- sqrt (1- n/N) * sd (agpop[,"acres92"]) / sqrt (n); true.se.ybar

```

### One SRS sampling

Draw a single random sample of size $n=300$ from the population and run our estimator function to see a single instance of survey inference.

```{r srs-one-sample}
##
# srs sampling
srs <- sample (1:N,n)
head(agpop [srs, ])
# get data of variable "acres92"
sdata <- agpop [srs, "acres92"]
# analysis
res_one <- srs_est(sdata, N, estimate = "mean", show.details = FALSE)
res_one$estimate |> format_est_gt("mean")

```

### Repeating SRS sampling 5000 times

To demonstrate the Central Limit Theorem and unbiasedness of the estimator, draw 5,000 independent samples, storing the estimates from each iteration. We plot the resulting sampling distribution against the true population mean.

*(Note: We use `show.details = FALSE` and pull out just `$estimate` here, so the loop skips building 5,000 working tables and stays fast.)*

```{r fig-srs-simulation-distribution, fig.height=10}
#| fig-cap: "Population distribution and sampling distribution of the SRS sample mean over 5000 repeated samples (n = 300)"
nres <- 5000 # number of repeated sampling
simulation.results <- matrix (0, nres, 4) # matrix recording repeated results
colnames(simulation.results) <- c( "Est.",   "S.E.",   "ci.low", "ci.upp")

for (i in 1:nres)
{
    srs <- sample (N, n)
   sdata <- agpop [srs, "acres92"]
   simulation.results [i,] <- srs_est (sdata, N, estimate = "mean", show.details = FALSE)$estimate
}

head(simulation.results)
# look at the distribution of sample mean
par (mfrow= c(2,2))
hist (agpop$acres92,main = "Population Distribution of acre92")
hist (simulation.results[,1], main = "Sampling Distribution of Sample Mean for acre92")
abline (v = ybarU, col = "red")
qqnorm (simulation.results[,1], main="QQ plot of Sample Mean"); qqline(simulation.results[,1])
boxplot (simulation.results[,1], main = "Boxplot of Sample Mean")
abline (h = ybarU, col = "red")

mean (simulation.results[,1])
ybarU
sd (simulation.results [,1])
true.se.ybar

```

### Empirical Coverage Rate of CIs

Verify the reliability of the 95% confidence intervals by checking whether the true population mean falls within the calculated bounds for each of the 5,000 samples. The empirical coverage rate should closely approximate 0.95. The plot highlights the first 100 intervals, flagging those that fail to capture the true mean in a distinct color.

```{r srs-coverage-setup}
library(dplyr)
library(plotrix)

# Convert to data frame (if it was a matrix) and use mutate to evaluate coverage
simulation.results <- as.data.frame(simulation.results) |>
  mutate(
    `Covered?` = as.numeric(ci.low < ybarU & ybarU < ci.upp)
  )

head(simulation.results)
```

```{r fig-srs-coverage-intervals, fig.height=6}
#| fig-cap: "First 100 simulated 95% confidence intervals for the SRS sample mean, with coverage failures highlighted"
par(mfrow=c(1,1))
plotCI(
  x = 1:100,
  y = simulation.results$Est.[1:100], 
  li = simulation.results$ci.low[1:100],
  ui = simulation.results$ci.upp[1:100],
  col = 2 - simulation.results$`Covered?`[1:100]
)
abline(h = ybarU, col = "blue")

# Empirical coverage rate
mean(simulation.results$`Covered?`)

```

## Shinylive App for Illustrating SRS Theory {#sec-app-srs}

The static plots above show a snapshot of 5,000 repeated samples. The app below
draws SRS samples one at a time, or continuously with **Play** until **Stop** or
until the requested number of samples has been reached.

The top panel shows the population of `acres92`: the grey curve is a kernel
density estimate and the grey points below the axis are the raw values, jittered
vertically. The axis is broken at twice the population mean; counties above the
cut are compressed, in rank order, into the narrow lane at the right, so that a
tail unit drawn into the sample is still visible. Points selected into the
current sample turn red.

The bottom panel accumulates the sample means, drawn as a density histogram so
its scale matches the orange curve overlaid on it: the theoretical sampling
distribution $N(\mu, \mathrm{SE}(\bar y))$ at the current sample size. Its window
is fixed at five theoretical standard errors for $n = 100$ on either side of
$\mu$, so the spread is comparable across sample sizes; for $n < 100$ some means
fall outside the window and are counted in the legend. The interval above the
histogram belongs to the current sample, drawn green when it covers $\mu$ and red
when it does not, and replaced at every draw. Both panels are centred on $\mu$, so the two dashed
lines are vertically aligned.

```{shinylive-r}
#| standalone: true
#| viewerHeight: 830

## file: app.R
library(shiny)

data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/acres92.csv"
download.file(data_url, "acres92.csv")
pop   <- read.csv("acres92.csv")$acres92
N     <- length(pop)
ybarU <- mean(pop)
S     <- sd(pop)

## ---- broken x-axis for the population panel --------------------------------
## The drawn region is [0, 2 * ybarU]: the visible scale [0, cut] plus a
## compressed lane holding the right tail. That total width puts the population
## mean exactly at the midpoint, matching the histogram panel below.
lane_frac <- 0.10
cut       <- 2 * ybarU / (1 + lane_frac)
lane_w    <- lane_frac * cut
xmax_top  <- cut + lane_w                      # == 2 * ybarU

tail_id <- which(pop > cut)
n_tail  <- length(tail_id)
px      <- pop                                 # plotting positions
if (n_tail > 0) {
  r <- rank(pop[tail_id], ties.method = "first") / (n_tail + 1)
  px[tail_id] <- cut + lane_w * (0.10 + 0.80 * r)
}

# fixed jitter band for the rug of raw population values
set.seed(42)
rug_y <- runif(N, -0.42, -0.08)

dens   <- density(pop, from = 0, to = cut)
dens_y <- dens$y / max(dens$y)

## ---- fixed window for the sampling distribution ----------------------------
se100  <- sqrt(1 - 100 / N) * S / sqrt(100)
half   <- 5 * se100
xlim_b <- ybarU + c(-1, 1) * half

MAR <- c(4.5, 4.5, 3, 1.5)   # identical in both panels, so the scales align

ui <- fluidPage(
  titlePanel("Simple random sampling: sampling distribution and CI coverage"),
  sidebarLayout(
    sidebarPanel(
      width = 3,
      sliderInput("n", "Sample size (n)", min = 10, max = 1000, value = 300, step = 10),
      sliderInput("conf", "Confidence level", min = 0.80, max = 0.99, value = 0.95, step = 0.01),
      actionButton("draw", "Draw new sample", class = "btn-primary", width = "100%"),
      br(), br(),
      fluidRow(
        column(6, actionButton("play", "Play", class = "btn-success", width = "100%")),
        column(6, actionButton("stop", "Stop", class = "btn-danger",  width = "100%"))
      ),
      br(),
      sliderInput("speed", "Samples per second", min = 1, max = 10, value = 3, step = 1),
      numericInput("nrep", "Stop after this many samples",
                   value = 500, min = 10, max = 20000, step = 50),
      actionButton("reset", "Reset", width = "100%")
    ),
    mainPanel(
      width = 9,
      plotOutput("popPlot",  height = "340px"),
      plotOutput("histPlot", height = "360px")
    )
  )
)

server <- function(input, output, session) {

  rv <- reactiveValues(
    idx = NULL, mean = NULL, ci = NULL, covered = NULL,
    means = numeric(0), hits = logical(0)
  )
  running <- reactiveVal(FALSE)

  # guard against a blank or absurd entry in the numeric box
  nrep <- reactive({
    v <- input$nrep
    if (is.null(v) || is.na(v) || v < 1) 1L else min(as.integer(v), 20000L)
  })

  draw_one <- function() {
    n    <- isolate(input$n)
    conf <- isolate(input$conf)
    idx  <- sample(N, n)
    s    <- pop[idx]
    m    <- mean(s)
    se   <- sqrt(1 - n / N) * sd(s) / sqrt(n)
    me   <- qt(1 - (1 - conf) / 2, df = n - 1) * se

    rv$idx     <- idx
    rv$mean    <- m
    rv$ci      <- c(m - me, m + me)
    rv$covered <- (m - me) <= ybarU && ybarU <= (m + me)
    rv$means   <- c(rv$means, m)
    rv$hits    <- c(rv$hits, rv$covered)
  }

  # one sample per click; Play repeats until Stop or until the cap is reached
  observeEvent(input$draw, { running(FALSE); draw_one() })
  observeEvent(input$play, running(TRUE))
  observeEvent(input$stop, running(FALSE))

  observe({
    if (!isTRUE(running())) return()
    if (length(isolate(rv$means)) >= isolate(nrep())) {
      running(FALSE)
      return()
    }
    invalidateLater(1000 / isolate(input$speed), session)
    isolate(draw_one())
  })

  # Reset clears the accumulated means and halts any run in progress
  reset_history <- function() {
    running(FALSE)
    rv$idx <- NULL; rv$mean <- NULL; rv$ci <- NULL; rv$covered <- NULL
    rv$means <- numeric(0); rv$hits <- logical(0)
  }
  observeEvent(input$reset, reset_history())
  observeEvent(input$n,     reset_history())
  observeEvent(input$conf,  reset_history())

  fmt <- function(x) format(round(x), big.mark = ",")

  output$popPlot <- renderPlot({
    par(mar = MAR, xaxs = "i")
    plot(dens$x, dens_y, type = "l", col = "grey40", lwd = 2,
         xlim = c(0, xmax_top), ylim = c(-0.62, 1.15),
         xlab = "acres92", ylab = "", xaxt = "n", yaxt = "n", bty = "n",
         main = "Population distribution with the current SRS sample")

    # raw values, jittered; tail units sit in the compressed lane
    points(px, rug_y, pch = 16, cex = 0.5, col = adjustcolor("grey60", 0.5))
    if (!is.null(rv$idx))
      points(px[rv$idx], rug_y[rv$idx], pch = 16, cex = 0.6,
             col = adjustcolor("red", 0.85))

    # axis up to the cut, then the break marker and the lane
    at <- pretty(c(0, cut), 6); at <- at[at <= cut]
    axis(1, at = at, labels = format(at, big.mark = ",", trim = TRUE))
    segments(cut, -0.62, cut, 1.05, col = "grey75", lty = 3)
    text(cut, -0.70, "//", col = "grey40", cex = 1.3, xpd = NA)
    text(cut + lane_w / 2, -0.70,
         sprintf("> %s\n(%d counties)", fmt(cut), n_tail),
         col = "grey40", cex = 0.75, xpd = NA)

    # current sample mean and interval
    if (!is.null(rv$mean)) {
      segments(rv$ci[1], -0.55, rv$ci[2], -0.55, col = "blue", lwd = 3)
      points(rv$mean, -0.55, pch = 19, cex = 1.3, col = "blue")
    }
    abline(v = ybarU, col = "black", lwd = 2, lty = 2)

    legend("topright", bty = "n", inset = c(0.12, 0),
           legend = c("Population", "Current sample", "Sample mean and CI", "True mean"),
           col = c("grey60", "red", "blue", "black"),
           pch = c(16, 16, 19, NA), lty = c(NA, NA, NA, 2))
  })

  output$histPlot <- renderPlot({
    par(mar = MAR, xaxs = "i")

    if (length(rv$means) == 0) {
      plot(1, type = "n", xlim = xlim_b, ylim = c(0, 1), bty = "n", yaxt = "n",
           xlab = "Sample mean of acres92", ylab = "",
           main = "Sampling distribution of the sample mean")
      abline(v = ybarU, col = "black", lwd = 2, lty = 2)
      text(ybarU, 0.5, "Click 'Draw new sample' or 'Play' to begin", col = "grey50")
      return(invisible())
    }

    # fixed bin width; breaks extended to cover means outside the fixed window
    bw <- 2 * half / 40
    lo <- min(xlim_b[1], min(rv$means)) - bw
    hi <- max(xlim_b[2], max(rv$means)) + bw
    h  <- hist(rv$means, breaks = seq(lo, hi, by = bw), plot = FALSE)

    # theoretical sampling distribution of the mean at the current sample size
    n         <- input$n
    se_theory <- sqrt(1 - n / N) * S / sqrt(n)
    curve_pk  <- dnorm(ybarU, mean = ybarU, sd = se_theory)

    vis <- h$density[h$mids >= xlim_b[1] & h$mids <= xlim_b[2]]
    top <- max(c(vis, curve_pk, 1e-6))

    plot(h, freq = FALSE, col = "skyblue", border = "white",
         xlim = xlim_b, ylim = c(0, top * 1.30),
         xlab = "Sample mean of acres92", ylab = "Density",
         main = "Sampling distribution of the sample mean")
    curve(dnorm(x, mean = ybarU, sd = se_theory), from = xlim_b[1], to = xlim_b[2],
          col = "darkorange", lwd = 2, add = TRUE)
    abline(v = ybarU, col = "black", lwd = 2, lty = 2)

    # transient interval, for the current sample only
    ci_col <- if (rv$covered) "forestgreen" else "red"
    y      <- top * 0.9
    segments(rv$ci[1], y, rv$ci[2], y, col = ci_col, lwd = 3)
    segments(rv$ci, y - top * 0.045, rv$ci, y + top * 0.045, col = ci_col, lwd = 3)
    points(rv$mean, y, pch = 19, cex = 1.3, col = ci_col)

    outside <- sum(rv$means < xlim_b[1] | rv$means > xlim_b[2])
    leg <- c(
      sprintf("True mean = %s", fmt(ybarU)),
      sprintf("Theoretical SE (n = %d) = %s", n, fmt(se_theory)),
      sprintf("Samples drawn = %d of %d", length(rv$means), nrep()),
      sprintf("Empirical coverage = %.1f%% (nominal %.0f%%)",
              100 * mean(rv$hits), 100 * input$conf),
      sprintf("Misses = %d of %d", sum(!rv$hits), length(rv$hits))
    )
    if (outside > 0)
      leg <- c(leg, sprintf("%d mean(s) beyond the axis", outside))
    legend("topright", bty = "n", cex = 0.95,
           legend = c("Theoretical Normal", leg),
           lty = c(1, rep(NA, length(leg))),
           lwd = c(2, rep(NA, length(leg))),
           col = c("darkorange", rep(NA, length(leg))))
  })
}

shinyApp(ui, server)

```


## Prediction-Based Theory for Simple Random Sampling {#sec-pred-srs}

### Setup {#sec-pred-setup}

Let $U=\{1,\dots,N\}$ index the finite population and let $Y_i$ denote the study variable. A sample $S\subset U$ of size $n$ is observed; write $R=U\setminus S$ for the $N-n$ non-sampled units. The targets are
$$
T=\sum_{i\in U}Y_i,\qquad \bar Y_U=T/N .
$$

The prediction approach treats $Y_1,\dots,Y_N$ as random variables generated by a model $\xi$, and treats $S$ as fixed. All expectations below are with respect to $\xi$, not with respect to the randomization distribution. The role of simple random sampling is only to guarantee that selection carries no information about $Y_i$, so that the model assumed for the sampled units also holds for the non-sampled ones.

Throughout, the working model is
$$
\xi:\quad E(Y_i)=\mu,\quad \operatorname{Var}(Y_i)=\sigma^2,\quad \operatorname{Cov}(Y_i,Y_j)=0\ (i\ne j).
$$ {#eq-model}

It is convenient to write the model errors as
$$
e_i=Y_i-\mu,\qquad E(e_i)=0,\quad \operatorname{Var}(e_i)=\sigma^2,\quad \operatorname{Cov}(e_i,e_j)=0\ (i\ne j),
$$ {#eq-errors}
and $\bar e_A=|A|^{-1}\sum_{i\in A}e_i$ for any index set $A$. Every derivation below proceeds by subtracting and adding $\mu$ so that the prediction error is expressed in the $e_i$, whose moments are given by @eq-errors.

Any predictor of $T$ has the form $\hat T=\sum_{i\in U}\hat Y_i$, and its accuracy is measured by the **mean squared prediction error**
$$
\operatorname{MSPE}(\hat T)=E\left[\left(\hat T-T\right)^{2}\right]
=E\left[\left(\sum_{i\in U}(\hat Y_i-Y_i)\right)^{2}\right].
$$ {#eq-mspe}

The name matters: $T$ is a random quantity to be predicted, not a fixed parameter to be estimated, so @eq-mspe measures prediction error rather than estimation error.

### Known $\mu$ and $\sigma^2$ {#sec-pred-known-mu}

Suppose $\mu$ and $\sigma^2$ are known and we set $\hat Y_i=\mu$ for every $i\in U$, ignoring the observed data. Then $\hat T=N\mu$, and subtracting and adding $\mu$ inside each term gives
$$
\hat T-T=\sum_{i\in U}\left(\mu-Y_i\right)=-\sum_{i\in U}\left(Y_i-\mu\right)=-\sum_{i\in U}e_i .
$$

The prediction error is a sum of zero-mean model errors, so $\hat T$ is model-unbiased, and
$$
\operatorname{MSPE}(\hat T)=E\left[\left(\sum_{i\in U}e_i\right)^{2}\right]
=\sum_{i\in U}E(e_i^2)+\sum_{i\ne j}E(e_ie_j)=N\sigma^2 ,
$$ {#eq-mspe-blind}
the cross-products vanishing by @eq-errors. For the population mean, $\hat{\bar y}_U=\mu$ gives
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\frac{\sigma^2}{N}.
$$

Two points deserve emphasis. First, the error is not zero even though the model is known exactly: $T$ is a realized sum of random variables, so it cannot be recovered from $\mu$ alone. This is the essential difference between predicting a finite-population quantity and estimating a superpopulation parameter. Second, $\sigma^2/N$ is negligible for large $N$, which is why the distinction is usually invisible in treatments that condition on the population.

The predictor $N\mu$ is nonetheless wasteful, since it predicts $Y_i$ for units whose values we have already seen. Replacing $\hat Y_i$ by $Y_i$ for $i\in S$ gives
$$
\hat T=\sum_{i\in S}Y_i+(N-n)\mu,\qquad
\hat T-T=\sum_{i\in R}\left(\mu-Y_i\right)=-\sum_{i\in R}e_i ,
$$
so that
$$
\operatorname{MSPE}(\hat T)=(N-n)\sigma^2 .
$$ {#eq-mspe-oracle}

The $n$ observed terms have simply dropped out. Expression @eq-mspe-oracle is the irreducible part of the problem: the unobserved units vary about their mean, and no predictor can remove that variation. The result of the next section is @eq-mspe-oracle plus a penalty for not knowing $\mu$.

### Unknown $\mu$: prediction by the sample mean {#sec-pred-sample-mean}

Now let $\mu$ be unknown and estimate it by $\bar y_S=n^{-1}\sum_{i\in S}Y_i$. Take
$$
\hat Y_i=
\begin{cases}
Y_i, & i\in S,\\
\bar y_S, & i\in R,
\end{cases}
\qquad\text{so that}\qquad
\hat T=\sum_{i\in S}Y_i+(N-n)\bar y_S=N\bar y_S .
$$ {#eq-expansion}

The familiar expansion estimator thus arises as a predictor, not as a design-unbiased estimator. Its prediction error involves only the non-sampled units, and subtracting and adding $\mu$ in each term separates the two sources of error:
$$
\hat T-T=\sum_{i\in R}\left(\bar y_S-Y_i\right)
=\sum_{i\in R}\Big\{\underbrace{(\bar y_S-\mu)}_{\bar e_S}-\underbrace{(Y_i-\mu)}_{e_i}\Big\}
=(N-n)\,\bar e_S-\sum_{i\in R}e_i .
$$ {#eq-error-decomp}

The first term in @eq-error-decomp is the error made in locating $\mu$ from the sample; the second is the intrinsic scatter of the unobserved units about $\mu$. Both have mean zero, so $\hat T$ is model-unbiased. Squaring and taking expectations,
$$
\operatorname{MSPE}(\hat T)=(N-n)^2E\left(\bar e_S^{\,2}\right)
-2(N-n)E\left(\bar e_S\sum_{i\in R}e_i\right)
+E\left[\left(\sum_{i\in R}e_i\right)^{2}\right].
$$

The three terms are evaluated separately. Since $E(\bar e_S^{\,2})=\sigma^2/n$, the first equals $(N-n)^2\sigma^2/n$. The cross term vanishes because $S$ and $R$ are disjoint, so every product $e_ie_j$ appearing in it has $i\ne j$. The third equals $(N-n)\sigma^2$, again by @eq-errors. Hence
$$
\operatorname{MSPE}(\hat T)=\frac{(N-n)^2}{n}\sigma^2+(N-n)\sigma^2
=(N-n)\sigma^2\left(\frac{N-n}{n}+1\right)
=\frac{N(N-n)}{n}\,\sigma^2 ,
$$ {#eq-mspe-total}
and, dividing by $N^2$,
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\frac{\sigma^2}{n}\left(1-\frac{n}{N}\right).
$$ {#eq-mspe-mean}

Expression @eq-mspe-mean is the classical simple-random-sampling variance, with the superpopulation variance $\sigma^2$ in place of the population variance $S^2$, obtained without any appeal to randomization. Three features are worth drawing out.

The two components of @eq-mspe-total have distinct meanings. The term $(N-n)\sigma^2$ is the irreducible variation identified in @eq-mspe-oracle; the term $(N-n)^2\sigma^2/n$ is the price of estimating $\mu$ from $n$ observations, and it alone shrinks as $n$ grows.

The finite-population correction is not an artifact of sampling without replacement. It emerges from the arithmetic of a census of $S$: the $n$ observed units contribute exactly zero to @eq-error-decomp, so only $N-n$ terms remain to be predicted. When $n=N$ the MSPE is zero because nothing is left to predict.

Finally, comparing @eq-mspe-total with @eq-mspe-oracle, the cost of not knowing $\mu$ is the factor $N/n$. For the population mean, the MSPE rises from $\sigma^2/N$ to approximately $\sigma^2/n$; the sample size, not the population size, governs precision.

A natural next step is to estimate $\sigma^2$ by the sample variance $s_S^2$ and to verify that $E(s_S^2)=\sigma^2$ under @eq-model, which yields the usual variance estimator and closes the loop with the design-based formulas.
