---
title: "Simple Random Sampling"
filters:
  - shinylive
execute:
  warning: false
  message: false
  cache: true
---


## Estimators for Simple Random Sampling {#sec-srs-formulas}

A simple random sample (SRS) of $n$ units is drawn **without replacement** from a
population of $N$ units, so that every subset of $n$ units is equally likely.
Write $y_i$ for the value of unit $i$, $\bar{y}_U = \frac{1}{N}\sum_{i=1}^N y_i$ for the
population mean, $t = N\bar{y}_U$ for the population total, and
$$
S^2 = \frac{1}{N-1}\sum_{i=1}^N (y_i - \bar{y}_U)^2
$$
for the population variance.

#### 1. Estimating the Population Mean {-}

The sample mean is unbiased for $\bar{y}_U$:
$$
\bar{y} = \frac{1}{n}\sum_{i=1}^n y_i .
$$
Its true variance and its estimate, which replaces $S^2$ by the sample variance $s^2$, are
$$
V(\bar{y}) = \left(1 - \frac{n}{N}\right)\frac{S^2}{n}, \qquad
\hat{V}(\bar{y}) = \left(1 - \frac{n}{N}\right)\frac{s^2}{n}, \qquad
s^2 = \frac{1}{n-1}\sum_{i=1}^n (y_i - \bar{y})^2 ,
$$
and $\mathrm{SE}(\bar{y}) = \sqrt{\hat{V}(\bar{y})}$. The factor $1 - n/N$ is the
**finite population correction** (fpc). It is close to $1$ when the sampling fraction $n/N$ is small.

#### 2. Estimating the Population Total {-}

$$
\hat{t} = N\bar{y}, \qquad
\hat{V}(\hat{t}) = N^2\,\hat{V}(\bar{y}) = N^2\left(1 - \frac{n}{N}\right)\frac{s^2}{n}, \qquad
\mathrm{SE}(\hat{t}) = N\,\mathrm{SE}(\bar{y}).
$$

#### 3. Estimating a Proportion {-}

For a 0/1 indicator $y_i$, the population mean is the proportion $p$, and the sample mean is the sample proportion $\hat{p}$.
Since then $s^2 = \frac{n}{n-1}\hat{p}(1-\hat{p})$,
$$
\hat{p} = \frac{1}{n}\sum_{i=1}^n y_i, \qquad
\hat{V}(\hat{p}) = \left(1 - \frac{n}{N}\right)\frac{\hat{p}(1 - \hat{p})}{n-1}.
$$

#### 4. Confidence Intervals {-}

By the central limit theorem for sample means, an approximate $95\%$ confidence interval is
$$
\bar{y} \pm t_{0.975,\,n-1}\,\mathrm{SE}(\bar{y}),
$$
and likewise for $\hat{t}$ and $\hat{p}$. With a skewed population, $n$ must be larger for the
normal approximation, and hence the nominal coverage, to hold.
The simulation and the app later in this chapter show this.

---

## Analysis of agsrs.csv Data

Throughout the agriculture examples in this book, county acreages are expressed
in **thousands of acres** --- the raw figures run into the millions, and every
mean, variance, and total below reads far more easily on that scale. The helper
`acres_in_thousands()` performs the conversion as the data is read.

### Step by step calculation without using a function

To evaluate a Simple Random Sample (SRS), we calculate the point estimate, the standard error (incorporating the Finite Population Correction factor), and the margin of error to construct a 95% confidence interval.

**Key Formulas:**

* **Sample Mean:** $\bar{y} = \frac{1}{n} \sum_{i=1}^n y_i$
* **Standard Error:** $\mathrm{SE}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left( \frac{s}{\sqrt{n}} \right)$
* **Margin of Error:** $\mathrm{ME} = t_{\alpha/2, n-1} \times \mathrm{SE}(\bar{y})$
* **Population Total:** $\hat{t} = N\bar{y}$

```{r srs-manual-calc}
## read survey data
agsrs <- read.csv ("data/agsrs.csv")
## work in thousands of acres, the unit used for every agriculture example here
agsrs[c("acres92", "acres87", "acres82")] <- agsrs[c("acres92", "acres87", "acres82")] / 1000
head(agsrs)

## extract the variable of interest
sdata <- agsrs$acres92
N <- 3078

## do calculation
n <- length (sdata)
ybar <- mean (sdata)
se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
mem <- qt (0.975, df = n - 1) * se.ybar

## return estimate vector for pop mean
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)

## return estimate vector for pop total
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem) * N

```

### Write a function for repeated use

#### A function for doing data analysis for srs sample

Wrap the previous manual calculations into a custom R function to quickly estimate means, standard errors, and confidence intervals for any variable of interest. To avoid repeating this code in every chapter, all estimating functions (and their `gt`-table formatting helpers) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

#### Apply srs_est to agsrs.csv data

*Import Data*

Load the survey sample data representing US counties.

```{r agsrs-load}
agsrs <- acres_in_thousands (read.csv ("data/agsrs.csv"))

```

*Estimating the mean of acre92*

Calculate the estimated average farm acreage per county using the custom function. The function's `$table` shows the row-level working values (truncated to the first 6 rows + Sum), and `$estimate` holds the final point estimate, SE, and CI, formatted with `format_est_gt()`.

```{r agsrs-mean-est}
res_mean <- srs_est(agsrs[,"acres92"], N = 3078, estimate = "mean")
res_mean$table
res_mean$estimate |> format_est_gt("mean")

```

*Estimating the total of acre92*

Set `estimate = "total"` to have the function itself scale the mean and confidence interval by the population size ($N = 3078$) to estimate the total farm acreage across all counties.

```{r agsrs-total-est}
res_total <- srs_est(agsrs[,"acres92"], N = 3078, estimate = "total", show.details = FALSE)
res_total$estimate |> format_est_gt("total")

```

*Estimating the proportion of counties with fewer than 200K acres for farming in 1992*

To estimate a proportion under SRS, create an indicator variable ($y_i = 1$ if true, $0$ if false). The mean of this indicator variable serves as the sample proportion ($\hat{p} = \bar{y}_{indicator}$).

```{r agsrs-prop-est}
acres92.is.fewer.200k <- as.numeric (agsrs[,"acres92"] < 200)   # 200 thousand acres
head(acres92.is.fewer.200k)

res_prop <- srs_est(acres92.is.fewer.200k, N = 3078, estimate = "mean", show.details = FALSE)
res_prop$estimate |> format_est_gt("mean")

```

*Estimating the total number of counties with fewer than 200K acres for farming in 1992*

Set `estimate = "total"` to estimate the absolute count of counties meeting the criteria.

```{r agsrs-prop-total-est}
res_prop_total <- srs_est(acres92.is.fewer.200k, N = 3078, estimate = "total", show.details = FALSE)
res_prop_total$estimate |> format_est_gt("total")

```

### Comparing with true value

Load the complete census (population) dataset to calculate the true parameters. This allows us to assess the accuracy of our previous SRS estimates against the exact population values.

```{r agpop-true-values}
agpop <- acres_in_thousands (read.csv ("data/agpop.csv"))
#true mean (thousands of acres)
mean (agpop[, "acres92"], na.rm = T)
# true total (thousands of acres)
sum (agpop[, "acres92"], na.rm = T)

# true proportion of counties with less than 200K acres for farming
mean (agpop[, "acres92"] < 200, na.rm = T)
# true number of counties with less than 200K acres for farming
sum (agpop[, "acres92"] < 200, na.rm = T)

```

## A Simulation Demonstration of SRS Inference

Load the full population data and filter out missing values to serve as a complete, known finite population for resampling experiments.

```{r agpop-load-clean}
# read population data, with acreages in thousands of acres
agpop <- acres_in_thousands (read.csv ("data/agpop.csv"))
# remove those counties with na
agpop <- subset( agpop, !is.na (acres92))

```

### True Values

Define the target sample size and compute the true population mean ($\mu$) and theoretical standard error ($\mathrm{SE}_{true}$) using the known population standard deviation ($S$).

**Theoretical Standard Error:**


$$\mathrm{SE}_{true}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left(\frac{S}{\sqrt{n}}\right)$$

```{r srs-true-values}
# sample size
n <- 300
# population size
N <- nrow (agpop); N

# true value of population mean
ybarU <- mean (agpop[,"acres92"]); ybarU
# true value of deviation of sample mean
true.se.ybar <- sqrt (1- n/N) * sd (agpop[,"acres92"]) / sqrt (n); true.se.ybar

```

### One SRS sampling

Draw a single random sample of size $n=300$ from the population and run our estimator function to see a single instance of survey inference.

```{r srs-one-sample}
##
# srs sampling
srs <- sample (1:N,n)
head(agpop [srs, ])
# get data of variable "acres92"
sdata <- agpop [srs, "acres92"]
# analysis
res_one <- srs_est(sdata, N, estimate = "mean", show.details = FALSE)
res_one$estimate |> format_est_gt("mean")

```

### Repeating SRS sampling 5000 times

To demonstrate the Central Limit Theorem and unbiasedness of the estimator, draw 5,000 independent samples, storing the estimates from each iteration. We plot the resulting sampling distribution against the true population mean.

*(Note: We use `show.details = FALSE` and pull out just `$estimate` here, so the loop skips building 5,000 working tables and stays fast.)*

```{r fig-srs-simulation-distribution, fig.height=10}
#| fig-cap: "Population distribution and sampling distribution of the SRS sample mean over 5000 repeated samples (n = 300)"
nres <- 5000 # number of repeated sampling
simulation.results <- matrix (0, nres, 4) # matrix recording repeated results
colnames(simulation.results) <- c( "Est.",   "S.E.",   "ci.low", "ci.upp")

for (i in 1:nres)
{
    srs <- sample (N, n)
   sdata <- agpop [srs, "acres92"]
   simulation.results [i,] <- srs_est (sdata, N, estimate = "mean", show.details = FALSE)$estimate
}

head(simulation.results)
# look at the distribution of sample mean
par (mfrow= c(2,2))
hist (agpop$acres92,main = "Population Distribution of acre92")
hist (simulation.results[,1], main = "Sampling Distribution of Sample Mean for acre92")
abline (v = ybarU, col = "red")
qqnorm (simulation.results[,1], main="QQ plot of Sample Mean"); qqline(simulation.results[,1])
boxplot (simulation.results[,1], main = "Boxplot of Sample Mean")
abline (h = ybarU, col = "red")

mean (simulation.results[,1])
ybarU
sd (simulation.results [,1])
true.se.ybar

```

### Empirical Coverage Rate of CIs

Verify the reliability of the 95% confidence intervals by checking whether the true population mean falls within the calculated bounds for each of the 5,000 samples. The empirical coverage rate should closely approximate 0.95. The plot highlights the first 100 intervals, flagging those that fail to capture the true mean in a distinct color.

```{r srs-coverage-setup}
library(dplyr)
library(plotrix)

# Convert to data frame (if it was a matrix) and use mutate to evaluate coverage
simulation.results <- as.data.frame(simulation.results) |>
  mutate(
    `Covered?` = as.numeric(ci.low < ybarU & ybarU < ci.upp)
  )

head(simulation.results)
```

```{r fig-srs-coverage-intervals, fig.height=6}
#| fig-cap: "First 100 simulated 95% confidence intervals for the SRS sample mean, with coverage failures highlighted"
par(mfrow=c(1,1))
plotCI(
  x = 1:100,
  y = simulation.results$Est.[1:100], 
  li = simulation.results$ci.low[1:100],
  ui = simulation.results$ci.upp[1:100],
  col = 2 - simulation.results$`Covered?`[1:100]
)
abline(h = ybarU, col = "blue")

# Empirical coverage rate
mean(simulation.results$`Covered?`)

```

## Shinylive App for Illustrating SRS Theory {#sec-app-srs}

The simulation for this section runs as a standalone app:
[Simple random sampling](app-srs.html){target="_blank"}.
It opens in a new tab, together with its documentation and the full
[list of apps](shinyliveapps_sampling.html){target="_blank"}.



## Prediction-Based Theory for Sampling (Optional) {#sec-pred-srs}

Sampling is a prediction problem. The sample $S$ plays the role of a training set, the non-sampled units $R=U\setminus S$ play the role of test cases, and a population total or mean is a sum of predictions for the test cases plus the observed training values. This section states the framework in general, with a model for every unit and a generic estimator of the model parameters, and then specializes it to simple random sampling, where the mean of every unit is the same and is estimated by $\bar y_S$.

### Setup {#sec-pred-setup}

Let $U=\{1,\dots,N\}$ index a finite population with study variable $Y_i$, let $S\subset U$ with $|S|=n$ denote the observed sample, and let $R=U\setminus S$ denote the $N-n$ non-sampled units. The quantities of interest are
$$
T=\sum_{i\in U}Y_i=\sum_{i\in S}Y_i+\sum_{i\in R}Y_i,\qquad \bar Y_U=T/N .
$$
Only the second sum is unknown, so the task is to predict $Y_i$ for each $i\in R$.

The prediction approach treats $Y_1,\dots,Y_N$ as random variables and $S$ as fixed. All expectations are taken with respect to the model $\xi$ rather than the randomization distribution. The sampling design enters only through the requirement that selection be uninformative, so that $\xi$ governs the units in $R$ as well as those in $S$. The general working model is
$$
\xi:\quad E(Y_i)=\mu_i(\theta),\quad \operatorname{Var}(Y_i)=\sigma_i^2,\quad
\operatorname{Cov}(Y_i,Y_j)=0\ (i\neq j),
$$ {#eq-model}
where $\theta$ is an unknown parameter and $\mu_i(\theta)$ is the mean function, for example $\mu_i(\theta)=x_i^{\top}\beta$ in regression. The model errors are
$$
e_i=Y_i-\mu_i(\theta),\qquad E(e_i)=0,\quad E(e_i^2)=\sigma_i^2,\quad E(e_ie_j)=0\ (i\neq j).
$$ {#eq-errors}

A prediction rule estimates $\theta$ from the sample by $\hat\theta=\hat\theta(Y_S)$ and predicts
$$
\hat Y_i=
\begin{cases}
Y_i, & i\in S,\\
\mu_i(\hat\theta), & i\in R,
\end{cases}
\qquad
\hat T=\sum_{i\in U}\hat Y_i=\sum_{i\in S}Y_i+\sum_{i\in R}\mu_i(\hat\theta).
$$ {#eq-predictor}
Observed values are kept as they are, and the fitted mean function is used only for the unobserved units. Because $\hat\theta$ is a function of $Y_S$ alone and the errors are uncorrelated, $\hat\theta$ is uncorrelated with $e_i$ for every $i\in R$. This is the only property of the estimator used below.

### Two sources of error {#sec-pred-two-sources}

**Individual prediction.** In machine learning a predictor is judged by its mean squared error on a test case. For $i\in R$ add and subtract $\mu_i(\theta)$:
$$
\hat Y_i-Y_i=\underbrace{\mu_i(\hat\theta)-\mu_i(\theta)}_{\delta_i}-\underbrace{\big(Y_i-\mu_i(\theta)\big)}_{e_i}.
$$ {#eq-error-individual}
The first part, $\delta_i$, is the error in the fitted mean caused by the uncertainty in $\hat\theta$, and the second part, $e_i$, is the gap between the unobserved $Y_i$ and its true mean. The cross term $E(\delta_ie_i)$ vanishes, so
$$
E\big(\hat Y_i-Y_i\big)^2=\underbrace{\sigma_i^2}_{\text{irreducible}}+\underbrace{E\big[\mu_i(\hat\theta)-\mu_i(\theta)\big]^2}_{\text{estimation error}} .
$$ {#eq-mse-individual}
The irreducible term is present even when $\theta$ is known, because no predictor removes the scatter of $Y_i$ about its mean. The estimation term is the mean squared error of $\mu_i(\hat\theta)$ as an estimator of $\mu_i(\theta)$ and decreases as the training sample grows. This is the usual decomposition of test error into noise and estimation variance, plus squared bias if the estimator is biased.

**The total.** The observed units contribute no error, so
$$
\hat T-T=\sum_{i\in R}(\hat Y_i-Y_i)=\sum_{i\in R}\delta_i-\sum_{i\in R}e_i .
$$ {#eq-error-decomp}
The two sources of error appear again, now summed over the test cases. Squaring, the cross term between $\sum_R\delta_i$ and $\sum_R e_i$ vanishes by the same argument as before, and the $e_i$ are uncorrelated, so
$$
\operatorname{MSPE}(\hat T)=E\big[(\hat T-T)^2\big]
=\underbrace{\sum_{i\in R}\sigma_i^2}_{\text{scatter of the unobserved }Y_i}
+\underbrace{E\Big[\Big(\sum_{i\in R}\delta_i\Big)^{2}\Big]}_{\text{uncertainty in }\hat\theta} .
$$ {#eq-mspe}
Dividing by $N^2$ gives $\operatorname{MSPE}(\hat{\bar y}_U)$.

Comparing @eq-mse-individual and @eq-mspe shows what changes when the target moves from a single case to the total. The irreducible parts simply add, $\sum_{i\in R}\sigma_i^2$. The estimation parts do not add, because the errors $\delta_i$ all come from the same $\hat\theta$ and are strongly correlated. Expanding the square gives
$$
E\Big[\Big(\sum_{i\in R}\delta_i\Big)^{2}\Big]=\sum_{i\in R}E(\delta_i^2)+\sum_{i\neq j\in R}E(\delta_i\delta_j),
$$
and the double sum is typically of larger order than the first. A total is a more aggregated target than an individual prediction, and its error is dominated by the shared error in $\hat\theta$, which does not average out across units.

### Specialization to simple random sampling {#sec-pred-sample-mean}

For simple random sampling there are no auxiliary variables, so every unit has the same mean,
$$
\mu_i(\theta)=\mu,\quad \sigma_i^2=\sigma^2,\qquad e_i=Y_i-\mu,\quad \bar e_A=|A|^{-1}\sum_{i\in A}e_i,
$$
and $\theta=\mu$ is estimated by the sample mean $\bar y_S=n^{-1}\sum_{i\in S}Y_i$. The predictor @eq-predictor is
$$
\hat Y_i=
\begin{cases}
Y_i, & i\in S,\\
\bar y_S, & i\in R,
\end{cases}
\qquad
\hat T=\sum_{i\in S}Y_i+(N-n)\bar y_S=N\bar y_S ,
$$ {#eq-expansion}
so the expansion estimator is a prediction rule that fills in every unobserved unit with the sample mean. Here the estimation error is common to all units, $\delta_i=\bar y_S-\mu=\bar e_S$, and @eq-error-decomp becomes
$$
\hat T-T=(N-n)\,\bar e_S-\sum_{i\in R}e_i .
$$ {#eq-error-srs}
Both terms have mean zero, so $\hat T$ is model-unbiased.

**Individual units.** From @eq-mse-individual, with $E(\bar e_S^{\,2})=\sigma^2/n$,
$$
E\big(\hat Y_i-Y_i\big)^2=\sigma^2+\frac{\sigma^2}{n}=\sigma^2\Big(1+\frac1n\Big),\qquad i\in R .
$$ {#eq-mse-srs-individual}

**The total.** Since $\delta_i=\bar e_S$ for all $i\in R$, the double sum in @eq-mspe is as large as it can be: $E[(\sum_R\delta_i)^2]=(N-n)^2E(\bar e_S^{\,2})=(N-n)^2\sigma^2/n$, whereas independent errors would give $(N-n)\sigma^2/n$. The scatter term is $(N-n)\sigma^2$, and therefore
$$
\operatorname{MSPE}(\hat T)=\frac{(N-n)^2}{n}\sigma^2+(N-n)\sigma^2
=\frac{N(N-n)}{n}\,\sigma^2 ,
$$ {#eq-mspe-total}
$$
\operatorname{MSPE}(\hat{\bar y}_U)=\frac{\sigma^2}{n}\left(1-\frac{n}{N}\right).
$$ {#eq-mspe-mean}

Expression @eq-mspe-mean is the classical simple-random-sampling variance with $\sigma^2$ in place of the population variance $S^2$, obtained without appeal to randomization. Four consequences follow.

1. The two terms of @eq-mspe-total are the two sources of error. The term $(N-n)\sigma^2$ is the irreducible scatter of the unobserved units; the term $(N-n)^2\sigma^2/n$ is the cost of estimating $\mu$ from $n$ observations and alone decreases with $n$.
2. The per-unit error $\sigma^2(1+1/n)$ is barely above $\sigma^2$ for moderate $n$, so the estimation error looks negligible for individual prediction. For the total it is not: the estimation error is shared by all $N-n$ unobserved units and accumulates to a factor $(N-n)^2$, whereas the scatter accumulates only to $N-n$. The ratio of the two terms is $(N-n)/n$.
3. The finite-population correction is not an artifact of sampling without replacement. The $n$ observed units contribute exactly zero to @eq-error-srs, leaving $N-n$ terms to predict; at $n=N$ the MSPE is zero.
4. If $\mu$ is known, $\delta_i=0$ and the predictor $\hat Y_i=\mu$ for $i\in R$ has $\operatorname{MSPE}(\hat T)=(N-n)\sigma^2$. This is the irreducible error, and the cost of not knowing $\mu$ is the factor $N/n$ applied to it. For the population mean the MSPE rises from $\sigma^2(N-n)/N^2$ to approximately $\sigma^2/n$, so precision is governed by the sample size rather than the population size.

Other sampling problems fit the same framework by changing $\mu_i(\theta)$: a ratio or regression estimator uses $\mu_i(\theta)=\beta x_i$ or $\beta_0+\beta_1x_i$, and a stratified estimator uses a separate mean $\mu_h$ in each stratum. Only $\hat\theta$ and the shared estimation error $\sum_R\delta_i$ change.

### Illustration {#sec-pred-demo}

@fig-srs-index displays one realization of a population under @eq-model with $\mu_i=\mu$. Since the labels are uninformative, the sample is placed on the left, shaded grey, and the units to be predicted on the right.

```{r}
#| echo: false
#| label: fig-srs-index
#| fig-cap: "One realized population under the model. The sample occupies the shaded region on the left; the remaining units are to be predicted. Horizontal lines: the model mean $\\mu$ (orange), the estimand $\\bar Y_U$ (red), and the predictor $\\bar y_S$ (blue, dashed)."
#| fig-width: 7.5
#| fig-height: 4.2

set.seed(20000)
N <- 100; n <- 10; mu <- 50; sigma <- 20

y <- rnorm(N, mean = mu, sd = sigma)   # one realization of Y_1, ..., Y_N
S <- 1:n                               # sample, placed on the left
R <- (n + 1):N                         # non-sampled units

ybar_S <- mean(y[S])                   # predictor of mu
ybar_U <- mean(y)                      # the estimand

ylim <- range(y) + c(-1, 1) * 0.10 * diff(range(y))
plot(seq_len(N), y, type = "n", ylim = ylim,
     xlab = "unit index", ylab = expression(y[i]))
rect(0.5, ylim[1], n + 0.5, ylim[2], col = "grey90", border = NA)
points(S, y[S], pch = 16, cex = 0.7)
points(R, y[R], pch = 1,  cex = 0.7, col = "grey45")
abline(h = mu,     col = "orange", lwd = 2)
abline(h = ybar_U, col = "red",    lwd = 2)
abline(h = ybar_S, col = "blue",   lwd = 2, lty = 2)
text(n / 2,       ylim[2], "observed (S)",        pos = 1, cex = 0.85)
text((n + N) / 2, ylim[2], "to be predicted (R)", pos = 1, cex = 0.85)
legend("bottomright", bty = "n", cex = 0.85, lwd = 2,
       col = c("orange", "red", "blue"), lty = c(1, 1, 2),
       legend = c(expression(mu), expression(bar(Y)[U]), expression(bar(y)[S])))
box()
```

The scatter of the open points about the orange line is the irreducible error $\sigma_i^2$ in @eq-mse-individual: it is the error of predicting an individual unit even with $\mu$ known. The gap between the orange and blue lines is the estimation error $\delta_i=\bar e_S$, with root mean square $\sigma/\sqrt{n}$, which is the same for every unit in $R$. The gap between the blue and red lines, scaled by $N$, is the realized prediction error of the total; @eq-mspe-total states that its root mean square is $\sigma\sqrt{N(N-n)/n}$.

```{r}
c(mu = mu, Ybar_U = ybar_U, ybar_S = ybar_S,
  error_total  = N * (ybar_S - ybar_U),
  rmspe_total  = sigma * sqrt(N * (N - n) / n))
```

Estimating $\sigma^2$ by the sample variance $s_S^2$, which satisfies $E(s_S^2)=\sigma^2$ under @eq-model, yields the usual variance estimator and recovers the design-based formulas.
