# Stratified Random Sampling

## Key Formulas for Stratified Sampling

Stratified random sampling divides a population of size $N$ into $H$ mutually exclusive strata. A simple random sample of size $n_h$ is then drawn from each stratum of size $N_h$. This design often yields higher precision than Simple Random Sampling (SRS).

#### 1. Stratified Population Mean Estimator {-}
The estimated population mean is a weighted average of the sample means from each stratum:
$$
\overline{y}_{str} = \sum_{h=1}^H \pi_h \overline{y}_h.
$$
Where $\pi_h = \frac{N_h}{N}$ is the stratum weight, and $\overline{y}_h$ is the sample mean of stratum $h$.

#### 2. Variance of the Stratified Mean {-}
The estimated variance accounts for the finite population correction in each stratum:
$$
\hat{V}(\overline{y}_{str}) = \sum_{h=1}^H \pi_h^2 \left(1 - \frac{n_h}{N_h}\right) \frac{s_h^2}{n_h}.
$$
Where $s_h^2$ is the sample variance in stratum $h$.

#### 3. Proportional Allocation {-}
Sample sizes are allocated proportional to the size of the stratum:
$$
n_h = n \left( \frac{N_h}{N} \right).
$$

#### 4. Neyman (Optimal) Allocation {-}
If surveying costs are equal across strata, Neyman allocation minimizes the variance of the estimator by sampling more heavily from larger and more highly variable strata:
$$
n_h = n \frac{N_h S_h}{\sum_{h=1}^H N_h S_h}
$$

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `str_est`, `str_est_data`, and `srs_est`, used below to compute the stratified mean, standard error, and confidence intervals) live in a single shared file, `samplingestimate.r`, which we source below. `str_est_data` computes these directly from a dataset, while `str_est` computes them from pre-calculated summary statistics.

```{r source-functions}
#| echo: false
library(latex2exp)
library(sampling)
library(dplyr)
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Analysis of "agstrat.csv" data

We will estimate the average 1992 farm acreage (`acres92`) across US counties, using geographic region (`region`) as the stratification variable.

### Importing agstrat.csv data

```{r agstrat-load}
agstrat <- read.csv("data/agstrat.csv")
head(agstrat) |> gt()
```
**Output Note:** The table above shows the sample survey data, including the stratum (region) and weights for each county.

### Spreadsheet calculation

#### Summarizing acre92 in each stratum

We first manually compute the sample size ($n_h$), standard deviation ($s_h$), and mean ($\overline{y}_h$) for each region, matching them with the known total counties in each region ($N_h$).

```{r agstrat-summary-table}
nh <- tapply (agstrat[, "acres92"], agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information (true population size per stratum)
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

data.frame(Nh, nh, ybarh, sh) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    Nh = md("$N_h$"),
    nh = md("$n_h$"),
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$")
  ) |>
  fmt_number(columns = c(ybarh, sh), decimals = 2)
```
**Output Note:** This displays the summary statistics per stratum. Note the significantly high variance ($s_h$) in the `r names(sh)[which.max(sh)]` region compared to the others.

#### Estimates

Rather than building the "spreadsheet-style" working table by hand, we call `str_est()`, whose `$table` shows the same intermediate stratum-level values ($N_h$, $n_h$, $\pi_h$, $\bar y_h$, $s_h^2$, $\pi_h\bar y_h$, $v_h$) needed for the final stratified mean and variance, with a Total row and footnotes for $\pi_h$ and $v_h$.

```{r agstrat-str-est-mean}
res_str_mean <- str_est(ybarh, sh, nh, Nh, estimate = "mean")
res_str_mean$table
```
**Output Note:** The table provides a detailed breakdown of how each individual stratum contributes to the final mean and variance.

Extracting `$estimate` gives our overall result.

```{r agstrat-mean-estimate}
res_str_mean$estimate |> format_est_gt("mean")
```
**Output Note:** The overall result indicates an estimated mean of `r format(round(res_str_mean$estimate["Est."], 0), big.mark=",")` acres, with a 95% confidence interval of [`r format(round(res_str_mean$estimate["ci.low"], 0), big.mark=",")`, `r format(round(res_str_mean$estimate["ci.upp"], 0), big.mark=",")`].

To find the population total (total acreage across all US counties), set `estimate = "total"`.

```{r agstrat-str-est-total}
## population total estimate
res_str_total <- str_est(ybarh, sh, nh, Nh, estimate = "total")
res_str_total$estimate |> format_est_gt("total")
```
**Output Note:** This scales the estimates up to the total area of `r format(round(res_str_total$estimate["Est."], 0), big.mark=",", scientific=FALSE)` acres.

### Using the function "str_est_data"

Instead of pre-calculating the summary stats, this function extracts everything directly from the dataset using the pre-assigned sampling weights.

```{r agstrat-weight-sum}
## In the function "str_est_data", we can find the stratum size with the variable "weights":
tapply(agstrat$weight, agstrat$region, sum)
```
**Output Note:** This reconstructs the total population sizes ($N_h$) per region.

```{r agstrat-acres92-est}
## if the dataset contains a variable "weight"
res_agstrat_acres92 <- str_est_data(agstrat, y="acres92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_acres92$table
res_agstrat_acres92$estimate |> format_est_gt("mean")
```
**Output Note:** Once again, this perfectly matches the manual calculation.

```{r agstrat-smallf92-est}
## estimating the mean of the number of small farms in 1992
res_agstrat_smallf92 <- str_est_data(agstrat, y="smallf92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_smallf92$table
res_agstrat_smallf92$estimate |> format_est_gt("mean")
```
**Output Note:** This demonstrates the function's flexibility: we can easily estimate a different variable (an average of `r round(res_agstrat_smallf92$estimate["Est."], 1)` small farms per county).

### Comparing with SRS estimate

To see the benefit of stratifying, we compare our stratified estimates against a Simple Random Sample approach on the same population.

```{r agsrs-compare-load}
agsrs <- read.csv("data/agsrs.csv")
res_agsrs_srs <- srs_est(agsrs[, "acres92"], N=3078, estimate = "mean", show.details = FALSE)
res_agsrs_srs$estimate |> format_est_gt("mean")
```
**Output Note:** The Simple Random Sample standard error is noticeably higher at `r format(round(res_agsrs_srs$estimate["S.E."], 0), big.mark=",")`.

```{r agstrat-srs-var-ratio}
## the ratio of estimated variance (Stratified SE / SRS SE)^2
ratio_var <- (res_agstrat_acres92$estimate["S.E."] / res_agsrs_srs$estimate["S.E."])^2
unname(ratio_var)
```
**Output Note:** The stratified variance is approximately `r round(ratio_var * 100, 1)`% of the SRS variance.

```{r agstrat-srs-var-reduction}
## percentage of reduction of variance of str estimates from that of SRS estimate
unname(1 - ratio_var)
```
**Output Note:** Stratification by region reduced the variance by `r round((1 - ratio_var) * 100, 1)`%. This is a substantial gain in precision for no extra sampling cost.

## Allocation of stratum sample size

Deciding how to distribute the total sample $n$ among the strata significantly impacts the precision of the estimator.

### Analyzing seals.csv collected with stratified sampling

This dataset investigates ringed seal holes in different zones. The researchers used Proportional Allocation, taking 20% of the areas in each zone.

```{r seals-load}
## load data
seals <- read.csv("data/seals.csv")

# survey data summary in each stratum
nh <- as.vector(table(seals[,"zone"]))
sh <- tapply(seals[, "holes"], seals[,"zone"], sd)
ybarh <- tapply(seals[, "holes"], seals[,"zone"], mean)
Nh <- c(68, 84, 48)

res_seals <- str_est(ybarh, sh, nh, Nh, estimate = "mean")
res_seals$table
```
**Output Note:** Zone `r names(sh)[which.max(sh)]` has the highest standard deviation ($s_h \approx `r round(max(sh), 1)`$), indicating a high degree of variability in seal density in that region.

```{r seals-mean-estimate}
res_seals$estimate |> format_est_gt("mean")
```
**Output Note:** This provides the estimated mean number of seal holes per square kilometer (`r round(res_seals$estimate["Est."], 1)`).

### Neyman allocation of stratum sample size

Because Zone 3 was highly variable, we could have achieved a lower overall variance if we had sampled it more heavily. Neyman allocation formally calculates this optimal distribution.

```{r seals-neyman-calc}
## assuming the cost of surveying each zone is the same
Ch <- rep(1, 3)
Sh <- sh ## estimate population Sh with sample sh
NhShDCh <- Nh * Sh / sqrt(Ch)

## the optimal allocation proportion (Lh) for estimating the population total or mean:
Lh <- NhShDCh / sum(NhShDCh)

data.frame(Ch, Nh, Sh, NhShDCh, Lh) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(NhShDCh = md("$N_h S_h / \\sqrt{C_h}$"), Lh = md("$L_h$ (Prop)"), Ch=md("$C_h$"), Sh=md("$S_h$"), Nh=md("$N_h$")) |>
  fmt_number(columns = c(Sh, NhShDCh, Lh), decimals = 3)
```
**Output Note:** This displays the optimal allocation proportions ($L_h$). Even though Zone 2 is a larger area ($N_h=84$), Zone 3's high variance means we should allocate the largest share (`r round(Lh["3"] * 100, 0)`%) of our sampling effort there.

## Simulation to Study the Efficiency of Stratified Sampling with Different Allocation

We will use the full `agpop.csv` dataset (acting as our true population) to run 2000 repeated samples. We will compare SRS, Proportional (P2S), and Optimal (Neyman) allocations.

### Using "Region" to Stratify

#### Read Population Data

```{r agpop-region-load}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```

#### Define Stratum Variable

```{r agpop-region-stratum}
agpop$stratum <- agpop$region
## reorder agpop for the ease of using strata of "sampling" (very important)
agpop <- agpop[order(agpop$stratum), ]
```

A stratification variable is most effective when it strongly explains the variance in the target variable. We check this via ANOVA and R-squared.

```{r agpop-region-anova}
boxplot(acres92 ~ stratum, data = agpop)
lm_region <- lm(agpop$acres92 ~ agpop$stratum)
anova(lm_region)
lm_region_gt <- format_lm_gt(lm_region)
lm_region_gt$fit
r_squared_region <- summary(lm_region)$r.squared
```
**Output Note:** The R-squared is `r round(r_squared_region, 3)`. The region variable only explains about `r round(r_squared_region * 100, 1)`% of the variance in acreage, which is helpful but not extremely strong.

#### Stratified Sampling with P2S allocation (Single Run)

```{r agpop-region-p2s-single}
# doing one stratified sampling
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

## doing stratified sampling
strsample <- strata(agpop, "stratum", size = nh, method = "srswor")

# collecting data on sampled counties
agstrat <- agpop[strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean")$estimate |>
  format_est_gt("mean")
```
**Output Note:** This provides a single point estimate for demonstration.

#### Repeat stratified sampling with P2S allocation 2000 times

```{r agpop-region-p2s-sim, cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300) ## make sure the order matches Nh

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r agpop-region-neyman-alloc}
Nh <- tapply(agpop$acres92, agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

data.frame(Nh, Sh, nh_opt) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = Sh, decimals = 2)
```
**Output Note:** This table compares the population sizes to the optimal sample sizes. Notice that the West region (high variance) gets a disproportionately larger sample under Neyman allocation.

```{r agpop-region-neyman-sim, cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r agpop-region-srs-sim, cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r agpop-region-compare-plot}
sim_results_str_region <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_region)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The plot visually demonstrates that Neyman allocation has the tightest spread around the true mean (the red line).

```{r agpop-region-compare-table}
sim_means <- sapply(sim_results_str_region, mean)
sim_var <- sapply(sim_results_str_region, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table summarizes the simulation. Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, while Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`% compared to Simple Random Sampling.

### Using "acres82" to Define Strata

What if we stratify by a variable highly correlated with our target, like the acreage from a decade prior (`acres82`)?

#### Read Population Data

```{r agpop-acres82-load}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] 
N <- nrow(agpop)
```

#### Define Stratum Variable with Quantiles of "acres82"
We convert the continuous `acres82` variable into four distinct categorical strata based on quantiles.

```{r agpop-acres82-stratum}
agpop$stratum <- cut(agpop$acres82,
                     breaks = quantile(agpop$acres82, probs = c(0,0.25,0.75,0.95,1)),
                     include.lowest = T)
levels(agpop$stratum) <- paste0("acres82", c("Q1", "Q2", "Q3", "Q4"))
agpop <- agpop[order(agpop$stratum), ]
```

Let's check the explanatory power of this new stratification variable.

```{r agpop-acres82-anova}
boxplot(acres92 ~ stratum, data = agpop)
lm_acres82 <- lm(agpop$acres92 ~ agpop$stratum)
lm_acres82_gt <- format_lm_gt(lm_acres82)
lm_acres82_gt$fit
r_squared_acres82 <- summary(lm_acres82)$r.squared
```
**Output Note:** The R-squared is `r round(r_squared_acres82, 3)`. The past acreage explains `r round(r_squared_acres82 * 100, 1)`% of the variance in 1992 acreage, making this a vastly superior stratification variable.

#### Stratified sampling with P2S allocation

```{r agpop-acres82-p2s-sim, cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r agpop-acres82-neyman-sim, cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r agpop-acres82-srs-sim, cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r agpop-acres82-compare-plot}
sim_results_str_acres82 <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_acres82)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The boxplot visually highlights a massive reduction in variance compared to SRS, especially for the Neyman allocation.

```{r agpop-acres82-compare-table}
sim_means <- sapply(sim_results_str_acres82, mean)
sim_var <- sapply(sim_results_str_acres82, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table shows dramatic results: Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, and Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`%. Stratifying by a highly correlated variable is immensely powerful.

## Interactive Demonstration: Stratification by `acres82`

The app below draws three samples of the same total size $n$ at every step: a
simple random sample, a stratified sample with proportional allocation, and a
stratified sample with Neyman allocation. The strata are the `acres82` quartile
groups defined above, labelled Q1 to Q4.

The top panel shows the population split into its four strata, each drawn as a
band of jittered points on a broken `acres92` axis; the radio buttons choose
which of the three samples is highlighted in red, and each band is labelled with
its $N_h$ and the $n_h$ that the chosen allocation assigns to it.

The bottom panel accumulates the sample means of all three designs on a common
axis, one row per design, with a violin summarising each cloud once enough draws
have accumulated. The bar above each row is the current confidence interval for
that design, green when it covers $\mu$ and red when it does not, and replaced at
every draw. The running empirical standard deviation, variance relative to SRS,
and coverage rate appear at the right of each row.

```{shinylive-r}
#| standalone: true
#| viewerHeight: 860

## file: app.R
library(shiny)

data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/acres92_stratum.csv"
download.file(data_url, "acres92_stratum.csv")
dat <- read.csv("acres92_stratum.csv")
dat <- dat[order(dat$stratum), ]

pop     <- dat$acres92
stratum <- factor(dat$stratum)
lev     <- sub("^acres82", "", levels(stratum))   # acres82Q1 -> Q1
H       <- length(lev)
N       <- length(pop)
ybarU   <- mean(pop)
S       <- sd(pop)

idx_h <- split(seq_len(N), stratum)      # population indices by stratum
Nh    <- sapply(idx_h, length)
Sh    <- sapply(idx_h, function(i) sd(pop[i]))
W     <- Nh / N

## ---- broken x-axis, shared by both panels ----------------------------------
## Drawn region [0, 2 * ybarU]: visible scale [0, cut] plus a compressed lane
## for the right tail, so the population mean sits at the midpoint of the panel.
lane_frac <- 0.10
cut       <- 2 * ybarU / (1 + lane_frac)
lane_w    <- lane_frac * cut
xmax_top  <- cut + lane_w

px      <- pop
tail_id <- which(pop > cut)
n_tail  <- length(tail_id)
if (n_tail > 0) {
  r <- rank(pop[tail_id], ties.method = "first") / (n_tail + 1)
  px[tail_id] <- cut + lane_w * (0.10 + 0.80 * r)
}

set.seed(42)
band  <- as.integer(stratum)
y_pop <- (H + 1 - band) + runif(N, -0.30, 0.30)   # Q1 in the top band

## ---- fixed window for the sampling distributions ---------------------------
se100  <- sqrt(1 - 100 / N) * S / sqrt(100)
xlim_b <- ybarU + c(-1, 1) * 5 * se100

METHOD <- c("SRS", "Proportional", "Neyman")
MCOL   <- c("grey35", "steelblue", "darkorange")
MAR    <- c(4.5, 7, 3, 1.5)   # identical in both panels, so the scales align

## allocation with at least 2 units per stratum, summing to n
alloc <- function(w, n) {
  nh <- pmin(pmax(2, round(w / sum(w) * n)), Nh)
  while (sum(nh) > n && any(nh > 2)) {
    i <- which.max(nh - 2); nh[i] <- nh[i] - 1
  }
  while (sum(nh) < n && any(nh < Nh)) {
    i <- which.max(Nh - nh); nh[i] <- nh[i] + 1
  }
  nh
}

## symmetric density polygon, drawn horizontally about y = at
violin <- function(x, at, col, w = 0.32) {
  if (length(x) < 10 || diff(range(x)) <= 0) return(invisible())
  d  <- density(x)
  hw <- w * d$y / max(d$y)
  polygon(c(d$x, rev(d$x)), c(at + hw, rev(at - hw)),
          col = adjustcolor(col, 0.35), border = col)
  m <- median(x)
  segments(m, at - w * 0.6, m, at + w * 0.6, col = col, lwd = 2)
}

ui <- fluidPage(
  titlePanel("Stratified sampling by acres82 quartiles, against simple random sampling"),
  sidebarLayout(
    sidebarPanel(
      width = 3,
      sliderInput("n", "Total sample size (n)", min = 40, max = 1000, value = 300, step = 10),
      sliderInput("conf", "Confidence level", min = 0.80, max = 0.99, value = 0.95, step = 0.01),
      radioButtons("show", "Highlight sample from", choiceNames = METHOD,
                   choiceValues = as.character(1:3), selected = "3"),
      actionButton("draw", "Draw new samples", class = "btn-primary", width = "100%"),
      br(), br(),
      fluidRow(
        column(6, actionButton("play", "Play", class = "btn-success", width = "100%")),
        column(6, actionButton("stop", "Stop", class = "btn-danger",  width = "100%"))
      ),
      br(),
      sliderInput("speed", "Samples per second", min = 1, max = 10, value = 3, step = 1),
      numericInput("nrep", "Stop after this many samples",
                   value = 500, min = 10, max = 20000, step = 50),
      actionButton("reset", "Reset", width = "100%")
    ),
    mainPanel(
      width = 9,
      plotOutput("popPlot",  height = "380px"),
      plotOutput("distPlot", height = "380px")
    )
  )
)

server <- function(input, output, session) {

  rv <- reactiveValues(
    idx = NULL,      # list of sampled indices, one element per design
    est = NULL,      # matrix 3 x 3: mean, lower, upper
    M   = NULL,      # accumulated means, one column per design
    HIT = NULL,      # accumulated coverage indicators
    jit = numeric(0)
  )
  running <- reactiveVal(FALSE)

  nrep <- reactive({
    v <- input$nrep
    if (is.null(v) || is.na(v) || v < 1) 1L else min(as.integer(v), 20000L)
  })

  nh_prop   <- reactive(alloc(Nh,      input$n))
  nh_neyman <- reactive(alloc(Nh * Sh, input$n))

  draw_one <- function() {
    n    <- isolate(input$n)
    conf <- isolate(input$conf)
    a    <- 1 - (1 - conf) / 2

    # simple random sample
    i_srs <- sample(N, n)
    y     <- pop[i_srs]
    m_srs <- mean(y)
    se    <- sqrt(1 - n / N) * sd(y) / sqrt(n)
    d_srs <- qt(a, df = n - 1) * se

    # stratified: ybar_str = sum W_h ybar_h, V = sum W_h^2 (1 - n_h/N_h) s_h^2 / n_h
    str_draw <- function(nh) {
      ii <- unlist(lapply(seq_len(H), function(h) sample(idx_h[[h]], nh[h])))
      yh <- split(pop[ii], rep(seq_len(H), nh))
      mh <- sapply(yh, mean)
      sh <- sapply(yh, sd)
      m  <- sum(W * mh)
      v  <- sum(W^2 * (1 - nh / Nh) * sh^2 / nh)
      c(m, qt(a, df = sum(nh) - H) * sqrt(v))
    }
    p <- str_draw(isolate(nh_prop()))
    q <- str_draw(isolate(nh_neyman()))

    take <- function(nh) unlist(lapply(seq_len(H), function(h) sample(idx_h[[h]], nh[h])))

    est <- rbind(c(m_srs, m_srs - d_srs, m_srs + d_srs),
                 c(p[1],  p[1]  - p[2],  p[1]  + p[2]),
                 c(q[1],  q[1]  - q[2],  q[1]  + q[2]))

    rv$idx <- list(i_srs, take(isolate(nh_prop())), take(isolate(nh_neyman())))
    rv$est <- est
    rv$M   <- rbind(rv$M,   est[, 1])
    rv$HIT <- rbind(rv$HIT, est[, 2] <= ybarU & ybarU <= est[, 3])
    rv$jit <- c(rv$jit, runif(1, -0.1, 0.1))
  }

  observeEvent(input$draw, { running(FALSE); draw_one() })
  observeEvent(input$play, running(TRUE))
  observeEvent(input$stop, running(FALSE))

  observe({
    if (!isTRUE(running())) return()
    if (NROW(isolate(rv$M)) >= isolate(nrep())) { running(FALSE); return() }
    invalidateLater(1000 / isolate(input$speed), session)
    isolate(draw_one())
  })

  reset_history <- function() {
    running(FALSE)
    rv$idx <- NULL; rv$est <- NULL; rv$M <- NULL; rv$HIT <- NULL
    rv$jit <- numeric(0)
  }
  observeEvent(input$reset, reset_history())
  observeEvent(input$n,     reset_history())
  observeEvent(input$conf,  reset_history())

  fmt <- function(x) format(round(x), big.mark = ",")

  output$popPlot <- renderPlot({
    j  <- as.integer(input$show)
    nh <- switch(j, rep(NA_integer_, H), nh_prop(), nh_neyman())

    par(mar = MAR, xaxs = "i")
    plot(NA, xlim = c(0, xmax_top), ylim = c(0.3, H + 0.9),
         xlab = "acres92", ylab = "", xaxt = "n", yaxt = "n", bty = "n",
         main = sprintf("Population by acres82 stratum, with the %s sample", METHOD[j]))

    for (h in seq_len(H)) {
      yc <- H + 1 - h
      rect(0, yc - 0.36, xmax_top, yc + 0.36,
           col = adjustcolor("grey92", 0.6), border = NA)
    }
    points(px, y_pop, pch = 16, cex = 0.5, col = adjustcolor("grey55", 0.5))
    if (!is.null(rv$idx))
      points(px[rv$idx[[j]]], y_pop[rv$idx[[j]]], pch = 16, cex = 0.6,
             col = adjustcolor("red", 0.85))

    axis(2, at = H:1, labels = lev, las = 1, tick = FALSE, line = -0.5)
    for (h in seq_len(H)) {
      lab <- if (j == 1) sprintf("N = %d", Nh[h])
             else        sprintf("N = %d,  n = %d", Nh[h], nh[h])
      mtext(lab, side = 2, at = H + 1 - h, line = 1.8, las = 1, cex = 0.72, col = "grey30")
    }

    at <- pretty(c(0, cut), 6); at <- at[at <= cut]
    axis(1, at = at, labels = format(at, big.mark = ",", trim = TRUE))
    segments(cut, 0.3, cut, H + 0.6, col = "grey60", lty = 3)
    text(cut, 0.14, "//", col = "grey40", cex = 1.3, xpd = NA)
    text(cut + lane_w / 2, 0.14, sprintf("> %s", fmt(cut)),
         col = "grey40", cex = 0.75, xpd = NA)
    abline(v = ybarU, col = "black", lwd = 2, lty = 2)
  })

  output$distPlot <- renderPlot({
    par(mar = MAR, xaxs = "i")
    plot(NA, xlim = xlim_b, ylim = c(0.4, 3.9), yaxt = "n", bty = "n",
         xlab = "Sample mean of acres92", ylab = "",
         main = "Sampling distributions of the three estimators")
    abline(v = ybarU, col = "black", lwd = 2, lty = 2)
    axis(2, at = 3:1, labels = METHOD, las = 1, tick = FALSE, line = -0.5)

    if (is.null(rv$M)) {
      text(ybarU, 2, "Click 'Draw new samples' or 'Play' to begin", col = "grey50")
      return(invisible())
    }

    k     <- NROW(rv$M)
    v_srs <- if (k > 1) var(rv$M[, 1]) else NA_real_

    for (j in 1:3) {
      yc <- 4 - j
      points(rv$M[, j], yc + rv$jit, pch = 16, cex = 0.45,
             col = adjustcolor(MCOL[j], 0.30))
      violin(rv$M[, j], at = yc, col = MCOL[j])

      # transient interval for the current draw
      ci_col <- if (rv$HIT[k, j]) "forestgreen" else "red"
      yb     <- yc + 0.42
      segments(rv$est[j, 2], yb, rv$est[j, 3], yb, col = ci_col, lwd = 3)
      segments(rv$est[j, 2:3], yb - 0.06, rv$est[j, 2:3], yb + 0.06, col = ci_col, lwd = 3)
      points(rv$est[j, 1], yb, pch = 19, cex = 1.1, col = ci_col)

      lab <- if (k > 1)
        sprintf("sd = %s   rel.var = %.3f   coverage = %.1f%%",
                fmt(sd(rv$M[, j])), var(rv$M[, j]) / v_srs, 100 * mean(rv$HIT[, j]))
      else
        sprintf("coverage = %.0f%%", 100 * mean(rv$HIT[, j]))
      text(xlim_b[2], yc - 0.34, lab, adj = c(1, 0), cex = 0.8, col = "grey25")
    }

    outside <- sum(rv$M < xlim_b[1] | rv$M > xlim_b[2])
    sub <- sprintf("n = %d per design   |   samples drawn = %d of %d   |   nominal %.0f%%",
                   input$n, k, nrep(), 100 * input$conf)
    if (outside > 0) sub <- paste0(sub, sprintf("   |   %d mean(s) beyond the axis", outside))
    mtext(sub, side = 3, line = -0.2, cex = 0.8, col = "grey30")
  })
}

shinyApp(ui, server)

```
