# Stratified Random Sampling

## Key Formulas for Stratified Sampling

Stratified random sampling divides a population of size $N$ into $H$ mutually exclusive strata. A simple random sample of size $n_h$ is then drawn from each stratum of size $N_h$. This design often yields higher precision than Simple Random Sampling (SRS).

#### 1. Stratified Population Mean Estimator {-}
The estimated population mean is a weighted average of the sample means from each stratum:
$$
\overline{y}_{str} = \sum_{h=1}^H W_h \overline{y}_h.
$$
Where $W_h = \frac{N_h}{N}$ is the stratum weight, and $\overline{y}_h$ is the sample mean of stratum $h$.

#### 2. Variance of the Stratified Mean {-}
The estimated variance accounts for the finite population correction in each stratum:
$$
\hat{V}(\overline{y}_{str}) = \sum_{h=1}^H W_h^2 \left(1 - \frac{n_h}{N_h}\right) \frac{s_h^2}{n_h}.
$$
Where $s_h^2$ is the sample variance in stratum $h$.

#### 3. Proportional Allocation {-}
Sample sizes are allocated proportional to the size of the stratum:
$$
n_h = n \left( \frac{N_h}{N} \right).
$$

#### 4. Neyman (Optimal) Allocation {-}
If surveying costs are equal across strata, Neyman allocation minimizes the variance of the estimator by sampling more heavily from larger and more highly variable strata:
$$
n_h = n \frac{N_h S_h}{\sum_{h=1}^H N_h S_h}
$$

---

## Functions and packages for Analyzing Data

The following code loads necessary libraries and defines custom functions to automate the calculation of the stratified mean, standard error, and confidence intervals. `str_mean_estimate_data` computes these directly from a dataset, while `str_mean_estimate` computes them from pre-calculated summary statistics. We also define helper functions to format our output using the `gt` package.

```{r}
#| code-fold: true
library(latex2exp)
library(sampling)
library(dplyr)
library(gt)

## this function finds statistical estimates given a dataset with sampling weight
# stratdata --- data.frame containing stratified sample
# y --- name of variable for which we want to estiamte population mean
# stratum --- name of variable that will be used as stratum variable
# weight --- name of variable indicating sampling weight
# post = TRUE is to indicate that we will do post-stratification analysis (ignoring it for the moment)
# note: from weights we can find Nh (see the code for formula)
str_mean_estimate_data <- function (stratdata, y, stratum, weight)
{
    ## compute stratum-wise data
    n <- nrow (stratdata)
    sh <- tapply (stratdata[, y], stratdata[,stratum], sd)
    ybarh <- tapply (stratdata[, y], stratdata[,stratum], mean)
    ## find population stratum size using sampling weight included in the data set
    Nh <- tapply (stratdata[, weight], stratdata[,stratum], sum)
    nh <- tapply (1:nrow(stratdata), stratdata[,stratum], length)
    ## find mean estimates
    N <- sum (Nh)
    Pi_h <- Nh/N
    ybar <- sum(ybarh * Pi_h)
    seybar <- sqrt(sum((1-nh/Nh)*Pi_h^2*sh^2/nh))
    mem <- 1.96 * seybar
    c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem)
    
}

## to find total, multiply N to the estimate returned by this function
## for poststratification, use nh = n * Nh/N
str_mean_estimate <- function (ybarh, sh, nh, Nh)
{
    N <- sum (Nh)
    Pi_h <- Nh/N
    ybar <- sum(ybarh * Pi_h)
    seybar <- sqrt(sum((1-nh/Nh)*Pi_h^2*sh^2/nh))
    mem <- 1.96 * seybar
    c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

## sdata --- a vector of original survey data
## N --- population size
## to find total, multiply N to the estimate returned by this function
srs_mean_est <- function (sdata, N = Inf)
{
    n <- length (sdata)
    ybar <- mean (sdata)
    se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
    mem <- qt (0.975, df = n - 1) * se.ybar
    c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}


# Helper function to format the output of stratified estimates using gt
format_str_gt <- function(est_vec, is_total = FALSE, is_prop = FALSE) {
  df <- as.data.frame(t(est_vec))
  
  if (is_total) {
    est_sym <- "$\\hat{t}_{str}$"
    se_sym <- "$\\mathrm{SE}(\\hat{t}_{str})$"
  } else if (is_prop) {
    est_sym <- "$\\hat{p}_{str}$"
    se_sym <- "$\\mathrm{SE}(\\hat{p}_{str})$"
  } else {
    est_sym <- "$\\bar{y}_{str}$"
    se_sym <- "$\\mathrm{SE}(\\bar{y}_{str})$"
  }
  
  df |>
    gt() |>
    cols_label(
      Est. = md(est_sym),
      S.E. = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(
      columns = everything(),
      decimals = 4
    ) |>
    tab_options(table.width = pct(60))
}

# Helper function to format the output of SRS estimates using gt
format_srs_gt <- function(est_vec, is_total = FALSE, is_prop = FALSE) {
  df <- as.data.frame(t(est_vec))
  
  if (is_total) {
    est_sym <- "$\\hat{t}$"
    se_sym <- "$\\mathrm{SE}(\\hat{t})$"
  } else if (is_prop) {
    est_sym <- "$\\hat{p}$"
    se_sym <- "$\\mathrm{SE}(\\hat{p})$"
  } else {
    est_sym <- "$\\bar{y}$"
    se_sym <- "$\\mathrm{SE}(\\bar{y})$"
  }
  
  df |>
    gt() |>
    cols_label(
      Est. = md(est_sym),
      S.E. = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(columns = everything(), decimals = 4) |>
    tab_options(table.width = pct(60))
}
```

## Analysis of "agstrat.csv" data

We will estimate the average 1992 farm acreage (`acres92`) across US counties, using geographic region (`region`) as the stratification variable.

### Importing agstrat.csv data

```{r}
agstrat <- read.csv("data/agstrat.csv")
head(agstrat) |> gt()
```
**Output Note:** The table above shows the sample survey data, including the stratum (region) and weights for each county.

### Spreadsheet calculation

#### Summarizing acre92 in each stratum

We first manually compute the sample size ($n_h$), standard deviation ($s_h$), and mean ($\overline{y}_h$) for each region, matching them with the known total counties in each region ($N_h$).

```{r}
nh <- tapply (agstrat[, "acres92"], agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information (true population size per stratum)
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

data.frame(Nh, nh, ybarh, sh) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    Nh = md("$N_h$"),
    nh = md("$n_h$"),
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$")
  ) |>
  fmt_number(columns = c(ybarh, sh), decimals = 2)
```
**Output Note:** This displays the summary statistics per stratum. Note the significantly high variance ($s_h$) in the `r names(sh)[which.max(sh)]` region compared to the others.

#### Estimates

Next, we build a "spreadsheet-style" table to calculate the intermediate values needed for the final stratified variance and mean, summing the results at the bottom. We use `gt` to properly format the mathematical column headers.

```{r}
N <- sum(Nh)
pi_h <- Nh/N
weighted_ybar_h <- ybarh * pi_h
var_ybarh <- (1-nh/Nh)*sh^2/nh
var_h <- var_ybarh * pi_h^2

workingtable <- data.frame(
  Nh = Nh, 
  nh = nh,  
  ybarh = ybarh, 
  sh = sh, 
  s2h = sh^2, 
  pi_h = pi_h,
  weighted_ybar_h = weighted_ybar_h,
  var_ybarh = var_ybarh,
  var_h = var_h
)

workingtable <- rbind(workingtable, colSums(workingtable))
row.names(workingtable)[5] <- "Sum"
workingtable$Region <- rownames(workingtable)

workingtable |>
  gt(rowname_col = "Region") |>
  cols_label(
    Nh = md("$N_h$"),
    nh = md("$n_h$"),
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$"),
    s2h = md("$s^2_h$"),
    pi_h = md("$\\pi_h = N_h/N$"),
    weighted_ybar_h = md("$\\bar{y}_h \\cdot \\pi_h$"),
    var_ybarh = md("$\\hat{V}(\\bar{y}_h) = (1 - n_h/N_h)s^2_h/n_h$"),
    var_h = md("$\\hat{V}(\\bar{y}_h) \\cdot \\pi^2_h$")
  ) |>
  fmt_number(
    columns = c(ybarh, sh, s2h, pi_h, weighted_ybar_h, var_ybarh, var_h),
    decimals = 4
  )
```
**Output Note:** The table provides a detailed breakdown of how each individual stratum contributes to the final mean and variance.

Extracting the final sum values gives our overall estimates.

```{r}
ybar <- sum(weighted_ybar_h)
var_ybar <- sum((1-nh/Nh)*pi_h^2*sh^2/nh)
seybar <- sqrt(var_ybar)
mem <- 1.96 * seybar

c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem) |> 
  format_str_gt()
```
**Output Note:** The overall result indicates an estimated mean of `r format(round(ybar, 0), big.mark=",")` acres, with a 95% confidence interval of [`r format(round(ybar - mem, 0), big.mark=",")`, `r format(round(ybar + mem, 0), big.mark=",")`].

### Using the function "str_mean_estimate"

We can bypass the manual table generation using our predefined function.

```{r}
nh <- tapply (agstrat[, "acres92"], agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

## strata mean estimate
str_mean_estimate(ybarh, sh, nh, Nh) |> 
  format_str_gt()
```
**Output Note:** This matches our manual spreadsheet calculation precisely.

To find the population total (total acreage across all US counties), multiply the mean and CI by $N$.

```{r}
## population total estimate 
(str_mean_estimate(ybarh, sh, nh, Nh) * sum(Nh)) |> 
  format_str_gt(is_total = TRUE)
```
**Output Note:** This successfully scales the estimates up to the total area of `r format(round(str_mean_estimate(ybarh, sh, nh, Nh)["Est."] * sum(Nh), 0), big.mark=",", scientific=FALSE)` acres.

### Using the function "str_mean_estimate_data"

Instead of pre-calculating the summary stats, this function extracts everything directly from the dataset using the pre-assigned sampling weights.

```{r}
## In the function "str_mean_estimate_data", we can find the stratum size with the variable "weights":
tapply(agstrat$weight, agstrat$region, sum)
```
**Output Note:** This reconstructs the total population sizes ($N_h$) per region.

```{r}
## if the dataset contains a variable "weight"
str_mean_estimate_data(agstrat, y="acres92", stratum="region", weight="weight") |> 
  format_str_gt()
```
**Output Note:** Once again, this perfectly matches the manual calculation.

```{r}
## estimating the mean of the number of small farms in 1992
str_mean_estimate_data(agstrat, y="smallf92", stratum="region", weight="weight") |> 
  format_str_gt()
```
**Output Note:** This demonstrates the function's flexibility: we can easily estimate a different variable (an average of `r round(str_mean_estimate_data(agstrat, "smallf92", "region", "weight")["Est."], 1)` small farms per county).

### Comparing with SRS estimate

To see the benefit of stratifying, we compare our stratified estimates against a Simple Random Sample approach on the same population.

```{r}
agsrs <- read.csv("data/agsrs.csv")
srs_mean_est(agsrs[, "acres92"], N=3078) |> 
  format_srs_gt()
```
**Output Note:** The Simple Random Sample standard error is noticeably higher at `r format(round(srs_mean_est(agsrs[, "acres92"], N=3078)["S.E."], 0), big.mark=",")`.

```{r}
## the ratio of estimated variance (Stratified SE / SRS SE)^2
ratio_var <- (str_mean_estimate_data(agstrat, "acres92", "region", "weight")["S.E."] / srs_mean_est(agsrs[, "acres92"], N=3078)["S.E."])^2
unname(ratio_var)
```
**Output Note:** The stratified variance is approximately `r round(ratio_var * 100, 1)`% of the SRS variance.

```{r}
## percentage of reduction of variance of str estimates from that of SRS estimate
unname(1 - ratio_var)
```
**Output Note:** Stratification by region reduced the variance by `r round((1 - ratio_var) * 100, 1)`%. This is a substantial gain in precision for no extra sampling cost.

## Allocation of stratum sample size

Deciding how to distribute the total sample $n$ among the strata significantly impacts the precision of the estimator.

### Analyzing seals.csv collected with stratified sampling

This dataset investigates ringed seal holes in different zones. The researchers used Proportional Allocation, taking 20% of the areas in each zone.

```{r}
## load data
seals <- read.csv("data/seals.csv")

# survey data summary in each stratum
nh <- as.vector(table(seals[,"zone"]))
sh <- tapply(seals[, "holes"], seals[,"zone"], sd)
ybarh <- tapply(seals[, "holes"], seals[,"zone"], mean)
Nh <- c(68, 84, 48)

data.frame(Nh, nh, ybarh, sh) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(ybarh, sh), decimals = 2)
```
**Output Note:** Zone `r names(sh)[which.max(sh)]` has the highest standard deviation ($s_h \approx `r round(max(sh), 1)`$), indicating a high degree of variability in seal density in that region.

```{r}
str_mean_estimate(ybarh, sh, nh, Nh) |> 
  format_str_gt()
```
**Output Note:** This provides the estimated mean number of seal holes per square kilometer (`r round(str_mean_estimate(ybarh, sh, nh, Nh)["Est."], 1)`).

### Neyman allocation of stratum sample size

Because Zone 3 was highly variable, we could have achieved a lower overall variance if we had sampled it more heavily. Neyman allocation formally calculates this optimal distribution.

```{r}
## assuming the cost of surveying each zone is the same
Ch <- rep(1, 3)
Sh <- sh ## estimate population Sh with sample sh
NhShDCh <- Nh * Sh / sqrt(Ch)

## the optimal allocation proportion (Lh) for estimating the population total or mean:
Lh <- NhShDCh / sum(NhShDCh)

data.frame(Ch, Nh, Sh, NhShDCh, Lh) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(NhShDCh = md("$N_h S_h / \\sqrt{C_h}$"), Lh = md("$L_h$ (Prop)"), Ch=md("$C_h$"), Sh=md("$S_h$"), Nh=md("$N_h$")) |>
  fmt_number(columns = c(Sh, NhShDCh, Lh), decimals = 3)
```
**Output Note:** This displays the optimal allocation proportions ($L_h$). Even though Zone 2 is a larger area ($N_h=84$), Zone 3's high variance means we should allocate the largest share (`r round(Lh["3"] * 100, 0)`%) of our sampling effort there.

## Simulation to Study the Efficiency of Stratified Sampling with Different Allocation

We will use the full `agpop.csv` dataset (acting as our true population) to run 2000 repeated samples. We will compare SRS, Proportional (P2S), and Optimal (Neyman) allocations.

### Using "Region" to Stratify

#### Read Population Data

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```

#### Define Stratum Variable

```{r}
agpop$stratum <- agpop$region
## reorder agpop for the ease of using strata of "sampling" (very important)
agpop <- agpop[order(agpop$stratum), ]
```

A stratification variable is most effective when it strongly explains the variance in the target variable. We check this via ANOVA and R-squared.

```{r}
boxplot(acres92 ~ stratum, data = agpop)
anova(lm(agpop$acres92 ~ agpop$stratum))
summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared
```
**Output Note:** The R-squared is `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared, 3)`. The region variable only explains about `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared * 100, 1)`% of the variance in acreage, which is helpful but not extremely strong.

#### Stratified Sampling with P2S allocation (Single Run)

```{r}
# doing one stratified sampling
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

## doing stratified sampling
strsample <- strata(agpop, "stratum", size = nh, method = "srswor")

# collecting data on sampled counties
agstrat <- agpop[strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_mean_estimate_data(agstrat, "acres92", "stratum", "weight") |> 
  format_str_gt()
```
**Output Note:** This provides a single point estimate for demonstration.

#### Repeat stratified sampling with P2S allocation 2000 times

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300) ## make sure the order matches Nh

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight")
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r}
Nh <- tapply(agpop$acres92, agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

data.frame(Nh, Sh, nh_opt) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = Sh, decimals = 2)
```
**Output Note:** This table compares the population sizes to the optimal sample sizes. Notice that the West region (high variance) gets a disproportionately larger sample under Neyman allocation.

```{r cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight")
}
```

#### Repeat simple random sampling 2000 times

```{r cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_mean_est(agsrs[, "acres92"], N = sum(Nh))
}
```

#### Compare the efficiency of different methods

```{r}
sim_results_str_region <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_region)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The plot visually demonstrates that Neyman allocation has the tightest spread around the true mean (the red line).

```{r}
sim_means <- sapply(sim_results_str_region, mean)
sim_var <- sapply(sim_results_str_region, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table summarizes the simulation. Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, while Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`% compared to Simple Random Sampling.

### Using "acres82" to Define Strata

What if we stratify by a variable highly correlated with our target, like the acreage from a decade prior (`acres82`)?

#### Read Population Data

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] 
N <- nrow(agpop)
```

#### Define Stratum Variable with Quantiles of "acres82"
We convert the continuous `acres82` variable into four distinct categorical strata based on quantiles.

```{r}
agpop$stratum <- cut(agpop$acres82,
                     breaks = quantile(agpop$acres82, probs = c(0,0.25,0.75,0.95,1)),
                     include.lowest = T)
agpop$stratum <- mapvalues(agpop$stratum, 
                           from = (sort(unique(agpop$stratum))), 
                           to = paste0("acres82", c("Q1", "Q2", "Q3", "Q4")))
agpop <- agpop[order(agpop$stratum), ]
```

Let's check the explanatory power of this new stratification variable.

```{r}
boxplot(acres92 ~ stratum, data = agpop)
summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared
```
**Output Note:** The R-squared is `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared, 3)`. The past acreage explains `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared * 100, 1)`% of the variance in 1992 acreage, making this a vastly superior stratification variable.

#### Stratified sampling with P2S allocation

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight")
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight")
}
```

#### Repeat simple random sampling 2000 times

```{r cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_mean_est(agsrs[, "acres92"], N = sum(Nh))
}
```

#### Compare the efficiency of different methods

```{r}
sim_results_str_acres82 <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_acres82)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The boxplot visually highlights a massive reduction in variance compared to SRS, especially for the Neyman allocation.

```{r}
sim_means <- sapply(sim_results_str_acres82, mean)
sim_var <- sapply(sim_results_str_acres82, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table shows dramatic results: Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, and Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`%. Stratifying by a highly correlated variable is immensely powerful.