# Stratified Random Sampling

## Key Formulas for Stratified Sampling

Stratified random sampling divides a population of size $N$ into $H$ mutually exclusive strata. A simple random sample of size $n_h$ is then drawn from each stratum of size $N_h$. This design often yields higher precision than Simple Random Sampling (SRS).

#### 1. Stratified Population Mean Estimator {-}
The estimated population mean is a weighted average of the sample means from each stratum:
$$
\overline{y}_{str} = \sum_{h=1}^H \pi_h \overline{y}_h.
$$
Where $\pi_h = \frac{N_h}{N}$ is the stratum weight, and $\overline{y}_h$ is the sample mean of stratum $h$.

#### 2. Variance of the Stratified Mean {-}
The estimated variance accounts for the finite population correction in each stratum:
$$
\hat{V}(\overline{y}_{str}) = \sum_{h=1}^H \pi_h^2 \left(1 - \frac{n_h}{N_h}\right) \frac{s_h^2}{n_h}.
$$
Where $s_h^2$ is the sample variance in stratum $h$.

#### 3. Proportional Allocation {-}
Sample sizes are allocated proportional to the size of the stratum:
$$
n_h = n \left( \frac{N_h}{N} \right).
$$

#### 4. Neyman (Optimal) Allocation {-}
If surveying costs are equal across strata, Neyman allocation minimizes the variance of the estimator by sampling more heavily from larger and more highly variable strata:
$$
n_h = n \frac{N_h S_h}{\sum_{h=1}^H N_h S_h}
$$

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `str_est`, `str_est_data`, and `srs_est`, used below to compute the stratified mean, standard error, and confidence intervals) live in a single shared file, `samplingestimate.r`, which we source below. `str_est_data` computes these directly from a dataset, while `str_est` computes them from pre-calculated summary statistics.

```{r source-functions}
#| echo: false
library(latex2exp)
library(sampling)
library(dplyr)
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Analysis of "agstrat.csv" data

We will estimate the average 1992 farm acreage (`acres92`) across US counties, using geographic region (`region`) as the stratification variable.

### Importing agstrat.csv data

```{r tbl-agstrat-sample-data}
#| tbl-cap: "Sample survey data for agstrat.csv, including stratum and weight"
agstrat <- read.csv("data/agstrat.csv")
head(agstrat) |> gt()
```
**Output Note:** The table above shows the sample survey data, including the stratum (region) and weights for each county.

### Spreadsheet calculation

#### Summarizing acre92 in each stratum

We first manually compute the sample size ($n_h$), standard deviation ($s_h$), and mean ($\overline{y}_h$) for each region, matching them with the known total counties in each region ($N_h$).

```{r tbl-agstrat-summary-by-region}
#| tbl-cap: "Sample summary statistics for acres92 by region stratum"
nh <- tapply (agstrat[, "acres92"], agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information (true population size per stratum)
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

data.frame(Nh, nh, ybarh, sh) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    Nh = md("$N_h$"),
    nh = md("$n_h$"),
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$")
  ) |>
  fmt_number(columns = c(ybarh, sh), decimals = 2)
```
**Output Note:** This displays the summary statistics per stratum. Note the significantly high variance ($s_h$) in the `r names(sh)[which.max(sh)]` region compared to the others.

#### Estimates

Rather than building the "spreadsheet-style" working table by hand, we call `str_est()`, whose `$table` shows the same intermediate stratum-level values ($N_h$, $n_h$, $\pi_h$, $\bar y_h$, $s_h^2$, $\pi_h\bar y_h$, $v_h$) needed for the final stratified mean and variance, with a Total row and footnotes for $\pi_h$ and $v_h$.

```{r agstrat-str-est-mean}
res_str_mean <- str_est(ybarh, sh, nh, Nh, estimate = "mean")
res_str_mean$table
```
**Output Note:** The table provides a detailed breakdown of how each individual stratum contributes to the final mean and variance.

Extracting `$estimate` gives our overall result.

```{r agstrat-mean-estimate}
res_str_mean$estimate |> format_est_gt("mean")
```
**Output Note:** The overall result indicates an estimated mean of `r format(round(res_str_mean$estimate["Est."], 0), big.mark=",")` acres, with a 95% confidence interval of [`r format(round(res_str_mean$estimate["ci.low"], 0), big.mark=",")`, `r format(round(res_str_mean$estimate["ci.upp"], 0), big.mark=",")`].

To find the population total (total acreage across all US counties), set `estimate = "total"`.

```{r agstrat-str-est-total}
## population total estimate
res_str_total <- str_est(ybarh, sh, nh, Nh, estimate = "total")
res_str_total$estimate |> format_est_gt("total")
```
**Output Note:** This scales the estimates up to the total area of `r format(round(res_str_total$estimate["Est."], 0), big.mark=",", scientific=FALSE)` acres.

### Using the function "str_est_data"

Instead of pre-calculating the summary stats, this function extracts everything directly from the dataset using the pre-assigned sampling weights.

```{r agstrat-weight-sum}
## In the function "str_est_data", we can find the stratum size with the variable "weights":
tapply(agstrat$weight, agstrat$region, sum)
```
**Output Note:** This reconstructs the total population sizes ($N_h$) per region.

```{r agstrat-acres92-est}
## if the dataset contains a variable "weight"
res_agstrat_acres92 <- str_est_data(agstrat, y="acres92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_acres92$table
res_agstrat_acres92$estimate |> format_est_gt("mean")
```
**Output Note:** Once again, this perfectly matches the manual calculation.

```{r agstrat-smallf92-est}
## estimating the mean of the number of small farms in 1992
res_agstrat_smallf92 <- str_est_data(agstrat, y="smallf92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_smallf92$table
res_agstrat_smallf92$estimate |> format_est_gt("mean")
```
**Output Note:** This demonstrates the function's flexibility: we can easily estimate a different variable (an average of `r round(res_agstrat_smallf92$estimate["Est."], 1)` small farms per county).

### Comparing with SRS estimate

To see the benefit of stratifying, we compare our stratified estimates against a Simple Random Sample approach on the same population.

```{r agsrs-compare-load}
agsrs <- read.csv("data/agsrs.csv")
res_agsrs_srs <- srs_est(agsrs[, "acres92"], N=3078, estimate = "mean", show.details = FALSE)
res_agsrs_srs$estimate |> format_est_gt("mean")
```
**Output Note:** The Simple Random Sample standard error is noticeably higher at `r format(round(res_agsrs_srs$estimate["S.E."], 0), big.mark=",")`.

```{r agstrat-srs-var-ratio}
## the ratio of estimated variance (Stratified SE / SRS SE)^2
ratio_var <- (res_agstrat_acres92$estimate["S.E."] / res_agsrs_srs$estimate["S.E."])^2
unname(ratio_var)
```
**Output Note:** The stratified variance is approximately `r round(ratio_var * 100, 1)`% of the SRS variance.

```{r agstrat-srs-var-reduction}
## percentage of reduction of variance of str estimates from that of SRS estimate
unname(1 - ratio_var)
```
**Output Note:** Stratification by region reduced the variance by `r round((1 - ratio_var) * 100, 1)`%. This is a substantial gain in precision for no extra sampling cost.

## Allocation of stratum sample size

Deciding how to distribute the total sample $n$ among the strata significantly impacts the precision of the estimator.

### Analyzing seals.csv collected with stratified sampling

This dataset investigates ringed seal holes in different zones. The researchers used Proportional Allocation, taking 20% of the areas in each zone.

```{r seals-load}
## load data
seals <- read.csv("data/seals.csv")

# survey data summary in each stratum
nh <- as.vector(table(seals[,"zone"]))
sh <- tapply(seals[, "holes"], seals[,"zone"], sd)
ybarh <- tapply(seals[, "holes"], seals[,"zone"], mean)
Nh <- c(68, 84, 48)

res_seals <- str_est(ybarh, sh, nh, Nh, estimate = "mean")
res_seals$table
```
**Output Note:** Zone `r names(sh)[which.max(sh)]` has the highest standard deviation ($s_h \approx `r round(max(sh), 1)`$), indicating a high degree of variability in seal density in that region.

```{r seals-mean-estimate}
res_seals$estimate |> format_est_gt("mean")
```
**Output Note:** This provides the estimated mean number of seal holes per square kilometer (`r round(res_seals$estimate["Est."], 1)`).

### Neyman allocation of stratum sample size

Because Zone 3 was highly variable, we could have achieved a lower overall variance if we had sampled it more heavily. Neyman allocation formally calculates this optimal distribution.

```{r tbl-seals-neyman-allocation}
#| tbl-cap: "Neyman-optimal allocation proportions for the seals survey zones"
## assuming the cost of surveying each zone is the same
Ch <- rep(1, 3)
Sh <- sh ## estimate population Sh with sample sh
NhShDCh <- Nh * Sh / sqrt(Ch)

## the optimal allocation proportion (Lh) for estimating the population total or mean:
Lh <- NhShDCh / sum(NhShDCh)

data.frame(Ch, Nh, Sh, NhShDCh, Lh) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(NhShDCh = md("$N_h S_h / \\sqrt{C_h}$"), Lh = md("$L_h$ (Prop)"), Ch=md("$C_h$"), Sh=md("$S_h$"), Nh=md("$N_h$")) |>
  fmt_number(columns = c(Sh, NhShDCh, Lh), decimals = 3)
```
**Output Note:** This displays the optimal allocation proportions ($L_h$). Even though Zone 2 is a larger area ($N_h=84$), Zone 3's high variance means we should allocate the largest share (`r round(Lh["3"] * 100, 0)`%) of our sampling effort there.

## Simulation to Study the Efficiency of Stratified Sampling with Different Allocation

We will use the full `agpop.csv` dataset (acting as our true population) to run 2000 repeated samples. We will compare SRS, Proportional (P2S), and Optimal (Neyman) allocations.

### Using "Region" to Stratify

#### Read Population Data

```{r agpop-region-load}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```

#### Define Stratum Variable

```{r agpop-region-stratum}
agpop$stratum <- agpop$region
## reorder agpop for the ease of using strata of "sampling" (very important)
agpop <- agpop[order(agpop$stratum), ]
```

A stratification variable is most effective when it strongly explains the variance in the target variable. We check this via ANOVA and R-squared.

```{r fig-agpop-region-boxplot}
#| fig-cap: "1992 farm acreage by region stratum"
boxplot(acres92 ~ stratum, data = agpop)
```

```{r agpop-region-anova}
lm_region <- lm(agpop$acres92 ~ agpop$stratum)
anova(lm_region)
lm_region_gt <- format_lm_gt(lm_region)
lm_region_gt$fit
r_squared_region <- summary(lm_region)$r.squared
```
**Output Note:** The R-squared is `r round(r_squared_region, 3)`. The region variable only explains about `r round(r_squared_region * 100, 1)`% of the variance in acreage, which is helpful but not extremely strong.

#### Stratified Sampling with P2S allocation (Single Run)

```{r agpop-region-p2s-single}
# doing one stratified sampling
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

## doing stratified sampling
strsample <- strata(agpop, "stratum", size = nh, method = "srswor")

# collecting data on sampled counties
agstrat <- agpop[strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean")$estimate |>
  format_est_gt("mean")
```
**Output Note:** This provides a single point estimate for demonstration.

#### Repeat stratified sampling with P2S allocation 2000 times

```{r agpop-region-p2s-sim, cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300) ## make sure the order matches Nh

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r tbl-agpop-region-neyman-allocation}
#| tbl-cap: "Population sizes and Neyman-optimal sample sizes by region stratum"
Nh <- tapply(agpop$acres92, agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

data.frame(Nh, Sh, nh_opt) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = Sh, decimals = 2)
```
**Output Note:** This table compares the population sizes to the optimal sample sizes. Notice that the West region (high variance) gets a disproportionately larger sample under Neyman allocation.

```{r agpop-region-neyman-sim, cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r agpop-region-srs-sim, cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r fig-agpop-region-compare-boxplot}
#| fig-cap: "Sampling distribution of the mean under SRS, proportional, and Neyman allocation (region strata, 2000 simulations)"
sim_results_str_region <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_region)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The plot visually demonstrates that Neyman allocation has the tightest spread around the true mean (the red line).

```{r tbl-agpop-region-compare-methods}
#| tbl-cap: "Simulated mean, variance, and variance reduction by allocation method (region strata)"
sim_means <- sapply(sim_results_str_region, mean)
sim_var <- sapply(sim_results_str_region, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table summarizes the simulation. Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, while Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`% compared to Simple Random Sampling.

### Using "acres82" to Define Strata

What if we stratify by a variable highly correlated with our target, like the acreage from a decade prior (`acres82`)?

#### Read Population Data

```{r agpop-acres82-load}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] 
N <- nrow(agpop)
```

#### Define Stratum Variable with Quantiles of "acres82"
We convert the continuous `acres82` variable into four distinct categorical strata based on quantiles.

```{r agpop-acres82-stratum}
agpop$stratum <- cut(agpop$acres82,
                     breaks = quantile(agpop$acres82, probs = c(0,0.25,0.75,0.95,1)),
                     include.lowest = T)
levels(agpop$stratum) <- paste0("acres82", c("Q1", "Q2", "Q3", "Q4"))
agpop <- agpop[order(agpop$stratum), ]
```

Let's check the explanatory power of this new stratification variable.

```{r fig-agpop-acres82-boxplot}
#| fig-cap: "1992 farm acreage by acres82 quartile stratum"
boxplot(acres92 ~ stratum, data = agpop)
```

```{r agpop-acres82-anova}
lm_acres82 <- lm(agpop$acres92 ~ agpop$stratum)
lm_acres82_gt <- format_lm_gt(lm_acres82)
lm_acres82_gt$fit
r_squared_acres82 <- summary(lm_acres82)$r.squared
```
**Output Note:** The R-squared is `r round(r_squared_acres82, 3)`. The past acreage explains `r round(r_squared_acres82 * 100, 1)`% of the variance in 1992 acreage, making this a vastly superior stratification variable.

#### Stratified sampling with P2S allocation

```{r agpop-acres82-p2s-sim, cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r agpop-acres82-neyman-sim, cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r agpop-acres82-srs-sim, cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r fig-agpop-acres82-compare-boxplot}
#| fig-cap: "Sampling distribution of the mean under SRS, proportional, and Neyman allocation (acres82 strata, 2000 simulations)"
sim_results_str_acres82 <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_acres82)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The boxplot visually highlights a massive reduction in variance compared to SRS, especially for the Neyman allocation.

```{r tbl-agpop-acres82-compare-methods}
#| tbl-cap: "Simulated mean, variance, and variance reduction by allocation method (acres82 strata)"
sim_means <- sapply(sim_results_str_acres82, mean)
sim_var <- sapply(sim_results_str_acres82, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table shows dramatic results: Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, and Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`%. Stratifying by a highly correlated variable is immensely powerful.

## Interactive Demonstration: Stratification by `acres82` or `region` {#sec-app-stratification}

The app below draws three samples of the same total size $n$ at every step: a
simple random sample, a stratified sample with proportional allocation, and a
stratified sample with Neyman allocation. It uses the full `agpop.csv`
population (with the `-99` missing-value codes recoded to `NA` and dropped),
and a radio button lets you choose the stratification variable: the `acres82`
quartile groups defined above (labelled Q1 to Q4), or the four `region`
categories (NC, NE, S, W) used earlier in this chapter.

The two panels sit side by side. The left panel shows the population split into
its strata along the x-axis, each drawn as a column of jittered points on a
broken `acres92` axis; the radio buttons choose which of the three samples is
highlighted in red, and each stratum is labelled with its $N_h$ and the $n_h$
that the chosen allocation assigns to it.

The right panel accumulates the sample means of all three designs on a common
`acres92` axis, one column per design (SRS, Proportional, Neyman), with a violin
summarising each cloud once enough draws have accumulated. The bar beside each
column is the current confidence interval for that design, green when it covers
$\mu$ and red when it does not, and replaced at every draw. The running CI
coverage rate and the variance reduction relative to SRS are shown as legends.

Play advances the simulation one draw at a time; Pause stops it; Skip jumps
straight to the final violin plots by drawing every remaining replication up to
the cap in a single step.


::: {#fig-strat-app}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 720

## file: app.R
library(shiny)

data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/data/agpop.csv"
download.file(data_url, "agpop.csv")
dat <- read.csv("agpop.csv")

## -99 is the missing-value code in agpop.csv; recode to NA and drop
## counties missing either the response (acres92) or the acres82 stratifier,
## so that the same population underlies both stratification choices below.
dat$acres92[dat$acres92 == -99] <- NA
dat$acres82[dat$acres82 == -99] <- NA
dat <- dat[!is.na(dat$acres92) & !is.na(dat$acres82), ]

pop   <- dat$acres92
N     <- length(pop)
ybarU <- mean(pop)
S     <- sd(pop)

## build the stratum factor, population index list, and per-stratum summaries
## for either stratification variable ("acres82" quartiles or "region"); x_pos
## jitters each unit about its stratum's integer position, 1 (left) to H (right)
make_strata <- function(method) {
  if (method == "acres82") {
    stratum <- cut(dat$acres82,
                    breaks = quantile(dat$acres82, probs = c(0, 0.25, 0.75, 0.95, 1)),
                    include.lowest = TRUE)
    levels(stratum) <- paste0("Q", seq_along(levels(stratum)))
  } else {
    stratum <- factor(dat$region)
  }
  H     <- nlevels(stratum)
  idx_h <- split(seq_len(N), stratum)
  Nh    <- sapply(idx_h, length)
  Sh    <- sapply(idx_h, function(i) sd(pop[i]))
  W     <- Nh / N

  set.seed(42)
  x_pos <- as.integer(stratum) + runif(N, -0.30, 0.30)

  list(lev = levels(stratum), H = H, idx_h = idx_h,
       Nh = Nh, Sh = Sh, W = W, x_pos = x_pos)
}

STRAT_CHOICES <- c("acres82 quartiles" = "acres82", "region" = "region")

## ---- broken y-axis, shared by the left panel --------------------------------
## Drawn region [0, 2 * ybarU]: visible scale [0, cut] plus a compressed lane
## for the right tail, so the population mean sits at the midpoint of the panel.
lane_frac <- 0.10
cut       <- 2 * ybarU / (1 + lane_frac)
lane_w    <- lane_frac * cut
ymax_top  <- cut + lane_w

py      <- pop                                 # plotting positions on the y axis
tail_id <- which(pop > cut)
n_tail  <- length(tail_id)
if (n_tail > 0) {
  r <- rank(pop[tail_id], ties.method = "first") / (n_tail + 1)
  py[tail_id] <- cut + lane_w * (0.10 + 0.80 * r)
}

## ---- fixed window for the sampling distributions ---------------------------
se100     <- sqrt(1 - 100 / N) * S / sqrt(100)
est_range <- ybarU + c(-1, 1) * 5 * se100

METHOD <- c("SRS", "Proportional", "Neyman")
MCOL   <- c("grey35", "steelblue", "darkorange")
MAR    <- c(6.5, 4.5, 3.0, 1.0)   # identical in both panels, so the scales align

## allocation with at least 2 units per stratum, summing to n
alloc <- function(w, n, Nh) {
  nh <- pmin(pmax(2, round(w / sum(w) * n)), Nh)
  while (sum(nh) > n && any(nh > 2)) {
    i <- which.max(nh - 2); nh[i] <- nh[i] - 1
  }
  while (sum(nh) < n && any(nh < Nh)) {
    i <- which.max(Nh - nh); nh[i] <- nh[i] + 1
  }
  nh
}

## symmetric density polygon, drawn vertically about x = at
violin <- function(x, at, col, w = 0.32) {
  if (length(x) < 10 || diff(range(x)) <= 0) return(invisible())
  d  <- density(x)
  hw <- w * d$y / max(d$y)
  polygon(c(at + hw, rev(at - hw)), c(d$x, rev(d$x)),
          col = adjustcolor(col, 0.35), border = col)
  m <- median(x)
  segments(at - w * 0.6, m, at + w * 0.6, m, col = col, lwd = 2)
}

pct <- function(v) if (is.na(v)) "--" else sprintf("%.1f%%", 100 * v)

## ---- inline SVG button icons (no font or icon-library dependency) ----------
svg_icon <- function(paths)
  HTML(sprintf(paste0('<svg width="16" height="16" viewBox="0 0 16 16" ',
                      'fill="currentColor" style="vertical-align:-2px">%s</svg>'), paths))
icon_play  <- svg_icon('<path d="M3 2v12l10-6z"/>')
icon_pause <- svg_icon('<rect x="3" y="2" width="4" height="12"/><rect x="9" y="2" width="4" height="12"/>')
icon_skip  <- svg_icon('<path d="M1 2v12l7-6zM8 2v12l6-6z"/><rect x="14" y="2" width="2" height="12"/>')

ui <- fluidPage(
  titlePanel("Stratified sampling (by acres82 quartiles or region), against simple random sampling"),
  sidebarLayout(
    sidebarPanel(
      width = 3,
      radioButtons("stratvar", "Stratification variable", choices = STRAT_CHOICES,
                   selected = "acres82"),
      sliderInput("n", "Total sample size (n)", min = 40, max = 1000, value = 300, step = 10),
      sliderInput("conf", "Confidence level", min = 0.80, max = 0.99, value = 0.95, step = 0.01),
      radioButtons("show", "Highlight sample from", choiceNames = METHOD,
                   choiceValues = as.character(1:3), selected = "3"),
      actionButton("draw", "Draw new samples", class = "btn-primary", width = "100%"),
      br(), br(),
      fluidRow(
        column(4, actionButton("play",  icon_play,  class = "btn-success", width = "100%",
                               title = "Play")),
        column(4, actionButton("pause", icon_pause, class = "btn-warning", width = "100%",
                               title = "Pause")),
        column(4, actionButton("skip",  icon_skip,  class = "btn-info",    width = "100%",
                               title = "Skip to the end"))
      ),
      br(),
      sliderInput("speed", "Samples per second", min = 1, max = 10, value = 3, step = 1),
      numericInput("nrep", "Stop after this many samples",
                   value = 500, min = 10, max = 20000, step = 50),
      actionButton("reset", "Reset", width = "100%")
    ),
    mainPanel(
      width = 9,
      plotOutput("mainplot", height = "480px")
    )
  )
)

server <- function(input, output, session) {

  rv <- reactiveValues(
    idx = NULL,      # list of sampled indices, one element per design (last draw)
    est = NULL,      # matrix 3 x 3: mean, lower, upper (last draw)
    M   = NULL,      # accumulated means, one column per design
    HIT = NULL,      # accumulated coverage indicators
    jit = numeric(0)
  )
  running <- reactiveVal(FALSE)

  nrep <- reactive({
    v <- input$nrep
    if (is.null(v) || is.na(v) || v < 1) 1L else min(as.integer(v), 20000L)
  })

  si <- reactive(make_strata(input$stratvar))

  strat_label <- reactive(names(STRAT_CHOICES)[STRAT_CHOICES == input$stratvar])

  nh_prop   <- reactive({ s <- si(); alloc(s$Nh,        input$n, s$Nh) })
  nh_neyman <- reactive({ s <- si(); alloc(s$Nh * s$Sh, input$n, s$Nh) })

  # draw k replications; every replication draws its own SRS, proportional, and
  # Neyman samples of total size n. Only the last replication's sampled indices
  # are kept, for highlighting in the left panel; "Skip" calls this with the
  # full remaining count so it jumps straight to the final violin plots.
  draw_many <- function(k) {
    n    <- isolate(input$n)
    conf <- isolate(input$conf)
    a    <- 1 - (1 - conf) / 2
    s    <- isolate(si())
    H     <- s$H
    idx_h <- s$idx_h
    Nh    <- s$Nh
    W     <- s$W
    npv   <- isolate(nh_prop())
    npy   <- isolate(nh_neyman())

    # stratified: ybar_str = sum W_h ybar_h, V = sum W_h^2 (1 - n_h/N_h) s_h^2 / n_h
    str_draw <- function(nh) {
      ii <- unlist(lapply(seq_len(H), function(h) sample(idx_h[[h]], nh[h])))
      yh <- split(pop[ii], rep(seq_len(H), nh))
      mh <- sapply(yh, mean)
      sh <- sapply(yh, sd)
      m  <- sum(W * mh)
      v  <- sum(W^2 * (1 - nh / Nh) * sh^2 / nh)
      c(m, qt(a, df = sum(nh) - H) * sqrt(v))
    }
    take <- function(nh) unlist(lapply(seq_len(H), function(h) sample(idx_h[[h]], nh[h])))

    Mnew   <- matrix(0, k, 3)
    HITnew <- matrix(FALSE, k, 3)
    jitnew <- numeric(k)

    for (i in seq_len(k)) {
      i_srs <- sample(N, n)
      y     <- pop[i_srs]
      m_srs <- mean(y)
      se    <- sqrt(1 - n / N) * sd(y) / sqrt(n)
      d_srs <- qt(a, df = n - 1) * se

      p <- str_draw(npv)
      q <- str_draw(npy)

      est <- rbind(c(m_srs, m_srs - d_srs, m_srs + d_srs),
                   c(p[1],  p[1]  - p[2],  p[1]  + p[2]),
                   c(q[1],  q[1]  - q[2],  q[1]  + q[2]))

      Mnew[i, ]   <- est[, 1]
      HITnew[i, ] <- est[, 2] <= ybarU & ybarU <= est[, 3]
      jitnew[i]   <- runif(1, -0.1, 0.1)

      if (i == k) {
        rv$idx <- list(i_srs, take(npv), take(npy))
        rv$est <- est
      }
    }
    rv$M   <- rbind(rv$M,   Mnew)
    rv$HIT <- rbind(rv$HIT, HITnew)
    rv$jit <- c(rv$jit, jitnew)
  }
  draw_one <- function() draw_many(1)

  # one sample per click; Play repeats until Pause or until the cap is reached
  observeEvent(input$draw,  { running(FALSE); draw_one() })
  observeEvent(input$play,  running(TRUE))
  observeEvent(input$pause, running(FALSE))

  # Skip: draw every remaining replication up to the cap in one go, landing
  # directly on the final violin plots of the three estimators
  observeEvent(input$skip, {
    running(FALSE)
    k <- nrep() - NROW(rv$M)
    if (k > 0) draw_many(k)
  })

  observe({
    if (!isTRUE(running())) return()
    if (NROW(isolate(rv$M)) >= isolate(nrep())) { running(FALSE); return() }
    invalidateLater(1000 / isolate(input$speed), session)
    isolate(draw_one())
  })

  reset_history <- function() {
    running(FALSE)
    rv$idx <- NULL; rv$est <- NULL; rv$M <- NULL; rv$HIT <- NULL
    rv$jit <- numeric(0)
  }
  observeEvent(input$reset,    reset_history())
  observeEvent(input$n,        reset_history())
  observeEvent(input$conf,     reset_history())
  observeEvent(input$stratvar, reset_history())

  fmt <- function(x) format(round(x), big.mark = ",")

  output$mainplot <- renderPlot({
    s  <- si()
    H  <- s$H; Nh <- s$Nh; lev <- s$lev; x_pos <- s$x_pos
    j  <- as.integer(input$show)
    nh <- switch(j, rep(NA_integer_, H), nh_prop(), nh_neyman())

    layout(matrix(1:2, nrow = 1), widths = c(1, 1))

    ## ---- left: population by stratum, with the highlighted sample ----
    par(mar = MAR, yaxs = "i")
    plot(NA, xlim = c(0.3, H + 0.9), ylim = c(0, ymax_top),
         xlab = "", ylab = "acres92", xaxt = "n", yaxt = "n", bty = "n",
         main = sprintf("Population by %s stratum, with the %s sample", strat_label(), METHOD[j]))

    for (h in seq_len(H)) {
      rect(h - 0.36, 0, h + 0.36, ymax_top,
           col = adjustcolor("grey92", 0.6), border = NA)
    }
    points(x_pos, py, pch = 16, cex = 0.5, col = adjustcolor("grey55", 0.5))
    if (!is.null(rv$idx))
      points(x_pos[rv$idx[[j]]], py[rv$idx[[j]]], pch = 16, cex = 0.6,
             col = adjustcolor("red", 0.85))

    axis(1, at = seq_len(H), labels = lev, tick = FALSE, line = -0.5)
    for (h in seq_len(H)) {
      lab <- if (j == 1) sprintf("N = %d", Nh[h])
             else        sprintf("N = %d\nn = %d", Nh[h], nh[h])
      mtext(lab, side = 1, at = h, line = 2.0, cex = 0.7, col = "grey30")
    }

    at <- pretty(c(0, cut), 6); at <- at[at <= cut]
    axis(2, at = at, labels = format(at, big.mark = ",", trim = TRUE), las = 1)
    segments(0.3, cut, H + 0.7, cut, col = "grey60", lty = 3)
    text(H + 0.65, cut + lane_w / 2, sprintf("compressed: > %s", fmt(cut)),
         adj = c(1, 0), col = "grey40", cex = 0.7)
    abline(h = ybarU, col = "black", lwd = 2, lty = 2)

    ## ---- right: sampling distributions of the three estimates, as violins ----
    par(mar = MAR, yaxs = "i")
    plot(NA, xlim = c(0.4, 3.6), ylim = est_range, xaxt = "n", bty = "n",
         xlab = "", ylab = "Sample mean of acres92",
         main = "Sampling distributions of the three estimators")
    abline(h = ybarU, col = "black", lwd = 2, lty = 2)
    axis(1, at = 1:3, labels = METHOD, tick = FALSE, line = -0.5)

    if (is.null(rv$M)) {
      text(2, ybarU, "Click 'Draw new samples' or Play to begin", col = "grey50")
      return(invisible())
    }

    k     <- NROW(rv$M)
    v_srs <- if (k > 1) var(rv$M[, 1]) else NA_real_

    for (jj in 1:3) {
      xc <- jj
      points(xc + rv$jit, rv$M[, jj], pch = 16, cex = 0.45,
             col = adjustcolor(MCOL[jj], 0.30))
      violin(rv$M[, jj], at = xc, col = MCOL[jj])

      # transient interval for the current draw
      ci_col <- if (rv$HIT[k, jj]) "forestgreen" else "red"
      xb     <- xc + 0.42
      segments(xb, rv$est[jj, 2], xb, rv$est[jj, 3], col = ci_col, lwd = 3)
      segments(xb - 0.06, rv$est[jj, 2:3], xb + 0.06, rv$est[jj, 2:3], col = ci_col, lwd = 3)
      points(xb, rv$est[jj, 1], pch = 19, cex = 1.1, col = ci_col)
    }

    legend("topleft", bty = "n", cex = 0.85, inset = c(0.02, 0.01),
           title = "95% CI coverage", fill = MCOL, border = NA,
           legend = paste(METHOD, vapply(1:3, function(jj) pct(mean(rv$HIT[, jj])), "")))

    if (k > 1) {
      legend("bottomleft", bty = "n", cex = 0.85, inset = c(0.02, 0.01),
             title = "Variance reduction vs. SRS", fill = MCOL[2:3], border = NA,
             legend = paste(METHOD[2:3],
                            vapply(2:3, function(jj) pct(1 - var(rv$M[, jj]) / v_srs), "")))
    }

    outside <- sum(rv$M < est_range[1] | rv$M > est_range[2])
    sub <- sprintf("n = %d per design   |   samples drawn = %d of %d   |   nominal %.0f%%",
                   input$n, k, nrep(), 100 * input$conf)
    if (outside > 0) sub <- paste0(sub, sprintf("   |   %d mean(s) beyond the axis", outside))
    mtext(sub, side = 3, line = -0.2, cex = 0.75, col = "grey30")
  })
}

shinyApp(ui, server)

```
**Shinyapp for Stratified Sampling**

:::