# Stratified Random Sampling

## Key Formulas for Stratified Sampling

Stratified random sampling divides a population of size $N$ into $H$ mutually exclusive strata. A simple random sample of size $n_h$ is then drawn from each stratum of size $N_h$. This design often yields higher precision than Simple Random Sampling (SRS).

#### 1. Stratified Population Mean Estimator {-}
The estimated population mean is a weighted average of the sample means from each stratum:
$$
\overline{y}_{str} = \sum_{h=1}^H W_h \overline{y}_h.
$$
Where $W_h = \frac{N_h}{N}$ is the stratum weight, and $\overline{y}_h$ is the sample mean of stratum $h$.

#### 2. Variance of the Stratified Mean {-}
The estimated variance accounts for the finite population correction in each stratum:
$$
\hat{V}(\overline{y}_{str}) = \sum_{h=1}^H W_h^2 \left(1 - \frac{n_h}{N_h}\right) \frac{s_h^2}{n_h}.
$$
Where $s_h^2$ is the sample variance in stratum $h$.

#### 3. Proportional Allocation {-}
Sample sizes are allocated proportional to the size of the stratum:
$$
n_h = n \left( \frac{N_h}{N} \right).
$$

#### 4. Neyman (Optimal) Allocation {-}
If surveying costs are equal across strata, Neyman allocation minimizes the variance of the estimator by sampling more heavily from larger and more highly variable strata:
$$
n_h = n \frac{N_h S_h}{\sum_{h=1}^H N_h S_h}
$$

---

## Functions and packages for Analyzing Data

The estimating functions used throughout this book (including `str_est`, `str_est_data`, and `srs_est`, used below to compute the stratified mean, standard error, and confidence intervals) live in a single shared file, `samplingestimate.r`, which we source below. `str_est_data` computes these directly from a dataset, while `str_est` computes them from pre-calculated summary statistics.

```{r}
#| echo: false
library(latex2exp)
library(sampling)
library(dplyr)
source("samplingestimate.r")
```

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

## Analysis of "agstrat.csv" data

We will estimate the average 1992 farm acreage (`acres92`) across US counties, using geographic region (`region`) as the stratification variable.

### Importing agstrat.csv data

```{r}
agstrat <- read.csv("data/agstrat.csv")
head(agstrat) |> gt()
```
**Output Note:** The table above shows the sample survey data, including the stratum (region) and weights for each county.

### Spreadsheet calculation

#### Summarizing acre92 in each stratum

We first manually compute the sample size ($n_h$), standard deviation ($s_h$), and mean ($\overline{y}_h$) for each region, matching them with the known total counties in each region ($N_h$).

```{r}
nh <- tapply (agstrat[, "acres92"], agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information (true population size per stratum)
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

data.frame(Nh, nh, ybarh, sh) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    Nh = md("$N_h$"),
    nh = md("$n_h$"),
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$")
  ) |>
  fmt_number(columns = c(ybarh, sh), decimals = 2)
```
**Output Note:** This displays the summary statistics per stratum. Note the significantly high variance ($s_h$) in the `r names(sh)[which.max(sh)]` region compared to the others.

#### Estimates

Rather than building the "spreadsheet-style" working table by hand, we call `str_est()`, whose `$table` shows the same intermediate stratum-level values ($N_h$, $n_h$, $\pi_h$, $\bar y_h$, $s_h^2$, $\pi_h\bar y_h$, $v_h$) needed for the final stratified mean and variance, with a Total row and footnotes for $\pi_h$ and $v_h$.

```{r}
res_str_mean <- str_est(ybarh, sh, nh, Nh, estimate = "mean")
res_str_mean$table
```
**Output Note:** The table provides a detailed breakdown of how each individual stratum contributes to the final mean and variance.

Extracting `$estimate` gives our overall result.

```{r}
res_str_mean$estimate |> format_est_gt("mean")
```
**Output Note:** The overall result indicates an estimated mean of `r format(round(res_str_mean$estimate["Est."], 0), big.mark=",")` acres, with a 95% confidence interval of [`r format(round(res_str_mean$estimate["ci.low"], 0), big.mark=",")`, `r format(round(res_str_mean$estimate["ci.upp"], 0), big.mark=",")`].

To find the population total (total acreage across all US counties), set `estimate = "total"`.

```{r}
## population total estimate
res_str_total <- str_est(ybarh, sh, nh, Nh, estimate = "total")
res_str_total$estimate |> format_est_gt("total")
```
**Output Note:** This scales the estimates up to the total area of `r format(round(res_str_total$estimate["Est."], 0), big.mark=",", scientific=FALSE)` acres.

### Using the function "str_est_data"

Instead of pre-calculating the summary stats, this function extracts everything directly from the dataset using the pre-assigned sampling weights.

```{r}
## In the function "str_est_data", we can find the stratum size with the variable "weights":
tapply(agstrat$weight, agstrat$region, sum)
```
**Output Note:** This reconstructs the total population sizes ($N_h$) per region.

```{r}
## if the dataset contains a variable "weight"
res_agstrat_acres92 <- str_est_data(agstrat, y="acres92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_acres92$table
res_agstrat_acres92$estimate |> format_est_gt("mean")
```
**Output Note:** Once again, this perfectly matches the manual calculation.

```{r}
## estimating the mean of the number of small farms in 1992
res_agstrat_smallf92 <- str_est_data(agstrat, y="smallf92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_smallf92$table
res_agstrat_smallf92$estimate |> format_est_gt("mean")
```
**Output Note:** This demonstrates the function's flexibility: we can easily estimate a different variable (an average of `r round(res_agstrat_smallf92$estimate["Est."], 1)` small farms per county).

### Comparing with SRS estimate

To see the benefit of stratifying, we compare our stratified estimates against a Simple Random Sample approach on the same population.

```{r}
agsrs <- read.csv("data/agsrs.csv")
res_agsrs_srs <- srs_est(agsrs[, "acres92"], N=3078, estimate = "mean", show.details = FALSE)
res_agsrs_srs$estimate |> format_est_gt("mean")
```
**Output Note:** The Simple Random Sample standard error is noticeably higher at `r format(round(res_agsrs_srs$estimate["S.E."], 0), big.mark=",")`.

```{r}
## the ratio of estimated variance (Stratified SE / SRS SE)^2
ratio_var <- (res_agstrat_acres92$estimate["S.E."] / res_agsrs_srs$estimate["S.E."])^2
unname(ratio_var)
```
**Output Note:** The stratified variance is approximately `r round(ratio_var * 100, 1)`% of the SRS variance.

```{r}
## percentage of reduction of variance of str estimates from that of SRS estimate
unname(1 - ratio_var)
```
**Output Note:** Stratification by region reduced the variance by `r round((1 - ratio_var) * 100, 1)`%. This is a substantial gain in precision for no extra sampling cost.

## Allocation of stratum sample size

Deciding how to distribute the total sample $n$ among the strata significantly impacts the precision of the estimator.

### Analyzing seals.csv collected with stratified sampling

This dataset investigates ringed seal holes in different zones. The researchers used Proportional Allocation, taking 20% of the areas in each zone.

```{r}
## load data
seals <- read.csv("data/seals.csv")

# survey data summary in each stratum
nh <- as.vector(table(seals[,"zone"]))
sh <- tapply(seals[, "holes"], seals[,"zone"], sd)
ybarh <- tapply(seals[, "holes"], seals[,"zone"], mean)
Nh <- c(68, 84, 48)

res_seals <- str_est(ybarh, sh, nh, Nh, estimate = "mean")
res_seals$table
```
**Output Note:** Zone `r names(sh)[which.max(sh)]` has the highest standard deviation ($s_h \approx `r round(max(sh), 1)`$), indicating a high degree of variability in seal density in that region.

```{r}
res_seals$estimate |> format_est_gt("mean")
```
**Output Note:** This provides the estimated mean number of seal holes per square kilometer (`r round(res_seals$estimate["Est."], 1)`).

### Neyman allocation of stratum sample size

Because Zone 3 was highly variable, we could have achieved a lower overall variance if we had sampled it more heavily. Neyman allocation formally calculates this optimal distribution.

```{r}
## assuming the cost of surveying each zone is the same
Ch <- rep(1, 3)
Sh <- sh ## estimate population Sh with sample sh
NhShDCh <- Nh * Sh / sqrt(Ch)

## the optimal allocation proportion (Lh) for estimating the population total or mean:
Lh <- NhShDCh / sum(NhShDCh)

data.frame(Ch, Nh, Sh, NhShDCh, Lh) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(NhShDCh = md("$N_h S_h / \\sqrt{C_h}$"), Lh = md("$L_h$ (Prop)"), Ch=md("$C_h$"), Sh=md("$S_h$"), Nh=md("$N_h$")) |>
  fmt_number(columns = c(Sh, NhShDCh, Lh), decimals = 3)
```
**Output Note:** This displays the optimal allocation proportions ($L_h$). Even though Zone 2 is a larger area ($N_h=84$), Zone 3's high variance means we should allocate the largest share (`r round(Lh["3"] * 100, 0)`%) of our sampling effort there.

## Simulation to Study the Efficiency of Stratified Sampling with Different Allocation

We will use the full `agpop.csv` dataset (acting as our true population) to run 2000 repeated samples. We will compare SRS, Proportional (P2S), and Optimal (Neyman) allocations.

### Using "Region" to Stratify

#### Read Population Data

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```

#### Define Stratum Variable

```{r}
agpop$stratum <- agpop$region
## reorder agpop for the ease of using strata of "sampling" (very important)
agpop <- agpop[order(agpop$stratum), ]
```

A stratification variable is most effective when it strongly explains the variance in the target variable. We check this via ANOVA and R-squared.

```{r}
boxplot(acres92 ~ stratum, data = agpop)
anova(lm(agpop$acres92 ~ agpop$stratum))
summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared
```
**Output Note:** The R-squared is `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared, 3)`. The region variable only explains about `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared * 100, 1)`% of the variance in acreage, which is helpful but not extremely strong.

#### Stratified Sampling with P2S allocation (Single Run)

```{r}
# doing one stratified sampling
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

## doing stratified sampling
strsample <- strata(agpop, "stratum", size = nh, method = "srswor")

# collecting data on sampled counties
agstrat <- agpop[strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean")$estimate |>
  format_est_gt("mean")
```
**Output Note:** This provides a single point estimate for demonstration.

#### Repeat stratified sampling with P2S allocation 2000 times

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300) ## make sure the order matches Nh

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r}
Nh <- tapply(agpop$acres92, agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

data.frame(Nh, Sh, nh_opt) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = Sh, decimals = 2)
```
**Output Note:** This table compares the population sizes to the optimal sample sizes. Notice that the West region (high variance) gets a disproportionately larger sample under Neyman allocation.

```{r cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r}
sim_results_str_region <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_region)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The plot visually demonstrates that Neyman allocation has the tightest spread around the true mean (the red line).

```{r}
sim_means <- sapply(sim_results_str_region, mean)
sim_var <- sapply(sim_results_str_region, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table summarizes the simulation. Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, while Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`% compared to Simple Random Sampling.

### Using "acres82" to Define Strata

What if we stratify by a variable highly correlated with our target, like the acreage from a decade prior (`acres82`)?

#### Read Population Data

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] 
N <- nrow(agpop)
```

#### Define Stratum Variable with Quantiles of "acres82"
We convert the continuous `acres82` variable into four distinct categorical strata based on quantiles.

```{r}
agpop$stratum <- cut(agpop$acres82,
                     breaks = quantile(agpop$acres82, probs = c(0,0.25,0.75,0.95,1)),
                     include.lowest = T)
levels(agpop$stratum) <- paste0("acres82", c("Q1", "Q2", "Q3", "Q4"))
agpop <- agpop[order(agpop$stratum), ]
```

Let's check the explanatory power of this new stratification variable.

```{r}
boxplot(acres92 ~ stratum, data = agpop)
summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared
```
**Output Note:** The R-squared is `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared, 3)`. The past acreage explains `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared * 100, 1)`% of the variance in 1992 acreage, making this a vastly superior stratification variable.

#### Stratified sampling with P2S allocation

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_est_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r}
sim_results_str_acres82 <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_acres82)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The boxplot visually highlights a massive reduction in variance compared to SRS, especially for the Neyman allocation.

```{r}
sim_means <- sapply(sim_results_str_acres82, mean)
sim_var <- sapply(sim_results_str_acres82, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table shows dramatic results: Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, and Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`%. Stratifying by a highly correlated variable is immensely powerful.