# Stratified Random Sampling

## Key Formulas for Stratified Sampling

Stratified random sampling divides a population of size $N$ into $H$ mutually exclusive strata. A simple random sample of size $n_h$ is then drawn from each stratum of size $N_h$. This design often yields higher precision than Simple Random Sampling (SRS).

#### 1. Stratified Population Mean Estimator {-}
The estimated population mean is a weighted average of the sample means from each stratum:
$$
\overline{y}_{str} = \sum_{h=1}^H W_h \overline{y}_h.
$$
Where $W_h = \frac{N_h}{N}$ is the stratum weight, and $\overline{y}_h$ is the sample mean of stratum $h$.

#### 2. Variance of the Stratified Mean {-}
The estimated variance accounts for the finite population correction in each stratum:
$$
\hat{V}(\overline{y}_{str}) = \sum_{h=1}^H W_h^2 \left(1 - \frac{n_h}{N_h}\right) \frac{s_h^2}{n_h}.
$$
Where $s_h^2$ is the sample variance in stratum $h$.

#### 3. Proportional Allocation {-}
Sample sizes are allocated proportional to the size of the stratum:
$$
n_h = n \left( \frac{N_h}{N} \right).
$$

#### 4. Neyman (Optimal) Allocation {-}
If surveying costs are equal across strata, Neyman allocation minimizes the variance of the estimator by sampling more heavily from larger and more highly variable strata:
$$
n_h = n \frac{N_h S_h}{\sum_{h=1}^H N_h S_h}
$$

---

## Functions and packages for Analyzing Data

The following code loads necessary libraries and defines custom functions to automate the calculation of the stratified mean, standard error, and confidence intervals. `str_mean_estimate_data` computes these directly from a dataset, while `str_mean_estimate` computes them from pre-calculated summary statistics. We also define helper functions to format our output using the `gt` package.

```{r}
#| code-fold: true
library(latex2exp)
library(sampling)
library(dplyr)
library(gt)

## Formats a gt table column as ordinary fixed-point, UNLESS its values are
## too small (< `small` in absolute value) or too large (>= `large`), in
## which case it switches that column to scientific notation instead.
fmt_auto <- function (gt_tbl, data, columns, decimals = 3, small = 1e-3, large = 1e5)
{
    for (col in columns)
    {
        x <- data[[col]]
        x <- x[is.finite (x) & x != 0]
        if (length (x) == 0) next
        if (any (abs (x) < small) || any (abs (x) >= large))
            gt_tbl <- fmt_scientific (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
        else
            gt_tbl <- fmt_number (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
    }
    gt_tbl
}

## Truncates a row-level working data.frame whose LAST row is a summary row
## to the first head_n data rows + "..." + the summary row.
truncate_working <- function (working, id_col, head_n = 6)
{
    n <- nrow (working) - 1
    if (n <= head_n + 1) return (working)

    dots <- working[1, ]
    dots[1, ] <- NA
    dots[1, id_col] <- "..."

    rbind (working[seq_len (head_n), ], dots, working[nrow (working), ])
}

## Given per-stratum summaries, compute the stratified (or post-stratified)
## mean or total.
## estimate --- "mean" (default) returns the stratified MEAN estimate;
##              "total" returns the stratified TOTAL estimate
## show.details --- if TRUE (default), also return a gt table showing the
##                  stratum-by-stratum working, with a Total row
## for poststratification, use nh = n * Nh/N
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
str_mean_estimate <- function (ybarh, sh, nh, Nh, estimate = c ("mean", "total"), show.details = TRUE)
{
    estimate <- match.arg (estimate)

    N    <- sum (Nh)
    Pi_h <- Nh / N
    v_h  <- (1 - nh / Nh) * Pi_h^2 * sh^2 / nh

    ybar   <- sum (ybarh * Pi_h)
    seybar <- sqrt (sum (v_h))

    if (estimate == "total")
    {
        est    <- N * ybar
        se.est <- N * seybar
    }
    else
    {
        est    <- ybar
        se.est <- seybar
    }
    mem          <- 1.96 * se.est
    estimate_vec <- c (Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return (list (estimate = estimate_vec, table = NULL))

    stratum <- names (Nh)
    if (is.null (stratum)) stratum <- seq_along (Nh)

    working <- data.frame (
        Stratum       = c (stratum, "Total"),
        Nh            = c (Nh, N),
        nh            = c (nh, sum (nh)),
        Pi_h          = c (Pi_h, sum (Pi_h)),
        ybarh         = c (ybarh, NA),
        sh2           = c (sh^2, NA),
        weighted_mean = c (Pi_h * ybarh, ybar),
        v_h           = c (v_h, seybar^2)
    )

    table <- gt (working) |>
        tab_header (title = "Stratified Mean Estimation: Working Table") |>
        cols_label (
            Stratum       = md ("Stratum"),
            Nh            = md ("$N_h$"),
            nh            = md ("$n_h$"),
            Pi_h          = md ("$\\pi_h$"),
            ybarh         = md ("$\\bar y_h$"),
            sh2           = md ("$s_h^2$"),
            weighted_mean = md ("$\\pi_h\\bar y_h$"),
            v_h           = md ("$v_h$")
        ) |>
        cols_width (c (Nh, nh) ~ px (60), everything () ~ px (110)) |>
        fmt_number (columns = c (Nh, nh), decimals = 0) |>
        fmt_number (columns = Pi_h, decimals = 2) |>
        fmt_auto (data = working, columns = c ("ybarh", "sh2", "weighted_mean", "v_h"), decimals = 2) |>
        sub_missing (missing_text = "") |>
        tab_style (
            style     = cell_text (weight = "bold"),
            locations = cells_body (rows = Stratum == "Total")
        ) |>
        tab_footnote (
            footnote  = md ("$\\pi_h = N_h/N$ is the stratum weight."),
            locations = cells_column_labels (columns = Pi_h)
        ) |>
        tab_footnote (
            footnote  = md ("$v_h = (1-n_h/N_h)\\,\\pi_h^2\\, s_h^2/n_h$ is stratum $h$'s contribution to $V(\\bar y_{str}) = \\sum_h v_h$."),
            locations = cells_column_labels (columns = v_h)
        )

    list (estimate = estimate_vec, table = table)
}

## this function finds statistical estimates given a dataset with sampling weight
# stratdata --- data.frame containing stratified sample
# y --- name of variable for which we want to estiamte population mean
# stratum --- name of variable that will be used as stratum variable
# weight --- name of variable indicating sampling weight
# note: from weights we can find Nh (see the code for formula)
# returns the same list ($estimate, $table) as str_mean_estimate(), which
# this function calls after deriving ybarh/sh/nh/Nh from the raw data
str_mean_estimate_data <- function (stratdata, y, stratum, weight, estimate = c ("mean", "total"), show.details = TRUE)
{
    ## compute stratum-wise data
    sh <- tapply (stratdata[, y], stratdata[,stratum], sd)
    ybarh <- tapply (stratdata[, y], stratdata[,stratum], mean)
    ## find population stratum size using sampling weight included in the data set
    Nh <- tapply (stratdata[, weight], stratdata[,stratum], sum)
    nh <- tapply (1:nrow(stratdata), stratdata[,stratum], length)

    str_mean_estimate (ybarh, sh, nh, Nh, estimate = estimate, show.details = show.details)
}

## sdata --- a vector of original survey data
## N --- population size
## estimate --- "mean" (default) returns the population MEAN estimate;
##              "total" returns the population TOTAL estimate (requires
##              finite N)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values, with a Sum row and a
##                  footnote showing s_y^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_mean_est <- function (sdata, N = Inf, estimate = c ("mean", "total"),
                          show.details = TRUE, col_labels = list (y = "y_i"))
{
    estimate <- match.arg (estimate)

    n <- length (sdata)
    ybar <- mean (sdata)
    dev <- sdata - ybar
    s2y <- sum (dev^2) / (n - 1)
    se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)

    if (estimate == "total")
    {
        if (!is.finite (N))
            stop ("N must be finite to compute estimate = \"total\"")
        est <- N * ybar
        se.est <- N * se.ybar
    }
    else
    {
        est <- ybar
        se.est <- se.ybar
    }
    mem <- qt (0.975, df = n - 1) * se.est
    estimate_vec <- c (Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return (list (estimate = estimate_vec, table = NULL))

    working <- data.frame (
        i    = c (seq_len (n), "Sum"),
        y    = c (sdata, sum (sdata)),
        dev  = c (dev, sum (dev)),
        dev2 = c (dev^2, sum (dev^2))
    )
    working <- truncate_working (working, id_col = "i")

    table <- gt (working) |>
        tab_header (title = "SRS Mean Estimation: Working Table", subtitle = md (sprintf ("$n = %d$", n))) |>
        cols_label (
            i    = md ("$i$"),
            y    = md (sprintf ("$%s$", col_labels$y)),
            dev  = md (sprintf ("$%s-\\bar y$", col_labels$y)),
            dev2 = md (sprintf ("$(%s-\\bar y)^2$", col_labels$y))
        ) |>
        cols_width (i ~ px (40), everything () ~ px (110)) |>
        fmt_auto (data = working, columns = c ("y", "dev", "dev2"), decimals = 3) |>
        sub_missing (missing_text = "...") |>
        tab_style (
            style     = cell_text (weight = "bold"),
            locations = cells_body (rows = i == "Sum")
        ) |>
        tab_footnote (
            footnote  = md (sprintf ("$s_y^2 = \\sum_i (%s-\\bar y)^2/(n-1) = %.4f$.", col_labels$y, s2y)),
            locations = cells_column_labels (columns = dev2)
        )

    list (estimate = estimate_vec, table = table)
}

# Helper function to format estimation outputs with appropriate mathematical headers
format_est_gt <- function(est_data, est_type = c("mean", "total", "ratio", "reg_mean", "reg_total", "ratio_mean", "domain_mean")) {
  est_type <- match.arg(est_type)

  sym_map <- list(
    "mean"        = c(est = "$\\bar{y}$",         se = "$\\mathrm{SE}(\\bar{y})$"),
    "total"       = c(est = "$\\hat{t}$",         se = "$\\mathrm{SE}(\\hat{t})$"),
    "ratio"       = c(est = "$\\hat{B}$",         se = "$\\mathrm{SE}(\\hat{B})$"),
    "reg_mean"    = c(est = "$\\bar{y}_{reg}$",   se = "$\\mathrm{SE}(\\bar{y}_{reg})$"),
    "reg_total"   = c(est = "$\\hat{t}_{reg}$",   se = "$\\mathrm{SE}(\\hat{t}_{reg})$"),
    "ratio_mean"  = c(est = "$\\bar{y}_{r}$",     se = "$\\mathrm{SE}(\\bar{y}_{r})$"),
    "domain_mean" = c(est = "$\\bar{y}_{d}$",     se = "$\\mathrm{SE}(\\bar{y}_{d})$")
  )

  est_sym <- sym_map[[est_type]]["est"]
  se_sym  <- sym_map[[est_type]]["se"]

  if (is.vector(est_data) && !is.list(est_data)) {
    df <- as.data.frame(t(unname(est_data)))
  } else {
    df <- as.data.frame(est_data)
  }

  colnames(df) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  use_stub <- !is.null(rownames(df)) && !all(rownames(df) == as.character(1:nrow(df)))

  df |>
    gt(rownames_to_stub = use_stub) |>
    cols_label(
      Est.   = md(est_sym),
      S.E.   = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(columns = everything(), decimals = 4) |>
    tab_options(table.width = pct(70))
}
```

## Analysis of "agstrat.csv" data

We will estimate the average 1992 farm acreage (`acres92`) across US counties, using geographic region (`region`) as the stratification variable.

### Importing agstrat.csv data

```{r}
agstrat <- read.csv("data/agstrat.csv")
head(agstrat) |> gt()
```
**Output Note:** The table above shows the sample survey data, including the stratum (region) and weights for each county.

### Spreadsheet calculation

#### Summarizing acre92 in each stratum

We first manually compute the sample size ($n_h$), standard deviation ($s_h$), and mean ($\overline{y}_h$) for each region, matching them with the known total counties in each region ($N_h$).

```{r}
nh <- tapply (agstrat[, "acres92"], agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information (true population size per stratum)
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

data.frame(Nh, nh, ybarh, sh) |> 
  gt(rownames_to_stub = TRUE) |>
  cols_label(
    Nh = md("$N_h$"),
    nh = md("$n_h$"),
    ybarh = md("$\\bar{y}_h$"),
    sh = md("$s_h$")
  ) |>
  fmt_number(columns = c(ybarh, sh), decimals = 2)
```
**Output Note:** This displays the summary statistics per stratum. Note the significantly high variance ($s_h$) in the `r names(sh)[which.max(sh)]` region compared to the others.

#### Estimates

Rather than building the "spreadsheet-style" working table by hand, we call `str_mean_estimate()`, whose `$table` shows the same intermediate stratum-level values ($N_h$, $n_h$, $\pi_h$, $\bar y_h$, $s_h^2$, $\pi_h\bar y_h$, $v_h$) needed for the final stratified mean and variance, with a Total row and footnotes for $\pi_h$ and $v_h$.

```{r}
res_str_mean <- str_mean_estimate(ybarh, sh, nh, Nh, estimate = "mean")
res_str_mean$table
```
**Output Note:** The table provides a detailed breakdown of how each individual stratum contributes to the final mean and variance.

Extracting `$estimate` gives our overall result.

```{r}
res_str_mean$estimate |> format_est_gt("mean")
```
**Output Note:** The overall result indicates an estimated mean of `r format(round(res_str_mean$estimate["Est."], 0), big.mark=",")` acres, with a 95% confidence interval of [`r format(round(res_str_mean$estimate["ci.low"], 0), big.mark=",")`, `r format(round(res_str_mean$estimate["ci.upp"], 0), big.mark=",")`].

To find the population total (total acreage across all US counties), set `estimate = "total"`.

```{r}
## population total estimate
res_str_total <- str_mean_estimate(ybarh, sh, nh, Nh, estimate = "total")
res_str_total$estimate |> format_est_gt("total")
```
**Output Note:** This scales the estimates up to the total area of `r format(round(res_str_total$estimate["Est."], 0), big.mark=",", scientific=FALSE)` acres.

### Using the function "str_mean_estimate_data"

Instead of pre-calculating the summary stats, this function extracts everything directly from the dataset using the pre-assigned sampling weights.

```{r}
## In the function "str_mean_estimate_data", we can find the stratum size with the variable "weights":
tapply(agstrat$weight, agstrat$region, sum)
```
**Output Note:** This reconstructs the total population sizes ($N_h$) per region.

```{r}
## if the dataset contains a variable "weight"
res_agstrat_acres92 <- str_mean_estimate_data(agstrat, y="acres92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_acres92$table
res_agstrat_acres92$estimate |> format_est_gt("mean")
```
**Output Note:** Once again, this perfectly matches the manual calculation.

```{r}
## estimating the mean of the number of small farms in 1992
res_agstrat_smallf92 <- str_mean_estimate_data(agstrat, y="smallf92", stratum="region", weight="weight", estimate = "mean")
res_agstrat_smallf92$table
res_agstrat_smallf92$estimate |> format_est_gt("mean")
```
**Output Note:** This demonstrates the function's flexibility: we can easily estimate a different variable (an average of `r round(res_agstrat_smallf92$estimate["Est."], 1)` small farms per county).

### Comparing with SRS estimate

To see the benefit of stratifying, we compare our stratified estimates against a Simple Random Sample approach on the same population.

```{r}
agsrs <- read.csv("data/agsrs.csv")
res_agsrs_srs <- srs_mean_est(agsrs[, "acres92"], N=3078, estimate = "mean")
res_agsrs_srs$table
res_agsrs_srs$estimate |> format_est_gt("mean")
```
**Output Note:** The Simple Random Sample standard error is noticeably higher at `r format(round(res_agsrs_srs$estimate["S.E."], 0), big.mark=",")`.

```{r}
## the ratio of estimated variance (Stratified SE / SRS SE)^2
ratio_var <- (res_agstrat_acres92$estimate["S.E."] / res_agsrs_srs$estimate["S.E."])^2
unname(ratio_var)
```
**Output Note:** The stratified variance is approximately `r round(ratio_var * 100, 1)`% of the SRS variance.

```{r}
## percentage of reduction of variance of str estimates from that of SRS estimate
unname(1 - ratio_var)
```
**Output Note:** Stratification by region reduced the variance by `r round((1 - ratio_var) * 100, 1)`%. This is a substantial gain in precision for no extra sampling cost.

## Allocation of stratum sample size

Deciding how to distribute the total sample $n$ among the strata significantly impacts the precision of the estimator.

### Analyzing seals.csv collected with stratified sampling

This dataset investigates ringed seal holes in different zones. The researchers used Proportional Allocation, taking 20% of the areas in each zone.

```{r}
## load data
seals <- read.csv("data/seals.csv")

# survey data summary in each stratum
nh <- as.vector(table(seals[,"zone"]))
sh <- tapply(seals[, "holes"], seals[,"zone"], sd)
ybarh <- tapply(seals[, "holes"], seals[,"zone"], mean)
Nh <- c(68, 84, 48)

res_seals <- str_mean_estimate(ybarh, sh, nh, Nh, estimate = "mean")
res_seals$table
```
**Output Note:** Zone `r names(sh)[which.max(sh)]` has the highest standard deviation ($s_h \approx `r round(max(sh), 1)`$), indicating a high degree of variability in seal density in that region.

```{r}
res_seals$estimate |> format_est_gt("mean")
```
**Output Note:** This provides the estimated mean number of seal holes per square kilometer (`r round(res_seals$estimate["Est."], 1)`).

### Neyman allocation of stratum sample size

Because Zone 3 was highly variable, we could have achieved a lower overall variance if we had sampled it more heavily. Neyman allocation formally calculates this optimal distribution.

```{r}
## assuming the cost of surveying each zone is the same
Ch <- rep(1, 3)
Sh <- sh ## estimate population Sh with sample sh
NhShDCh <- Nh * Sh / sqrt(Ch)

## the optimal allocation proportion (Lh) for estimating the population total or mean:
Lh <- NhShDCh / sum(NhShDCh)

data.frame(Ch, Nh, Sh, NhShDCh, Lh) |> 
  gt(rownames_to_stub = TRUE) |> 
  cols_label(NhShDCh = md("$N_h S_h / \\sqrt{C_h}$"), Lh = md("$L_h$ (Prop)"), Ch=md("$C_h$"), Sh=md("$S_h$"), Nh=md("$N_h$")) |>
  fmt_number(columns = c(Sh, NhShDCh, Lh), decimals = 3)
```
**Output Note:** This displays the optimal allocation proportions ($L_h$). Even though Zone 2 is a larger area ($N_h=84$), Zone 3's high variance means we should allocate the largest share (`r round(Lh["3"] * 100, 0)`%) of our sampling effort there.

## Simulation to Study the Efficiency of Stratified Sampling with Different Allocation

We will use the full `agpop.csv` dataset (acting as our true population) to run 2000 repeated samples. We will compare SRS, Proportional (P2S), and Optimal (Neyman) allocations.

### Using "Region" to Stratify

#### Read Population Data

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```

#### Define Stratum Variable

```{r}
agpop$stratum <- agpop$region
## reorder agpop for the ease of using strata of "sampling" (very important)
agpop <- agpop[order(agpop$stratum), ]
```

A stratification variable is most effective when it strongly explains the variance in the target variable. We check this via ANOVA and R-squared.

```{r}
boxplot(acres92 ~ stratum, data = agpop)
anova(lm(agpop$acres92 ~ agpop$stratum))
summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared
```
**Output Note:** The R-squared is `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared, 3)`. The region variable only explains about `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared * 100, 1)`% of the variance in acreage, which is helpful but not extremely strong.

#### Stratified Sampling with P2S allocation (Single Run)

```{r}
# doing one stratified sampling
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

## doing stratified sampling
strsample <- strata(agpop, "stratum", size = nh, method = "srswor")

# collecting data on sampled counties
agstrat <- agpop[strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_mean_estimate_data(agstrat, "acres92", "stratum", "weight", estimate = "mean")$estimate |>
  format_est_gt("mean")
```
**Output Note:** This provides a single point estimate for demonstration.

#### Repeat stratified sampling with P2S allocation 2000 times

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300) ## make sure the order matches Nh

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r}
Nh <- tapply(agpop$acres92, agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

data.frame(Nh, Sh, nh_opt) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = Sh, decimals = 2)
```
**Output Note:** This table compares the population sizes to the optimal sample sizes. Notice that the West region (high variance) gets a disproportionately larger sample under Neyman allocation.

```{r cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_mean_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r}
sim_results_str_region <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_region)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The plot visually demonstrates that Neyman allocation has the tightest spread around the true mean (the red line).

```{r}
sim_means <- sapply(sim_results_str_region, mean)
sim_var <- sapply(sim_results_str_region, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table summarizes the simulation. Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, while Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`% compared to Simple Random Sampling.

### Using "acres82" to Define Strata

What if we stratify by a variable highly correlated with our target, like the acreage from a decade prior (`acres82`)?

#### Read Population Data

```{r}
agpop <- read.csv("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] 
N <- nrow(agpop)
```

#### Define Stratum Variable with Quantiles of "acres82"
We convert the continuous `acres82` variable into four distinct categorical strata based on quantiles.

```{r}
agpop$stratum <- cut(agpop$acres82,
                     breaks = quantile(agpop$acres82, probs = c(0,0.25,0.75,0.95,1)),
                     include.lowest = T)
agpop$stratum <- mapvalues(agpop$stratum, 
                           from = (sort(unique(agpop$stratum))), 
                           to = paste0("acres82", c("Q1", "Q2", "Q3", "Q4")))
agpop <- agpop[order(agpop$stratum), ]
```

Let's check the explanatory power of this new stratification variable.

```{r}
boxplot(acres92 ~ stratum, data = agpop)
summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared
```
**Output Note:** The R-squared is `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared, 3)`. The past acreage explains `r round(summary(lm(agpop$acres92 ~ agpop$stratum))$r.squared * 100, 1)`% of the variance in 1992 acreage, making this a vastly superior stratification variable.

#### Stratified sampling with P2S allocation

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum(Nh)*300)

nres <- 2000
str_p2s_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_p2s_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat stratified sampling with optimal allocation 2000 times

```{r cache=TRUE}
Nh <- tapply(1:nrow(agpop), agpop$stratum, length) 
Sh <- tapply(agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum(Nh*Sh) * 300)

nres <- 2000
str_neyman_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    strsample <- strata(agpop, "stratum", size = nh_opt, method = "srswor")
    agstrat <- agpop[strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    str_neyman_simulated[i,] <- str_mean_estimate_data(agstrat, "acres92", "stratum", "weight", estimate = "mean", show.details = FALSE)$estimate
}
```

#### Repeat simple random sampling 2000 times

```{r cache=TRUE}
nres <- 2000
srs_simulated <- matrix(0, nres, 4)
for (i in 1:nres) {
    srs <- sample(sum(Nh), sum(nh))
    agsrs <- agpop[srs, ]
    srs_simulated[i,] <- srs_mean_est(agsrs[, "acres92"], N = sum(Nh), estimate = "mean", show.details = FALSE)$estimate
}
```

#### Compare the efficiency of different methods

```{r}
sim_results_str_acres82 <- data.frame(
  "SRS" = srs_simulated[,1], 
  "Prop2size" = str_p2s_simulated[,1], 
  "Neyman" = str_neyman_simulated[,1]
)
boxplot(sim_results_str_acres82)
abline(h = mean(agpop$acres92), col = "red")
```
**Output Note:** The boxplot visually highlights a massive reduction in variance compared to SRS, especially for the Neyman allocation.

```{r}
sim_means <- sapply(sim_results_str_acres82, mean)
sim_var <- sapply(sim_results_str_acres82, var)
sim_var_relative <- sim_var / sim_var[1]
sim_var_reduction <- 1 - sim_var_relative

data.frame(
  "Mean" = sim_means, 
  "Variance" = sim_var, 
  "Relative Variance" = sim_var_relative, 
  "Percentage of Variance Reduction" = sim_var_reduction, 
  check.names = FALSE
) |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_number(columns = c(Mean, Variance), decimals = 1) |> 
  fmt_percent(columns = c(`Percentage of Variance Reduction`), decimals = 2) |> 
  fmt_number(columns = c(`Relative Variance`), decimals = 3)
```
**Output Note:** The table shows dramatic results: Proportional allocation reduces variance by `r round(sim_var_reduction["Prop2size"] * 100, 1)`%, and Neyman allocation reduces it by `r round(sim_var_reduction["Neyman"] * 100, 1)`%. Stratifying by a highly correlated variable is immensely powerful.