# Unequal Probability Sampling

## Key Concepts and Formulas

Unequal Probability Sampling (UPS) improves estimation precision by sampling units with probabilities proportional to their size (PPS). If a cluster or county is larger, it has a higher chance of being included in the sample. This approach prevents extreme variance when cluster sizes vary drastically.

**1. UPS With Replacement (UPSWR)**
When sampling with replacement, we draw $n$ units. A single unit $i$ has a selection probability $\psi_i$ on each draw. 

*   **Total Estimator (Hansen-Hurwitz):** $\hat{t}_{pwr} = \frac{1}{n} \sum_{i=1}^n \frac{y_i}{\psi_i}$
*   **Ratio Estimator (Mean):** $\hat{\bar{y}}_{r} = \frac{\sum_{i=1}^n \frac{y_i}{\psi_i}}{\sum_{i=1}^n \frac{M_i}{\psi_i}}$

**2. UPS Without Replacement (UPSWO)**
When sampling without replacement, the inclusion probability for unit $i$ across all draws is $\pi_i$.

*   **Total Estimator (Horvitz-Thompson):** $\hat{t}_{HT} = \sum_{i=1}^n \frac{y_i}{\pi_i}$

---

## Functions and Packages

The estimating functions used throughout this book (including `ratio_est`, `reg_est`, `srs_est`, `upswr_total`, `upswr_ratio`, `cluster_upswr_ratio`, `cluster_upswo_ratio`, and the `format_est_gt` table-formatting helper used below) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions, message=FALSE, warning=FALSE}
#| echo: false
library(dplyr)
library(latex2exp)
source("samplingestimate.r")
```

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

## Analysis of a dataset on student study hours collected by UPSWR

We first manually define a small dataset describing hours spent studying, where the probability of sampling a specific student group ($\psi_i$) was proportional to the number of students in that group ($M_i$).

```{r studyhrs-data}
# 1. Define the raw data
psi <- c(24, 100, 100, 76, 44) / 647
Mi <- c(24, 100, 100, 76, 44)
hrs <- c(75, 203, 203, 191, 168)

# Display the data cleanly
data.frame(Mi, psi, hrs) |> 
  gt() |> 
  cols_label(
    Mi = md("$M_i$"),
    psi = md("$\\psi_i$"),
    hrs = md("**Total Hours ($y_i$)**")
  ) |> 
  fmt_number(columns = psi, decimals = 3)

plot(hrs ~ Mi, main = "Study Hours vs Number of Students", 
     xlab = "Number of Students (Mi)", ylab = "Study Hours (hrs)")
abline(lm(hrs ~ Mi), col = "blue")
```

**Estimate the total and mean study hours:**

```{r studyhrs-upswr-est}
# Estimate the total hours spent by all students
res_hrs_total <- upswr_total(total = hrs, psi = psi, show.details = FALSE)
res_hrs_total$estimate |> format_est_gt("total")

# Estimate the mean hours spent by each student
res_hrs_ratio <- upswr_ratio(total = hrs, M = Mi, psi = psi, estimate = "mean")
res_hrs_ratio$table
res_hrs_ratio$estimate |> format_est_gt("ratio_mean")
```

## Analysis of `statepop.csv` collected with one-stage UPSWR

The file `statepop.csv` contains data from an unequal-probability sample of 100 U.S. counties. Counties were chosen with probabilities proportional to their populations. The total U.S. population is $M_0 = 255,077,536$. Because sampling was done **with replacement**, large counties like Los Angeles occur multiple times in the sample.

```{r statepop-load}
statepop <- read.csv("data/statepop.csv")

# Constants
M0 <- 255077536 
popn <- statepop$popn 
psi <- popn / M0
phys <- statepop$phys

# Display a subset of the raw data in a scrollable box
statepop |> 
  select(state, county, popn, phys) |> 
  mutate(psi = popn / M0) |> 
  gt() |> 
  tab_header(title = "U.S. County Data (Sample)") |> 
  tab_options(container.height = px(300), container.overflow.y = TRUE)

plot(popn, phys, log = "xy", 
     main = "Physicians vs Population (Log Scale)", 
     xlab = "Population", ylab = "Physicians")
```

**Estimate the Total and Per Capita Physicians:**

```{r statepop-upswr-est}
# Estimate the total number of physicians in the USA
res_phys_total <- upswr_total(phys, psi, show.details = FALSE)
res_phys_total$estimate |> format_est_gt("total")

# Estimate the number of physicians per capita (Ratio)
res_phys_ratio <- upswr_ratio(phys, popn, psi, estimate = "mean")
res_phys_ratio$table
res_phys_ratio$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The total estimator heavily leverages the selection probabilities to estimate that there are roughly `r round(res_phys_total$estimate["Est."])` physicians in the US.

### Compare with an estimate based on an SRS sample

Let's compare our UPSWR results against a naive Simple Random Sample of 100 counties.

```{r counties-srs-compare}
# Read SRS sample, treating -99 as NA
counties <- read.csv("data/counties.csv", na.strings = "-99")
totpop <- 255077536
N <- 3141 # Total number of counties

# 1. Naive SRS estimate for total
res_counties_srs <- srs_est(counties$physician, N = N, estimate = "total", show.details = FALSE)
res_counties_srs$estimate |> format_est_gt("total")

# 2. Ratio estimate for total using SRS data
res_counties_ratio <- ratio_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, estimate = "total")
res_counties_ratio$table
res_counties_ratio$estimate |> format_est_gt("total")

# 3. Regression estimate for total using SRS data
res_counties_reg <- reg_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, estimate = "total")
res_counties_reg$table
res_counties_reg$estimate |> format_est_gt("reg_total")
```
**Output Note:** The true value is **532,638**. Notice how poorly the naive SRS estimate performs (predicting over 800,000) because it ignores county sizes. Both the SRS Ratio and SRS Regression methods perform much better by incorporating auxiliary population data.

## Analysis of `studenthrs.csv` collected with two-stage UPSWR

Here, we sample classes (Primary Sampling Units) with replacement based on class size, and then subsample students within them to ask about their study hours.

```{r studenthrs-2stage-upswr}
studenthrs <- read.table("data/studentshrs.txt", header = TRUE)

# Calculate sampling probability (cpsi) based on known total students (647)
studenthrs <- studenthrs |> 
  mutate(cpsi = class_size / 647)

# Run the 2-stage UPSWR function
res_hrs <- cluster_upswr_ratio(
  data = studenthrs,
  sid = "sid",
  csize = "class_size",
  cpsi = "cpsi",
  yvar = "hrs",
  estimate = "mean"
)

# Display the cluster-level working table
res_hrs$table
```

**Estimate the overall mean study hours:**

```{r studenthrs-mean-est}
res_hrs$estimate |> format_est_gt("ratio_mean")
```

## Simulation Studies for UPSWR using `agpop` data

Let's simulate 2000 random samples to see how UPSWR empirically compares to Simple Random Sampling. 

```{r agpop-upswr-sim, message=FALSE, warning=FALSE, cache=TRUE}
# Read population data
agpop <- read.csv("data/agpop.csv", na.strings = "-99")
agpop <- agpop |> filter(!is.na(acres92), !is.na(acres87))

N <- nrow(agpop)
total_acres92 <- sum(agpop$acres92)
total_acres87 <- sum(agpop$acres87)
acres87 <- agpop$acres87

# Storage vectors for simulation
est_srs300 <- rep(0, 2000)
est_srs300_reg_acres87 <- rep(0, 2000)
est_ups300_acres87 <- rep(0, 2000)

for (i in 1:2000) {
  # SRS Simulation
  srs300 <- sample(1:N, size = 300)
  acres92_srs300 <- agpop$acres92[srs300]
  acres87_srs300 <- agpop$acres87[srs300]

  est_srs300[i] <- srs_est(acres92_srs300, N = N, estimate = "total", show.details = FALSE)$estimate[1]
  est_srs300_reg_acres87[i] <- reg_est(acres92_srs300, acres87_srs300,
                                           xbarU = total_acres87/N, N = N, estimate = "total", show.details = FALSE)$estimate[1]

  # UPSWR Simulation (Probability proportional to acres87)
  ups300 <- sample(1:N, size = 300, prob = acres87, replace = TRUE)
  acres92_ups300 <- agpop$acres92[ups300]

  psi <- (acres87 / sum(acres87))[ups300]
  est_ups300_acres87[i] <- upswr_total(acres92_ups300, psi, show.details = FALSE)$estimate[1]
}
```

### Visualizing Estimator Variance

```{r upswr-sim-plot, fig.height=8, fig.width=8}
par(mfrow = c(2,1))

# Compare Naive SRS, Regression SRS, and UPSWR
boxplot(data.frame(est_srs300, est_srs300_reg_acres87, est_ups300_acres87),
        main = "Comparing SRS, SRS-Regression, and UPSWR",
        names = c("SRS Naive", "SRS Regression", "UPSWR"),
        col = c("lightblue", "lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)

# Zoom in on the two high-precision estimators
boxplot(data.frame(est_srs300_reg_acres87, est_ups300_acres87),
        main = "Zoomed: SRS-Regression vs. UPSWR",
        names = c("SRS Regression", "UPSWR"),
        col = c("lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)
```
**Output Note:** The red line represents the true total acreage. The naive SRS estimator has massive variance. Both SRS Regression and UPSWR drastically reduce variance, centering tightly around the true mean.

## Analysis of a dataset on student study hours collected by UPSWO 

When sampling **Without Replacement** with unequal probabilities, we rely on the inclusion probabilities ($\pi_k$) rather than single-draw probabilities ($\psi_i$). 

```{r classpps-upswo}
classpps <- read.csv("data/classpps.csv")

res_upswo <- cluster_upswo_ratio(
  data = classpps,
  cname = "class",
  csize = "clssize",
  cpik = "clssize", # Assuming inclusion prob is proportional to class size
  yvar = "hours",
  estimate = "mean"
)

# Display the cluster-level working table
res_upswo$table
```

**Estimate the overall mean study hours (Horvitz-Thompson ratio style):**

```{r classpps-upswo-est}
res_upswo$estimate |> format_est_gt("ratio_mean")
```