# Unequal Probability Sampling

## Key Concepts and Formulas

Unequal Probability Sampling (UPS) improves estimation precision by sampling units with probabilities proportional to their size (PPS). If a cluster or county is larger, it has a higher chance of being included in the sample. This approach prevents extreme variance when cluster sizes vary drastically.

**1. UPS With Replacement (UPSWR)**
When sampling with replacement, we draw $n$ units. A single unit $i$ has a selection probability $\psi_i$ on each draw. 

*   **Total Estimator (Hansen-Hurwitz):** $\hat{t}_{pwr} = \frac{1}{n} \sum_{i=1}^n \frac{y_i}{\psi_i}$
*   **Ratio Estimator (Mean):** $\hat{\bar{y}}_{r} = \frac{\sum_{i=1}^n \frac{y_i}{\psi_i}}{\sum_{i=1}^n \frac{M_i}{\psi_i}}$

**2. UPS Without Replacement (UPSWO)**
When sampling without replacement, the inclusion probability for unit $i$ across all draws is $\pi_i$.

*   **Total Estimator (Horvitz-Thompson):** $\hat{t}_{HT} = \sum_{i=1}^n \frac{y_i}{\pi_i}$

---

## Functions and Packages

We consolidate our custom estimation functions here, modifying them to return structured lists for cluster summaries where applicable. We also include the `format_est_gt` helper to render all final estimates with mathematical LaTeX headers.

```{r message=FALSE, warning=FALSE}
#| code-fold: true
library(dplyr)
library(gt)
library(latex2exp)

# ---------------------------------------------------------
# Shared gt-table helpers
# ---------------------------------------------------------

## Formats a gt table column as ordinary fixed-point, UNLESS its values are
## too small (< `small` in absolute value) or too large (>= `large`), in
## which case it switches that column to scientific notation instead.
fmt_auto <- function (gt_tbl, data, columns, decimals = 3, small = 1e-3, large = 1e5)
{
    for (col in columns)
    {
        x <- data[[col]]
        x <- x[is.finite (x) & x != 0]
        if (length (x) == 0) next
        if (any (abs (x) < small) || any (abs (x) >= large))
            gt_tbl <- fmt_scientific (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
        else
            gt_tbl <- fmt_number (gt_tbl, columns = tidyselect::all_of (col), decimals = decimals)
    }
    gt_tbl
}

## Formats a single scalar for inline formula display (fixed-point with
## thousands separators, switching to scientific notation for very small or
## very large magnitudes, mirroring fmt_auto()) --- used to show worked-example
## steps such as \bar y = sum/n with the actual numbers plugged in.
fmt_num <- function (x, decimals = 3, small = 1e-3, large = 1e5)
{
    use_sci <- is.finite (x) & x != 0 & (abs (x) < small | abs (x) >= large)
    ifelse (use_sci,
            formatC (x, format = "e", digits = decimals),
            formatC (x, format = "f", digits = decimals, big.mark = ","))
}

## Truncates a row-level working data.frame whose LAST row is a summary row
## to the first head_n data rows + "..." + the summary row.
truncate_working <- function (working, id_col, head_n = 6)
{
    n <- nrow (working) - 1
    if (n <= head_n + 1) return (working)

    dots <- working[1, ]
    dots[1, ] <- NA
    dots[1, id_col] <- "..."

    rbind (working[seq_len (head_n), ], dots, working[nrow (working), ])
}

# ---------------------------------------------------------
# Base Estimators
# ---------------------------------------------------------

## ydata --- observations of the variable of interest
## xdata --- observations of the auxiliary variable
## xbarU --- population mean of the auxiliary variable; required when
##           estimate is "mean" or "total"; ignored when estimate = "model"
## N --- population size
## estimate --- "mean" (default) returns the ratio estimate of the
##              population MEAN of y, Bhat*xbarU; "total" returns the ratio
##              estimate of the population TOTAL of y (requires finite N);
##              "model" returns the raw ratio coefficient Bhat = ybar/xbar
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values, with a Sum row and a
##                  footnote showing Bhat and s_e^2; NULL otherwise
## col_labels --- named list (y, x, yhat) of RAW latex for the working-table
##                columns, so this function can be reused (e.g. by
##                upswr_ratio_est()) with context-appropriate labels
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_ratio_est <- function(ydata, xdata, xbarU = NULL, N = Inf,
                           estimate = c("mean", "total", "model"), show.details = TRUE,
                           col_labels = list(y = "y_i", x = "x_i", yhat = "\\hat y_i")) {
  estimate <- match.arg(estimate)

  n <- length(xdata)
  xbar <- mean(xdata)
  ybar <- mean(ydata)
  B_hat <- ybar / xbar
  yhat <- B_hat * xdata
  e <- ydata - yhat
  var_e <- sum(e^2) / (n - 1)
  sd_B_hat <- sqrt((1 - n/N) * var_e / n) / xbar

  if (estimate == "model") {
      est <- B_hat
      sd_est <- sd_B_hat
  } else {
      if (is.null(xbarU))
          stop("xbarU (population mean of x) is required when estimate is \"mean\" or \"total\"")
      if (estimate == "total") {
          if (!is.finite(N))
              stop("N must be finite to compute estimate = \"total\"")
          est <- B_hat * xbarU * N
          sd_est <- sd_B_hat * xbarU * N
      } else {
          est <- B_hat * xbarU
          sd_est <- sd_B_hat * xbarU
      }
  }

  mem <- qt(0.975, df = n - 1) * sd_est
  estimate_vec <- c(Est. = est, S.E. = sd_est, ci.low = est - mem, ci.upp = est + mem)

  if (!show.details)
      return(list(estimate = estimate_vec, table = NULL))

  working <- data.frame(
      i    = c(seq_len(n), "Sum"),
      y    = c(ydata, sum(ydata)),
      x    = c(xdata, sum(xdata)),
      yhat = c(yhat, sum(yhat)),
      e    = c(e, sum(e)),
      e2   = c(e^2, sum(e^2))
  )
  working <- truncate_working(working, id_col = "i")

  table <- gt(working) |>
      tab_header(title = "Ratio Estimation: Working Table", subtitle = md(sprintf("$n = %d$", n))) |>
      cols_label(
          i    = md("$i$"),
          y    = md(sprintf("$%s$", col_labels$y)),
          x    = md(sprintf("$%s$", col_labels$x)),
          yhat = md(sprintf("$%s$", col_labels$yhat)),
          e    = md("$e_i$"),
          e2   = md("$e_i^2$")
      ) |>
      cols_width(i ~ px(40), everything() ~ px(100)) |>
      fmt_auto(data = working, columns = c("y", "x", "yhat", "e", "e2"), decimals = 3) |>
      sub_missing(missing_text = "...") |>
      tab_style(
          style     = cell_text(weight = "bold"),
          locations = cells_body(rows = i == "Sum")
      ) |>
      tab_footnote(
          footnote  = md(sprintf("$\\hat B = \\sum_i %s / \\sum_i %s = %.4f$, $%s = \\hat B\\, %s$, $s_e^2 = %.4f$.",
                                  col_labels$y, col_labels$x, B_hat, col_labels$yhat, col_labels$x, var_e)),
          locations = cells_column_labels(columns = yhat)
      ) |>
      tab_source_note(
          source_note = md(if (estimate == "model")
              sprintf("**Point estimate:** $\\hat B = \\dfrac{\\sum_i %s}{\\sum_i %s} = \\dfrac{%s}{%s} = %s$",
                      col_labels$y, col_labels$x, fmt_num(sum(ydata)), fmt_num(sum(xdata)), fmt_num(B_hat))
          else if (estimate == "total")
              sprintf("**Point estimate:** $\\hat t = \\hat B\\,\\bar x_U\\,N = %s \\times %s \\times %s = %s$",
                      fmt_num(B_hat), fmt_num(xbarU), fmt_num(N, 0), fmt_num(est))
          else
              sprintf("**Point estimate:** $\\bar y_r = \\hat B\\,\\bar x_U = %s \\times %s = %s$",
                      fmt_num(B_hat), fmt_num(xbarU), fmt_num(est)))
      ) |>
      tab_source_note(
          source_note = md(if (estimate == "model")
              (if (is.finite(N))
                  sprintf("**Standard error:** $\\mathrm{SE}(\\hat B) = \\dfrac{1}{\\bar x}\\sqrt{\\left(1-\\dfrac{n}{N}\\right)\\dfrac{s_e^2}{n}} = \\dfrac{1}{%s}\\sqrt{\\left(1-\\dfrac{%d}{%d}\\right)\\dfrac{%s}{%d}} = %s$",
                          fmt_num(xbar), n, N, fmt_num(var_e), n, fmt_num(sd_B_hat))
              else
                  sprintf("**Standard error:** $\\mathrm{SE}(\\hat B) = \\dfrac{1}{\\bar x}\\sqrt{\\dfrac{s_e^2}{n}} = \\dfrac{1}{%s}\\sqrt{\\dfrac{%s}{%d}} = %s$",
                          fmt_num(xbar), fmt_num(var_e), n, fmt_num(sd_B_hat)))
          else if (estimate == "total")
              sprintf("**Standard error:** $\\mathrm{SE}(\\hat t) = \\mathrm{SE}(\\hat B)\\,\\bar x_U\\,N = %s \\times %s \\times %s = %s$",
                      fmt_num(sd_B_hat), fmt_num(xbarU), fmt_num(N, 0), fmt_num(sd_est))
          else
              sprintf("**Standard error:** $\\mathrm{SE}(\\bar y_r) = \\mathrm{SE}(\\hat B)\\,\\bar x_U = %s \\times %s = %s$",
                      fmt_num(sd_B_hat), fmt_num(xbarU), fmt_num(sd_est)))
      )

  list(estimate = estimate_vec, table = table)
}

## ydata --- observations of the variable of interest
## xdata --- observations of the auxiliary variable
## xbarU --- population mean of the auxiliary variable; required when
##           estimate is "mean" or "total"; ignored when estimate = "model"
## N --- population size
## estimate --- "mean" (default) returns the regression estimate of the
##              population MEAN of y; "total" returns the population TOTAL
##              of y (requires finite N); "model" returns the raw fitted
##              slope Bhat1 itself (with its own SE from the lm fit)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values, with a Sum row and a
##                  footnote showing Bhat0, Bhat1, and s_e^2
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_reg_est <- function(ydata, xdata, xbarU = NULL, N = Inf,
                         estimate = c("mean", "total", "model"), show.details = TRUE) {
  estimate <- match.arg(estimate)

  n <- length(ydata)
  lmfit <- lm(ydata ~ xdata)
  Bhat <- lmfit$coefficients
  yhat <- lmfit$fitted.values
  e <- lmfit$residuals
  SSe <- sum(e^2) / (n - 2)

  if (estimate == "model") {
      est <- unname(Bhat[2])
      sd_est <- summary(lmfit)$coefficients[2, "Std. Error"]
  } else {
      if (is.null(xbarU))
          stop("xbarU (population mean of x) is required when estimate is \"mean\" or \"total\"")
      yhat_reg <- unname(Bhat[1] + Bhat[2] * xbarU)
      se_yhat_reg <- sqrt((1-n/N) * SSe / n)
      if (estimate == "total") {
          if (!is.finite(N)) stop("N must be finite to compute estimate = \"total\"")
          est <- yhat_reg * N
          sd_est <- se_yhat_reg * N
      } else {
          est <- yhat_reg
          sd_est <- se_yhat_reg
      }
  }

  mem <- qt(0.975, df = n - 2) * sd_est
  estimate_vec <- c(Est. = est, S.E. = sd_est, ci.low = est - mem, ci.upp = est + mem)

  if (!show.details)
      return(list(estimate = estimate_vec, table = NULL))

  Sxx <- sum((xdata - mean(xdata))^2)
  Sxy <- sum((xdata - mean(xdata)) * (ydata - mean(ydata)))

  working <- data.frame(
      i    = c(seq_len(n), "Sum"),
      y    = c(ydata, sum(ydata)),
      x    = c(xdata, sum(xdata)),
      yhat = c(yhat, sum(yhat)),
      e    = c(e, sum(e)),
      e2   = c(e^2, sum(e^2))
  )
  working <- truncate_working(working, id_col = "i")

  table <- gt(working) |>
      tab_header(title = "Regression Estimation: Working Table", subtitle = md(sprintf("$n = %d$", n))) |>
      cols_label(
          i = md("$i$"), y = md("$y_i$"), x = md("$x_i$"),
          yhat = md("$\\hat y_i$"), e = md("$e_i$"), e2 = md("$e_i^2$")
      ) |>
      cols_width(i ~ px(40), everything() ~ px(100)) |>
      fmt_auto(data = working, columns = c("y", "x", "yhat", "e", "e2"), decimals = 3) |>
      sub_missing(missing_text = "...") |>
      tab_style(style = cell_text(weight = "bold"), locations = cells_body(rows = i == "Sum")) |>
      tab_footnote(
          footnote  = md(sprintf("$\\hat B_0 = %.4f$, $\\hat B_1 = %.4f$, $\\hat y_i = \\hat B_0 + \\hat B_1 x_i$, $s_e^2 = %.4f$.",
                                  Bhat[1], Bhat[2], SSe)),
          locations = cells_column_labels(columns = yhat)
      ) |>
      tab_source_note(
          source_note = md(if (estimate == "model")
              sprintf("**Point estimate:** $\\hat B_1 = \\dfrac{\\sum_i(x_i-\\bar x)(y_i-\\bar y)}{\\sum_i(x_i-\\bar x)^2} = \\dfrac{%s}{%s} = %s$",
                      fmt_num(Sxy), fmt_num(Sxx), fmt_num(unname(Bhat[2])))
          else if (estimate == "total")
              sprintf("**Point estimate:** $\\hat t_{reg} = N\\bar y_{reg} = %s \\times %s = %s$",
                      fmt_num(N, 0), fmt_num(yhat_reg), fmt_num(est))
          else
              sprintf("**Point estimate:** $\\bar y_{reg} = \\hat B_0+\\hat B_1\\bar x_U = %s + %s \\times %s = %s$",
                      fmt_num(Bhat[1]), fmt_num(Bhat[2]), fmt_num(xbarU), fmt_num(yhat_reg)))
      ) |>
      tab_source_note(
          source_note = md(if (estimate == "model")
              sprintf("**Standard error:** $\\mathrm{SE}(\\hat B_1) = \\sqrt{\\dfrac{s_e^2}{\\sum_i(x_i-\\bar x)^2}} = \\sqrt{\\dfrac{%s}{%s}} = %s$",
                      fmt_num(SSe), fmt_num(Sxx), fmt_num(sd_est))
          else if (estimate == "total")
              sprintf("**Standard error:** $\\mathrm{SE}(\\hat t_{reg}) = N\\times\\mathrm{SE}(\\bar y_{reg}) = %s \\times %s = %s$",
                      fmt_num(N, 0), fmt_num(se_yhat_reg), fmt_num(sd_est))
          else if (is.finite(N))
              sprintf("**Standard error:** $\\mathrm{SE}(\\bar y_{reg}) = \\sqrt{\\left(1-\\dfrac{n}{N}\\right)\\dfrac{s_e^2}{n}} = \\sqrt{\\left(1-\\dfrac{%d}{%d}\\right)\\dfrac{%s}{%d}} = %s$",
                      n, N, fmt_num(SSe), n, fmt_num(se_yhat_reg))
          else
              sprintf("**Standard error:** $\\mathrm{SE}(\\bar y_{reg}) = \\sqrt{\\dfrac{s_e^2}{n}} = \\sqrt{\\dfrac{%s}{%d}} = %s$",
                      fmt_num(SSe), n, fmt_num(se_yhat_reg)))
      )

  list(estimate = estimate_vec, table = table)
}

## sdata --- a vector of original survey data
## N --- population size
## estimate --- "mean" (default) returns the population MEAN estimate;
##              "total" returns the population TOTAL estimate (requires
##              finite N)
## show.details --- if TRUE (default), also return a gt table of the
##                  row-level working values, with a Sum row and a
##                  footnote showing s_y^2; NULL otherwise
## returns list ($estimate = c(Est., S.E., ci.low, ci.upp), $table)
srs_mean_est <- function(sdata, N = Inf, estimate = c("mean", "total"),
                          show.details = TRUE, col_labels = list(y = "y_i")) {
    estimate <- match.arg(estimate)

    n <- length(sdata)
    ybar <- mean(sdata)
    dev <- sdata - ybar
    s2y <- sum(dev^2) / (n - 1)
    se.ybar <- sqrt((1 - n / N)) * sd(sdata) / sqrt(n)

    if (estimate == "total") {
        if (!is.finite(N)) stop("N must be finite to compute estimate = \"total\"")
        est <- N * ybar
        se.est <- N * se.ybar
    } else {
        est <- ybar
        se.est <- se.ybar
    }
    mem <- qt(0.975, df = n - 1) * se.est
    estimate_vec <- c(Est. = est, S.E. = se.est, ci.low = est - mem, ci.upp = est + mem)

    if (!show.details)
        return(list(estimate = estimate_vec, table = NULL))

    working <- data.frame(
        i    = c(seq_len(n), "Sum"),
        y    = c(sdata, sum(sdata)),
        dev  = c(dev, sum(dev)),
        dev2 = c(dev^2, sum(dev^2))
    )
    working <- truncate_working(working, id_col = "i")

    table <- gt(working) |>
        tab_header(title = "SRS Mean Estimation: Working Table", subtitle = md(sprintf("$n = %d$", n))) |>
        cols_label(
            i = md("$i$"), y = md(sprintf("$%s$", col_labels$y)),
            dev = md(sprintf("$%s-\\bar y$", col_labels$y)),
            dev2 = md(sprintf("$(%s-\\bar y)^2$", col_labels$y))
        ) |>
        cols_width(i ~ px(40), everything() ~ px(110)) |>
        fmt_auto(data = working, columns = c("y", "dev", "dev2"), decimals = 3) |>
        sub_missing(missing_text = "...") |>
        tab_style(style = cell_text(weight = "bold"), locations = cells_body(rows = i == "Sum")) |>
        tab_footnote(
            footnote  = md(sprintf("$s_y^2 = \\sum_i (%s-\\bar y)^2/(n-1) = %.4f$.", col_labels$y, s2y)),
            locations = cells_column_labels(columns = dev2)
        ) |>
        tab_source_note(
            source_note = md(if (estimate == "total")
                sprintf("**Point estimate:** $\\hat t = N\\bar y = %s \\times %s = %s$",
                        fmt_num(N, 0), fmt_num(ybar), fmt_num(est))
            else
                sprintf("**Point estimate:** $\\bar y = \\dfrac{\\sum_i %s}{n} = \\dfrac{%s}{%d} = %s$",
                        col_labels$y, fmt_num(sum(sdata)), n, fmt_num(ybar)))
        ) |>
        tab_source_note(
            source_note = md(if (estimate == "total")
                sprintf("**Standard error:** $\\mathrm{SE}(\\hat t) = N \\times \\mathrm{SE}(\\bar y) = %s \\times %s = %s$",
                        fmt_num(N, 0), fmt_num(se.ybar), fmt_num(se.est))
            else if (is.finite(N))
                sprintf("**Standard error:** $\\mathrm{SE}(\\bar y) = \\sqrt{1-\\dfrac{n}{N}}\\,\\dfrac{s}{\\sqrt n} = \\sqrt{1-\\dfrac{%d}{%d}}\\,\\dfrac{%s}{\\sqrt{%d}} = %s$",
                        n, N, fmt_num(sqrt(s2y)), n, fmt_num(se.ybar))
            else
                sprintf("**Standard error:** $\\mathrm{SE}(\\bar y) = \\dfrac{s}{\\sqrt n} = \\dfrac{%s}{\\sqrt{%d}} = %s$",
                        fmt_num(sqrt(s2y)), n, fmt_num(se.ybar)))
        )

    list(estimate = estimate_vec, table = table)
}

# ---------------------------------------------------------
# UPS Estimators
# ---------------------------------------------------------

## Total estimation for datasets collected with UPS with Replacement
## (psi-estimator / Hansen-Hurwitz).
## estimate --- "total" (default) is this function's whole purpose, computed
##              as the SAMPLE MEAN of the expanded pseudo-values t_i/psi_i
##              (a with-replacement HT-type property); "mean" instead
##              divides that total by N (requires finite N)
upswr_total_est <- function(total, psi, N = Inf, estimate = c("total", "mean"), show.details = TRUE) {
  estimate <- match.arg(estimate)
  res <- srs_mean_est(total/psi, N = N, estimate = "mean", show.details = show.details,
                       col_labels = list(y = "t_i/\\psi_i"))
  if (estimate == "mean") {
      if (!is.finite(N)) stop("N must be finite to compute estimate = \"mean\" (mean total per psu)")
      res$estimate <- res$estimate / N
  }
  res
}

## Ratio estimation for datasets collected with UPS with Replacement.
## estimate --- "mean" (default) or "model" both return the raw ratio
##              Bhat = sum(t_i/psi_i)/sum(M_i/psi_i) --- that raw B IS the
##              population MEAN per element; "total" returns the population
##              TOTAL of y, Bhat*Mtotal_U (requires Mtotal_U, the population
##              total of the size variable)
upswr_ratio_est <- function(total, M, psi, N = Inf, estimate = c("mean", "total", "model"),
                             show.details = TRUE, Mtotal_U = NULL) {
  estimate <- match.arg(estimate)

  if (estimate == "total") {
      if (is.null(Mtotal_U))
          stop("Mtotal_U (population total of the size variable) is required when estimate = \"total\"")
      srs_ratio_est(total/psi, M/psi, xbarU = Mtotal_U, N = N, estimate = "mean", show.details = show.details,
                    col_labels = list(y = "t_i/\\psi_i", x = "M_i/\\psi_i", yhat = "\\hat B\\,M_i/\\psi_i"))
  } else {
      srs_ratio_est(total/psi, M/psi, N = N, estimate = "model", show.details = show.details,
                    col_labels = list(y = "t_i/\\psi_i", x = "M_i/\\psi_i", yhat = "\\hat B\\,M_i/\\psi_i"))
  }
}

## Cluster UPS with Replacement: aggregates to the cluster level (ybari, Mi,
## psi per sampled cluster), then delegates to upswr_ratio_est(), so the
## working table and estimate/total logic are shared with the base function.
## data --- data frame holding the sample
## sid --- variable recording cluster (psu) identity
## csize --- variable recording cluster (psu) population size
## cpsi --- variable recording the cluster's selection probability psi_i
## yvar --- variable of interest (element level)
## estimate, Mtotal_U --- see upswr_ratio_est()
cluster_upswr_ratio_est <- function(data, sid, csize, cpsi, yvar, N = Inf,
                                     estimate = c("mean", "total", "model"),
                                     show.details = TRUE, Mtotal_U = NULL) {
    estimate <- match.arg(estimate)

    clust <- data[, sid]
    ydata <- data[, yvar]

    ybari <- tapply(ydata, clust, mean)
    Mi <- tapply(data[,csize], clust, function(x) x[1])
    psi <- tapply(data[,cpsi], clust, function(x) x[1])
    t_hat_cls <- ybari * Mi

    upswr_ratio_est(t_hat_cls, Mi, psi, N = N, estimate = estimate,
                     show.details = show.details, Mtotal_U = Mtotal_U)
}

## Cluster UPS without Replacement: same idea as cluster_upswr_ratio_est(),
## but dividing by the cluster's inclusion probability pik instead of psi.
## data --- data frame holding the sample
## cname --- variable recording cluster identity
## csize --- variable recording cluster population size
## cpik --- variable recording the cluster's inclusion probability pi_k
## yvar --- variable of interest (element level)
cluster_upswo_ratio_est <- function(data, cname, csize, cpik, yvar, N = Inf,
                                     estimate = c("mean", "total", "model"),
                                     show.details = TRUE, Mtotal_U = NULL) {
  estimate <- match.arg(estimate)

  clust <- data[, cname]
  ydata <- data[, yvar]

  ybari <- tapply(ydata, clust, mean)
  Mi <- tapply(data[,csize], clust, function(x) x[1])
  pik <- tapply(data[,cpik], clust, function(x) x[1])
  t_hat_cls <- ybari * Mi

  if (estimate == "total") {
      if (is.null(Mtotal_U))
          stop("Mtotal_U (population total of the size variable) is required when estimate = \"total\"")
      srs_ratio_est(t_hat_cls/pik, Mi/pik, xbarU = Mtotal_U, N = N, estimate = "mean", show.details = show.details,
                    col_labels = list(y = "\\hat t_i/\\pi_i", x = "M_i/\\pi_i", yhat = "\\hat B\\,M_i/\\pi_i"))
  } else {
      srs_ratio_est(t_hat_cls/pik, Mi/pik, N = N, estimate = "model", show.details = show.details,
                    col_labels = list(y = "\\hat t_i/\\pi_i", x = "M_i/\\pi_i", yhat = "\\hat B\\,M_i/\\pi_i"))
  }
}

# ---------------------------------------------------------
# GT Formatting Helper
# ---------------------------------------------------------

format_est_gt <- function(est_data, est_type = c("mean", "total", "ratio", "reg_mean", "reg_total", "ratio_mean", "domain_mean")) {
  est_type <- match.arg(est_type)

  sym_map <- list(
    "mean"        = c(est = "$\\bar{y}$",         se = "$SE(\\bar{y})$"),
    "total"       = c(est = "$\\hat{t}$",         se = "$SE(\\hat{t})$"),
    "ratio"       = c(est = "$\\hat{B}$",         se = "$SE(\\hat{B})$"),
    "reg_mean"    = c(est = "$\\bar{y}_{reg}$",   se = "$SE(\\bar{y}_{reg})$"),
    "reg_total"   = c(est = "$\\hat{t}_{reg}$",   se = "$SE(\\hat{t}_{reg})$"),
    "ratio_mean"  = c(est = "$\\bar{y}_{r}$",     se = "$SE(\\bar{y}_{r})$"),
    "domain_mean" = c(est = "$\\bar{y}_{d}$",     se = "$SE(\\bar{y}_{d})$")
  )

  est_sym <- sym_map[[est_type]]["est"]
  se_sym  <- sym_map[[est_type]]["se"]

  if (is.vector(est_data) && !is.list(est_data)) {
    df <- as.data.frame(t(unname(est_data)))
  } else {
    df <- as.data.frame(est_data)
  }

  colnames(df) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  use_stub <- !is.null(rownames(df)) && !all(rownames(df) == as.character(1:nrow(df)))

  df |>
    gt(rownames_to_stub = use_stub) |>
    cols_label(
      Est.   = md(est_sym),
      S.E.   = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(columns = everything(), decimals = 4) |>
    tab_options(table.width = pct(70))
}
```

## Analysis of a dataset on student study hours collected by UPSWR

We first manually define a small dataset describing hours spent studying, where the probability of sampling a specific student group ($\psi_i$) was proportional to the number of students in that group ($M_i$).

```{r}
# 1. Define the raw data
psi <- c(24, 100, 100, 76, 44) / 647
Mi <- c(24, 100, 100, 76, 44)
hrs <- c(75, 203, 203, 191, 168)

# Display the data cleanly
data.frame(Mi, psi, hrs) |> 
  gt() |> 
  cols_label(
    Mi = md("$M_i$"),
    psi = md("$\\psi_i$"),
    hrs = md("**Total Hours ($y_i$)**")
  ) |> 
  fmt_number(columns = psi, decimals = 3)

plot(hrs ~ Mi, main = "Study Hours vs Number of Students", 
     xlab = "Number of Students (Mi)", ylab = "Study Hours (hrs)")
abline(lm(hrs ~ Mi), col = "blue")
```

**Estimate the total and mean study hours:**

```{r}
# Estimate the total hours spent by all students
res_hrs_total <- upswr_total_est(total = hrs, psi = psi)
res_hrs_total$table
res_hrs_total$estimate |> format_est_gt("total")

# Estimate the mean hours spent by each student
res_hrs_ratio <- upswr_ratio_est(total = hrs, M = Mi, psi = psi, estimate = "mean")
res_hrs_ratio$table
res_hrs_ratio$estimate |> format_est_gt("ratio_mean")
```

## Analysis of `statepop.csv` collected with one-stage UPSWR

The file `statepop.csv` contains data from an unequal-probability sample of 100 U.S. counties. Counties were chosen with probabilities proportional to their populations. The total U.S. population is $M_0 = 255,077,536$. Because sampling was done **with replacement**, large counties like Los Angeles occur multiple times in the sample.

```{r}
statepop <- read.csv("data/statepop.csv")

# Constants
M0 <- 255077536 
popn <- statepop$popn 
psi <- popn / M0
phys <- statepop$phys

# Display a subset of the raw data in a scrollable box
statepop |> 
  select(state, county, popn, phys) |> 
  mutate(psi = popn / M0) |> 
  gt() |> 
  tab_header(title = "U.S. County Data (Sample)") |> 
  tab_options(container.height = px(300), container.overflow.y = TRUE)

plot(popn, phys, log = "xy", 
     main = "Physicians vs Population (Log Scale)", 
     xlab = "Population", ylab = "Physicians")
```

**Estimate the Total and Per Capita Physicians:**

```{r}
# Estimate the total number of physicians in the USA
res_phys_total <- upswr_total_est(phys, psi)
res_phys_total$table
res_phys_total$estimate |> format_est_gt("total")

# Estimate the number of physicians per capita (Ratio)
res_phys_ratio <- upswr_ratio_est(phys, popn, psi, estimate = "mean")
res_phys_ratio$table
res_phys_ratio$estimate |> format_est_gt("ratio_mean")
```
**Output Note:** The total estimator heavily leverages the selection probabilities to estimate that there are roughly `r round(res_phys_total$estimate["Est."])` physicians in the US.

### Compare with an estimate based on an SRS sample

Let's compare our UPSWR results against a naive Simple Random Sample of 100 counties.

```{r}
# Read SRS sample, treating -99 as NA
counties <- read.csv("data/counties.csv", na.strings = "-99")
totpop <- 255077536
N <- 3141 # Total number of counties

# 1. Naive SRS estimate for total
res_counties_srs <- srs_mean_est(counties$physician, N = N, estimate = "total")
res_counties_srs$table
res_counties_srs$estimate |> format_est_gt("total")

# 2. Ratio estimate for total using SRS data
res_counties_ratio <- srs_ratio_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, estimate = "total")
res_counties_ratio$table
res_counties_ratio$estimate |> format_est_gt("total")

# 3. Regression estimate for total using SRS data
res_counties_reg <- srs_reg_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, estimate = "total")
res_counties_reg$table
res_counties_reg$estimate |> format_est_gt("reg_total")
```
**Output Note:** The true value is **532,638**. Notice how poorly the naive SRS estimate performs (predicting over 800,000) because it ignores county sizes. Both the SRS Ratio and SRS Regression methods perform much better by incorporating auxiliary population data.

## Analysis of `studenthrs.csv` collected with two-stage UPSWR

Here, we sample classes (Primary Sampling Units) with replacement based on class size, and then subsample students within them to ask about their study hours.

```{r}
studenthrs <- read.table("data/studentshrs.txt", header = TRUE)

# Calculate sampling probability (cpsi) based on known total students (647)
studenthrs <- studenthrs |> 
  mutate(cpsi = class_size / 647)

# Run the 2-stage UPSWR function
res_hrs <- cluster_upswr_ratio_est(
  data = studenthrs,
  sid = "sid",
  csize = "class_size",
  cpsi = "cpsi",
  yvar = "hrs",
  estimate = "mean"
)

# Display the cluster-level working table
res_hrs$table
```

**Estimate the overall mean study hours:**

```{r}
res_hrs$estimate |> format_est_gt("ratio_mean")
```

## Simulation Studies for UPSWR using `agpop` data

Let's simulate 2000 random samples to see how UPSWR empirically compares to Simple Random Sampling. 

```{r message=FALSE, warning=FALSE, cache=TRUE}
# Read population data
agpop <- read.csv("data/agpop.csv", na.strings = "-99")
agpop <- agpop |> filter(!is.na(acres92), !is.na(acres87))

N <- nrow(agpop)
total_acres92 <- sum(agpop$acres92)
total_acres87 <- sum(agpop$acres87)
acres87 <- agpop$acres87

# Storage vectors for simulation
est_srs300 <- rep(0, 2000)
est_srs300_reg_acres87 <- rep(0, 2000)
est_ups300_acres87 <- rep(0, 2000)

for (i in 1:2000) {
  # SRS Simulation
  srs300 <- sample(1:N, size = 300)
  acres92_srs300 <- agpop$acres92[srs300]
  acres87_srs300 <- agpop$acres87[srs300]

  est_srs300[i] <- srs_mean_est(acres92_srs300, N = N, estimate = "total", show.details = FALSE)$estimate[1]
  est_srs300_reg_acres87[i] <- srs_reg_est(acres92_srs300, acres87_srs300,
                                           xbarU = total_acres87/N, N = N, estimate = "total", show.details = FALSE)$estimate[1]

  # UPSWR Simulation (Probability proportional to acres87)
  ups300 <- sample(1:N, size = 300, prob = acres87, replace = TRUE)
  acres92_ups300 <- agpop$acres92[ups300]

  psi <- (acres87 / sum(acres87))[ups300]
  est_ups300_acres87[i] <- upswr_total_est(acres92_ups300, psi, show.details = FALSE)$estimate[1]
}
```

### Visualizing Estimator Variance

```{r fig.height=8, fig.width=8}
par(mfrow = c(2,1))

# Compare Naive SRS, Regression SRS, and UPSWR
boxplot(data.frame(est_srs300, est_srs300_reg_acres87, est_ups300_acres87),
        main = "Comparing SRS, SRS-Regression, and UPSWR",
        names = c("SRS Naive", "SRS Regression", "UPSWR"),
        col = c("lightblue", "lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)

# Zoom in on the two high-precision estimators
boxplot(data.frame(est_srs300_reg_acres87, est_ups300_acres87),
        main = "Zoomed: SRS-Regression vs. UPSWR",
        names = c("SRS Regression", "UPSWR"),
        col = c("lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)
```
**Output Note:** The red line represents the true total acreage. The naive SRS estimator has massive variance. Both SRS Regression and UPSWR drastically reduce variance, centering tightly around the true mean.

## Analysis of a dataset on student study hours collected by UPSWO 

When sampling **Without Replacement** with unequal probabilities, we rely on the inclusion probabilities ($\pi_k$) rather than single-draw probabilities ($\psi_i$). 

```{r}
classpps <- read.csv("data/classpps.csv")

res_upswo <- cluster_upswo_ratio_est(
  data = classpps,
  cname = "class",
  csize = "clssize",
  cpik = "clssize", # Assuming inclusion prob is proportional to class size
  yvar = "hours",
  estimate = "mean"
)

# Display the cluster-level working table
res_upswo$table
```

**Estimate the overall mean study hours (Horvitz-Thompson ratio style):**

```{r}
res_upswo$estimate |> format_est_gt("ratio_mean")
```