# Unequal Probability Sampling

## Estimation for UPS {#sec-ups-formulas}

Unequal Probability Sampling (UPS) improves estimation precision by sampling units with probabilities proportional to their size (PPS). If a cluster or county is larger, it has a higher chance of being included in the sample. This approach prevents extreme variance when cluster sizes vary drastically.

**Notation.** The population has $N$ units (clusters, counties, ...). Unit $i$ has

* a value $t_i$ of the study variable. For a cluster, $t_i$ is its total; for a county, it is the county's value. The population total is $t = \sum_{i=1}^N t_i$, and the mean per unit is $\bar{t}_U = t/N$;
* a size $M_i$ (e.g. the number of elements in a cluster), with total $M = \sum_{i=1}^N M_i$ and mean size $\bar{M} = M/N$. The mean per element is then $\bar{y}_U = t/M$;
* possibly another auxiliary variable $x_i$ with known total $t_x = \sum_{i=1}^N x_i$ and mean $\bar{x}_U = t_x/N$. The size variable is the special case
  $$
  x_i = M_i, \qquad t_x = M, \qquad \bar{x}_U = \bar{M},
  $$
  and every formula below is written first for a general $x$, then in this form.

We write $t_i$ rather than $y_i$ so that $\bar{t}_U$, the mean per unit, is kept
apart from $\bar{y}_U$, the mean per unit of size that ratio estimators target.

**1. General Formulas for UPS With Replacement (UPSWR)**

We make $n$ independent draws. On each draw, unit $i$ is selected with probability
$\psi_i$, where $\sum_{i=1}^N \psi_i = 1$. A unit may be selected more than once, and it then appears in the sample once per draw.

* **Hansen–Hurwitz estimator of the total.** Each draw gives an unbiased estimate
  $u_i = t_i/\psi_i$ of $t$. The HH estimator is their average, and the SE comes from
  their sample standard deviation $s_u$:
  $$
  \hat{t}_{\mathrm{HH}} = \frac{1}{n}\sum_{i=1}^n \frac{t_i}{\psi_i}, \qquad
  \widehat{\mathrm{SE}}(\hat{t}_{\mathrm{HH}}) = \frac{s_u}{\sqrt{n}}, \qquad
  s_u^2 = \frac{1}{n-1}\sum_{i=1}^n \left(\frac{t_i}{\psi_i} - \hat{t}_{\mathrm{HH}}\right)^2 .
  $$
  Dividing by $N$ gives the HH estimator of the mean per unit, and its SE:
  $$
  \hat{\bar{t}}_{\mathrm{HH}} = \frac{\hat{t}_{\mathrm{HH}}}{N} = \frac{1}{n}\sum_{i=1}^n \frac{t_i}{N\psi_i}, \qquad
  \widehat{\mathrm{SE}}(\hat{\bar{t}}_{\mathrm{HH}}) = \frac{\widehat{\mathrm{SE}}(\hat{t}_{\mathrm{HH}})}{N}.
  $$

* **Hansen–Hurwitz ratio estimator.** Dividing the HH estimator of $t$ by the HH
  estimator of the total of an auxiliary variable $x$ gives
  $$
  \hat{\bar{y}}_{\mathrm{HH},r} = \frac{\hat{t}_{\mathrm{HH}}}{\hat{t}_{x,\mathrm{HH}}}
  = \frac{\sum_{i=1}^n t_i/\psi_i}{\sum_{i=1}^n x_i/\psi_i},
  $$
  which estimates $t/t_x$. Multiplying by $\bar{x}_U$ or $t_x$ gives estimators of $\bar{t}_U$ or $t$.
  Its SE is that of an ordinary ratio estimator, applied to the pairs
  $(t_i/\psi_i,\; x_i/\psi_i)$. Note that $\hat{\bar{y}}_{\mathrm{HH},r}$ itself never uses
  $t_x$ --- only the $\psi_i$; the total $t_x$ enters only when converting the ratio to
  $\bar{t}_U$ or $t$.

**2. The PPS Special Case**

PPS is the general design with the auxiliary variable taken to *be* the size, and the
selection probabilities taken proportional to it:
$$
x_i = M_i, \qquad t_x = M, \qquad \bar{x}_U = \bar{M}, \qquad
\psi_i = \frac{x_i}{t_x} = \frac{M_i}{M} .
$$

* **The ratio estimator collapses to an average.** Every draw now has
  $x_i/\psi_i = t_x$, a constant, so the denominator of the ratio estimator is exactly
  $n\,t_x = nM$ and
  $$
  \hat{\bar{y}}_{\mathrm{HH},r}
  = \frac{\sum_{i=1}^n t_i/\psi_i}{\sum_{i=1}^n x_i/\psi_i}
  = \frac{1}{n}\sum_{i=1}^n \frac{t_i}{x_i}
  = \frac{1}{n}\sum_{i=1}^n \frac{t_i}{M_i}
  = \frac{1}{n}\sum_{i=1}^n \bar{y}_i ,
  $$
  the plain average of the sampled means per element $\bar{y}_i = t_i/M_i$. Its SE
  comes from the sample variance $s_{\bar{y}}^2$ of the $\bar{y}_i$:
  $$
  \widehat{\mathrm{SE}}(\hat{\bar{y}}_{\mathrm{HH},r}) = \frac{s_{\bar{y}}}{\sqrt{n}} ,
  \qquad
  s_{\bar{y}}^2 = \frac{1}{n-1}\sum_{i=1}^n \left(\bar{y}_i - \hat{\bar{y}}_{\mathrm{HH},r}\right)^2 .
  $$

* **Converting to a total or a per-unit mean** uses the general factors $t_x$ and
  $\bar{x}_U$, which here are $M$ and $\bar{M}$:
  $$
  \hat{t}_{\mathrm{HH}} = t_x\,\hat{\bar{y}}_{\mathrm{HH},r} = M\,\hat{\bar{y}}_{\mathrm{HH},r},
  \qquad
  \hat{\bar{t}}_{\mathrm{HH}} = \bar{x}_U\,\hat{\bar{y}}_{\mathrm{HH},r}
  = \bar{M}\,\hat{\bar{y}}_{\mathrm{HH},r} = \frac{M}{N}\,\hat{\bar{y}}_{\mathrm{HH},r} .
  $$

So under PPS the ratio step adds nothing new: the design already uses $M_i$ fully, and
HH is most precise when $t_i/\psi_i$ is nearly constant --- that is, when $t_i$ is
nearly proportional to $M_i$.

**Comparison with the SRS Ratio Estimator of $\bar{t}_U$**

Under a simple random sample of $n$ units drawn without replacement, the usual estimators of $\bar{t}_U$ are
$$
\hat{\bar{t}} = \frac{1}{n}\sum_{i=1}^n t_i, \qquad
\hat{\bar{t}}_r = \hat{B}\,\bar{x}_U, \quad \hat{B} = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n x_i},
$$
with SEs $\sqrt{(1 - n/N)\,s_t^2/n}$ and $(\bar{x}_U/\bar{x})\sqrt{(1 - n/N)\,s_e^2/n}$, where $e_i = t_i - \hat{B}x_i$.
UPSWR and the SRS ratio estimator both exploit a size variable, but in different ways:
UPSWR uses it at the **design** stage, while the ratio estimator uses it at the **estimation** stage.

A direct comparison of the SRS ratio estimator with UPSWR, both using the same size
variable, is given with the app in @sec-ups-app.

**3. UPS Without Replacement (UPSWO)**

When sampling without replacement, the inclusion probability of unit $i$ across all draws is $\pi_i$.

* **Horvitz–Thompson estimator of the total:**
  $$
  \hat{t}_{HT} = \sum_{i=1}^n \frac{t_i}{\pi_i}.
  $$




## Functions and Packages

The estimating functions used throughout this book (including `ratio_est`, `reg_est`, `srs_est`, `upswr_total`, `upswr_ratio`, `cluster_upswr_ratio`, `cluster_upswo_ratio`, and the `format_est_gt` table-formatting helper used below) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions, message=FALSE, warning=FALSE}
#| echo: false
library(dplyr)
library(latex2exp)
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

## Analysis of a dataset on student study hours collected by UPSWR

We first manually define a small dataset describing hours spent studying, where the probability of sampling a specific student group ($\psi_i$) was proportional to the number of students in that group ($M_i$).

```{r tbl-studyhrs-raw-data}
#| tbl-cap: "Raw data: student study hours by group, with selection probabilities"
# 1. Define the raw data
psi <- c(24, 100, 100, 76, 44) / 647
Mi <- c(24, 100, 100, 76, 44)
hrs <- c(75, 203, 203, 191, 168)

# Display the data cleanly
data.frame(Mi, psi, hrs) |> 
  gt() |> 
  cols_label(
    Mi = md("$M_i$"),
    psi = md("$\\psi_i$"),
    hrs = md("**Total Hours ($y_i$)**")
  ) |> 
  fmt_number(columns = psi, decimals = 3)
```

```{r fig-studyhrs-vs-Mi}
#| fig-cap: "Study hours vs. number of students, by group"
plot(hrs ~ Mi, main = "Study Hours vs Number of Students", 
     xlab = "Number of Students (Mi)", ylab = "Study Hours (hrs)")
abline(lm(hrs ~ Mi), col = "blue")
```

**Estimate the total and mean study hours:**

```{r studyhrs-upswr-est}
# Estimate the total hours spent by all students
res_hrs_total <- upswr_total(total = hrs, psi = psi, show.details = TRUE)
res_hrs_total$table
res_hrs_total$estimate |> format_est_gt("hh_total")

# Estimate the mean hours spent by each student (SE and CI scale down too);
# under PPS, t_HH / M is exactly the HH ratio estimator
(res_hrs_total$estimate / 647) |> format_est_gt("hh_ratio_mean")


# Estimate the mean hours spent by each student
res_hrs_ratio <- upswr_ratio(total = hrs, M = Mi, psi = psi, estimate = "mean")
res_hrs_ratio$table
res_hrs_ratio$estimate |> format_est_gt("hh_ratio_mean")
```

## Analysis of `statepop.csv` collected with one-stage UPSWR

The file `statepop.csv` contains data from an unequal-probability sample of 100 U.S. counties. Counties were chosen with probabilities proportional to their populations. The total U.S. population is $M = 255,077,536$. Because sampling was done **with replacement**, large counties like Los Angeles occur multiple times in the sample.

```{r tbl-statepop-sample-data}
#| tbl-cap: "Sample of U.S. county population and physician counts"
statepop <- read.csv("data/statepop.csv")

# Constants
M    <- 255077536
popn <- statepop$popn
psi  <- popn / M
phys <- statepop$phys

# Paged display of the raw data: printing a data frame under `df-print: paged`
# draws one page at a time instead of a long scrolling box
statepop |>
  select(state, county, popn, phys) |>
  mutate(psi = popn / M)
```

```{r fig-statepop-phys-vs-popn}
#| fig-cap: "Physicians vs. population (log-log scale), sample of U.S. counties"
plot(popn, phys, log = "xy", 
     main = "Physicians vs Population (Log Scale)", 
     xlab = "Population", ylab = "Physicians")
```

**Estimating physicians three ways.** One sample answers three different questions,
using the three estimators of @sec-ups-formulas: the population total $t$, the mean per
unit $\bar{t}_U$ (physicians per county), and the mean per element $\bar{y}_U$
(physicians per capita).

*Total physicians in the US,* $\hat{t}_{\mathrm{HH}}$:

```{r statepop-upswr-total}
res_phys_total <- upswr_total(phys, psi, show.details = FALSE)
res_phys_total$estimate |> format_est_gt("hh_total")
```
**Output Note:** Each draw contributes $\mathrm{phys}_i/\psi_i$, an unbiased estimate of
the national total on its own; their average puts about
`r format(round(res_phys_total$estimate["Est."]), big.mark = ",")` physicians in the US,
with a 95% interval of [`r format(round(res_phys_total$estimate["ci.low"]), big.mark = ",")`,
`r format(round(res_phys_total$estimate["ci.upp"]), big.mark = ",")`].

*Physicians per county,* $\hat{\bar{t}}_{\mathrm{HH}} = \hat{t}_{\mathrm{HH}}/N$:

```{r statepop-upswr-per-county}
N <- 3141   # counties in the US
res_phys_county <- upswr_total(phys, psi, N = N, estimate = "mean", show.details = FALSE)
res_phys_county$estimate |> format_est_gt("hh_mean")
```
**Output Note:** This is the same estimate divided by the known $N$, so its standard error
shrinks by the same factor and the two statements carry identical information. It averages
over *counties*, each weighted equally, so the many small rural counties pull it down to
about `r round(res_phys_county$estimate["Est."])` physicians per county.

*Physicians per capita,* $\hat{\bar{y}}_{\mathrm{HH},r}$:

```{r statepop-upswr-per-capita}
res_phys_ratio <- upswr_ratio(phys, popn, psi, estimate = "mean")
res_phys_ratio$table
res_phys_ratio$estimate |> format_est_gt("hh_ratio_mean")
```
**Output Note:** Here the size variable is the auxiliary one ($x_i = \mathrm{popn}_i$) and
the design is PPS on it, so this is exactly the collapse of Section 2: every
$x_i/\psi_i$ equals $M$, and the ratio estimator reduces to the plain average of the
county rates $\mathrm{phys}_i/\mathrm{popn}_i$. The working table shows those pairs. The
estimate is about `r signif(res_phys_ratio$estimate["Est."], 3)` physicians per person, i.e.
roughly `r round(res_phys_ratio$estimate["Est."] * 1e5)` per 100,000 people --- a
population-weighted figure, unlike the per-county average above.

### Compare with an estimate based on an SRS sample

Let's compare our UPSWR results against a naive Simple Random Sample of 100 counties.

```{r counties-srs-compare}
# Read SRS sample, treating -99 as NA
counties <- read.csv("data/counties.csv", na.strings = "-99")
totpop <- 255077536
N <- 3141 # Total number of counties

# 1. Naive SRS estimate for total
res_counties_srs <- srs_est(counties$physician, N = N, estimate = "total", show.details = FALSE)
res_counties_srs$estimate |> format_est_gt("total")

# 2. Ratio estimate for total using SRS data
res_counties_ratio <- ratio_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, estimate = "total")
res_counties_ratio$table
res_counties_ratio$estimate |> format_est_gt("total")

# 3. Regression estimate for total using SRS data
res_counties_reg <- reg_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, estimate = "total")
res_counties_reg$table
res_counties_reg$estimate |> format_est_gt("reg_total")
```
**Output Note:** The true value is **532,638**. Notice how poorly the naive SRS estimate performs (predicting over 800,000) because it ignores county sizes. Both the SRS Ratio and SRS Regression methods perform much better by incorporating auxiliary population data.

## Analysis of `studenthrs.csv` collected with two-stage UPSWR

Here, we sample classes (Primary Sampling Units) with replacement based on class size, and then subsample students within them to ask about their study hours.

```{r studenthrs-2stage-upswr}
studenthrs <- read.table("data/studentshrs.txt", header = TRUE)

# Calculate sampling probability (cpsi) based on known total students (647)
studenthrs <- studenthrs |> 
  mutate(cpsi = class_size / 647)

# Run the 2-stage UPSWR function
res_hrs <- cluster_upswr_ratio(
  data = studenthrs,
  sid = "sid",
  csize = "class_size",
  cpsi = "cpsi",
  yvar = "hrs",
  estimate = "mean"
)

# Display the cluster-level working table
res_hrs$table
```

**Estimate the overall mean study hours:**

```{r studenthrs-mean-est}
res_hrs$estimate |> format_est_gt("hh_ratio_mean")
```

## Simulation Studies for UPSWR using `agpop` data

Let's simulate 2000 random samples to see how UPSWR empirically compares to Simple Random Sampling. 

```{r agpop-upswr-sim, message=FALSE, warning=FALSE, cache=TRUE}
# Read population data
agpop <- acres_in_thousands(read.csv("data/agpop.csv", na.strings = "-99"))
agpop <- agpop |> filter(!is.na(acres92), !is.na(acres87))

N <- nrow(agpop)
total_acres92 <- sum(agpop$acres92)
total_acres87 <- sum(agpop$acres87)
acres87 <- agpop$acres87

# Storage vectors for simulation
est_srs300 <- rep(0, 2000)
est_srs300_reg_acres87 <- rep(0, 2000)
est_ups300_acres87 <- rep(0, 2000)

for (i in 1:2000) {
  # SRS Simulation
  srs300 <- sample(1:N, size = 300)
  acres92_srs300 <- agpop$acres92[srs300]
  acres87_srs300 <- agpop$acres87[srs300]

  est_srs300[i] <- srs_est(acres92_srs300, N = N, estimate = "total", show.details = FALSE)$estimate[1]
  est_srs300_reg_acres87[i] <- reg_est(acres92_srs300, acres87_srs300,
                                           xbarU = total_acres87/N, N = N, estimate = "total", show.details = FALSE)$estimate[1]

  # UPSWR Simulation (Probability proportional to acres87)
  ups300 <- sample(1:N, size = 300, prob = acres87, replace = TRUE)
  acres92_ups300 <- agpop$acres92[ups300]

  psi <- (acres87 / sum(acres87))[ups300]
  est_ups300_acres87[i] <- upswr_total(acres92_ups300, psi, show.details = FALSE)$estimate[1]
}
```

### Visualizing Estimator Variance

```{r fig-upswr-sim-comparison, fig.height=8, fig.width=8}
#| fig-cap: "Estimator variance: naive SRS vs. SRS regression vs. UPSWR (2000 simulations), full and zoomed views"
par(mfrow = c(2,1))

# Compare Naive SRS, Regression SRS, and UPSWR
boxplot(data.frame(est_srs300, est_srs300_reg_acres87, est_ups300_acres87),
        main = "Comparing SRS, SRS-Regression, and UPSWR",
        names = c("SRS Naive", "SRS Regression", "UPSWR"),
        col = c("lightblue", "lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)

# Zoom in on the two high-precision estimators
boxplot(data.frame(est_srs300_reg_acres87, est_ups300_acres87),
        main = "Zoomed: SRS-Regression vs. UPSWR",
        names = c("SRS Regression", "UPSWR"),
        col = c("lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)
```
**Output Note:** The red line represents the true total acreage. The naive SRS estimator has massive variance. Both SRS Regression and UPSWR drastically reduce variance, centering tightly around the true mean.


## Shinylive App Comparing SRS and UPSWR {#sec-ups-app}

The app compares two ways of using a size variable $x_i$ to estimate the mean per
unit $\bar{t}_U = t/N$ (here $t_i$ is `acres92` for county $i$): at the **estimation**
stage (SRS + ratio) or at the **design** stage (UPSWR with $\psi_i = x_i/t_x$). Choosing the
same variable for $M_i$ and $x_i$ in the app gives exactly this comparison. The plain SRS
average is included as a baseline. Throughout, write
$$
B = \frac{t}{t_x}, \qquad r_i = \frac{t_i}{x_i}, \qquad e_i = t_i - B x_i ,
$$
so $e_i$ is unit $i$'s residual from the line through the origin with the population slope.
Note that $\sum_{i=1}^N e_i = t - B t_x = 0$.

**SRS + ratio.** Use $x_i$ only at the estimation stage:
$$
\hat{\bar{t}}_{r} = \hat{B}\,\bar{x}_U, \quad \hat{B} = \frac{\sum_{i=1}^n t_i}{\sum_{i=1}^n x_i}, \qquad
\widehat{\mathrm{SE}}(\hat{\bar{t}}_{r}) = \frac{\bar{x}_U}{\bar{x}}\sqrt{\left(1-\frac{n}{N}\right)\frac{s_e^2}{n}}, \qquad
s_e^2 = \frac{1}{n-1}\sum_{i=1}^n \left(t_i - \hat{B}x_i\right)^2 .
$$

**UPSWR + ratio.** Use $x_i$ at the design stage, with $\psi_i = x_i/t_x$. Every draw
then has $x_i/\psi_i = t_x$. So the HH ratio estimator, scaled by $\bar{x}_U$, is exactly
the HH estimator of the mean, $\hat{t}_{\mathrm{HH}}/N$, and it reduces to $\bar{x}_U$
times the average of the sampled ratios $r_i$:
$$
\hat{\bar{t}}_{\mathrm{HH}} = \bar{x}_U\,\hat{\bar{y}}_{\mathrm{HH},r}
= \bar{x}_U\,\frac{\sum_{i=1}^n t_i/\psi_i}{\sum_{i=1}^n x_i/\psi_i}
= \frac{\hat{t}_{\mathrm{HH}}}{N}
= \bar{x}_U\,\frac{1}{n}\sum_{i=1}^n r_i ,
$$
$$
\widehat{\mathrm{SE}}(\hat{\bar{t}}_{\mathrm{HH}}) = \bar{x}_U\,\frac{s_r}{\sqrt{n}}, \qquad
s_r^2 = \frac{1}{n-1}\sum_{i=1}^n \left(r_i - \frac{1}{n}\sum_{j=1}^n r_j\right)^2 .
$$


::: {#fig-ups-app}

```{shinylive-r}
#| label: upswr-app
#| standalone: true
#| viewerHeight: 860

library(shiny)

## ---------------------------------------------------------------- population
data_url <- "https://raw.githubusercontent.com/longhaiSK/sampling/main/data/agpop.csv"

raw <- tryCatch({
  download.file(data_url, "agpop.csv")
  read.csv("agpop.csv")
}, error = function(e) NULL)

if (is.null(raw)) {            # fallback so the app runs without network access
  set.seed(1)
  Nf <- 3000
  x1 <- rgamma(Nf, shape = 1.2, scale = 1.8e5)
  raw <- data.frame(
    acres92 = pmax(0, 0.93 * x1 + rnorm(Nf, 0, 0.5 * x1^0.85)),
    acres87 = x1,
    farms92 = pmax(1, round(300 + x1 / 6000 + rnorm(Nf, 0, 250)))
  )
}

raw <- raw[is.finite(raw$acres92) & raw$acres92 != -99, ]
yv  <- raw$acres92
N   <- length(yv)

## every numeric column other than y is a candidate auxiliary variable, both
## for the UPS sizes M and for the ratio estimator's x;
## -99 marks a missing entry and is replaced by the mean of the rest
num <- vapply(raw, is.numeric, logical(1))
X <- raw[, num & names(raw) != "acres92", drop = FALSE]
X <- as.data.frame(lapply(X, function(v) {
  v[!is.finite(v) | v == -99] <- NA
  v[is.na(v)] <- mean(v, na.rm = TRUE)
  v
}))
keep <- vapply(X, function(v) all(is.finite(v)) && all(v >= 0) && var(v) > 0,
               logical(1))
X <- X[, keep, drop = FALSE]
xvars <- names(X)
xdef  <- if ("acres87" %in% xvars) "acres87" else xvars[1]
mdef  <- xdef

ybarU <- mean(yv)
Vy    <- var(yv)                     # S^2 of y, for the SRS average
ZCRIT <- 1.96                        # nominal 95% intervals

set.seed(11)                         # display-only vertical jitter, fixed once
ypl <- yv + rnorm(N, 0, 0.004 * diff(range(yv)))
rgy <- range(ypl)

## ------------------------------------------------------------ estimator specs
EST <- c("srs", "ratio", "hh")
LAB <- c(srs = "SRS average", ratio = "SRS ratio", hh = "UPSWR HH")
COL <- c(srs = "#e07b39", ratio = "#2ca02c", hh = "#9467bd")
PCH <- c(srs = 17, ratio = 16, hh = 15)

col_samp <- "#d62728"                # missed intervals and the population mean
col_pts  <- "blue"                   # sampled observations
col_true <- "black"
col_pop  <- "#9aa0a655"
col_band <- "#eef3f8"

## point size grows with the per-draw selection probability; the scale is
## shared by both designs, so SRS points (all 1/N) show what "equal" looks like
psize <- function(p, pmax) 0.25 + 2.4 * sqrt(p / pmax)

## ------------------------------------------------------- broken-axis mapping
## "top": linear from the minimum up to a cut, then the sparse upper tail
##        squeezed into a thin strip (SRS view: every unit is equally likely).
## "low": the many small values squeezed into a thin strip at the bottom and
##        the large values spread out (UPSWR view: large units dominate).
BRK_THIN <- 0.14
BRK_GAP  <- 0.025
BRK_WIDE <- 1 - BRK_THIN - BRK_GAP

## SRS view: the population mean falls at mid-height
cut_at_centre <- function(rmin, rmax, m) {
  cut <- rmin + (m - rmin) * BRK_WIDE / 0.5
  if (!is.finite(cut) || cut >= rmax) rmin + 0.95 * (rmax - rmin) else cut
}

## UPSWR view: squeeze the values that PPS reaches with only `share` of its
## draws, i.e. the lower `share` of the psi-weighted distribution
cut_psi_share <- function(v, psi, share = 0.2) {
  o <- order(v)
  v[o][which(cumsum(psi[o]) >= share)[1]]
}

mk_break <- function(rmin, rmax, cut, mode = c("top", "low")) {
  mode <- match.arg(mode)
  eps  <- 1e-9 * (rmax - rmin)
  cut  <- min(max(cut, rmin + eps), rmax - eps)
  lo_h <- if (mode == "top") BRK_WIDE else BRK_THIN
  hi_h <- 1 - lo_h - BRK_GAP
  tf <- function(v) {
    v <- pmin(pmax(v, rmin), rmax)
    ifelse(v <= cut, (v - rmin) / (cut - rmin) * lo_h,
           lo_h + BRK_GAP + (v - cut) / (rmax - cut) * hi_h)
  }
  list(tf = tf, rmin = rmin, rmax = rmax, cut = cut, mode = mode,
       gap = c(lo_h, lo_h + BRK_GAP))
}

mk_plain <- function(rmin, rmax) {
  list(tf = function(v) (pmin(pmax(v, rmin), rmax) - rmin) / (rmax - rmin),
       rmin = rmin, rmax = rmax, cut = NA, gap = NULL)
}

## counts stay as counts, acreages are shown in thousands
unit_div <- function(rmax) if (rmax >= 1e5) 1000 else 1
unit_lab <- function(div) if (div == 1000) " (thousands)" else ""

axis_at <- function(bk, side, div, rg, nt) {
  v <- pretty(rg, nt)
  v <- v[v >= rg[1] & v <= rg[2]]
  if (!length(v)) return(invisible())
  d <- if (diff(rg) / div < 40) 1 else 0
  axis(side, at = bk$tf(v), labels = formatC(v / div, format = "f", digits = d,
                                             big.mark = ","), las = 1, cex.axis = 0.75)
}

draw_axis <- function(bk, side, div) {
  if (is.null(bk$gap)) {
    axis_at(bk, side, div, c(bk$rmin, bk$rmax), 5)
  } else {
    nt <- if (bk$mode == "top") c(5, 2) else c(2, 5)
    axis_at(bk, side, div, c(bk$rmin, bk$cut), nt[1])
    axis_at(bk, side, div, c(bk$cut + 1e-9, bk$rmax), nt[2])
  }
}

mask_gap <- function(bk, side) {
  if (is.null(bk$gap)) return(invisible())
  usr <- par("usr")
  g <- bk$gap
  m <- mean(g)
  if (side == 2) {
    rect(usr[1], g[1], usr[2], g[2], col = "white", border = NA)
    d <- 0.010 * (usr[2] - usr[1])
    segments(usr[1] - d, c(m - 0.012, m + 0.004), usr[1] + d,
             c(m - 0.004, m + 0.012), xpd = NA, lwd = 1.2)
  } else {
    rect(g[1], usr[3], g[2], usr[4], col = "white", border = NA)
    d <- 0.010 * (usr[4] - usr[3])
    segments(c(m - 0.012, m + 0.004), usr[3] - d, c(m - 0.004, m + 0.012),
             usr[3] + d, xpd = NA, lwd = 1.2)
  }
}

## interval drawn as a capped vertical bar with the estimate marked on it;
## the bar turns red when the interval misses the population mean
ci_bar <- function(x, est, ci, col, pch, covered, tf = identity, cap = 0.012) {
  bcol <- if (covered) col else col_samp
  segments(x, tf(ci[1]), x, tf(ci[2]), col = bcol, lwd = 2)
  segments(x - cap, tf(ci), x + cap, tf(ci), col = bcol, lwd = 2)
  points(x, tf(est), pch = pch, col = col, cex = 1.7)
}

## symmetric density polygon, drawn vertically about x = at
violin <- function(x, at, col, w = 0.32) {
  if (length(x) < 10 || diff(range(x)) <= 0) return(invisible())
  d  <- density(x)
  hw <- w * d$y / max(d$y)
  polygon(c(at + hw, rev(at - hw)), c(d$x, rev(d$x)),
          col = adjustcolor(col, 0.35), border = col)
  m <- median(x)
  segments(at - w * 0.6, m, at + w * 0.6, m, col = col, lwd = 2)
}

pct <- function(v) if (is.na(v)) "--" else sprintf("%.1f%%", 100 * v)

## ---- inline SVG button icons (no font or icon-library dependency) ----------
svg_icon <- function(paths)
  HTML(sprintf(paste0('<svg width="16" height="16" viewBox="0 0 16 16" ',
                      'fill="currentColor" style="vertical-align:-2px">%s</svg>'), paths))
icon_play  <- svg_icon('<path d="M3 2v12l10-6z"/>')
icon_pause <- svg_icon('<rect x="3" y="2" width="4" height="12"/><rect x="9" y="2" width="4" height="12"/>')
icon_skip  <- svg_icon('<path d="M1 2v12l7-6zM8 2v12l6-6z"/><rect x="14" y="2" width="2" height="12"/>')

## ---------------------------------------------------------------------- ui
ui <- fluidPage(
  titlePanel("Shinylive App for UPSWR"),
  tags$style(HTML(
    ".well {padding: 10px 12px; margin-bottom: 10px;}
     .form-group {margin-bottom: 6px;}
     .checkbox, .radio {margin-top: 2px; margin-bottom: 2px;}
     .irs {margin-bottom: 0;}
     .btn {margin-bottom: 6px;}"
  )),
  wellPanel(
    fluidRow(
      column(
        3,
        selectInput("mvar", HTML("Aux. var. for UPS (M<sub>i</sub>)"),
                    choices = xvars, selected = mdef),
        selectInput("xvar", HTML("Aux. var. for estimation (x<sub>i</sub>)"),
                    choices = xvars, selected = xdef),
        checkboxInput("sync", HTML("Same variable (M<sub>i</sub> = x<sub>i</sub>)"), TRUE),
        checkboxInput("brk", "Break the axes", TRUE)
      ),
      column(
        3,
        checkboxGroupInput("show", "Estimators",
                           choices = c("SRS + average" = "srs",
                                       "SRS + ratio" = "ratio",
                                       "UPSWR + HH ratio" = "hh"),
                           selected = EST),
        sliderInput("n", "Sample size (n)", 10, 500, 100, step = 10)
      ),
      column(
        3,
        radioButtons("design", "Show sample from",
                     choices = c("SRS" = "srs", "UPSWR" = "ups"),
                     selected = "ups", inline = TRUE),
        sliderInput("speed", "Samples per second", min = 1, max = 10, value = 5, step = 1),
        numericInput("nrep", "Stop after this many samples",
                     value = 500, min = 10, max = 20000, step = 50)
      ),
      column(
        3,
        actionButton("draw", "Draw new samples", class = "btn-primary", width = "100%"),
        br(), br(),
        fluidRow(
          column(4, actionButton("play",  icon_play,  class = "btn-success", width = "100%",
                                 title = "Play")),
          column(4, actionButton("pause", icon_pause, class = "btn-warning", width = "100%",
                                 title = "Pause")),
          column(4, actionButton("skip",  icon_skip,  class = "btn-info",    width = "100%",
                                 title = "Skip to the end"))
        ),
        actionButton("reset", "Reset", width = "100%"),
        div(style = "font-size: 90%;", textOutput("counter"))
      )
    )
  ),
  plotOutput("mainplot", height = "520px")
)

## ------------------------------------------------------------------- server
server <- function(input, output, session) {

  rv <- reactiveValues(s_srs = NULL, s_ups = NULL, n = 100, ests = NULL)
  running <- reactiveVal(FALSE)

  nrep <- reactive({
    v <- input$nrep
    if (is.null(v) || is.na(v) || v < 1) 1L else min(as.integer(v), 20000L)
  })

  ## everything that depends on the two auxiliary variables: M sets the UPS
  ## selection probabilities, x feeds the SRS ratio estimator
  aux <- reactive({
    mv  <- X[[input$mvar]]
    xv  <- X[[input$xvar]]
    psi <- mv / sum(mv)                  # per-draw selection probabilities
    B   <- sum(yv) / sum(xv)
    pos <- psi > 0
    list(m = mv, x = xv, psi = psi, B = B, xbar = mean(xv),
         ## per-draw variances, so the SE of each estimator with n draws is
         ## sqrt(fpc * V / n) for the SRS ones and sqrt(V / n) for HH
         V = c(srs = Vy, ratio = var(yv - B * xv),
               hh = sum(psi[pos] * (yv[pos] / psi[pos] - sum(yv))^2) / N^2),
         ## horizontal axis of each view: x for SRS, M for UPSWR
         ax = list(
           top = list(v = xv, rg = range(xv), slope = B, name = input$xvar,
                      cut = cut_at_centre(min(xv), max(xv), mean(xv)),
                      cuty = cut_at_centre(rgy[1], rgy[2], ybarU)),
           low = list(v = mv, rg = range(mv), slope = sum(yv) / sum(mv),
                      name = input$mvar, cut = cut_psi_share(mv, psi),
                      cuty = cut_psi_share(ypl, psi))))
  })

  se_theory <- function(a, n)
    sqrt(c(srs = 1 - n / N, ratio = 1 - n / N, hh = 1) * a$V / n)

  ## one replication: an SRS (without replacement) feeding the average and the
  ## ratio estimator, and an independent PPS-with-replacement sample feeding HH
  one_draw <- function(a, n) {
    s   <- sample.int(N, n)
    u   <- sample.int(N, n, replace = TRUE, prob = a$psi)
    ys  <- yv[s]
    xs  <- a$x[s]
    fpc <- 1 - n / N

    Bhat <- if (sum(xs) > 0) sum(ys) / sum(xs) else 0
    z    <- yv[u] / a$psi[u] / N          # each draw's estimate of the mean

    est <- c(srs   = mean(ys),
             ratio = Bhat * a$xbar,
             hh    = mean(z))
    se  <- c(srs   = sqrt(fpc * var(ys) / n),
             ratio = (a$xbar / mean(xs)) * sqrt(fpc * var(ys - Bhat * xs) / n),
             hh    = sd(z) / sqrt(n))

    e <- c(setNames(est, paste0("est_", EST)),
           setNames(est - ZCRIT * se, paste0("lo_", EST)),
           setNames(est + ZCRIT * se, paste0("hi_", EST)),
           setNames(as.numeric(abs(est - ybarU) <= ZCRIT * se), paste0("cov_", EST)),
           Bhat = Bhat, Bhh = mean(yv[u] / a$m[u]))
    list(s = s, u = u, e = e)
  }

  ## draw k replications; only the last one's samples are kept for the left
  ## panel. "Skip" calls this with the full remaining count.
  draw_many <- function(k) {
    a <- isolate(aux())
    n <- isolate(input$n)
    E <- vector("list", k)
    for (i in seq_len(k)) {
      d <- one_draw(a, n)
      E[[i]] <- d$e
    }
    rv$s_srs <- d$s
    rv$s_ups <- d$u
    rv$n     <- n
    rv$ests  <- rbind(rv$ests, do.call(rbind, E))
  }
  draw_one <- function() draw_many(1)

  ## a fresh history always starts from one sample, so the plot is never empty
  reset_history <- function() {
    running(FALSE)
    rv$ests <- NULL
    draw_one()
  }
  observeEvent(list(input$n, input$mvar, input$xvar), reset_history(), ignoreInit = FALSE)

  ## keep M and x the same while "Same variable" is ticked: a change to either
  ## box is copied to the other, and ticking the box copies M to x
  sync_to <- function(id, v) {
    if (isTRUE(input$sync) && !identical(input[[id]], v))
      updateSelectInput(session, id, selected = v)
  }
  observeEvent(input$mvar, sync_to("xvar", input$mvar))
  observeEvent(input$xvar, sync_to("mvar", input$xvar))
  observeEvent(input$sync, sync_to("xvar", input$mvar))
  observeEvent(input$reset, reset_history())

  ## one sample per click; Play repeats until Pause or until the cap is reached
  observeEvent(input$draw,  { running(FALSE); draw_one() })
  observeEvent(input$play,  running(TRUE))
  observeEvent(input$pause, running(FALSE))
  observeEvent(input$skip, {
    running(FALSE)
    k <- nrep() - NROW(rv$ests)
    if (k > 0) draw_many(k)
  })

  observe({
    if (!isTRUE(running())) return()
    if (NROW(isolate(rv$ests)) >= isolate(nrep())) { running(FALSE); return() }
    invalidateLater(1000 / isolate(input$speed), session)
    isolate(draw_one())
  })

  output$counter <- renderText(
    sprintf("%d of %d samples", NROW(rv$ests), nrep())
  )

  output$mainplot <- renderPlot({
    req(rv$s_srs)
    sel <- EST[EST %in% input$show]
    validate(need(length(sel) > 0, "Select at least one estimator."))

    a    <- aux()
    n    <- rv$n
    M    <- rv$ests
    nrep <- NROW(M)
    cur  <- M[nrep, ]
    ups  <- identical(input$design, "ups")
    s    <- if (ups) rv$s_ups else rv$s_srs
    xpos <- setNames(0.02 + 0.04 * (seq_along(sel) - 1), sel)   # CI bar positions

    ## violin window, wide enough for the least precise estimator shown
    half <- 4 * max(se_theory(a, n)[sel])
    lo <- ybarU - half
    hi <- ybarU + half

    md <- if (ups) "low" else "top"
    h  <- a$ax[[md]]
    if (isTRUE(input$brk)) {
      bky <- mk_break(rgy[1], rgy[2], h$cuty, md)
      bkx <- mk_break(h$rg[1], h$rg[2], h$cut, md)
    } else {
      bky <- mk_plain(rgy[1], rgy[2])
      bkx <- mk_plain(h$rg[1], h$rg[2])
    }
    tfy  <- bky$tf
    tfx  <- bkx$tf
    divy <- unit_div(rgy[2])
    divx <- unit_div(h$rg[2])

    xg <- seq(h$rg[1], h$rg[2], length.out = 1500)
    draw_line <- function(yl, ...) {          # stop the line at the frame
      yl[yl < rgy[1] | yl > rgy[2]] <- NA
      lines(tfx(xg), tfy(yl), ...)
    }

    ## point sizes: per-draw selection probability under the design shown
    pmax_ <- max(a$psi)
    cexp  <- if (ups) psize(a$psi, pmax_) else rep(psize(1 / N, pmax_), N)

    layout(matrix(1:2, nrow = 1), widths = c(1, 1))

    ## ---- left: population sized by selection probability, with the sample ----
    par(mar = c(4.2, 4.8, 3.0, 1.0))
    plot(NA, xlim = c(0, 1), ylim = c(0, 1), axes = FALSE,
         xlab = paste0(if (ups) "M: " else "x: ", h$name, unit_lab(divx)),
         ylab = paste0("t: acres92", unit_lab(divy)),
         main = sprintf("Population and %s sample, n = %d",
                        if (ups) "UPSWR" else "SRS", n))
    rect(0, tfy(lo), 1, tfy(hi), col = col_band, border = NA)
    points(tfx(h$v), tfy(ypl), pch = 16, cex = cexp, col = col_pop)

    draw_line(h$slope * xg, col = col_true, lwd = 1.8)
    if (ups) {
      if ("hh" %in% sel)
        draw_line(cur[["Bhh"]] * xg, col = COL["hh"], lwd = 2, lty = 2)
    } else {
      if ("ratio" %in% sel)
        draw_line(cur[["Bhat"]] * xg, col = COL["ratio"], lwd = 2, lty = 2)
      if ("srs" %in% sel)
        abline(h = tfy(cur[["est_srs"]]), col = COL["srs"], lwd = 2, lty = 2)
    }
    ## a unit drawn more than once is drawn once, ringed by its draw count
    tab <- table(s)
    su  <- as.integer(names(tab))
    points(tfx(h$v[su]), tfy(ypl[su]), pch = 16, cex = cexp[su], col = col_pts)
    rep_u <- su[tab > 1]
    if (length(rep_u))
      text(tfx(h$v[rep_u]), tfy(ypl[rep_u]), labels = tab[tab > 1],
           pos = 4, cex = 0.8, col = col_pts, font = 2)

    abline(h = tfy(ybarU), col = col_samp, lty = 3)
    for (k in sel) ci_bar(xpos[[k]], cur[[paste0("est_", k)]],
                          c(cur[[paste0("lo_", k)]], cur[[paste0("hi_", k)]]),
                          COL[[k]], PCH[[k]], cur[[paste0("cov_", k)]] == 1, tfy)

    mask_gap(bky, 2); mask_gap(bkx, 1)
    box(); draw_axis(bky, 2, divy); draw_axis(bkx, 1, divx)

    leg <- data.frame(l = c(if (ups) "size ~ psi = M / M0" else "size ~ 1/N (equal)",
                            "sampled t"),
                      c = c("grey55", col_pts), lt = NA, p = 16, w = NA)
    leg <- rbind(leg, data.frame(l = if (ups) "true t = (t/M0) M" else "true t = (t/tx) x",
                                 c = col_true, lt = 1, p = NA, w = 1.8))
    if (ups && "hh" %in% sel) leg <- rbind(leg, data.frame(
      l = "HH slope mean(t/M)", c = COL[["hh"]], lt = 2, p = NA, w = 2))
    if (!ups && "ratio" %in% sel) leg <- rbind(leg, data.frame(
      l = "ratio fit", c = COL[["ratio"]], lt = 2, p = NA, w = 2))
    if (!ups && "srs" %in% sel) leg <- rbind(leg, data.frame(
      l = "sample average", c = COL[["srs"]], lt = 2, p = NA, w = 2))
    leg <- rbind(leg, data.frame(l = "population mean", c = col_samp, lt = 3, p = NA, w = 1))
    legend("topleft", bty = "n", cex = 0.9, inset = c(0.15, 0.01),
           legend = leg$l, col = leg$c, lty = leg$lt, pch = leg$p, lwd = leg$w)

    ## ---- right: sampling distributions as vertical violins, with current CIs ----
    par(mar = c(4.2, 4.8, 3.0, 1.0))
    k_n <- length(sel)
    at  <- setNames(seq_len(k_n), sel)
    plot(NA, xlim = c(0.4, k_n + 0.6), ylim = c(lo, hi), axes = FALSE,
         xlab = "", ylab = paste0("estimate of mean t (acres92)", unit_lab(divy)),
         main = "Sampling distributions of the estimates")
    usr <- par("usr")
    rect(usr[1], usr[3], usr[2], usr[4], col = col_band, border = NA)
    abline(h = ybarU, col = col_samp, lty = 3)
    for (k in sel) {
      violin(M[, paste0("est_", k)], at[[k]], COL[[k]])
      ci <- c(cur[[paste0("lo_", k)]], cur[[paste0("hi_", k)]])
      bc <- if (cur[[paste0("cov_", k)]] == 1) COL[[k]] else col_samp
      segments(at[[k]], ci[1], at[[k]], ci[2], col = bc, lwd = 2.5)
      segments(at[[k]] - 0.08, ci, at[[k]] + 0.08, ci, col = bc, lwd = 2.5)
      points(at[[k]], cur[[paste0("est_", k)]], pch = PCH[[k]], col = COL[[k]], cex = 1.7)
    }
    box()
    vy <- pretty(c(lo, hi), 5)
    vy <- vy[vy >= lo & vy <= hi]
    axis(2, at = vy, labels = formatC(vy / divy, format = "f",
                                      digits = if ((hi - lo) / divy < 40) 1 else 0,
                                      big.mark = ","), las = 1, cex.axis = 0.75)
    axis(1, at = at, labels = LAB[sel], tick = FALSE, cex.axis = 0.9)

    cov_txt <- function(k) if (nrep >= 10) pct(mean(M[, paste0("cov_", k)])) else "--"
    legend("topleft", bty = "n", cex = 0.9, inset = c(0.02, 0.01),
           title = "95% CI coverage", fill = COL[sel], border = NA,
           legend = paste(LAB[sel], vapply(sel, cov_txt, "")))

    ## reductions are relative to the SRS average, which is always simulated
    cmp <- setdiff(sel, "srs")
    if (length(cmp)) {
      se_th <- se_theory(a, n)
      red <- lapply(cmp, function(k) {
        mc <- if (nrep > 5) 1 - var(M[, paste0("est_", k)]) / var(M[, "est_srs"]) else NA
        th <- 1 - se_th[[k]]^2 / se_th[["srs"]]^2
        list(txt = sprintf("%s %s (theory %s)", LAB[[k]], pct(mc), pct(th)),
             neg = if (is.na(mc)) th < 0 else mc < 0)
      })
      legend("bottomleft", bty = "n", cex = 0.9, inset = c(0.02, 0.01),
             title = "Variance reduction vs. SRS average", title.col = "black",
             fill = COL[cmp], border = NA,
             legend = vapply(red, `[[`, "", "txt"),
             text.col = ifelse(vapply(red, `[[`, TRUE, "neg"), col_samp, "black"))
    }
  })
}

shinyApp(ui, server)
```

**Shinylive App Comparing SRS (average and ratio) with UPSWR (Hansen–Hurwitz)**

:::



### Why the residuals $e_i$ drive both variances

Both estimators are $\bar{x}_U$ times an
estimate of the slope $B$. SRS uses a *ratio of sample totals*, with every unit equally
likely. UPSWR uses a *mean of ratios*, over a sample tilted towards large $x_i$. Their
errors can be written in terms of the $e_i$ of the sampled units:

* **UPSWR.** Each draw contributes $\bar{x}_U r_i$, and its error is exactly a scaled residual:
  $$
  \bar{x}_U r_i - \bar{t}_U = \bar{x}_U\,(r_i - B) = \bar{x}_U\,\frac{t_i - Bx_i}{x_i} = \frac{\bar{x}_U}{x_i}\,e_i .
  $$
  A draw lands on unit $i$ with probability $\psi_i = x_i/t_x$, so these errors average to
  $\sum_i \psi_i\,\bar{x}_U e_i/x_i = \sum_i e_i/N = 0$; this is why HH is unbiased. The $n$ draws are
  independent, so the variance is $1/n$ times the $\psi$-weighted mean square of the draw errors:
  $$
  V(\hat{\bar{t}}_{\mathrm{HH}}) = \frac{1}{n}\sum_{i=1}^N \psi_i \left(\frac{\bar{x}_U}{x_i}\,e_i\right)^2
  = \frac{1}{n}\sum_{i=1}^N \frac{x_i}{N\bar{x}_U}\cdot\frac{\bar{x}_U^2}{x_i^2}\,e_i^2
  = \frac{1}{n}\cdot\frac{1}{N}\sum_{i=1}^N e_i^2\,\frac{\bar{x}_U}{x_i} .
  $$
  A unit is drawn in proportion to $x_i$, but its error is scaled by $1/x_i$. It squares to
  $1/x_i^2$, so each unit's net contribution is proportional to $e_i^2/x_i$.

* **SRS + ratio.** Since $\sum_{i=1}^n t_i - B\sum_{i=1}^n x_i = \sum_{i=1}^n e_i$,
  $$
  \hat{\bar{t}}_{r} - \bar{t}_U = \bar{x}_U\,(\hat{B} - B) = \frac{\bar{x}_U}{\bar{x}}\,\bar{e},
  \qquad \bar{e} = \frac{1}{n}\sum_{i=1}^n e_i ,
  $$
  so the error is (up to $\bar{x}_U/\bar{x} \approx 1$) the SRS mean of the residuals. Their
  population mean is $0$, so
  $$
  V(\hat{\bar{t}}_{r}) \approx \left(1-\frac{n}{N}\right)\frac{1}{n}\cdot\frac{1}{N-1}\sum_{i=1}^N e_i^2 .
  $$

Ignoring the finite population correction (and taking $N - 1 \approx N$), the two
variances differ only by the weight $\bar{x}_U/x_i$ on each squared residual:
$$
V(\hat{\bar{t}}_{r}) \approx \frac{1}{n}\cdot\frac{1}{N}\sum_{i=1}^N e_i^2,
\qquad
V(\hat{\bar{t}}_{\mathrm{HH}}) = \frac{1}{n}\cdot\frac{1}{N}\sum_{i=1}^N e_i^2\,\frac{\bar{x}_U}{x_i}.
$$
Under UPSWR, units larger than average count less and smaller ones count more. UPSWR
therefore wins when the large residuals belong to the units with $x_i > \bar{x}_U$, i.e.
when the spread of $t_i$ about the line grows with $x_i$, as it does for county acreages.
When the spread is constant in $x_i$, the SRS ratio estimator does better. The app's
"Variance reduction vs. SRS average" legend reports both variances, simulated and
theoretical, relative to the plain SRS average $\hat{\bar{t}}$.

## Analysis of a dataset on student study hours collected by UPSWO 

When sampling **Without Replacement** with unequal probabilities, we rely on the inclusion probabilities ($\pi_k$) rather than single-draw probabilities ($\psi_i$). 

```{r classpps-upswo}
classpps <- read.csv("data/classpps.csv")

res_upswo <- cluster_upswo_ratio(
  data = classpps,
  cname = "class",
  csize = "clssize",
  cpik = "clssize", # Assuming inclusion prob is proportional to class size
  yvar = "hours",
  estimate = "mean"
)

# Display the cluster-level working table
res_upswo$table
```

**Estimate the overall mean study hours (Horvitz-Thompson ratio style):**

```{r classpps-upswo-est}
res_upswo$estimate |> format_est_gt("ratio_mean")
```


