# Unequal Probability Sampling

## Key Concepts and Formulas

Unequal Probability Sampling (UPS) improves estimation precision by sampling units with probabilities proportional to their size (PPS). If a cluster or county is larger, it has a higher chance of being included in the sample. This approach prevents extreme variance when cluster sizes vary drastically.

**1. UPS With Replacement (UPSWR)**
When sampling with replacement, we draw $n$ units. A single unit $i$ has a selection probability $\psi_i$ on each draw. 

*   **Total Estimator (Hansen-Hurwitz):** $\hat{t}_{pwr} = \frac{1}{n} \sum_{i=1}^n \frac{y_i}{\psi_i}$
*   **Ratio Estimator (Mean):** $\hat{\bar{y}}_{r} = \frac{\sum_{i=1}^n \frac{y_i}{\psi_i}}{\sum_{i=1}^n \frac{M_i}{\psi_i}}$

**2. UPS Without Replacement (UPSWO)**
When sampling without replacement, the inclusion probability for unit $i$ across all draws is $\pi_i$.

*   **Total Estimator (Horvitz-Thompson):** $\hat{t}_{HT} = \sum_{i=1}^n \frac{y_i}{\pi_i}$

---

## Functions and Packages

We consolidate our custom estimation functions here, modifying them to return structured lists for cluster summaries where applicable. We also include the `format_est_gt` helper to render all final estimates with mathematical LaTeX headers.

```{r message=FALSE, warning=FALSE}
#| code-fold: true
library(dplyr)
library(gt)
library(latex2exp)

# ---------------------------------------------------------
# Base Estimators
# ---------------------------------------------------------

srs_ratio_est <- function(ydata, xdata, N = Inf) { 
  n <- length(xdata)
  xbar <- mean(xdata)
  ybar <- mean(ydata)
  B_hat <- ybar / xbar
  d <- ydata - B_hat * xdata
  var_d <- sum(d^2) / (n - 1)
  sd_B_hat <- sqrt((1 - n/N) * var_d / n) / xbar
  mem <- qt(0.975, df = n - 1) * sd_B_hat
  
  output <- c(B_hat, sd_B_hat, B_hat - mem, B_hat + mem)
  names(output) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  output
}

srs_reg_est <- function(ydata, xdata, xbarU, N = Inf, est.total = FALSE) {
  n <- length(ydata)
  lmfit <- lm(ydata ~ xdata)
  Bhat <- lmfit$coefficients
  efit <- lmfit$residuals
  SSe <- sum(efit^2) / (n - 2) 
  yhat_reg <- Bhat[1] + Bhat[2] * xbarU
  se_yhat_reg <- sqrt((1-n/N) * SSe / n)
  mem <- qt(0.975, df = n - 2) * se_yhat_reg
  
  output <- c(yhat_reg, se_yhat_reg, yhat_reg - mem, yhat_reg + mem)
  
  if (est.total) {
      if(!is.finite(N)) stop("N must be finite for estimating population total")
      output <- output * N
  }
  names(output) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  output
}

srs_mean_est <- function(sdata, N = Inf) {
  n <- length(sdata)
  ybar <- mean(sdata)
  se.ybar <- sqrt((1 - n / N)) * sd(sdata) / sqrt(n)  
  mem <- qt(0.975, df = n - 1) * se.ybar
  output <- c(ybar, se.ybar, ybar - mem, ybar + mem)
  names(output) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  output
}

# ---------------------------------------------------------
# UPS Estimators
# ---------------------------------------------------------

# Total estimation for datasets collected with UPS with Replacement
upswr_total_est <- function(total, psi) { 
  srs_mean_est(total/psi, N = Inf)
}

# Ratio estimation for datasets collected with UPS with Replacement
upswr_ratio_est <- function(total, M, psi) {
  srs_ratio_est(total/psi, M/psi, N = Inf)
}

# Cluster UPS with Replacement
cluster_upswr_ratio_est <- function(data, sid, csize, cpsi, yvar) {
    clust <- data[, sid]
    ydata <- data[, yvar]  
    
    ybari <- tapply(ydata, clust, mean)
    Mi <- tapply(data[,csize], clust, function(x) x[1])
    psi <- tapply(data[,cpsi], clust, function(x) x[1])
    t_hat_cls <- ybari * Mi 
    
    summary_df <- data.frame(
        Cluster = names(Mi),
        Mi = unname(Mi),
        psi = unname(psi),
        ybari = unname(ybari),
        ti_hat = unname(t_hat_cls),
        ti_psi = unname(t_hat_cls/psi),
        Mi_psi = unname(Mi/psi)
    )
    
    est <- srs_ratio_est(t_hat_cls/psi, Mi/psi)
    list(summary = summary_df, estimate = est)
}

# Cluster UPS without Replacement
cluster_upswo_ratio_est <- function(data, cname, csize, cpik, yvar) {
  clust <- data[, cname]
  ydata <- data[, yvar]  
  
  ybari <- tapply(ydata, clust, mean)
  Mi <- tapply(data[,csize], clust, function(x) x[1])
  pik <- tapply(data[,cpik], clust, function(x) x[1])
  t_hat_cls <- ybari * Mi 
  
  summary_df <- data.frame(
        Cluster = names(Mi),
        Mi = unname(Mi),
        pik = unname(pik),
        ybari = unname(ybari),
        ti_hat = unname(t_hat_cls)
  )
  
  est <- srs_ratio_est(t_hat_cls/pik, Mi/pik)
  list(summary = summary_df, estimate = est)
}

# ---------------------------------------------------------
# GT Formatting Helper
# ---------------------------------------------------------

format_est_gt <- function(est_data, est_type = c("mean", "total", "ratio", "reg_mean", "reg_total", "ratio_mean", "domain_mean")) {
  est_type <- match.arg(est_type)
  
  sym_map <- list(
    "mean"        = c(est = "$\\bar{y}$",         se = "$SE(\\bar{y})$"),
    "total"       = c(est = "$\\hat{t}$",         se = "$SE(\\hat{t})$"),
    "ratio"       = c(est = "$\\hat{B}$",         se = "$SE(\\hat{B})$"),
    "reg_mean"    = c(est = "$\\bar{y}_{reg}$",   se = "$SE(\\bar{y}_{reg})$"),
    "reg_total"   = c(est = "$\\hat{t}_{reg}$",   se = "$SE(\\hat{t}_{reg})$"),
    "ratio_mean"  = c(est = "$\\bar{y}_{r}$",     se = "$SE(\\bar{y}_{r})$"),
    "domain_mean" = c(est = "$\\bar{y}_{d}$",     se = "$SE(\\bar{y}_{d})$")
  )
  
  est_sym <- sym_map[[est_type]]["est"]
  se_sym  <- sym_map[[est_type]]["se"]
  
  if (is.vector(est_data) && !is.list(est_data)) {
    df <- as.data.frame(t(unname(est_data)))
  } else {
    df <- as.data.frame(est_data)
  }
  
  colnames(df) <- c("Est.", "S.E.", "ci.low", "ci.upp")
  use_stub <- !is.null(rownames(df)) && !all(rownames(df) == as.character(1:nrow(df)))
  
  df |>
    gt(rownames_to_stub = use_stub) |>
    cols_label(
      Est.   = md(est_sym),
      S.E.   = md(se_sym),
      ci.low = md("95% CI Lower"),
      ci.upp = md("95% CI Upper")
    ) |>
    fmt_number(columns = everything(), decimals = 4) |>
    tab_options(table.width = pct(70))
}
```

## Analysis of a dataset on student study hours collected by UPSWR

We first manually define a small dataset describing hours spent studying, where the probability of sampling a specific student group ($\psi_i$) was proportional to the number of students in that group ($M_i$).

```{r}
# 1. Define the raw data
psi <- c(24, 100, 100, 76, 44) / 647
Mi <- c(24, 100, 100, 76, 44)
hrs <- c(75, 203, 203, 191, 168)

# Display the data cleanly
data.frame(Mi, psi, hrs) |> 
  gt() |> 
  cols_label(
    Mi = md("$M_i$"),
    psi = md("$\\psi_i$"),
    hrs = md("**Total Hours ($y_i$)**")
  ) |> 
  fmt_number(columns = psi, decimals = 3)

plot(hrs ~ Mi, main = "Study Hours vs Number of Students", 
     xlab = "Number of Students (Mi)", ylab = "Study Hours (hrs)")
abline(lm(hrs ~ Mi), col = "blue")
```

**Estimate the total and mean study hours:**

```{r}
# Estimate the total hours spent by all students
upswr_total_est(total = hrs, psi = psi) |> 
  format_est_gt("total")

# Estimate the mean hours spent by each student
upswr_ratio_est(total = hrs, M = Mi, psi = psi) |> 
  format_est_gt("ratio_mean")
```

## Analysis of `statepop.csv` collected with one-stage UPSWR

The file `statepop.csv` contains data from an unequal-probability sample of 100 U.S. counties. Counties were chosen with probabilities proportional to their populations. The total U.S. population is $M_0 = 255,077,536$. Because sampling was done **with replacement**, large counties like Los Angeles occur multiple times in the sample.

```{r}
statepop <- read.csv("data/statepop.csv")

# Constants
M0 <- 255077536 
popn <- statepop$popn 
psi <- popn / M0
phys <- statepop$phys

# Display a subset of the raw data in a scrollable box
statepop |> 
  select(state, county, popn, phys) |> 
  mutate(psi = popn / M0) |> 
  gt() |> 
  tab_header(title = "U.S. County Data (Sample)") |> 
  tab_options(container.height = px(300), container.overflow.y = TRUE)

plot(popn, phys, log = "xy", 
     main = "Physicians vs Population (Log Scale)", 
     xlab = "Population", ylab = "Physicians")
```

**Estimate the Total and Per Capita Physicians:**

```{r}
# Estimate the total number of physicians in the USA
upswr_total_est(phys, psi) |> 
  format_est_gt("total")

# Estimate the number of physicians per capita (Ratio)
upswr_ratio_est(phys, popn, psi) |> 
  format_est_gt("ratio_mean")
```
**Output Note:** The total estimator heavily leverages the selection probabilities to estimate that there are roughly `r round(upswr_total_est(phys, psi)["Est."])` physicians in the US.

### Compare with an estimate based on an SRS sample

Let's compare our UPSWR results against a naive Simple Random Sample of 100 counties.

```{r}
# Read SRS sample, treating -99 as NA
counties <- read.csv("data/counties.csv", na.strings = "-99")
totpop <- 255077536
N <- 3141 # Total number of counties

# 1. Naive SRS estimate for total
(srs_mean_est(counties$physician, N = N) * N) |> 
  format_est_gt("total")

# 2. Ratio estimate for total using SRS data
(srs_ratio_est(counties$physician, counties$totpop, N = N) * totpop) |> 
  format_est_gt("total")

# 3. Regression estimate for total using SRS data
srs_reg_est(counties$physician, counties$totpop, xbarU = totpop/N, N = N, est.total = TRUE) |> 
  format_est_gt("reg_total")
```
**Output Note:** The true value is **532,638**. Notice how poorly the naive SRS estimate performs (predicting over 800,000) because it ignores county sizes. Both the SRS Ratio and SRS Regression methods perform much better by incorporating auxiliary population data.

## Analysis of `studenthrs.csv` collected with two-stage UPSWR

Here, we sample classes (Primary Sampling Units) with replacement based on class size, and then subsample students within them to ask about their study hours.

```{r}
studenthrs <- read.table("data/studentshrs.txt", header = TRUE)

# Calculate sampling probability (cpsi) based on known total students (647)
studenthrs <- studenthrs |> 
  mutate(cpsi = class_size / 647)

# Run the 2-stage UPSWR function
res_hrs <- cluster_upswr_ratio_est(
  data = studenthrs, 
  sid = "sid", 
  csize = "class_size", 
  cpsi = "cpsi", 
  yvar = "hrs"
)

# Display the cluster-level summary
res_hrs$summary |> 
  gt() |> 
  tab_header(title = "Cluster-Level Summary (2-Stage UPSWR)") |>
  cols_label(
    Cluster = md("**Sample ID**"),
    Mi = md("$M_i$"),
    psi = md("$\\psi_i$"),
    ybari = md("$\\bar{y}_i$"),
    ti_hat = md("$\\hat{t}_i$"),
    ti_psi = md("$\\hat{t}_i/\\psi_i$"),
    Mi_psi = md("$M_i/\\psi_i$")
  ) |> 
  fmt_number(columns = c(psi, ybari, ti_hat, ti_psi, Mi_psi), decimals = 3) |>
  tab_options(container.height = px(300), container.overflow.y = TRUE)
```

**Estimate the overall mean study hours:**

```{r}
res_hrs$estimate |> format_est_gt("ratio_mean")
```

## Simulation Studies for UPSWR using `agpop` data

Let's simulate 2000 random samples to see how UPSWR empirically compares to Simple Random Sampling. 

```{r message=FALSE, warning=FALSE, cache=TRUE}
# Read population data
agpop <- read.csv("data/agpop.csv", na.strings = "-99")
agpop <- agpop |> filter(!is.na(acres92), !is.na(acres87))

N <- nrow(agpop)
total_acres92 <- sum(agpop$acres92)
total_acres87 <- sum(agpop$acres87)
acres87 <- agpop$acres87

# Storage vectors for simulation
est_srs300 <- rep(0, 2000)
est_srs300_reg_acres87 <- rep(0, 2000)
est_ups300_acres87 <- rep(0, 2000)

for (i in 1:2000) {
  # SRS Simulation
  srs300 <- sample(1:N, size = 300)
  acres92_srs300 <- agpop$acres92[srs300]
  acres87_srs300 <- agpop$acres87[srs300]
  
  est_srs300[i] <- (srs_mean_est(acres92_srs300, N = N) * N)[1] 
  est_srs300_reg_acres87[i] <- srs_reg_est(acres92_srs300, acres87_srs300, 
                                           total_acres87/N, N = N, est.total = TRUE)[1]
  
  # UPSWR Simulation (Probability proportional to acres87)
  ups300 <- sample(1:N, size = 300, prob = acres87, replace = TRUE)
  acres92_ups300 <- agpop$acres92[ups300]
  
  psi <- (acres87 / sum(acres87))[ups300]
  est_ups300_acres87[i] <- upswr_total_est(acres92_ups300, psi)[1] 
}
```

### Visualizing Estimator Variance

```{r fig.height=8, fig.width=8}
par(mfrow = c(2,1))

# Compare Naive SRS, Regression SRS, and UPSWR
boxplot(data.frame(est_srs300, est_srs300_reg_acres87, est_ups300_acres87),
        main = "Comparing SRS, SRS-Regression, and UPSWR",
        names = c("SRS Naive", "SRS Regression", "UPSWR"),
        col = c("lightblue", "lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)

# Zoom in on the two high-precision estimators
boxplot(data.frame(est_srs300_reg_acres87, est_ups300_acres87),
        main = "Zoomed: SRS-Regression vs. UPSWR",
        names = c("SRS Regression", "UPSWR"),
        col = c("lightgreen", "salmon"))
abline(h = total_acres92, col = "red", lwd = 2)
```
**Output Note:** The red line represents the true total acreage. The naive SRS estimator has massive variance. Both SRS Regression and UPSWR drastically reduce variance, centering tightly around the true mean.

## Analysis of a dataset on student study hours collected by UPSWO 

When sampling **Without Replacement** with unequal probabilities, we rely on the inclusion probabilities ($\pi_k$) rather than single-draw probabilities ($\psi_i$). 

```{r}
classpps <- read.csv("data/classpps.csv")

res_upswo <- cluster_upswo_ratio_est(
  data = classpps, 
  cname = "class", 
  csize = "clssize", 
  cpik = "clssize", # Assuming inclusion prob is proportional to class size
  yvar = "hours"
)

# Display the cluster-level summary
res_upswo$summary |> 
  gt() |> 
  tab_header(title = "Cluster-Level Summary (UPS Without Replacement)") |>
  cols_label(
    Cluster = md("**Class**"),
    Mi = md("$M_i$"),
    pik = md("$\\pi_k$"),
    ybari = md("$\\bar{y}_i$"),
    ti_hat = md("$\\hat{t}_i$")
  ) |> 
  fmt_number(columns = c(ybari, ti_hat), decimals = 2)
```

**Estimate the overall mean study hours (Horvitz-Thompson ratio style):**

```{r}
res_upswo$estimate |> format_est_gt("ratio_mean")
```