---
title: "STAT 348 Sampling Techniques"
subtitle: "Introduction to Unequal Probability Sampling"
pagetitle: "UPS-Slides"
format: banner_slides-revealjs
engine: knitr
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center"
)
```

# Introduction to Unequal Probability Sampling

Textbook sections:

- 6.1--6.3 in Lohr's book (not covering UPS without replacement)
- 8.9 and 9.7 in Scheaffer et al., titled "Cluster Sampling with Probabilities
  Proportional to Size"

## Introductory Example

O'Brien et al. (1995) sampled nursing home residents in the Philadelphia area
to determine residents' preferences on life-sustaining treatments (CPR,
hospital transfer, feeding tubes). The target population was all residents of
294 licensed nursing homes, with 37,652 beds total — before sampling, only the
number of *beds* was known, not the number of residents.

In a cluster sample with **equal** probabilities, a nursing home with 20 beds
is just as likely to be chosen as one with 1000 beds — wasting the fact that
we already know home size, and that resident counts (and hence answers) are
tied to home size.

## A Toy Example

A town has four supermarkets, ranging in size from $100\,\text{m}^2$ to
$1000\,\text{m}^2$. We want to estimate total sales across the four stores
last month, but can only afford to sample **one** store. Let $\psi_i = P(\text{Store } i \text{ selected})$.

## Scheme 1: Probabilities Proportional to Size (PPS)

Since Store D accounts for $10/16$ of the total floor area, we sample it with
probability $10/16$:

| Store | Size ($M_i$) | $\psi_i$ | $t_i$ (thousands) |
|:---:|---:|---:|---:|
| A | 100 | $1/16$ | 11 |
| B | 200 | $2/16$ | 20 |
| C | 300 | $3/16$ | 24 |
| D | 1000 | $10/16$ | 245 |
| **Total** | **1600** | **1** | **300** |

We could realize this design with 16 numbered cards: card 1 $\to$ A; cards 2--3
$\to$ B; cards 4--6 $\to$ C; cards 7--16 $\to$ D.

## Estimation for Scheme 1 {.smaller}

Since only one store is sampled, define the **Hansen-Hurwitz** estimator
$\hat t_\psi = t_i/\psi_i$ for the selected store $i$. The **Sum** row gives
$E[\hat t_\psi]$ and $V[\hat t_\psi]$ directly:

:::: {.columns}

:::{.column style="width:60%;"}

| Sample | $\psi_i$ | $t_i$ | $\hat t_\psi$ | $\psi_i\hat t_\psi$ | $(\hat t_\psi-t)^2$ | $\psi_i(\hat t_\psi-t)^2$ |
|:---:|---:|---:|---:|---:|---:|---:|
| $\{A\}$ | $1/16$ | 11 | 176 | 11 | 15,376 | 961 |
| $\{B\}$ | $2/16$ | 20 | 160 | 20 | 19,600 | 2,450 |
| $\{C\}$ | $3/16$ | 24 | 128 | 24 | 29,584 | 5,547 |
| $\{D\}$ | $10/16$ | 245 | 392 | 245 | 8,464 | 5,290 |
| **Sum** | **1** | **300** | | **300** | | **14,248** |

:::

:::{.column style="width:40%;"}

$$
\begin{aligned}
E[\hat t_\psi] &= \sum_{i=1}^N \psi_i\hat t_\psi = 300 = t \\[4pt]
V[\hat t_\psi] &= \sum_{i=1}^N \psi_i(\hat t_\psi-t)^2 = 14{,}248
\end{aligned}
$$

$\hat t_\psi$ is always unbiased, because in general
$E[\hat t_\psi]=\sum_{i=1}^N \psi_i\dfrac{t_i}{\psi_i}=t$.

:::

::::

## Scheme 2: SRS with Simple (Unbiased) Estimation {.smaller}

Now suppose instead we take an SRS of size 1, so $\psi_i=1/4$ for every store
(and $1/\psi_i=4=N$). The **Sum** row gives $V[\hat t_{\text{SRS}}]$ directly:

:::: {.columns}

:::{.column style="width:60%;"}

| Sample | $\psi_i$ | $t_i$ | $\hat t_\psi$ | $(\hat t_\psi-t)^2$ | $\psi_i(\hat t_\psi-t)^2$ |
|:---:|---:|---:|---:|---:|---:|
| $\{A\}$ | $1/4$ | 11 | 44 | 65,536 | 16,384 |
| $\{B\}$ | $1/4$ | 20 | 80 | 48,400 | 12,100 |
| $\{C\}$ | $1/4$ | 24 | 96 | 41,616 | 10,404 |
| $\{D\}$ | $1/4$ | 245 | 980 | 462,400 | 115,600 |
| **Sum** | **1** | | | | **154,488** |

:::

:::{.column style="width:40%;"}

$$
\begin{aligned}
E[\hat t_{\text{SRS}}] &= 300 = t \\[4pt]
V[\hat t_{\text{SRS}}] &= \sum_{i} \psi_i(\hat t_\psi-t)^2 = 154{,}488
\end{aligned}
$$

$\hat t_{\text{SRS}}$ is still unbiased, but this is over ten times larger
than the PPS design's variance of 14,248!

:::

::::

The PPS
design uses **auxiliary information**: store size is known and related to
sales, and the design exploits this. If Store D were selected under SRS, its
huge sales total gets multiplied by $N=4$ (since it represents only itself),
producing a wildly unstable estimate.

## Scheme 2: SRS with Ratio Estimation {style="font-size: 32px"}

Alternatively, keep the SRS of size 1 but use *ratio* estimation with $x_i=M_i$
(known total $M_0=1600$), $\hat t_r = \dfrac{t_i}{M_i}\cdot M_0$. The **Sum**
row gives $\sum(\hat t_r-t)^2$ directly:

:::: {.columns}

:::{.column style="width:55%;"}

| Store | $M_i$ | $t_i$ | $\hat t_r$ | $\hat t_r-t$ | $(\hat t_r-t)^2$ |
|:---:|---:|---:|---:|---:|---:|
| A | 100 | 11 | 176 | $-124$ | 15,376 |
| B | 200 | 20 | 160 | $-140$ | 19,600 |
| C | 300 | 24 | 128 | $-172$ | 29,584 |
| D | 1000 | 245 | 392 | $92$ | 8,464 |
| **Sum** | **1600** | **300** | **856** | | **73,024** |

:::

:::{.column style="width:45%;"}

$$
\begin{aligned}
E(\hat t_r) &= \frac{856}{4} = 214 \quad(\ne 300,\ \text{biased!}) \\[4pt]
\text{MSE}(\hat t_r) &= \frac{\sum(\hat t_r-t)^2}{4} = \frac{73{,}024}{4} = 18{,}256
\end{aligned}
$$

:::

::::

## Ratio Estimation vs. Unequal Probability Sampling

Notice the four $\hat t_r$ values (176, 160, 128, 392) are *numerically
identical* to the PPS estimates $\hat t_\psi$ from Scheme 1! But averaged with
the **equal** SRS probabilities ($1/4$ each) rather than the **size-based**
PPS probabilities, they give a biased result (214, not 300).

- **Ratio estimation** is an *estimating* method that exploits the correlation
  between $t_i$ and $M_i$ *after* an SRS is drawn.
- **Unequal probability sampling** is a *sampling design* that exploits the
  same correlation *before* the sample is drawn, by choosing $\psi_i\propto M_i$.

# Unequal Probability Sampling with Replacement (UPSWR)

## How to Do Unequal Probability Sampling?

Suppose $n>1$ and we sample **with replacement**: sampling with replacement
means the selection probabilities do not change after the first draw. Let
$$
\psi_i = P(\text{select unit } i \text{ on any given draw}).
$$

The idea: draw $n$ psus with replacement (some psus may be drawn more than
once). For each draw, estimate the population total using the single-draw
estimator from before; average the $n$ (independent) estimates.

A common choice, **probabilities proportional to size (PPS)**:
$$
\psi_i = \frac{M_i}{\sum_{k=1}^N M_k}
$$

## Drawing Probabilities in Practice

```{r}
#| label: fig-pps-line
#| fig-width: 8
#| fig-height: 2.6
#| out-width: "85%"
set.seed(1)
psi <- c(0.10, 0.22, 0.15, 0.28, 0.25)
labels <- LETTERS[1:5]
cum <- c(0, cumsum(psi))
draws <- c(0.06, 0.35, 0.55, 0.83)

par(mar = c(3, 0.5, 0.5, 0.5))
plot(NA, xlim = c(0, 1), ylim = c(0, 1), xaxt = "n", yaxt = "n",
     xlab = "", ylab = "", bty = "n")
cols <- rep(c("#dbe9fb", "#c3dbf7"), length.out = 5)
rect(cum[-6], 0.3, cum[-1], 0.7, col = cols, border = "black", lwd = 1.5)
text((cum[-6] + cum[-1]) / 2, 0.5, labels, cex = 1.4)
axis(1, at = cum, labels = round(cum, 2), cex.axis = 0.9)
points(draws, rep(0.5, 4), pch = 17, col = "firebrick", cex = 1.8)
text(draws, rep(0.82, 4), sprintf("u=%.2f", draws), col = "firebrick", cex = 0.85)
```

Divide $[0,1]$ into segments of length $\psi_1,\ldots,\psi_N$. Draw a random
number $u\sim\text{Uniform}(0,1)$ (red triangles); whichever segment $u$ falls
into determines the selected unit. In R: `sample(1:N, size = n, prob = psi, replace = TRUE)`.

## Notation for the Sample $\mathcal R$

Because we sample with replacement, the sample may contain the same psu more
than once. Let $\mathcal R$ (not $\mathcal S$) denote the multiset of $n$ draws,
**including repeats**. For example, with the supermarket population, a sample
of size $n=10$ might be
$$
\mathcal R = \{D,D,D,C,C,D,C,A,B,B\}.
$$

## The Hansen-Hurwitz Estimator

Let $u_i = t_i/\psi_i$. We saw that for a single draw, $u_i$ is an unbiased
estimator of $t$. Sampling $n$ psus with replacement gives $n$ independent
copies, which we average:
$$
\hat t_\psi = \frac1n \sum_{i\in\mathcal R} \frac{t_i}{\psi_i} = \frac1n\sum_{i\in\mathcal R} u_i = \bar u
$$

The estimated variance:
$$
\hat V(\hat t_\psi) = \frac1n\cdot\frac{1}{n-1}\sum_{i\in\mathcal R}\Big(\frac{t_i}{\psi_i}-\hat t_\psi\Big)^2
$$

## Why $\hat t_\psi$ Is Unbiased: the IID View

Let $u_j^*$ denote the value of $u_i=t_i/\psi_i$ for the $j$th *draw* (not the
$j$th population unit) — so $\hat t_\psi = \frac1n\sum_{j=1}^n u_j^*=\bar u^*$.
Since draws are independent and identically distributed,
$$
E(u_j^*) = \sum_{i=1}^N u_i\psi_i = \sum_{i=1}^N \frac{t_i}{\psi_i}\psi_i = \sum_{i=1}^N t_i = t
$$
$$
V(u_j^*) = \sum_{i=1}^N \psi_i(u_i-t)^2 = \sum_{i=1}^N \psi_i\Big(\frac{t_i}{\psi_i}-t\Big)^2
$$

Since $\hat t_\psi$ is the mean of $n$ IID copies of $u_j^*$:
$$
E(\hat t_\psi) = E(u_j^*) = t, \qquad
V(\hat t_\psi) = \frac{V(u_j^*)}{n} = \frac1n\sum_{i=1}^N \psi_i\Big(\frac{t_i}{\psi_i}-t\Big)^2
$$

::: {.callout-note}
There is **no finite population correction** — the $n$ draws are independent
(with replacement), unlike SRS without replacement.
:::

## Ratio Estimation for the ssu-Level Mean

We estimate the population mean $\bar y_U = t/M_0$ by
$$
\hat{\bar y}_\psi = \frac{\hat t_\psi}{\hat M_{0\psi}}, \qquad
\hat M_{0\psi} = \frac1n\sum_{i\in\mathcal R}\frac{M_i}{\psi_i}
$$

Using Chapter 4's ratio-estimation results with $y_i=t_i/\psi_i$ and
$x_i=M_i/\psi_i$:
$$
\hat V(\hat{\bar y}_\psi) = \frac{1}{(\hat M_{0\psi})^2}\cdot\frac1n\cdot\frac{1}{n-1}\sum_{i\in\mathcal R}\Big(\frac{t_i}{\psi_i}-\hat{\bar y}_\psi\frac{M_i}{\psi_i}\Big)^2
$$

Again, no fpc is needed, since sampled psus are drawn independently.

# Example: College Statistics Classes

## Setup

A college has 15 introductory statistics classes, with a total of 647
students. We sample $n=5$ classes **with replacement**, with probability
proportional to $M_i$ (class size), so $\psi_i = M_i/647$. We then survey
every student in each sampled class about hours spent studying last week.

| Class | $M_i$ | $\psi_i$ |
|:---:|---:|---:|
| 1 | 44 | 0.0680 |
| 5 | 76 | 0.1175 |
| 12 | 24 | 0.0371 |
| 14 | 100 | 0.1546 |
| $\vdots$ | $\vdots$ | $\vdots$ |
| **Total** | **647** | **1** |

Classes are drawn using cumulative ranges of $M_i$ (Lahiri's method): class 1
gets numbers 1--44, class 2 gets 45--77, and so on up to class 15.

## Sample and Estimate of the Total {.smaller}

Suppose the sample (with repeats) is $\mathcal R = \{12, 14, 14, 5, 1\}$, with
class totals $t_i$ (total hours studying, summed over all students). Writing
$u_i=t_i/\psi_i$, the **Sum** row gives $\sum u_i$ and $\sum(u_i-\hat t_\psi)^2$
directly:

:::: {.columns}

:::{.column style="width:60%;"}

| Class | $\psi_i$ | $t_i$ | $u_i=t_i/\psi_i$ | $u_i-\hat t_\psi$ | $(u_i-\hat t_\psi)^2$ |
|:---:|---:|---:|---:|---:|---:|
| 12 | $24/647$ | 75 | 2021.875 | 272.861 | 74,452.9 |
| 14 | $100/647$ | 203 | 1313.410 | $-435.604$ | 189,751.2 |
| 14 | $100/647$ | 203 | 1313.410 | $-435.604$ | 189,751.2 |
| 5 | $76/647$ | 191 | 1626.013 | $-123.001$ | 15,129.3 |
| 1 | $44/647$ | 168 | 2470.364 | 721.349 | 520,344.8 |
| **Sum** | | | **8745.07** | | **989,429.3** |

:::

:::{.column style="width:40%;"}

$$
\begin{aligned}
\hat t_\psi &= \frac{\sum u_i}{n} = \frac{8745.07}{5} = 1749.014 \\[4pt]
s^2 &= \frac{\sum(u_i-\hat t_\psi)^2}{n-1} = \frac{989{,}429.3}{4} = 247{,}357.3 \\[4pt]
\text{SE}(\hat t_\psi) &= \sqrt{\frac{s^2}{n}} = \sqrt{\frac{247{,}357.3}{5}} = 222.42
\end{aligned}
$$

:::

::::

## Estimate of Hours per Student

Since $\psi_i = M_i/647$ exactly, $M_i/\psi_i = 647$ for every sampled class,
so $\hat M_{0\psi} = M_0 = 647$ exactly. The average hours studied per student:
$$
\begin{aligned}
\hat{\bar y}_\psi &= \frac{1749.014}{647} = 2.70 \text{ hours} \\[4pt]
\text{SE}(\hat{\bar y}_\psi) &= \frac{222.42}{647} = 0.34 \text{ hours}
\end{aligned}
$$

# One-Stage Cluster Sampling with PPS

## A Useful Simplification

When $\psi_i = M_i/M_0$ (PPS), then $u_i = t_i/\psi_i = t_i\cdot(M_0/M_i)=\bar y_i\cdot M_0$,
where $\bar y_i=t_i/M_i$ is the $i$th cluster's own mean. So:
$$
\hat{\bar y}_{\text{pps}} = \frac{\hat t_\psi}{M_0} = \frac1n\sum_{i\in\mathcal R} \bar y_i
$$
$$
\text{SE}(\hat{\bar y}_{\text{pps}}) = \frac{\text{SE}(\hat t_\psi)}{M_0}
 = \sqrt{\frac1n\cdot\frac{\sum_{i\in\mathcal R}(\bar y_i-\hat{\bar y}_{\text{pps}})^2}{n-1}}
$$

Under PPS, we don't need to weight the cluster means by $M_i/\sum M_i$ at all
— they can simply be averaged!

## Example: College Classes, Revisited with the PPS Shortcut

:::: {.columns}

:::{.column style="width:50%;"}

| Class | $\bar y_i = t_i/M_i$ |
|:---:|---:|
| 12 | $75/24 = 3.13$ |
| 14 | $203/100 = 2.03$ |
| 14 | $203/100 = 2.03$ |
| 5 | $191/76 = 2.51$ |
| 1 | $168/44 = 3.81$ |
| **Sum** | **13.51** |

:::

:::{.column style="width:50%;"}

$$
\begin{aligned}
\hat{\bar y}_{\text{pps}} &= \frac{\sum \bar y_i}{5} = \frac{13.51}{5} = 2.70 \\[4pt]
\text{SE}(\hat{\bar y}_{\text{pps}}) &= 0.34
\end{aligned}
$$

Identical to the weighted calculation on the previous slide — but computed
directly as the mean and SE of the five class averages.

:::

::::

# Review: SRS-of-Clusters vs. PPS

## Ratio Estimation for Cluster Sampling with SRS

Recall from cluster sampling: if psus are instead drawn by an ordinary SRS
(not PPS with replacement), the ratio estimator is
$$
\hat{\bar y}_r = \frac{\sum_{i\in\mathcal S} \hat t_i}{\sum_{i\in\mathcal S} M_i} = \frac{\sum_{i\in\mathcal S} M_i\bar y_i}{\sum_{i\in\mathcal S} M_i}
$$
with estimated variance (including a finite-population correction)
$$
\hat V(\hat{\bar y}_r) = \frac{1}{\bar M^2}\Big(1-\frac nN\Big)\frac{s_r^2}{n} + \frac{1}{nN\bar M^2}\sum_{i\in\mathcal S}M_i^2\Big(1-\frac{m_i}{M_i}\Big)\frac{s_i^2}{m_i}
$$

## Key Contrast

|  | SRS of psus + ratio estimation | PPS with replacement |
|:---|:---|:---|
| Selection | equal probability | $\psi_i \propto M_i$ |
| Point estimator | $\hat{\bar y}_r$ (ratio, slightly biased) | $\hat{\bar y}_\psi$ (Hansen-Hurwitz ratio, still approx.) |
| Finite population correction | yes, $(1-n/N)$ | **no** — draws are independent |
| Weighting of cluster means | by $M_i/\sum M_i$ | **no weighting needed** if $\psi_i\propto M_i$ |

Both approaches exploit the same correlation between $t_i$ and $M_i$, but PPS
builds it into the *design*, giving the cleanest formulas when $\psi_i\propto M_i$
exactly.

# Two-Stage Cluster Sampling with Replacement

## Setup

Now suppose that, instead of observing every ssu in a selected psu, we take a
subsample. Let $Q_i$ be the number of times psu $i$ occurs in $\mathcal R$; if
$Q_i>0$, we draw a **different, independent** subsample each time psu $i$ is
selected (unlike one-stage sampling, where a repeated psu just contributes the
same observed total again). For the $j$th subsample from psu $i$
($j=1,\ldots,Q_i$), let $\hat t_{ij}$ estimate $t_i$, with $E[\hat t_{ij}]=t_i$.

## Estimators for the Total

$$
\hat t_\psi = \frac1n \sum_{i=1}^N\sum_{j=1}^{Q_i} \frac{\hat t_{ij}}{\psi_i}
$$
$$
\hat V(\hat t_\psi) = \frac1n\cdot\frac{1}{n-1}\sum_{i=1}^N\sum_{j=1}^{Q_i}\Big(\frac{\hat t_{ij}}{\psi_i}-\hat t_\psi\Big)^2
$$

(If $Q_i=0$ for some psu, that psu simply contributes no terms.) Because
sampling is with replacement, it's possible to draw more than one independent
subsample from the same psu — so this variance estimator automatically
captures **both** the between-psu variability and the within-psu subsampling
variability.

## Estimator for the ssu-Level Mean

$$
\hat{\bar y}_\psi = \frac{\hat t_\psi}{\hat M_{0\psi}}, \qquad
\hat M_{0\psi} = \frac1n\sum_{i\in\mathcal R}\frac{M_i}{\psi_i}
$$
$$
\hat V(\hat{\bar y}_\psi) = \frac{1}{(\hat M_{0\psi})^2}\cdot\frac1n\cdot\frac{1}{n-1}\sum_{i=1}^N\sum_{j=1}^{Q_i}\Big(\frac{\hat t_{ij}}{\psi_i}-\hat{\bar y}_\psi\frac{M_i}{\psi_i}\Big)^2
$$

If an SRS of $m_i$ ssus is taken within psu $i$, the overall sampling weight
for ssu $j$ in psu $i$ becomes
$$
w_{ij} = \frac{1}{n\psi_i}\cdot\frac{M_i}{m_i}
$$

## Example: Hours Spent Studying, Subsampled {.smaller}

Instead of surveying every student, we now subsample 5 students per class
(class 14 is selected twice, with a *different* subsample of 5 each time).
Writing $u_i=\hat t_i/\psi_i$, the **Sum** row gives $\sum u_i$ and
$\sum(u_i-\hat t_\psi)^2$ directly:

:::: {.columns}

:::{.column style="width:65%;"}

| Class | $M_i$ | $\psi_i$ | $\bar y_i$ | $\hat t_i$ | $u_i$ | $e_i=u_i-\hat t_\psi$ | $e^2_i$ |
|:--:|---:|---:|---:|---:|---:|-------:|-----:|
| 12 | 24 | 0.0371 | 2.4 | 57.6 | 1552.8 | $-64.7$ | 4,186.1 |
| 14 | 100 | 0.1546 | 1.6 | 160.0 | 1035.2 | $-582.3$ | 339,073.3 |
| 14 | 100 | 0.1546 | 2.0 | 200.0 | 1294.0 | $-323.5$ | 104,652.2 |
| 5 | 76 | 0.1175 | 2.8 | 212.8 | 1811.6 | $194.1$ | 37,674.8 |
| 1 | 44 | 0.0680 | 3.7 | 162.8 | 2393.9 | $776.4$ | 602,797.0 |
| **Sum** | | | | | **8087.5** | 0| **1,088,383** |

:::

:::{.column style="width:35%;"}

$$
\begin{aligned}
\hat t_\psi &= \frac{8087.5}{5} = 1617.5 \\[4pt]
s^2 &= \frac{1{,}088{,}383}{4} = 272{,}095.8 \\[4pt]
\text{SE}(\hat t_\psi) &= \sqrt{\frac{s^2}{5}} = 233.28\\[4pt]
\hat{\bar y}_\psi &= \frac{1617.5}{647} = 2.5\text{ h} \\[4pt]
\text{SE}(\hat{\bar y}_\psi) &= \frac{233.28}{647} = 0.36\text{ h}
\end{aligned}
$$

:::

::::

Compared with the full-census example ($\hat{\bar y}_\psi=2.70$,
$\text{SE}=0.34$), subsampling gives a similar point estimate at a somewhat
higher SE — the price of observing fewer students per class.

# Weights in Unequal-Probability Sampling

## General Weight Formula

For sampling **without** replacement, the weight is the reciprocal of the
inclusion probability, $1/E[Z_i]$. For sampling **with** replacement, we use
the *first-stage* weight
$$
w_i = \frac{1}{\text{expected number of hits}} = \frac{1}{E[Q_i]} = \frac{1}{n\psi_i}
$$

For one-stage cluster sampling (every ssu observed each time psu $i$ is hit),
$w_{ij}=w_i=1/(n\psi_i)$ for every ssu $j$ in psu $i$, and
$$
\hat t_\psi = \sum_{i\in\mathcal R}\sum_{j=1}^{M_i} w_{ij}y_{ij}, \qquad
\hat{\bar y}_\psi = \frac{\sum_{i\in\mathcal R}\sum_{j=1}^{M_i} w_{ij}y_{ij}}{\sum_{i\in\mathcal R}\sum_{j=1}^{M_i} w_{ij}}
$$

For two-stage sampling with an SRS of $m_i$ ssus per selected psu, the weight
picks up an extra factor $M_i/m_i$ for the subsampling stage, exactly as in
ordinary two-stage cluster sampling.
