---
title: "Simple Random Sampling"
filters:
  - shinylive
execute:
  warning: false
  message: false
  cache: true
---


## Estimators for Simple Random Sampling {#sec-srs-formulas}

A simple random sample (SRS) of $n$ units is drawn **without replacement** from a
population of $N$ units, so that every subset of $n$ units is equally likely.
Write $y_i$ for the value of unit $i$, $\bar{y}_U = \frac{1}{N}\sum_{i=1}^N y_i$ for the
population mean, $t = N\bar{y}_U$ for the population total, and
$$
S^2 = \frac{1}{N-1}\sum_{i=1}^N (y_i - \bar{y}_U)^2
$$
for the population variance.

#### 1. Estimating the Population Mean {-}

The sample mean is unbiased for $\bar{y}_U$:
$$
\bar{y} = \frac{1}{n}\sum_{i=1}^n y_i .
$$
Its true variance and its estimate, which replaces $S^2$ by the sample variance $s^2$, are
$$
V(\bar{y}) = \left(1 - \frac{n}{N}\right)\frac{S^2}{n}, \qquad
\hat{V}(\bar{y}) = \left(1 - \frac{n}{N}\right)\frac{s^2}{n}, \qquad
s^2 = \frac{1}{n-1}\sum_{i=1}^n (y_i - \bar{y})^2 ,
$$
and $\mathrm{SE}(\bar{y}) = \sqrt{\hat{V}(\bar{y})}$. The factor $1 - n/N$ is the
**finite population correction** (fpc). It is close to $1$ when the sampling fraction $n/N$ is small.

#### 2. Estimating the Population Total {-}

$$
\hat{t} = N\bar{y}, \qquad
\hat{V}(\hat{t}) = N^2\,\hat{V}(\bar{y}) = N^2\left(1 - \frac{n}{N}\right)\frac{s^2}{n}, \qquad
\mathrm{SE}(\hat{t}) = N\,\mathrm{SE}(\bar{y}).
$$

#### 3. Estimating a Proportion {-}

For a 0/1 indicator $y_i$, the population mean is the proportion $p$, and the sample mean is the sample proportion $\hat{p}$.
Since then $s^2 = \frac{n}{n-1}\hat{p}(1-\hat{p})$,
$$
\hat{p} = \frac{1}{n}\sum_{i=1}^n y_i, \qquad
\hat{V}(\hat{p}) = \left(1 - \frac{n}{N}\right)\frac{\hat{p}(1 - \hat{p})}{n-1}.
$$

#### 4. Confidence Intervals {-}

By the central limit theorem for sample means, an approximate $95\%$ confidence interval is
$$
\bar{y} \pm t_{0.975,\,n-1}\,\mathrm{SE}(\bar{y}),
$$
and likewise for $\hat{t}$ and $\hat{p}$. With a skewed population, $n$ must be larger for the
normal approximation, and hence the nominal coverage, to hold.
The simulation and the app later in this chapter show this.

---

## Analysis of agsrs.csv Data

Throughout the agriculture examples in this book, county acreages are expressed
in **thousands of acres** --- the raw figures run into the millions, and every
mean, variance, and total below reads far more easily on that scale. The helper
`acres_in_thousands()` performs the conversion as the data is read.

### Step by step calculation without using a function

To evaluate a Simple Random Sample (SRS), we calculate the point estimate, the standard error (incorporating the Finite Population Correction factor), and the margin of error to construct a 95% confidence interval.

**Key Formulas:**

* **Sample Mean:** $\bar{y} = \frac{1}{n} \sum_{i=1}^n y_i$
* **Standard Error:** $\mathrm{SE}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left( \frac{s}{\sqrt{n}} \right)$
* **Margin of Error:** $\mathrm{ME} = t_{\alpha/2, n-1} \times \mathrm{SE}(\bar{y})$
* **Population Total:** $\hat{t} = N\bar{y}$

```{r srs-manual-calc}
## read survey data
agsrs <- read.csv ("data/agsrs.csv")
## work in thousands of acres, the unit used for every agriculture example here
agsrs[c("acres92", "acres87", "acres82")] <- agsrs[c("acres92", "acres87", "acres82")] / 1000
head(agsrs)

## extract the variable of interest
sdata <- agsrs$acres92
N <- 3078

## do calculation
n <- length (sdata)
ybar <- mean (sdata)
se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
mem <- qt (0.975, df = n - 1) * se.ybar

## return estimate vector for pop mean
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)

## return estimate vector for pop total
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem) * N

```

### Write a function for repeated use

#### A function for doing data analysis for srs sample

Wrap the previous manual calculations into a custom R function to quickly estimate means, standard errors, and confidence intervals for any variable of interest. To avoid repeating this code in every chapter, all estimating functions (and their `gt`-table formatting helpers) live in a single shared file, `samplingestimate.r`, which we source below.

```{r source-functions}
#| echo: false
source("samplingestimate.r")
```

::: {.content-visible when-format="html"}

<details>
<summary>Show the shared estimating functions (`samplingestimate.r`)</summary>

```{r print-functions}
#| echo: false
#| results: asis
cat("```r\n")
cat(readLines("samplingestimate.r"), sep = "\n")
cat("\n```\n")
```

</details>

:::

#### Apply srs_est to agsrs.csv data

*Import Data*

Load the survey sample data representing US counties.

```{r agsrs-load}
agsrs <- acres_in_thousands (read.csv ("data/agsrs.csv"))

```

*Estimating the mean of acre92*

Calculate the estimated average farm acreage per county using the custom function. The function's `$table` shows the row-level working values (truncated to the first 6 rows + Sum), and `$estimate` holds the final point estimate, SE, and CI, formatted with `format_est_gt()`.

```{r agsrs-mean-est}
res_mean <- srs_est(agsrs[,"acres92"], N = 3078, estimate = "mean")
res_mean$table
res_mean$estimate |> format_est_gt("mean")

```

*Estimating the total of acre92*

Set `estimate = "total"` to have the function itself scale the mean and confidence interval by the population size ($N = 3078$) to estimate the total farm acreage across all counties.

```{r agsrs-total-est}
res_total <- srs_est(agsrs[,"acres92"], N = 3078, estimate = "total", show.details = FALSE)
res_total$estimate |> format_est_gt("total")

```

*Estimating the proportion of counties with fewer than 200K acres for farming in 1992*

To estimate a proportion under SRS, create an indicator variable ($y_i = 1$ if true, $0$ if false). The mean of this indicator variable serves as the sample proportion ($\hat{p} = \bar{y}_{indicator}$).

```{r agsrs-prop-est}
acres92.is.fewer.200k <- as.numeric (agsrs[,"acres92"] < 200)   # 200 thousand acres
head(acres92.is.fewer.200k)

res_prop <- srs_est(acres92.is.fewer.200k, N = 3078, estimate = "mean", show.details = FALSE)
res_prop$estimate |> format_est_gt("mean")

```

*Estimating the total number of counties with fewer than 200K acres for farming in 1992*

Set `estimate = "total"` to estimate the absolute count of counties meeting the criteria.

```{r agsrs-prop-total-est}
res_prop_total <- srs_est(acres92.is.fewer.200k, N = 3078, estimate = "total", show.details = FALSE)
res_prop_total$estimate |> format_est_gt("total")

```

### Comparing with true value

Load the complete census (population) dataset to calculate the true parameters. This allows us to assess the accuracy of our previous SRS estimates against the exact population values.

```{r agpop-true-values}
agpop <- acres_in_thousands (read.csv ("data/agpop.csv"))
#true mean (thousands of acres)
mean (agpop[, "acres92"], na.rm = T)
# true total (thousands of acres)
sum (agpop[, "acres92"], na.rm = T)

# true proportion of counties with less than 200K acres for farming
mean (agpop[, "acres92"] < 200, na.rm = T)
# true number of counties with less than 200K acres for farming
sum (agpop[, "acres92"] < 200, na.rm = T)

```

## A Simulation Demonstration of SRS Inference

Load the full population data and filter out missing values to serve as a complete, known finite population for resampling experiments.

```{r agpop-load-clean}
# read population data, with acreages in thousands of acres
agpop <- acres_in_thousands (read.csv ("data/agpop.csv"))
# remove those counties with na
agpop <- subset( agpop, !is.na (acres92))

```

### True Values

Define the target sample size and compute the true population mean ($\mu$) and theoretical standard error ($\mathrm{SE}_{true}$) using the known population standard deviation ($S$).

**Theoretical Standard Error:**


$$\mathrm{SE}_{true}(\bar{y}) = \sqrt{1 - \frac{n}{N}} \left(\frac{S}{\sqrt{n}}\right)$$

```{r srs-true-values}
# sample size
n <- 300
# population size
N <- nrow (agpop); N

# true value of population mean
ybarU <- mean (agpop[,"acres92"]); ybarU
# true value of deviation of sample mean
true.se.ybar <- sqrt (1- n/N) * sd (agpop[,"acres92"]) / sqrt (n); true.se.ybar

```

### One SRS sampling

Draw a single random sample of size $n=300$ from the population and run our estimator function to see a single instance of survey inference.

```{r srs-one-sample}
##
# srs sampling
srs <- sample (1:N,n)
head(agpop [srs, ])
# get data of variable "acres92"
sdata <- agpop [srs, "acres92"]
# analysis
res_one <- srs_est(sdata, N, estimate = "mean", show.details = FALSE)
res_one$estimate |> format_est_gt("mean")

```

### Repeating SRS sampling 5000 times

To demonstrate the Central Limit Theorem and unbiasedness of the estimator, draw 5,000 independent samples, storing the estimates from each iteration. We plot the resulting sampling distribution against the true population mean.

*(Note: We use `show.details = FALSE` and pull out just `$estimate` here, so the loop skips building 5,000 working tables and stays fast.)*

```{r fig-srs-simulation-distribution, fig.height=10}
#| fig-cap: "Population distribution and sampling distribution of the SRS sample mean over 5000 repeated samples (n = 300)"
nres <- 5000 # number of repeated sampling
simulation.results <- matrix (0, nres, 4) # matrix recording repeated results
colnames(simulation.results) <- c( "Est.",   "S.E.",   "ci.low", "ci.upp")

for (i in 1:nres)
{
    srs <- sample (N, n)
   sdata <- agpop [srs, "acres92"]
   simulation.results [i,] <- srs_est (sdata, N, estimate = "mean", show.details = FALSE)$estimate
}

head(simulation.results)
# look at the distribution of sample mean
par (mfrow= c(2,2))
hist (agpop$acres92,main = "Population Distribution of acre92")
hist (simulation.results[,1], main = "Sampling Distribution of Sample Mean for acre92")
abline (v = ybarU, col = "red")
qqnorm (simulation.results[,1], main="QQ plot of Sample Mean"); qqline(simulation.results[,1])
boxplot (simulation.results[,1], main = "Boxplot of Sample Mean")
abline (h = ybarU, col = "red")

mean (simulation.results[,1])
ybarU
sd (simulation.results [,1])
true.se.ybar

```

### Empirical Coverage Rate of CIs

Verify the reliability of the 95% confidence intervals by checking whether the true population mean falls within the calculated bounds for each of the 5,000 samples. The empirical coverage rate should closely approximate 0.95. The plot highlights the first 100 intervals, flagging those that fail to capture the true mean in a distinct color.

```{r srs-coverage-setup}
library(dplyr)
library(plotrix)

# Convert to data frame (if it was a matrix) and use mutate to evaluate coverage
simulation.results <- as.data.frame(simulation.results) |>
  mutate(
    `Covered?` = as.numeric(ci.low < ybarU & ybarU < ci.upp)
  )

head(simulation.results)
```

```{r fig-srs-coverage-intervals, fig.height=6}
#| fig-cap: "First 100 simulated 95% confidence intervals for the SRS sample mean, with coverage failures highlighted"
par(mfrow=c(1,1))
plotCI(
  x = 1:100,
  y = simulation.results$Est.[1:100], 
  li = simulation.results$ci.low[1:100],
  ui = simulation.results$ci.upp[1:100],
  col = 2 - simulation.results$`Covered?`[1:100]
)
abline(h = ybarU, col = "blue")

# Empirical coverage rate
mean(simulation.results$`Covered?`)

```

## Shinylive App for Illustrating SRS Theory {#sec-app-srs}

The simulation for this section runs as a standalone app:
[Simple random sampling](../shinylives_sampling/app-srs.html){target="_blank"}.
It opens in a new tab, together with its documentation and the full
[list of apps](../shinylives_sampling/index.html){target="_blank"}.
